### 2024-05-30 Comparison.Rmd
First upload and start of adding data

---
title: "Template Title"
author: "Niklas Hausmann"
date: "Today"
output: pdf_document
---
```{r Libraries,echo=FALSE,message = FALSE,warning = FALSE}

knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)

{
  pacman::p_load(
    here,
    janitor,
    tidyverse,
    cowplot,ggx,ggpubr,
    patchwork,
    RColorBrewer)

  theme_set(theme_cowplot())
}

source("DTW_Script.R")

files <- list.files(path = here("data"), pattern = "\\.csv$", full.names = TRUE)

MgCa <- files %>%
  set_names() %>%
  map_dfr(~read_csv(.x, col_names = c("x", "y", "z", "mg_ca", "std", "relstd")), .id = "sampleID") %>%
  mutate(sampleID = tools::file_path_sans_ext(basename(sampleID))) %>%
  mutate(sampleID = str_replace(sampleID, "_analyzed_data", "")) %>%
  mutate(sampleID = str_replace(sampleID, "^(.*?A\\d+).*", "\\1")) %>% 
  mutate(sampleID = str_replace(sampleID, "-A10", "")) %>% 
  group_by(sampleID) %>% 
  mutate(dist=row_number()*0.01) %>% 
  ungroup() %>% 
  filter(!(sampleID=="ND-1016-4"&dist>6.5)) %>% 
  filter(relstd<0.1)

D18O <- read_csv(here("data","Isotopes","d18O.csv"))

```

# Plots

```{r}
#| label: fig-timeline
#| fig-cap: Timeline of recent earthquakes on La Palma
#| fig-alt: An event plot of the years of the last 8 eruptions on La Palma.
#| fig-height: 1.5
#| fig-width: 6

P1 <- MgCa %>% 
  filter(std<0.1) %>% 
  filter(mg_ca<0.7) %>% 
  filter(dist<10) %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path(col="firebrick")+
  xlim(0,10)+
  facet_wrap(~sampleID,scales = "free_y",nrow=1)

P2 <- D18O %>% 
  filter(sampleID %in% MgCa$sampleID) %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path(col="cornflowerblue")+
  scale_y_reverse()+
  facet_wrap(~sampleID,scales = "free_y",nrow=1)

plot_compare <- P1/P2

```

# Compare

```{r}
D18O %>%
  filter(sampleID %in% MgCa$sampleID) %>%
  bind_rows(
    MgCa %>% filter(std < 0.1) %>%  filter(mg_ca < 0.7) %>%  filter(dist < 10)%>% select(sampleID, dist, mg_ca)
  ) %>%
  mutate(d18O = d18O * (-0.05)) %>%
  pivot_longer(values_to = "value",
               names_to = "proxy",
               cols = c(d18O, mg_ca)) %>%
  na.omit() %>%
  ggplot() +
  aes(dist, value, col = proxy) +
  geom_path() +
  facet_wrap( ~ sampleID, scales = "free")

```

```{r}
Samples <- unique(MgCa$sampleID)

# Initialize an empty list to store the results of shell_dtw for each sample
dtw_results <- list()

Corr_data <- tibble()

for (sample in Samples) {
  temp_D18O <- D18O %>% 
    filter(sampleID == sample) %>% 
    select(d18O) %>% 
    rename(d18o = d18O) %>% 
    mutate(sample_seq = row_number())

  temp_MgCa <- MgCa %>% 
    filter(sampleID == sample) %>% 
    filter(!(sampleID=="ND-1016-4"&dist<0.3)) %>% 
    select(mg_ca) %>% 
    mutate(sample_seq = row_number())

  # Apply shell_dtw to the temp dataframes and store the result in the list
  dtw_results[[sample]] <- shell_dtw_data(temp_D18O, temp_MgCa)

   # Add the sampleID column to temp_data
  temp_data <- dtw_results[[sample]]$data %>% 
    mutate(sampleID = sample)

  # Append temp_data to Corr_data
  Corr_data <- bind_rows(Corr_data, temp_data)
}

```

# Show Correlation data

```{r}

P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +

  scale_colour_manual(values = c("#007f7f", "#d73027", "#ffc100","black"))+
  scale_fill_manual(values = c("#007f7f", "#d73027", "#ffc100","black"))
# Expanded colour palette with 6 additional colours
# Updated colour palette with 4 new replacements for the last colours
# Refined colour palette with 3 additional colours
# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")

P_individual <- P_all+
  facet_wrap(~sampleID,ncol=2)+#ylim(0.15,NA)+
  ggpubr::stat_regline_equation(label.x = 1.5, label.y = 0.25,
    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
    formula = y~x
  )

Plot_Correlations <- P_all/P_individual+plot_layout(heights = c(0.5,0.5))

ggsave(plot=Plot_Correlations, here("Figures", "Correlations.png"), units = "cm", width = 29.7, height = 20.7)
```

# Look at ND-1016-4
```{r}

sample <- "ND-1016-4"

temp_D18O <- D18O %>% 
    filter(sampleID == sample) %>% 
    rename(d18o = d18O) %>% 
    ggplot() +
  aes(dist, d18o) +
  scale_y_reverse()+
  geom_path(col="turquoise")

temp_MgCa <- MgCa %>% 
    filter(sampleID == sample) %>% 
    filter(dist>0.3) %>%
    ggplot()+
  aes(dist,mg_ca)+
  geom_path(col="coral")

temp_MgCa/temp_D18O
```

### 2024-05-30 Manuscript.qmd
First upload and start of adding data

---
title: "Manuscript"
format: html
output:
  units: cm
---

```{r}
#| label: Libray and datasets
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)
# Packages
  pacman::p_load(
    here,
    janitor,
    tidyverse,
    cowplot,ggx,ggpubr,
    patchwork,
    RColorBrewer)

  theme_set(theme_cowplot())

# Scripts
source("DTW_Script.R")

# Files

MgCa <- read_csv(here("data","MgCa.csv")) %>% 
  arrange(sampleID)
D18O <- read_csv(here("data","d18O.csv")) %>% 
    arrange(sampleID)

# Check overlaps
# Find common sample IDs
intersect(MgCa$sampleID, D18O$sampleID)

# Find sample IDs only in MgCa
setdiff(MgCa$sampleID, D18O$sampleID)

# Find sample IDs only in D18O
setdiff(D18O$sampleID, MgCa$sampleID)

```

```{r}
#| label: Raw Data
#| fig-cap: Raw Mg/Ca and δ^18^O Data
#| fig-height: 30
#| fig-width: 8

P1 <- MgCa %>% 
  filter(std<0.1) %>% 
  # filter(mg_ca<0.7) %>% 
  # filter(dist<10) %>% 
ggplot()+
  aes(dist,mg_ca)+
  geom_path(col="firebrick")+
  # xlim(0,10)+
  facet_wrap(~sampleID,ncol=1,scales = "free")

P2 <- D18O %>% 
  filter(sampleID %in% MgCa$sampleID) %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path(col="cornflowerblue")+
  scale_y_reverse()+
  facet_wrap(~sampleID,scales = "free",ncol=1)

P1|P2

```

### 2024-05-30 including QG2-7246.Rmd
First upload and start of adding data

---
title: "Template Title"
author: "Niklas Hausmann"
date: "Today"
output: pdf_document
---

# Libraries 
```{r Libraries,echo=FALSE,message = FALSE,warning = FALSE}
knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)

{
  pacman::p_load(
    here,
    janitor,
    tidyverse,
    cowplot,ggx,
    patchwork,
    RColorBrewer)

  theme_set(theme_cowplot())
}
```

# Data

```{r}
QGShell <- read_csv(file=here("data","QG1-7246-1-A30Centre_analyzed_data.csv"), col_names = c("x", "y", "z", "mg_ca", "std", "relstd")) %>%
  mutate(sampleID = "QG1-7246-1") %>%
  filter(relstd<0.1,
         x>1.3) %>% 
  mutate(x=round(x,2)) %>% 
  group_by(y) %>% 
  mutate(mg_ca=mean(mg_ca),
         std=mean(std),
         n=n()) %>% 
  ungroup() %>% 
  filter(n>1) %>% 
  select(sampleID,everything(),-n) %>% 
  arrange(y) %>% 
  mutate(dist=row_number()*0.03)

QGShell %>% 
  ggplot()+
  aes(y,mg_ca)+
  geom_line()

```

```{r}

write_csv(bind_rows(MgCa,QGShell),file=here("data","MgCa.csv"))

```

### 2024-05-30 DTW_Script.R

@@ -1,0 +1,90 @@
+shell_dtw_data <- function(D18o, MgCa, show.dtw.plot = FALSE, plot_alignment = FALSE, return_data = TRUE) {
+  require(tidyverse)
+  require(dtw)
+  require(ggplotify)
+  
+  if (is.null(show.dtw.plot)) 
+    show.dtw.plot <- FALSE
+  
+  plot_alignment_local <- plot_alignment
+  data_to_return <- NULL
+  
+  suppressMessages({
+    suppressWarnings({
+      D18o_warp <- D18o %>% mutate(d18o = scale(d18o)) %>%
+        pull(d18o)
+      MgCa_warp <- MgCa %>% mutate(mg_ca = scale(mg_ca)) %>%
+        pull(mg_ca)
+      alignment <- dtw(D18o_warp * -1,
+                       MgCa_warp,
+                       keep = TRUE,
+                       step.pattern = symmetric2)
+      Aligned_d18o <- D18o %>% mutate(x = 1:length(D18o$d18o),
+                                      proxy = "d18o") %>% rename(value = d18o) %>%
+        select(x, value, proxy)
+      Aligned_mg_ca <-
+        MgCa %>% pull(mg_ca) %>% .[alignment$index2] %>%
+        as.data.frame() %>% clean_names() %>% rename(value = x) %>%
+        cbind(x = alignment$index1) %>% mutate(proxy = "mg_ca")
+      
+      # Code for plotting alignment data
+      if (plot_alignment) {
+        plot(alignment,
+             type = "twoway",
+             offset = 10)
+      }
+      
+      Data_aligned <- rbind(Aligned_d18o, Aligned_mg_ca) %>%
+        group_by(x, proxy) %>% summarise(value = mean(value)) %>%
+        ungroup() %>% mutate(value = round(value + 1e-04 *
+                                             row_number(), 5)) %>% group_by(proxy, value) %>%
+        summarise(x = mean(x))
+      lmp <- function(modelobject) {
+        if (class(modelobject) != "lm")
+          stop("Not an object of class 'lm' ")
+        f <- summary(modelobject)$fstatistic
+        p <- pf(f[1], f[2], f[3], lower.tail = F)
+        attributes(p) <- NULL
+        return(p)
+      }
+      corr_data <-
+        Data_aligned %>% pivot_wider(names_from = "proxy",
+                                     values_from = "value") %>% select(d18o, mg_ca)
[…]

### 2024-05-30 Manuscript.qmd
added author info
Cleaned Data for dtw

@@
 format: html
+author: Niklas Hausmann
 output:
@@
 # Files  
-
-MgCa <- read_csv(here("data","MgCa.csv")) %>% 
-  arrange(sampleID)
-D18O <- read_csv(here("data","d18O.csv")) %>% 
-    arrange(sampleID)
+read_csv(D18O,here("data","D18O.csv"))
+read_csv(MgCa,here("data","MgCa.csv"))
 
@@
 # Find common sample IDs
-intersect(MgCa$sampleID, D18O$sampleID)
+intersect(MgCa$sampleID, NMgCa$sampleID)
 
 # Find sample IDs only in MgCa
-setdiff(MgCa$sampleID, D18O$sampleID)
+setdiff(MgCa$sampleID, NMgCa$sampleID)
 
 # Find sample IDs only in D18O
-setdiff(D18O$sampleID, MgCa$sampleID)
+setdiff(NMgCa$sampleID, MgCa$sampleID)
 
@@
 ```
+
+```{r}
+#| label: DTW Analysis
+
+Samples <- unique(MgCa$sampleID)
+
+# Initialize an empty list to store the results of shell_dtw for each sample
+dtw_results <- list()
+
+Corr_data <- tibble()
+
+sample <- Samples[6]
+
+# for (sample in Samples) {
+  temp_D18O <- D18O %>% 
+    filter(sampleID == sample) %>% 
+    select(d18O) %>% 
+    rename(d18o = d18O) %>% 
+    mutate(sample_seq = row_number())
+
+  temp_MgCa <- MgCa %>% 
+    filter(sampleID == sample) %>% 
+    filter(!(sampleID=="ND-1016-4"&dist<0.3)) %>% 
+    select(mg_ca) %>% 
+    mutate(sample_seq = row_number())
+
+  # Apply shell_dtw to the temp dataframes and store the result in the list
+  dtw_results[[sample]] <- shell_dtw_data(temp_D18O, temp_MgCa)
+
+   # Add the sampleID column to temp_data
+  temp_data <- dtw_results[[sample]]$data %>% 
+    mutate(sampleID = sample)
+
+  # Append temp_data to Corr_data
+  Corr_data <- bind_rows(Corr_data, temp_data)
+# }
+
+```
+
+```{r}
+#| label: DTW Results 
+#| fig-cap: 
+#| fig-height:
+#| fig-width:
+
+P_all <- Corr_data %>% 
+  pivot_wider(values_from = "value",names_from = "proxy") %>% 
+  ggplot()+
+  aes(d18o,mg_ca,col=sampleID)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
+  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
+  # stat_regline_equation(label.x = 2, #label.y = 32
+  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+  #   formula = y~x
+  # ) +
+  scale_colour_manual(values = c("#007f7f", "#d73027", "#ffc100", "black", "#1f78b4", "#33a02c", "#ff7f00", "#6a3d9a", "#e74c3c", "#2ecc71", "#f39c12")) +
+scale_fill_manual(values = c("#007f7f", "#d73027", "#ffc100", "black", "#1f78b4", "#33a02c", "#ff7f00", "#6a3d9a", "#e74c3c", "#2ecc71", "#f39c12"))
+# Expanded colour palette with 6 additional colours
+# Updated colour palette with 4 new replacements for the last colours
+# Refined colour palette with 3 additional colours
+# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
+
+P_individual <- P_all+
+  facet_wrap(~sampleID,ncol=2)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 1.5, label.y = 0.25,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
+    formula = y~x
+  )
+
+P_all/P_individual+plot_layout(heights = c(0.5,0.5))
+
+```

### 2024-05-30 DTW_Script.R

@@ -36,7 +36,7 @@ shell_dtw_data <- function(D18o, MgCa, show.dtw.plot = FALSE, plot_alignment = F
       
       Data_aligned <- rbind(Aligned_d18o, Aligned_mg_ca) %>%
         group_by(x, proxy) %>% summarise(value = mean(value)) %>%
-        ungroup() %>% mutate(value = round(value + 1e-04 *
+        ungroup() %>% mutate(value = round(value + 1e-05 *
                                              row_number(), 5)) %>% group_by(proxy, value) %>%
         summarise(x = mean(x))
       lmp <- function(modelobject) {

### 2024-05-30 Comparison.Rmd
reshaped plots

@@
 
-for (sample in Samples) {
+# for (sample in Samples) {
   temp_D18O <- D18O %>% 
@@
   # Apply shell_dtw to the temp dataframes and store the result in the list
-  dtw_results[[sample]] <- shell_dtw_data(temp_D18O, temp_MgCa)
+  dtw_results[[sample]] <- shell_dtw_data(temp_D18O, temp_MgCa,plot_alignment = FALSE)
 
@@
   Corr_data <- bind_rows(Corr_data, temp_data)
-}
+# }
 
@@
 
-  scale_colour_manual(values = c("#007f7f", "#d73027", "#ffc100","black"))+
-  scale_fill_manual(values = c("#007f7f", "#d73027", "#ffc100","black"))
+  scale_colour_manual(values = c("#007f7f", "#d73027", "#ffc100","black", "black"))+
+  scale_fill_manual(values = c("#007f7f", "#d73027", "#ffc100","black", "black"))
 # Expanded colour palette with 6 additional colours
@@
 
-ggsave(plot=Plot_Correlations, here("Figures", "Correlations.png"), units = "cm", width = 29.7, height = 20.7)
+# ggsave(plot=Plot_Correlations, here("Figures", "Correlations.png"), units = "cm", width = 29.7, height = 20.7)
 ```
@@
 
-sample <- "ND-1016-4"
+# sample <- "ND-1016-4"
 

### 2024-05-30 Manuscript.qmd
reshaped plots

--- 2024-05-30 Manuscript.qmd
@@
 
-MgCa <- read_csv(here("data","MgCa.csv")) %>% 
-  arrange(sampleID)
-D18O <- read_csv(here("data","d18O.csv")) %>% 
-    arrange(sampleID)
+D18O <- read_csv(here("data","D18O.csv"))
+MgCa <- read_csv(here("data","MgCa.csv"))
 
@@
 ```
+
+```{r}
+#| label: DTW Analysis
+
+Samples <- unique(MgCa$sampleID)
+
+# Initialize an empty list to store the results of shell_dtw for each sample
+dtw_results <- list()
+
+Corr_data <- tibble()
+
+sample <- Samples[6]
+
+for (sample in Samples) {
+  temp_D18O <- D18O %>% 
+    filter(sampleID == sample) %>% 
+    select(d18O) %>% 
+    rename(d18o = d18O) %>% 
+    mutate(sample_seq = row_number())
+
+  temp_MgCa <- MgCa %>% 
+    filter(sampleID == sample) %>% 
+    filter(!(sampleID=="ND-1016-4"&dist<0.3)) %>% 
+    select(mg_ca) %>% 
+    mutate(sample_seq = row_number())
+
+  # Apply shell_dtw to the temp dataframes and store the result in the list
+  dtw_results[[sample]] <- shell_dtw_data(temp_D18O, temp_MgCa)
+
+   # Add the sampleID column to temp_data
+  temp_data <- dtw_results[[sample]]$data %>% 
+    mutate(sampleID = sample)
+
+  # Append temp_data to Corr_data
+  Corr_data <- bind_rows(Corr_data, temp_data)
+}
+
+```
+
+```{r}
+#| label: Correlation Graphs 
+#| fig-cap: Correlation graphs for all specimens
+#| fig-height: 20
+#| fig-width: 10
+
+P_all <- Corr_data %>% 
+  pivot_wider(values_from = "value",names_from = "proxy") %>% 
+  mutate(region = substr(sampleID, 1, 1)) %>% 
+  ggplot()+
+  aes(d18o,mg_ca,col=sampleID)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
+  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
+  # stat_regline_equation(label.x = 2, #label.y = 32
+  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+  #   formula = y~x
+  # ) +
+  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063")) +
+scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063"))#+
+  # facet_wrap(~region,ncol=2,scales = "free")
+
+# Expanded colour palette with 6 additional colours
+# Updated colour palette with 4 new replacements for the last colours
+# Refined colour palette with 3 additional colours
+# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
+
+P_individual <- P_all+ylim(0,0.8)+
+  facet_wrap(~sampleID,ncol=3)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 0.65,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label,p.value), sep = "~~~~")),show.legend = FALSE,
+    formula = y~x
+  )+
+  theme(
+    strip.background = element_rect(fill = "white", colour = NA),
+    strip.text = element_text(colour = "black"),
+    panel.border = element_rect(colour="black")
+  )
+
+P_all/P_individual+plot_layout(heights = c(0.5,0.5))
+
+```

### 2024-05-30 Manuscript.tex
reshaped plots

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Manuscript},
  pdfauthor={Niklas Hausmann},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Manuscript}
\author{Niklas Hausmann}
\date{}

\begin{document}
\maketitle

\begin{verbatim}
 [1] "QG1-7188-1" "QG1-7189-2" "QG2-1061-1" "QG2-1064-1" "QG2-7180-1"
 [6] "QG2-7180-2" "QG1-7246-1" "ND-1016-3"  "ND-1016-4"  "NM-1016-1" 
[11] "NM-1016-3" 
\end{verbatim}

\begin{verbatim}
character(0)
\end{verbatim}

\begin{verbatim}
[1] "ND-1016-1" "NM-1016-5"
\end{verbatim}

\begin{figure}[H]

{\centering \includegraphics{Manuscript_files/figure-pdf/Raw Data-1.pdf}

}

\caption{Raw Mg/Ca and δ\textsuperscript{18}O Data}

\end{figure}%

\begin{figure}[H]

{\centering \includegraphics{Manuscript_files/figure-pdf/Correlation Graphs-1.pdf}

}

\caption{Correlation graphs for all specimens}

\end{figure}%

\end{document}

### 2024-05-30 Manuscript.qmd
changed stats

@@
   ggpubr::stat_regline_equation(label.x = 0.5, label.y = 0.65,
-    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label,p.value), sep = "~~~~")),show.legend = FALSE,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
     formula = y~x

### 2024-05-30 Manuscript.qmd
added pdf template

@@
 title: "Manuscript"
-format: html
-author: Niklas Hausmann
+subtitle: A Short Subtitle
+author:
+  - name: Niklas Hausmann
+    affiliations:
+      - name: Leibniz Zentrum für Archäologie
+        address: Ludwig-Lindenschmit-Forum 1
+        city: Mainz
+        country: Germany
+        postal-code: 55116
+    orcid: 0000-0000-0000-0000
+    email: niklas@palaeo.eu
+    attributes:
+        corresponding: true
+abstract: |
+  This document is a template demonstrating the Aps format.
+keywords: 
+  - keyword1
+  - keyword2
+date: last-modified
+bibliography: bibliography.bib
+format:
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Palaeo^3^
+      formatting: preprint
+      model: 3p
+      cite-style: authoryear
 output:
@@
 
-# Check overlaps
-# Find common sample IDs
-intersect(MgCa$sampleID, D18O$sampleID)
-
-# Find sample IDs only in MgCa
-setdiff(MgCa$sampleID, D18O$sampleID)
-
-# Find sample IDs only in D18O
-setdiff(D18O$sampleID, MgCa$sampleID)
+# # Check overlaps
+# # Find common sample IDs
+# intersect(MgCa$sampleID, D18O$sampleID)
+# 
+# # Find sample IDs only in MgCa
+# setdiff(MgCa$sampleID, D18O$sampleID)
+# 
+# # Find sample IDs only in D18O
+# setdiff(D18O$sampleID, MgCa$sampleID)
 

### 2024-05-30 Manuscript.tex
added pdf template

@@
 \documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+  super,
+  preprint,
+  3p]{elsarticle}
 
@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 \makeatother
+\journal{Palaeo3}
 \ifLuaTeX
@@
 \fi
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-num}
 \usepackage{bookmark}
@@
   pdfauthor={Niklas Hausmann},
+  pdfkeywords={keyword1, keyword2},
   colorlinks=true,
@@
 
-\title{Manuscript}
-\author{Niklas Hausmann}
-\date{}
+\setlength{\parindent}{6pt}
+\begin{document}
 
-\begin{document}
-\maketitle
+\begin{frontmatter}
+\title{Manuscript \\\large{A Short Subtitle} }
+\author[1]{Niklas Hausmann%
+\corref{cor1}%
+}
+ \ead{niklas@palaeo.eu}
+
+\affiliation[1]{organization={Leibniz Zentrum für
+Archäologie},addressline={Ludwig-Lindenschmit-Forum
+1},city={Mainz},country={Germany},countrysep={,},postcode={55116},postcodesep={}}
+
+\cortext[cor1]{Corresponding author}
+
+\begin{abstract}
+This document is a template demonstrating the Aps format.
+\end{abstract}
+
+\begin{keyword}
+    keyword1 \sep 
+    keyword2
+\end{keyword}
+\end{frontmatter}
 
@@
 
+  \bibliography{bibliography.bib}
+
 \end{document}

### 2024-05-30 Manuscript.qmd
added data

@@
 ---
-title: "Manuscript"
-subtitle: A Short Subtitle
+title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
 author:
@@
         postal-code: 55116
-    orcid: 0000-0000-0000-0000
+    orcid: 0000–0002–4381–8334
     email: niklas@palaeo.eu
@@
       corresponding: true
+  - name: Donna Surge
+    affiliations:
+      - name: University of North Carolina
+        address: 104 South Road, 225 Geology Building
+        city: Chapel Hill, NC
+        country: US
+        postal-code: 27599-3315
+  - name: Francisco Zangrando
+    affiliations:
+      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
+        address: Avenida Maipú 305
+        city: Ushuaia
+        country: Argentina
+        postal-code: V9410BJA
+  - name: Angelica Tivoli
+    affiliations:
+      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
+        address: Avenida Maipú 305
+        city: Ushuaia
+        country: Argentina
+        postal-code: V9410BJA
+  - name: Ivan Briz-Godino
+    affiliations:
+      - name: 
+        address: 
+        city: 
+        country: Spain
+        postal-code: 
 abstract: |
@@
 
-```{r}
-#| label: Libray and datasets
+# Introduction {#Introduction}
+
+```{r Libray and datasets}
 #| echo: FALSE
@@
 
-```{r}
-#| label: Raw Data
-#| fig-cap: Raw Mg/Ca and δ^18^O Data
-#| fig-height: 30
-#| fig-width: 8
+# Methods {#Methods}
 
-P1 <- MgCa %>% 
-  filter(std<0.1) %>% 
-  # filter(mg_ca<0.7) %>% 
-  # filter(dist<10) %>% 
-ggplot()+
-  aes(dist,mg_ca)+
-  geom_path(col="firebrick")+
-  # xlim(0,10)+
-  facet_wrap(~sampleID,ncol=1,scales = "free")
-
-P2 <- D18O %>% 
-  filter(sampleID %in% MgCa$sampleID) %>% 
-  ggplot()+
-  aes(dist,d18O)+
-  geom_path(col="cornflowerblue")+
-  scale_y_reverse()+
-  facet_wrap(~sampleID,scales = "free",ncol=1)
-
-P1|P2
-
-```
-
-```{r}
-#| label: DTW Analysis
+```{r DTW Analysis}
 
@@
 
-```{r}
-#| label: Correlation Graphs 
+# Results {#Results}
+
+```{r Correlation Graphs }
 #| fig-cap: Correlation graphs for all specimens
@@
 ```
+
+| Species            | Locality       | Specimen | Correlation R^2^ | Study                                                          |
+|-------------------|---------------|---------------|---------------|---------------|
+| *Patella depressa* | Northern Spain | LAN541   | 0.87             | [**10.3390/app11072959**](https://doi.org/10.3390/app11072959) |
+|                    |                | LAN545   | 0.86             |                                                                |
+|                    |                | LAN554   | 0.78             |                                                                |
+|                    |                | LAN559   | 0.82             |                                                                |
+| Patella caerulea   | Croatia        | ISTPC1   | 0.9              | 10.1038/s41598-019-39959-9                                     |
+|                    |                | ISTPC2   | 0.84             |                                                                |
+|                    | Crete          | AF1911A  | 0.91[^1]         |                                                                |
+|                    |                | AF3003A  | 0.92[^2]         |                                                                |
+|                    | Israel         | AKKPC2   | 0.96             |                                                                |
+|                    |                | AKKPC3   | 0.89             |                                                                |
+|                    |                | FRMPC1   | 0.84             |                                                                |
+|                    |                | FRMPC2   | 0.96             |                                                                |
+|                    | Libya          | MO31A    | 0.83             |                                                                |
+|                    |                | MP64A    | 0.33             |                                                                |
+|                    |                | MP67A    | 0.96             |                                                                |
+|                    |                | MP68A    | 0.81             |                                                                |
+|                    | Malta          | MA10     | 0.82             |                                                                |
+|                    | Tunisia        | TUNPC1   | 0.81             |                                                                |
+|                    |                | TUNPC2   | 0.78             |                                                                |
+|                    | Turkey         | ANTPC1   | 0.95             |                                                                |
+|                    |                | ANTPC2   | 0.93             |                                                                |
+|                    |                | KIZPC1   | 0.94             |                                                                |
+|                    |                | KIZPC2   | 0.86             |                                                                |
+|                    |                |          |                  |                                                                |
+|                    |                |          |                  |                                                                |
+|                    |                |          |                  |                                                                |
+|                    |                |          |                  |                                                                |
+|                    |                |          |                  |                                                                |
+|                    |                |          |                  |                                                                |
+|                    |                |          |                  |                                                                |
+|                    |                |          |                  |                                                                |
+
+: Overview of comparative correlations {#tab:correlations}
+
+[^1]: SST only, no other geochemical data available
+
+[^2]: SST only, no geochemical data available

### 2024-05-30 Manuscript.tex
added data

@@
 \documentclass[
-  super,
+  authoryear,
   preprint,
@@
 \usepackage[]{natbib}
-\bibliographystyle{elsarticle-num}
+\bibliographystyle{elsarticle-harv}
 \usepackage{bookmark}
@@
 \hypersetup{
-  pdftitle={Manuscript},
-  pdfauthor={Niklas Hausmann},
+  pdftitle={Interspecies comparisons of Mg/Ca ratios in limpet shells},
+  pdfauthor={Niklas Hausmann; Donna Surge; Francisco Zangrando; Angelica Tivoli; Ivan Briz-Godino},
   pdfkeywords={keyword1, keyword2},
@@
 \begin{frontmatter}
-\title{Manuscript \\\large{A Short Subtitle} }
+\title{Interspecies comparisons of Mg/Ca ratios in limpet shells}
 \author[1]{Niklas Hausmann%
@@
  \ead{niklas@palaeo.eu} 
+\author[2]{Donna Surge%
+%
+}
+
+\author[3]{Francisco Zangrando%
+%
+}
+
+\author[3]{Angelica Tivoli%
+%
+}
+
+\author[4]{Ivan Briz-Godino%
+%
+}
 
@@
 1},city={Mainz},country={Germany},countrysep={,},postcode={55116},postcodesep={}}
+\affiliation[2]{organization={University of North
+Carolina},addressline={104 South Road, 225 Geology
+Building},city={Chapel Hill,
+NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
+\affiliation[3]{organization={CONICET (Consejo Nacional de
+Investigaciones Científicas y Técnicas)},addressline={Avenida Maipú
+305},city={Ushuaia},country={Argentina},countrysep={,},postcode={V9410BJA},postcodesep={}}
+\affiliation[4]{organization={},country={Spain},countrysep={,},postcodesep={}}
 
@@
 \end{frontmatter}
-
-\begin{verbatim}
- [1] "QG1-7188-1" "QG1-7189-2" "QG2-1061-1" "QG2-1064-1" "QG2-7180-1"
- [6] "QG2-7180-2" "QG1-7246-1" "ND-1016-3"  "ND-1016-4"  "NM-1016-1" 
-[11] "NM-1016-3" 
-\end{verbatim}
-
-\begin{verbatim}
-character(0)
-\end{verbatim}
-
-\begin{verbatim}
-[1] "ND-1016-1" "NM-1016-5"
-\end{verbatim}
 

### 2024-06-14 Manuscript.qmd
Added ORK-LT5

@@
 
-```{r Libray and datasets}
+```{r Library and datasets}
 #| echo: FALSE
@@
 # Scripts
-source("DTW_Script.R")
+# source("DTW_Script.R")
+source("DTW_Script-update.R")
 
@@
 MgCa <- read_csv(here("data","MgCa.csv"))
+
+Ferguson_data <- read_csv(here("data","Ferguson-data.csv")) %>% 
+  na.omit()
 
@@
 
+This paper is a short reassessment of Atlantic limpet shells and their correlations to Mg/Ca ratios.
+
+In previous studies, various limpet specimens (Patella vulgata, Nacella magellanica and Nacella odifosdio) have been studied using Mg/Ca with mixed results.
+While Patella vulgata along the Spanish shoreline has since then repeatedly produced reliable correlations between sst and Mg/Ca ratios, this is not the case for other species.
+
+In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Some of these...
+
 # Methods {#Methods}
@@
 
-sample <- Samples[6]
+# sample <- Samples[12]
 
@@
     filter(!(sampleID=="ND-1016-4"&dist<0.3)) %>%
+
     select(mg_ca) %>% 
@@
   # ) +
-  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063")) +
-scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063"))#+
+  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
+scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
   # facet_wrap(~region,ncol=2,scales = "free")
@@
 
+# Discussion
+
+```{r Ferguson Data}
+#| fig-cap: Correlation graphs for Ferguson et al. specimens
+#| fig-height: 20
+#| fig-width: 10
+
+Ferg_all <-
+  Ferguson_data %>% 
+  ggplot()+
+  aes(d18o,mg_ca,col=shell)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
+  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
+  # stat_regline_equation(label.x = 2, #label.y = 32
+  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+  #   formula = y~x
+  # ) +
+  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
+scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))
+
+# Expanded colour palette with 6 additional colours
+# Updated colour palette with 4 new replacements for the last colours
+# Refined colour palette with 3 additional colours
+# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
+
+Ferg_individual <- Ferg_all+
+  facet_wrap(~shell,ncol=2)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 1, label.y = 26,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
+    formula = y~x
+  )+
+  theme(
+    strip.background = element_rect(fill = "white", colour = NA),
+    strip.text = element_text(colour = "black"),
+    panel.border = element_rect(colour="black")
+  )
+
+Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
+
+```
+
 | Species            | Locality       | Specimen | Correlation R^2^ | Study                                                                                                                              |
-|-------------------|---------------|---------------|---------------|---------------|
+|-------------|-------------|-------------|-------------|----------------------|
 | *Patella depressa* | Northern Spain | LAN541   | 0.87             | [**10.3390/app11072959**](https://doi.org/10.3390/app11072959)                                                                     |
@@
 |                    |                | LAN559   | 0.82             |                                                                                                                                    |
-| Patella caerulea   | Croatia        | ISTPC1   | 0.9              | 10.1038/s41598-019-39959-9                                     |
+| *Patella caerulea* | Croatia        | ISTPC1   | 0.9              | 10.1038/s41598-019-39959-9                                                                                                         |
 |                    |                | ISTPC2   | 0.84             |                                                                                                                                    |
@@
 |                    |                | KIZPC2   | 0.86             |                                                                                                                                    |
-|                    |                |          |                  |                                                                |
-|                    |                |          |                  |                                                                |
-|                    |                |          |                  |                                                                |
-|                    |                |          |                  |                                                                |
+| Patella rustica    | Gibraltar      | JL1      | 0.02             | [doi.org/10.1016/j.epsl.2011.05.054](https://doi.org/10.1016/j.epsl.2011.05.054 "Persistent link using digital object identifier") |
+|                    |                | JL2      | 0.8 (0.79)       |                                                                                                                                    |
+| Patella caerulea   | Gibraltar      | JM00     | 0.69 (0.79)      |                                                                                                                                    |
+|                    |                | JM30     | 0.83 (0.79)      |                                                                                                                                    |
 |                    |                |          |                  |                                                                                                                                    |
@@
 [^2]: SST only, no geochemical data available
+
+## Reanalysis of ORK-LT5
+
+```{r}
+
+P1 <-
+  MgCa %>% 
+  filter(sampleID=="ORK_LT5") %>% 
+  # filter(std<0.1) %>% 
+  # filter(mg_ca<0.7) %>% 
+  # filter(dist<20) %>% 
+ggplot()+
+  aes(dist,mg_ca)+
+  geom_path(col="firebrick")+
+  # xlim(0,10)+
+  facet_wrap(~sampleID,scales = "free",nrow=1)
+
+P2 <-
+  D18O %>% 
+  filter(sampleID=="ORK_LT5") %>% 
+  ggplot()+
+  aes(dist,d18O)+
+  geom_path(col="cornflowerblue")+
+  scale_y_reverse()+
+  facet_wrap(~sampleID,scales = "free",nrow=1)
+
+plot_compare <- P1/P2
+
+plot_compare
+
+```
+
+```{r subsample}
+
+# Function to create the desired subsampled tibble
+subsample_data <- function(data, window_size = 0.05, step_size = 0.15) {
+  max_dist <- max(data$dist)
+  subsample_intervals <- seq(0, max_dist, by = step_size)
+
+  subsampled_data <- subsample_intervals %>%
+    purrr::map_df(~ {
+      window_start <- .x
+      window_end <- .x + window_size
+
+      data_in_window <- data %>%
+        filter(dist >= window_start & dist < window_end)
+
+      if (nrow(data_in_window) > 0) {
+        summarised_data <- data_in_window %>%
+          summarise(across(c(x, y, z, mg_ca, std, relstd, dist), mean, .names = "mean_{.col}"))
+        summarised_data <- summarised_data %>%
+          mutate(window_start = window_start, window_end = window_end)
+        return(summarised_data)
+      } else {
+        return(tibble())
+      }
+    })
+
+  return(subsampled_data)
+}
+
+# Apply the function to your data
+MgCa_filtered <- MgCa %>%
+  filter(sampleID == "ORK_LT5")
+
+MgCa_subsampled_150 <- subsample_data(MgCa_filtered,step_size = 0.15)
+
+MgCa_subsampled_300 <-  subsample_data(MgCa_filtered,step_size = 0.30)
+
+P3.1 <-
+  MgCa_subsampled_150 %>% 
+ggplot()+
+  aes(mean_dist,mean_mg_ca)+
+  geom_path(col="firebrick2")
+
+P3.2 <-
+  MgCa_subsampled_300 %>% 
+ggplot()+
+  aes(mean_dist,mean_mg_ca)+
+  geom_path(col="firebrick4")
+
+P1/P3.1/P3.2
+
+```

### 2024-06-17 Manuscript.qmd
transitioned to new repository name

@@
 keywords: 
-  - keyword1
-  - keyword2
+  - Sclerochronology
+  - Limpets
+  - Elemental Ratio
+  - Mg/Ca
 date: last-modified
@@
 
-This paper is a short reassessment of Atlantic limpet shells and their correlations to Mg/Ca ratios.
+This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures.
+While recent studies of particularly *Patella* sp.
+in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].
 
@@
 | Species            | Locality       | Specimen | Correlation R^2^        | Study                                                                                                                              |
-|-------------|-------------|-------------|-------------|----------------------|
+|-----------|-----------|-----------|-----------|----------------------------|
 | *Patella depressa* | Northern Spain | LAN541   | 0.87                    | [**10.3390/app11072959**](https://doi.org/10.3390/app11072959)                                                                     |
@@
 |                    |                | JM30     | 0.83 (0.79)             |                                                                                                                                    |
-|                    |                |          |                  |                                                                                                                                    |
+| Patella vulgata    | Orkney         | ORK-LT5  | not reported, here 0.88 | doi.org/10.1016/j.palaeo.2016.10.021 and this study                                                                                |
 |                    |                |          |                         |                                                                                                                                    |

### 2024-06-17 Manuscript.tex
transitioned to new repository name

@@
   pdfauthor={Niklas Hausmann; Donna Surge; Francisco Zangrando; Angelica Tivoli; Ivan Briz-Godino},
-  pdfkeywords={keyword1, keyword2},
+  pdfkeywords={Sclerochronology, Limpets, Elemental Ratio, Mg/Ca},
   colorlinks=true,
@@
 \begin{keyword}
-    keyword1 \sep 
-    keyword2
+    Sclerochronology \sep Limpets \sep Elemental Ratio \sep 
+    Mg/Ca
 \end{keyword}
@@
 
-\begin{figure}[H]
+\section{Introduction}\label{Introduction}
 
-{\centering \includegraphics{Manuscript_files/figure-pdf/Raw Data-1.pdf}
+This study provides a short reassessment of the use of Magnesium to
+Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea
+surface temperatures. While recent studies of particularly
+\emph{Patella} sp. in the Mediterranean and Southwest Europe have
+provided promising results
+\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
 
-}
+In previous studies, various limpet specimens (Patella vulgata, Nacella
+magellanica and Nacella odifosdio) have been studied using Mg/Ca with
+mixed results. While Patella vulgata along the Spanish shoreline has
+since then repeatedly produced reliable correlations between sst and
+Mg/Ca ratios, this is not the case for other species.
 
-\caption{Raw Mg/Ca and δ\textsuperscript{18}O Data}
+In this study, we present elemental maps of various such species
+together with stable oxygen isotope values for some of the specimens.
+Some of these\ldots{}
 
-\end{figure}%
+\section{Methods}\label{Methods}
+
+\section{Results}\label{Results}
 
@@
 
+\section{Discussion}\label{discussion}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{Manuscript_files/figure-pdf/Ferguson Data-1.pdf}
+
+}
+
+\caption{Correlation graphs for Ferguson et al.~specimens}
+
+\end{figure}%
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1528}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3889}}@{}}
+\caption{Overview of comparative correlations
+\{\#tab:correlations\}}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Species
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Locality
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Specimen
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Correlation R\textsuperscript{2}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Study
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Species
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Locality
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Specimen
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Correlation R\textsuperscript{2}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Study
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+\emph{Patella depressa} & Northern Spain & LAN541 & 0.87 &
+\href{https://doi.org/10.3390/app11072959}{\textbf{10.3390/app11072959}} \\
+& & LAN545 & 0.86 & \\
+& & LAN554 & 0.78 & \\
+& & LAN559 & 0.82 & \\
+\emph{Patella caerulea} & Croatia & ISTPC1 & 0.9 &
+10.1038/s41598-019-39959-9 \\
+& & ISTPC2 & 0.84 & \\
+& Crete & AF1911A & 0.91\footnote{SST only, no other geochemical data
+  available} & \\
+& & AF3003A & 0.92\footnote{SST only, no geochemical data available}
+& \\
+& Israel & AKKPC2 & 0.96 & \\
+& & AKKPC3 & 0.89 & \\
+& & FRMPC1 & 0.84 & \\
+& & FRMPC2 & 0.96 & \\
+& Libya & MO31A & 0.83 & \\
+& & MP64A & 0.33 & \\
+& & MP67A & 0.96 & \\
+& & MP68A & 0.81 & \\
+& Malta & MA10 & 0.82 & \\
+& Tunisia & TUNPC1 & 0.81 & \\
+& & TUNPC2 & 0.78 & \\
+& Turkey & ANTPC1 & 0.95 & \\
+& & ANTPC2 & 0.93 & \\
+& & KIZPC1 & 0.94 & \\
+& & KIZPC2 & 0.86 & \\
+Patella rustica & Gibraltar & JL1 & 0.02 &
+\href{https://doi.org/10.1016/j.epsl.2011.05.054}{doi.org/10.1016/j.epsl.2011.05.054} \\
+& & JL2 & 0.8 (0.79) & \\
+Patella caerulea & Gibraltar & JM00 & 0.69 (0.79) & \\
+& & JM30 & 0.83 (0.79) & \\
+Patella vulgata & Orkney & ORK-LT5 & not reported, here 0.88 &
+doi.org/10.1016/j.palaeo.2016.10.021 and this study \\
+& & & & \\
+& & & & \\
+& & & & \\
+\end{longtable}
+
+\subsection{Reanalysis of ORK-LT5}\label{reanalysis-of-ork-lt5}
+
+\includegraphics{Manuscript_files/figure-pdf/unnamed-chunk-1-1.pdf}
+
+\includegraphics{Manuscript_files/figure-pdf/subsample-1.pdf}
+
   \bibliography{bibliography.bib}

### 2024-06-17 DTW_Script-update.R

@@ -18,7 +18,7 @@ shell_dtw_data <- function(D18o, MgCa, show.dtw.plot = FALSE, plot_alignment = F
       alignment <- dtw(D18o_warp * -1,
                        MgCa_warp,
                        keep = TRUE,
-                       step.pattern = symmetric1)
+                       step.pattern = symmetric2)
       Aligned_d18o <- D18o %>% mutate(x = 1:length(D18o$d18o),
                                       proxy = "d18o") %>% rename(value = d18o) %>%
         select(x, value, proxy)

### 2024-06-17 Manuscript.qmd
Added structure and Graniero data

@@
 abstract: |
-  This document is a template demonstrating the Aps format.
+  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures.
+While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species.
+*Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results.
+In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens.
+Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios.
+By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P.
+vulgata*, *N.
+deaureata*, and *N.
+magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e.
+annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
 keywords: 
@@
 
-This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures.
-While recent studies of particularly *Patella* sp.
-in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].
+While recent studies of particularly *Patella* sp. in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].
 
-In previous studies, various limpet specimens (Patella vulgata, Nacella magellanica and Nacella odifosdio) have been studied using Mg/Ca with mixed results.
-While Patella vulgata along the Spanish shoreline has since then repeatedly produced reliable correlations between sst and Mg/Ca ratios, this is not the case for other species.
+# Materials and Methods {#Methods}
 
-In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Some of these...
+## Limpet specimens
 
-# Methods {#Methods}
+### Modern specimens
+
+### Archaeological specimens
+
+## Oxygen isotopes
+
+## Mg/Ca ratios
+
+# Results {#Results}
+
+## Patella vulgata
+
+## Nacella deaureata
+
+## Nacella magellanica
 
@@
 
-# Results {#Results}
+#
 
@@
 
+## Other correlations
+
 ```{r Ferguson Data}
@@
 | Species            | Locality       | Specimen | Correlation R^2^        | Study                                                                                                                              |
-|-----------|-----------|-----------|-----------|----------------------------|
+|-------------|-------------|-------------|-------------|----------------------|
 | *Patella depressa* | Northern Spain | LAN541   | 0.87                    | [**10.3390/app11072959**](https://doi.org/10.3390/app11072959)                                                                     |
@@
 
-## Reanalysis of ORK-LT5
+## Comparison of ORK-LT5
 
@@
 ```
+
+```{r}
+# This is data extracted from Fig. 6E of Graniero et al. 2016
+GranieroDF <- tibble::tribble(
+                ~dist, ~mg_ca,
+                  0.1,     19,
+                  0.2,   20.6,
+                  0.4,   23.8,
+                  0.6,   20.4,
+                    1,   19.1,
+                  1.3,   14.9,
+                  1.6,     16,
+                  1.9,   17.8,
+                  2.3,   15.8,
+                  2.8,   17.3,
+                  3.1,   13.8,
+                  3.3,   24.9,
+                  3.7,   16.1,
+                    4,   15.6,
+                  4.3,   18.4,
+                  4.6,   15.5,
+                  5.2,   18.6,
+                  5.2,   21.2,
+                  5.5,   24.1,
+                  5.8,   18.7,
+                  6.2,   19.3,
+                  6.3,   15.5,
+                  6.7,   16.7,
+                  6.9,   15.3,
+                  7.3,   17.9,
+                  7.6,   13.7,
+                    8,   11.8,
+                  8.2,   16.2,
+                  8.5,   18.1,
+                  8.9,     18,
+                  9.1,   19.7,
+                  9.4,   20.8,
+                  9.7,   19.6,
+                   10,   22.2,
+                 10.2,   20.2,
+                 10.5,   19.3,
+                 10.8,   22.8,
+                 11.2,   18.7,
+                 11.8,   22.1
+                )
+
+```
+
+<!--# Maybe get the actual data from the image itself. -->

### 2024-06-23 Manuscript.qmd
Intro

@@
 
-While recent studies of particularly *Patella* sp. in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].
+Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source.
+They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@shackleton1973; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
+Determining past sea surface temperature (SST) change largely relies on the measurement of δ^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than δ^18^O-values, which are also affected by changes in salinity.
+In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].
+
+Problematically, links between Mg/Ca ratios and SST changes in limpets have also been shown to be unreliable [@Graniero2015-zv] and even where they were successful, anomalous patterns in some specimens had to be filtered out, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].
+
+While recent research particularly of *Patella* sp.
+in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].
+There remains unclarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].
 
@@
 ### Modern specimens
+
+The analysed specimens, their origin and studies with background information can be found in Table 1.
+
+| Context        | Study              | Species          | Location                                          | Sample ID     | Previous analyses                                           |
+|------------|------------|------------|------------|------------|------------|
+| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Bahía Cambaceres Exterior - Tierra del Fuego (AR) | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
+|                |                    |                  |                                                   | ND-1016-4     |                                                             |
+|                |                    | *N. magellanica* |                                                   | NM-1016-1     |                                                             |
+|                |                    |                  |                                                   | NM-1016-3     |                                                             |
+|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - Westray (UK)                      | ORK-LT5       | Stable oxygen and carbon isotope analysis; Mg, Li, Sr, Ca,. |
+| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - Westray (UK)                      | QG1-7188-1    | Stable oxygen and carbon isotope analysis                   |
+|                |                    |                  |                                                   | QG1-7189-2    |                                                             |
+|                |                    |                  |                                                   | QG2-1061-1    |                                                             |
+|                |                    |                  |                                                   | QG2-1064-1    |                                                             |
+|                |                    |                  |                                                   | QG2-7180-1    |                                                             |
+|                |                    |                  |                                                   | QG2-7180-2    |                                                             |
+|                |                    |                  |                                                   | QG1-7246-1    |                                                             |
+|                | This study         | *N. magellanica* | Heshkaia 35                                       | Heshkaia_35-1 | *none*                                                      |
+|                |                    | *N. deaureata*   |                                                   | Heshkaia_35-2 |                                                             |
+|                |                    | *N. deaureata*   |                                                   | Heshkaia_35-3 |                                                             |
+|                |                    | *N. magellanica* |                                                   | Heshkaia_35-4 |                                                             |
+|                |                    | *N. magellanica* |                                                   | Heshkaia_35-5 |                                                             |
+|                |                    | *N. magellanica* |                                                   | Heshkaia_35-6 |                                                             |
+|                |                    | *N. magellanica* |                                                   | Heshkaia_35-7 |                                                             |
+|                |                    | *N. deaureata*   |                                                   | Heshkaia_35-8 |                                                             |
+|                |                    | *N. magellanica* |                                                   | Heshkaia_35-9 |                                                             |
+
+: Table 1 Overview of the modern and archaeological limpet specimens analysed in this study {#tab:specimens}
 
@@
 | Species       | Locality       | Specimen | Correlation R^2^        | Study                             |
-|-------------|-------------|-------------|-------------|----------------------|
-| *Patella depressa* | Northern Spain | LAN541   | 0.87                    | [**10.3390/app11072959**](https://doi.org/10.3390/app11072959)                                                                     |
+|---------------|---------------|---------------|---------------|---------------|
+| *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@garcía-escárzaga2021]           |
 |               |                | LAN545   | 0.86                    |                                   |
@@
 |               |                | LAN559   | 0.82                    |                                   |
-| *Patella caerulea* | Croatia        | ISTPC1   | 0.9                     | 10.1038/s41598-019-39959-9                                                                                                         |
+| *P. caerulea* | Croatia        | ISTPC1   | 0.9                     | [@Hausmann2019-fi]                |
 |               |                | ISTPC2   | 0.84                    |                                   |
@@
 |               |                | KIZPC2   | 0.86                    |                                   |
-| Patella rustica    | Gibraltar      | JL1      | 0.02                    | [doi.org/10.1016/j.epsl.2011.05.054](https://doi.org/10.1016/j.epsl.2011.05.054 "Persistent link using digital object identifier") |
+| *P. rustica*  | Gibraltar      | JL1      | 0.02                    | [@Ferguson2011-zl]                |
 |               |                | JL2      | 0.8 (0.79)              |                                   |
-| Patella caerulea   | Gibraltar      | JM00     | 0.69 (0.79)             |                                                                                                                                    |
+| *P. caerulea* | Gibraltar      | JM00     | 0.69 (0.79)             |                                   |
 |               |                | JM30     | 0.83 (0.79)             |                                   |
-| Patella vulgata    | Orkney         | ORK-LT5  | not reported, here 0.88 | doi.org/10.1016/j.palaeo.2016.10.021 and this study                                                                                |
+| *P. vulgata*  | Orkney         | ORK-LT5  | not reported, here 0.88 | [@Graniero2017-io] and this study |
 |               |                |          |                         |                                   |
@@
 
-: Overview of comparative correlations {#tab:correlations}
+: Table 2: Overview of comparative correlations {#tab:correlations}
 

### 2024-06-23 Manuscript.tex
Intro

--- 2024-05-30 Manuscript.tex
@@
 \begin{abstract}
-This document is a template demonstrating the Aps format.
+This study provides a short reassessment of the use of Magnesium to
+Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea
+surface temperatures. While \emph{Patella vulgata} along the Spanish
+shoreline has since then repeatedly produced reliable correlations
+between sea surface temperature and Mg/Ca ratios, this is not the case
+for other patelloid species. \emph{Patella vulgata} and \emph{Nacella
+deaureata} have been studied using Mg/Ca with mixed or contrary results.
+In this study, we present elemental maps of various such species
+together with stable oxygen isotope values for some of the specimens.
+Our dataset also includes specimens that were previously unsuccessful in
+providing significant correlations between δ\textsuperscript{18}O and
+Mg/Ca ratios. By reassessing these previous specimens and including a
+wider range of modern and archaeological samples from three patelloid
+species (\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N.
+magellanica}) we further add to the growing set of evidence for the
+reliable use of Mg/Ca ratios to detect palaeotemperature change and
+serve as a means to determine ontogenetic age and season of capture as
+well as to reveal locations of interest within the growth record
+(i.e.~annual temperature minima and maxima) for the targeted analysis
+using δ\textsuperscript{18}O or clumped oxygen isotope analysis.
 \end{abstract}
@@
 
-\begin{figure}[H]
+\section{Introduction}\label{Introduction}
 
-{\centering \includegraphics{Manuscript_files/figure-pdf/Raw Data-1.pdf}
+Limpet shells are commonly found within archaeological sites and and
+past shorelines due to their robust carbonate structure and the
+long-term use of limpets as a marine food source. They have successfully
+been studied in the past in the context of coastal subsistence
+economies, site occupation on seasonal
+\citep{shackleton1973, Parker2018-wf, Bosch2018-ud} or long-term scale
+\citep{Ortiz2015-mr}, as well as palaeotemperature
+\citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
+Determining past sea surface temperature (SST) change largely relies on
+the measurement of δ\textsuperscript{18}O-values within the calcitic
+parts of the shell, but attempts have been made to also use elemental
+ratios, such as magnesium to calcium (Mg/Ca), to have an alternative
+measure, that potentially provides a more accurate SST estimate than
+δ\textsuperscript{18}O-values, which are also affected by changes in
+salinity. In addition, the data acquisition of elemental ratios ---
+either through laser-ablation-isotope-ratio-mass-spectrometry
+(LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much
+faster and cost-effective, increasing the number of specimens that can
+be studied overall \citep{Durham2017-fh, Hausmann2023-ih}.
 
-}
+Problematically, links between Mg/Ca ratios and SST changes in limpets
+have also been shown to be unreliable \citep{Graniero2015-zv} and even
+where they were successful, anomalous patterns in some specimens had to
+be filtered out, reducing the overall robustness of the results of those
+successful studies \citep{Ferguson2011-zl}.
 
-\caption{Raw Mg/Ca and δ\textsuperscript{18}O Data}
+While recent research particularly of \emph{Patella} sp. in the
+Mediterranean and Southwest Europe have provided promising results
+\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
+There remains unclarity for Atlantic limpet species, particularly since
+it was last shown here, that they are not reliable recorders of
+palaeotemperature \citep{Graniero2015-zv}.
 
-\end{figure}%
+\section{Materials and Methods}\label{Methods}
+
+\subsection{Limpet specimens}\label{limpet-specimens}
+
+\subsubsection{Modern specimens}\label{modern-specimens}
+
+The analysed specimens, their origin and studies with background
+information can be found in Table 1.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
+\caption{Table 1 Overview of the modern and archaeological limpet
+specimens analysed in this study \{\#tab:specimens\}}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Context
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Study
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Species
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Location
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Sample ID
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Previous analyses
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Context
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Study
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Species
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Location
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Sample ID
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Previous analyses
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Bahía
+Cambaceres Exterior - Tierra del Fuego (AR) & ND-1016-3 & Stable oxygen
+and carbon isotope analysis \\
+& & & & ND-1016-4 & \\
+& & \emph{N. magellanica} & & NM-1016-1 & \\
+& & & & NM-1016-3 & \\
+& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - Westray
+(UK) & ORK-LT5 & Stable oxygen and carbon isotope analysis; Mg, Li, Sr,
+Ca,. \\
+Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
+Bay - Westray (UK) & QG1-7188-1 & Stable oxygen and carbon isotope
+analysis \\
+& & & & QG1-7189-2 & \\
+& & & & QG2-1061-1 & \\
+& & & & QG2-1064-1 & \\
+& & & & QG2-7180-1 & \\
+& & & & QG2-7180-2 & \\
+& & & & QG1-7246-1 & \\
+& This study & \emph{N. magellanica} & Heshkaia 35 & Heshkaia\_35-1 &
+\emph{none} \\
+& & \emph{N. deaureata} & & Heshkaia\_35-2 & \\
+& & \emph{N. deaureata} & & Heshkaia\_35-3 & \\
+& & \emph{N. magellanica} & & Heshkaia\_35-4 & \\
+& & \emph{N. magellanica} & & Heshkaia\_35-5 & \\
+& & \emph{N. magellanica} & & Heshkaia\_35-6 & \\
+& & \emph{N. magellanica} & & Heshkaia\_35-7 & \\
+& & \emph{N. deaureata} & & Heshkaia\_35-8 & \\
+& & \emph{N. magellanica} & & Heshkaia\_35-9 & \\
+\end{longtable}
+
+\subsubsection{Archaeological specimens}\label{archaeological-specimens}
+
+\subsection{Oxygen isotopes}\label{oxygen-isotopes}
+
+\subsection{Mg/Ca ratios}\label{mgca-ratios}
+
+\section{Results}\label{Results}
+
+\subsection{Patella vulgata}\label{patella-vulgata}
+
+\subsection{Nacella deaureata}\label{nacella-deaureata}
+
+\subsection{Nacella magellanica}\label{nacella-magellanica}
+
+\section{}\label{section}
 
@@
 
+\section{Discussion}\label{discussion}
+
+\subsection{Other correlations}\label{other-correlations}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{Manuscript_files/figure-pdf/Ferguson Data-1.pdf}
+
+}
+
+\caption{Correlation graphs for Ferguson et al.~specimens}
+
+\end{figure}%
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
+\caption{Table 2: Overview of comparative correlations
+\{\#tab:correlations\}}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Species
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Locality
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Specimen
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Correlation R\textsuperscript{2}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Study
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Species
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Locality
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Specimen
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Correlation R\textsuperscript{2}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Study
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+\emph{P. depressa} & Northern Spain & LAN541 & 0.87 &
+\citep{garcía-escárzaga2021} \\
+& & LAN545 & 0.86 & \\
+& & LAN554 & 0.78 & \\
+& & LAN559 & 0.82 & \\
+\emph{P. caerulea} & Croatia & ISTPC1 & 0.9 & \citep{Hausmann2019-fi} \\
+& & ISTPC2 & 0.84 & \\
+& Crete & AF1911A & 0.91\footnote{SST only, no other geochemical data
+  available} & \\
+& & AF3003A & 0.92\footnote{SST only, no geochemical data available}
+& \\
+& Israel & AKKPC2 & 0.96 & \\
+& & AKKPC3 & 0.89 & \\
+& & FRMPC1 & 0.84 & \\
+& & FRMPC2 & 0.96 & \\
+& Libya & MO31A & 0.83 & \\
+& & MP64A & 0.33 & \\
+& & MP67A & 0.96 & \\
+& & MP68A & 0.81 & \\
+& Malta & MA10 & 0.82 & \\
+& Tunisia & TUNPC1 & 0.81 & \\
+& & TUNPC2 & 0.78 & \\
+& Turkey & ANTPC1 & 0.95 & \\
+& & ANTPC2 & 0.93 & \\
+& & KIZPC1 & 0.94 & \\
+& & KIZPC2 & 0.86 & \\
+\emph{P. rustica} & Gibraltar & JL1 & 0.02 & \citep{Ferguson2011-zl} \\
+& & JL2 & 0.8 (0.79) & \\
+\emph{P. caerulea} & Gibraltar & JM00 & 0.69 (0.79) & \\
+& & JM30 & 0.83 (0.79) & \\
+\emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
+\citep{Graniero2017-io} and this study \\
+& & & & \\
+& & & & \\
+& & & & \\
+\end{longtable}
+
+\subsection{Comparison of ORK-LT5}\label{comparison-of-ork-lt5}
+
+\includegraphics{Manuscript_files/figure-pdf/unnamed-chunk-1-1.pdf}
+
+\includegraphics{Manuscript_files/figure-pdf/subsample-1.pdf}
+
   \bibliography{bibliography.bib}

### 2024-06-23 Manuscript.qmd
finished intro

@@
 There remains unclarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].
+
+In this study we will repeat and expand the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies.
+To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS.
+LIBS allows us to carry out 2D imaging of entire shell sections, which helps us to navigate the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.
+
+Relying and adding onto published datasets, allowed us to simultaneously avoid costs for new high-resolution δ^18^O-data, use real-world examples, and also provide pilot-data for areas of existing research interest.
+Generally establishing the usefulness and reliabilty of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.
 

### 2024-06-23 Manuscript.tex
finished intro

@@
 
+In this study we will repeat and expand the analysis of Atlantic limpets
+to determine the reliability of Mg/Ca ratios as palaeotemperature
+proxies. To do this, we sampled a set of previously published and
+unpublished limpet specimens dating to modern and archaeological
+contexts using LIBS. LIBS allows us to carry out 2D imaging of entire
+shell sections, which helps us to navigate the complex elemental
+structure of the shell and better separate the external,
+temperature-related changes from the internal and less understood
+factors that influence the Mg/Ca ratio.
+
+Relying and adding onto published datasets, allowed us to simultaneously
+avoid costs for new high-resolution δ\textsuperscript{18}O-data, use
+real-world examples, and also provide pilot-data for areas of existing
+research interest. Generally establishing the usefulness and reliabilty
+of Mg/Ca as SST proxy in limpets should help to provide a platform for
+future research and an important stepping stone to better understand
+elemental ratios in other marine mollusc shells.
+
 \section{Materials and Methods}\label{Methods}

### 2024-06-23 Manuscript.qmd
added temp data again

@@
 
-The analysed specimens, their origin and studies with background information can be found in Table 1.
+The analysed specimens, their origin and respective studies with research background information can be found in Table 1.
+Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io].
+While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.
+
+There are two main locations that we sourced specimens from: the island of West Ray in Orkney (UK) and the Beagle Channel in Tierra del Fuego (Argentina).
+A modern shell on West Ray was collected in Rack Wick Bay in August 2009, which experiences virtually no freshwater input.
+The temperature
 
@@
 |------------|------------|------------|------------|------------|------------|
-| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Bahía Cambaceres Exterior - Tierra del Fuego (AR) | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
+| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay - Tierra del Fuego (AR)          | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
 |                |                    |                  |                                                 | ND-1016-4     |                                                             |
@@
 |                |                    |                  |                                                 | QG1-7246-1    |                                                             |
-|                | This study         | *N. magellanica* | Heshkaia 35                                       | Heshkaia_35-1 | *none*                                                      |
+|                | This study         | *N. magellanica* | Heshkaia 35 in Moat Bay - Tierra del Fuego (AR) | Heshkaia_35-1 | *none*                                                      |
 |                |                    | *N. deaureata*   |                                                 | Heshkaia_35-2 |                                                             |
@@
 : Table 1 Overview of the modern and archaeological limpet specimens analysed in this study {#tab:specimens}
+
+```{r SST Data}
+#| fig-cap: SST values averaged over the last 20 years. 
+#| fig-height: 10
+#| fig-width: 10
+
+# Read the CSV files, skipping the comments
+df1 <- read_csv(here("data","Westray-SST.csv"), comment = "#")
+df2 <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")
+
+# Combine the datasets with new names and add year and month columns
+combined_df <- bind_rows(df1 %>% mutate(dataset = "Westray"),
+                         df2 %>% mutate(dataset = "Tierra del Fuego")) %>%
+  mutate(analysed_sst = analysed_sst - 273.15,
+         time = as_datetime(time),
+         year = year(time),
+         month = month(time, label = TRUE)) %>%
+  group_by(dataset, year, month) %>%
+  summarise(mean_sst = mean(analysed_sst, na.rm = TRUE),
+            sd_sst = sd(mean_sst, na.rm = TRUE))
+
+# Plot the data
+combined_df %>% 
+    group_by(dataset, month) %>%
+  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
+            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
+ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
+  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
+  geom_path() +
+  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
+scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
+  labs(title = "Monthly Sea Surface Temperature",
+       x = "Year-Month",
+       y = "Mean Analysed SST (°C)") +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1),
+    legend.position = c(0.1, 0.9))+  
+  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) +
+  labs(colour="",fill="")
+
+```
 

### 2024-06-23 Manuscript.tex
added citation

@@
 
-The analysed specimens, their origin and studies with background
-information can be found in Table 1.
+The analysed specimens, their origin and respective studies with
+research background information can be found in Table 1. Here we will
+briefly summarise their contextual information, which can be accessed in
+more detail at the respective studies
+\citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While those
+studies also included other specimens, their accessibility or state of
+preservation did not all lend them to be re-analysed.
+
+There are two main locations that we sourced specimens from: the island
+of West Ray in Orkney (UK) and the Beagle Channel in Tierra del Fuego
+(Argentina). A modern shell on West Ray was collected in Rack Wick Bay
+in August 2009, which experiences virtually no freshwater input. The
+temperature
 
@@
 \endlastfoot
-Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Bahía
-Cambaceres Exterior - Tierra del Fuego (AR) & ND-1016-3 & Stable oxygen
-and carbon isotope analysis \\
+Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
+- Tierra del Fuego (AR) & ND-1016-3 & Stable oxygen and carbon isotope
+analysis \\
 & & & & ND-1016-4 & \\
@@
 & & & & QG1-7246-1 & \\
-& This study & \emph{N. magellanica} & Heshkaia 35 & Heshkaia\_35-1 &
-\emph{none} \\
+& This study & \emph{N. magellanica} & Heshkaia 35 in Moat Bay - Tierra
+del Fuego (AR) & Heshkaia\_35-1 & \emph{none} \\
 & & \emph{N. deaureata} & & Heshkaia\_35-2 & \\
@@
 \end{longtable}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{Manuscript_files/figure-pdf/SST Data-1.pdf}
+
+}
+
+\caption{SST values averaged over the last 20 years. SST values were
+sourced from \citep{europeanunion-copernicusmarineservice2015}}
+
+\end{figure}%
 

### 2024-06-23 Manuscript.qmd
added citation
Modern shells

--- 2024-06-23 Manuscript.qmd
@@
 
-The analysed specimens, their origin and studies with background information can be found in Table 1.
+The analysed specimens, their origin and respective studies with research background information can be found in [Table 1](#Table_1).
+Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io].
+While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.
 
+::: {#Table_1}
 | Context        | Study              | Species          | Location                                        | Sample ID     | Previous analyses                                           |
 |------------|------------|------------|------------|------------|------------|
-| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Bahía Cambaceres Exterior - Tierra del Fuego (AR) | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
+| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay - Tierra del Fuego (AR)          | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
 |                |                    |                  |                                                 | ND-1016-4     |                                                             |
@@
 
-: Table 1 Overview of the modern and archaeological limpet specimens analysed in this study {#tab:specimens}
+: Overview of the modern and archaeological limpet specimens analysed in this study.
+:::
+
+```{r SST Data}
+
+# Read the CSV files, skipping the comments
+Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
+TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")
+
+# Combine the datasets with new names and add year and month columns
+SST_df <- bind_rows(Westray_SST %>% mutate(dataset = "Westray - Orkney"),
+                         TdF_SST %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
+  mutate(analysed_sst = analysed_sst - 273.15,
+         time = as_datetime(time),
+         year = year(time),
+         month = month(time, label = TRUE)) %>%
+  group_by(dataset, year, month) %>%
+  summarise(mean_sst = mean(analysed_sst, na.rm = TRUE),
+            sd_sst = sd(analysed_sst, na.rm = TRUE))
+
+WestrayMin <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
+  pull(mean_sst) %>% min() %>% round(1)
+WestrayMax <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
+  pull(mean_sst) %>% max() %>% round(1)
+
+CamBayMin <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
+  pull(mean_sst) %>% min() %>% round(1)
+CamBayMax <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
+  pull(mean_sst) %>% max() %>% round(1)
+
+```
+
+There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK).
+Four specimens were collected from the Beagle Channel in Cambaceres Bay.
+The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih].
+SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.
+
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009.
+Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu.
+The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).
+
+```{r SST plot}
+#| fig-cap: SST values averaged from  years. SST values were sourced from [@europeanunion-copernicusmarineservice2015]
+#| fig-height: 10
+#| fig-width: 10
+#| label: fig-SSTs
+
+# Plot the data
+SST_df %>% 
+    group_by(dataset, month) %>%
+  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
+            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
+ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
+  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
+  geom_path() +
+  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
+scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
+  labs(title = "Monthly Sea Surface Temperature",
+       x = "Month",
+       y = "Mean Analysed SST (°C)") +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1),
+    legend.position = c(0.1, 0.9))+  
+  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) +
+  labs(colour="",fill="")
+
+```
 

### 2024-06-23 Manuscript.tex
Modern shells

@@
 The analysed specimens, their origin and respective studies with
-research background information can be found in Table 1. Here we will
-briefly summarise their contextual information, which can be accessed in
-more detail at the respective studies
+research background information can be found in \hyperref[Table_1]{Table
+1}. Here we will briefly summarise their contextual information, which
+can be accessed in more detail at the respective studies
 \citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While those
@@
 
-There are two main locations that we sourced specimens from: the island
-of West Ray in Orkney (UK) and the Beagle Channel in Tierra del Fuego
-(Argentina). A modern shell on West Ray was collected in Rack Wick Bay
-in August 2009, which experiences virtually no freshwater input. The
-temperature
-
+\phantomsection\label{Table_1}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
-\caption{Table 1 Overview of the modern and archaeological limpet
-specimens analysed in this study \{\#tab:specimens\}}\tabularnewline
+\caption{Overview of the modern and archaeological limpet specimens
+analysed in this study.}\tabularnewline
 \toprule\noalign{}
@@
 
-\begin{figure}[H]
+There are two main locations that we sourced specimens from: the Beagle
+Channel in Tierra del Fuego (Argentina) and the island of West Ray in
+Orkney (UK). Four specimens were collected from the Beagle Channel in
+Cambaceres Bay. The area receives around 570 mm precipitation but the
+channel waters are predominantly influenced by marine currents stayed
+around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
+August to 10ºC in February.
 
-{\centering \includegraphics{Manuscript_files/figure-pdf/SST Data-1.pdf}
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
+August 2009. Westray lies about 70 km north of the Scottish mainland and
+experiences virtually no freshwater input. The annual SST ranges from
+6.3ºC in March to 13.8ºC in August (Figure~\ref{fig-SSTs}).
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-SSTs-1.pdf}
 
@@
 
-\caption{SST values averaged over the last 20 years. SST values were
+\caption{\label{fig-SSTs}SST values averaged from years. SST values were
 sourced from \citep{europeanunion-copernicusmarineservice2015}}

### 2024-06-23 Manuscript.qmd
added modern isotope data

@@
 
-D18O <- read_csv(here("data","D18O.csv"))
+D18O <- read_csv(here("data","D18O.csv")) %>% 
+  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern"))
+
 MgCa <- read_csv(here("data","MgCa.csv"))
@@
 On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009.
-Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu.
+Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
 The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).
@@
 #| fig-cap: SST values averaged from  years. SST values were sourced from [@europeanunion-copernicusmarineservice2015]
-#| fig-height: 10
-#| fig-width: 10
+#| fig-height: 5
+#| fig-width: 5
 #| label: fig-SSTs
@@
 
+All shells show multi-year quasi-sinusoidal records of SST change (@fig-Mod_iso), with *Nacella* sp.
+shells living 2–3 years and *Patella v*.
+over 5 years.
+The δ^18^O values of the Orcadian shell have not been directly compared to the local SST, but accurately reflect the seasonal changes expected with one large anomaly in the second summer of recorded growth.
+<!--# ADD FIGURE TO THESE ISOTOPE VALUES -->
+
+```{r Modern Isotope data}
+#| fig-cap: Isotope values of modern limpet shells from Cambaceres Bay (*a-f*) and Westray (*g*).
+#| fig-height: 10
+#| fig-width: 10
+#| label: fig-Mod_iso
+# Filter data for Modern period
+modern_data <- D18O %>% filter(period == "Modern")
+
+# Get unique sample IDs
+sample_ids <- unique(modern_data$sampleID)
+
+# Initialise an empty list to store plots
+plots <- list()
+
+# Loop over each sampleID to create individual plots
+for (i in seq_along(sample_ids)) {
+  sample_id <- sample_ids[i]
+  plot <- modern_data %>%
+    filter(sampleID == sample_id) %>%
+    ggplot(aes(dist, d18O)) +
+    geom_path() +
+    geom_point() +
+    xlim(0,15)+
+    labs(
+      x = "Distance from margin in mm",
+      y = expression(delta^18*O~("\u2030")),
+      subtitle = sample_id,
+      col = "",
+      fill = ""
+    ) +
+    theme(plot.subtitle = element_text(hjust = 0.5))
+
+  # Store plot in the list
+  plots[[i]] <- plot
+}
+
+# Combine plots using patchwork and add letters
+wrap_plots(plots,nrow=4) + plot_annotation(tag_levels = 'a')
+
+```
+
 ### Archaeological specimens
+
+In total, we studied 16 archaeological specimens.
+Of those, seven come from Quoygrew on Westray dating to 900 – 1200 CE.
+These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba].
+and nine from the site Hesk
 

### 2024-06-23 Manuscript.tex
added modern isotope data

@@
 Cambaceres Bay. The area receives around 570 mm precipitation but the
-channel waters are predominantly influenced by marine currents stayed
-around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
-August to 10ºC in February.
+channel waters are predominantly influenced by marine currents and in
+the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu
+\citep{Nicastro2020-ih}. SSTs range from 4.7ºC in August to 10ºC in
+February.
 
@@
 August 2009. Westray lies about 70 km north of the Scottish mainland and
-experiences virtually no freshwater input. The annual SST ranges from
-6.3ºC in March to 13.8ºC in August (Figure~\ref{fig-SSTs}).
+experiences virtually no freshwater input with salinity values of 34.0
+-- 34.5 psu \citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in
+March to 13.8ºC in August (Figure~\ref{fig-SSTs}).
 
@@
 
+All shells show multi-year quasi-sinusoidal records of SST change
+(Figure~\ref{fig-Mod_iso}), with \emph{Nacella} sp. shells living 2--3
+years and \emph{Patella v}. over 5 years. The δ\textsuperscript{18}O
+values of the Orcadian shell have not been directly compared to the
+local SST, but accurately reflect the seasonal changes expected with one
+large anomaly in the second summer of recorded growth.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Mod_iso-1.pdf}
+
+}
+
+\caption{\label{fig-Mod_iso}Isotope values of modern limpet shells from
+Cambaceres Bay (\emph{a-f}) and Westray (\emph{g}).}
+
+\end{figure}%
+
 \subsubsection{Archaeological specimens}\label{archaeological-specimens}
+
+In total, we studied 16 archaeological specimens. Of those, seven come
+from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
+previously been studied to characterise seasonal temperature change
+during the Medieval Climate Anomaly and the Little Ice Age
+\citep{Surge2012-ba}. and nine from the site Hesk
 

### 2024-06-23 Manuscript.qmd
Added info

@@
 There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK).
-Four specimens were collected from the Beagle Channel in Cambaceres Bay.
+Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016.
 The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih].
@@
 
-All shells show multi-year quasi-sinusoidal records of SST change (@fig-Mod_iso), with *Nacella* sp.
+### Archaeological specimens
+
+In total, we studied 16 archaeological specimens.
+Of those, seven come from Quoygrew on Westray dating to 900 – 1200 CE.
+These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba].
+Together they show a range of lifespans, with some comparatively long for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to time-averaging.
+That said, the majority are under 5 years old.
+These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].
+
+Nine additional shells have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp.
+shells.
+These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells.
+
+## Methods {#methods}
+
+### Oxygen isotopes
+
+High-quality oxygen isotope values (δ^18^O) were acquired from previous publications on *P.
+vulgata* [@Surge2012-ba; @Graniero2017-io], *N.
+deaureata* and *N.
+magellanica* [@Nicastro2020-ih], where detailed descriptions of the stable isotope analysis methodology can be found.
+In short, all shells were sectioned along the main growth axis to expose the internal growth structure.
+From the exposed section carbonate powder samples were acquired using a micromilling approach with resolutions of 100–200 µm and targeting the calcitic M+2 layer of the shells.
+Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.
+
+### Mg/Ca ratios
+
+Mg/Ca elemental imaging was carried out using Laser Induced Breakdown Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany) using previously published methods [@Hausmann2023-ih].
+These involve the ablation of the shell section using an infrared (1064 nm) laser pulse (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume.
+This plume emits light, which is measured using a synchronised spectrometer.
+The resulting light spectrum quantifies the emission lines of magnesium () and calcium () to determine their intensity ratio.
+While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g.
+[@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa].
+
+# Results {#Results}
+
+### Modern Specimens
+
+```{r Modern iso ranges}
+# Ranges <- D18O %>% 
+#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
+# group_by(sampleID,period) %>% 
+#   mutate(range=max(d18O)-min(d18O),
+#          range=round(range,2)) %>% 
+#   select(period, sampleID,range) %>% unique()
+# 
+# Modern_Ranges <- Ranges %>% 
+#   filter(period=="Modern") %>% 
+#   arrange(range)
+
+#Interestingly, the overall annual range in δ18O-values on Orkney (~r Modern_Ranges$range[3]‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp.
+shells (r min(Modern_Ranges$range)‰ to r max(Modern_Ranges$range)‰) despite the different SST ranges.
+
+```
+
+All modern shells show multi-year quasi-sinusoidal records of SST change (@fig-Mod_iso), with *Nacella* sp.
 shells living 2–3 years and *Patella v*.
 over 5 years.
-The δ^18^O values of the Orcadian shell have not been directly compared to the local SST, but accurately reflect the seasonal changes expected with one large anomaly in the second summer of recorded growth.
-<!--# ADD FIGURE TO THESE ISOTOPE VALUES -->
+The δ^18^O values of the Orcadian shell have not been directly compared to the local SST, but accurately reflect the seasonal changes expected barring one extreme, anomalous value in the second summer of recorded growth.
 
-```{r Modern Isotope data}
+```{r Modern Isotope plot}
 #| fig-cap: Isotope values of modern limpet shells from Cambaceres Bay (*a-f*) and Westray (*g*).
@@
 
-### Archaeological specimens
+### Archaeological Specimens
 
-In total, we studied 16 archaeological specimens.
-Of those, seven come from Quoygrew on Westray dating to 900 – 1200 CE.
-These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba].
-and nine from the site Hesk
+```{r Archaeological Isotope plot}
+#| fig-cap: Isotope values of Archaeological limpet shells from Quoygrew
+#| fig-height: 10
+#| fig-width: 10
+#| label: fig-Arch_iso
+# Filter data for Modern period
+arch_data <- D18O %>% filter(period == "Archaeological")
 
-## Oxygen isotopes
+# Get unique sample IDs
+sample_ids <- unique(arch_data$sampleID)
 
-## Mg/Ca ratios
+# Initialise an empty list to store plots
+plots <- list()
 
-# Results {#Results}
+# Loop over each sampleID to create individual plots
+for (i in seq_along(sample_ids)) {
+  sample_id <- sample_ids[i]
+  plot <- arch_data %>%
+    filter(sampleID == sample_id) %>%
+    ggplot(aes(dist, d18O)) +
+    geom_path() +
+    geom_point() +
+    # xlim(0,20)+
+    labs(
+      x = "Distance from margin in mm",
+      y = expression(delta^18*O~("\u2030")),
+      subtitle = sample_id,
+      col = "",
+      fill = ""
+    ) +
+    theme(plot.subtitle = element_text(hjust = 0.5))
+
+  # Store plot in the list
+  plots[[i]] <- plot
+}
+
+# Combine plots using patchwork and add letters
+wrap_plots(plots,nrow=4) + plot_annotation(tag_levels = 'a')
+
+```
 

### 2024-06-23 Manuscript.tex
Added info

@@
 Orkney (UK). Four specimens were collected from the Beagle Channel in
-Cambaceres Bay. The area receives around 570 mm precipitation but the
-channel waters are predominantly influenced by marine currents and in
-the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu
-\citep{Nicastro2020-ih}. SSTs range from 4.7ºC in August to 10ºC in
-February.
+Cambaceres Bay in October 2016. The area receives around 570 mm
+precipitation but the channel waters are predominantly influenced by
+marine currents and in the period of October 2015 to October 2016 stayed
+around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
+August to 10ºC in February.
 
@@
 
-All shells show multi-year quasi-sinusoidal records of SST change
+\subsubsection{Archaeological specimens}\label{archaeological-specimens}
+
+In total, we studied 16 archaeological specimens. Of those, seven come
+from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
+previously been studied to characterise seasonal temperature change
+during the Medieval Climate Anomaly and the Little Ice Age
+\citep{Surge2012-ba}. Together they show a range of lifespans, with some
+comparatively long for their size (QG1-7246-1, 12 years within 16 mm of
+growth) making them particularly susceptible to time-averaging. That
+said, the majority are under 5 years old. These shells are an ideal
+comparative dataset from an archaeological assemblage to compare to the
+modern shell ORK\_LT5 \citep{Graniero2017-io}.
+
+Nine additional shells have been selected for LIBS analysis only, to
+provide a comparative dataset to the modern \emph{Nacella} sp. shells.
+These have not been analysed for δ\textsuperscript{18}O-values, but
+rather as a test to see whether patterns visible in the modern Mg/Ca
+ratio distributions are also preserved and apparent in the
+archaeological shells.
+
+\subsection{Methods}\label{methods}
+
+\subsubsection{Oxygen isotopes}\label{oxygen-isotopes}
+
+High-quality oxygen isotope values (δ\textsuperscript{18}O) were
+acquired from previous publications on \emph{P. vulgata}
+\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
+magellanica} \citep{Nicastro2020-ih}, where detailed descriptions of the
+stable isotope analysis methodology can be found. In short, all shells
+were sectioned along the main growth axis to expose the internal growth
+structure. From the exposed section carbonate powder samples were
+acquired using a micromilling approach with resolutions of 100--200 µm
+and targeting the calcitic M+2 layer of the shells. Analytical precision
+of the isotope values was consistently at 0.05--0.10‰ and values are
+reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee
+Belemnite) standard.
+
+\subsubsection{Mg/Ca ratios}\label{mgca-ratios}
+
+Mg/Ca elemental imaging was carried out using Laser Induced Breakdown
+Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany)
+using previously published methods \citep{Hausmann2023-ih}. These
+involve the ablation of the shell section using an infrared (1064 nm)
+laser pulse (1--2 mJ, 100 Hz) onto a surface area of 20--30 µm to
+generate a plasma plume. This plume emits light, which is measured using
+a synchronised spectrometer. The resulting light spectrum quantifies the
+emission lines of magnesium () and calcium () to determine their
+intensity ratio. While these two peaks alone do not represent the molar
+concentrations of both elements (as opposed to calibration free LIBS;
+e.g. \citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
+correlated to the molar concentration \citep{Hausmann2017-oa}.
+
+\section{Results}\label{Results}
+
+\subsubsection{Modern Specimens}\label{modern-specimens-1}
+
+All modern shells show multi-year quasi-sinusoidal records of SST change
 (Figure~\ref{fig-Mod_iso}), with \emph{Nacella} sp. shells living 2--3
@@
 values of the Orcadian shell have not been directly compared to the
-local SST, but accurately reflect the seasonal changes expected with one
-large anomaly in the second summer of recorded growth.
+local SST, but accurately reflect the seasonal changes expected barring
+one extreme, anomalous value in the second summer of recorded growth.
 
@@
 
-\subsubsection{Archaeological specimens}\label{archaeological-specimens}
+\subsubsection{Archaeological
+Specimens}\label{archaeological-specimens-1}
 
-In total, we studied 16 archaeological specimens. Of those, seven come
-from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
-previously been studied to characterise seasonal temperature change
-during the Medieval Climate Anomaly and the Little Ice Age
-\citep{Surge2012-ba}. and nine from the site Hesk
+\begin{figure}
 
-\subsection{Oxygen isotopes}\label{oxygen-isotopes}
+\centering{
 
-\subsection{Mg/Ca ratios}\label{mgca-ratios}
+\includegraphics{Manuscript_files/figure-pdf/fig-Arch_iso-1.pdf}
 
-\section{Results}\label{Results}
+}
+
+\caption{\label{fig-Arch_iso}Isotope values of Archaeological limpet
+shells from Quoygrew}
+
+\end{figure}%
 

### 2024-06-23 Manuscript.qmd
bib clean

@@
 Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source.
-They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@shackleton1973; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
+They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
 Determining past sea surface temperature (SST) change largely relies on the measurement of δ^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than δ^18^O-values, which are also affected by changes in salinity.
@@
 |---------------|---------------|---------------|---------------|---------------|
-| *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@garcía-escárzaga2021]           |
+| *P. depressa* | Northern Spain | LAN541   | 0.87                    |                                   |
 |               |                | LAN545   | 0.86                    |                                   |

### 2024-06-23 Manuscript.qmd
fixed bib

@@
 ```{r SST plot}
-#| fig-cap: SST values averaged from  years. SST values were sourced from [@europeanunion-copernicusmarineservice2015]
+#| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
 #| fig-height: 5
@@
 |---------------|---------------|---------------|---------------|---------------|
-| *P. depressa* | Northern Spain | LAN541   | 0.87                    |                                   |
+| *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
 |               |                | LAN545   | 0.86                    |                                   |

### 2024-06-23 Manuscript.tex
fixed bib

@@
 economies, site occupation on seasonal
-\citep{shackleton1973, Parker2018-wf, Bosch2018-ud} or long-term scale
-\citep{Ortiz2015-mr}, as well as palaeotemperature
+\citep{Shackleton1973-ij, Parker2018-wf, Bosch2018-ud} or long-term
+scale \citep{Ortiz2015-mr}, as well as palaeotemperature
 \citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
@@
 \caption{\label{fig-SSTs}SST values averaged from years. SST values were
-sourced from \citep{europeanunion-copernicusmarineservice2015}}
+sourced from \citet{Good2020-nl}}
 
@@
 \emph{P. depressa} & Northern Spain & LAN541 & 0.87 &
-\citep{garcía-escárzaga2021} \\
+\citep{Garcia-Escarzaga2021-ij} \\
 & & LAN545 & 0.86 & \\

### 2024-06-24 Manuscript.qmd
added elemental background

@@
 bibliography: bibliography.bib
-format:
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Palaeo^3^
-      formatting: preprint
-      model: 3p
-      cite-style: authoryear
+format: html
+
+  # elsevier-pdf:
+  #   keep-tex: true
+  #   journal:
+  #     name: Palaeo^3^
+  #     formatting: preprint
+  #     model: 3p
+  #     cite-style: authoryear
 output:
@@
 
-Problematically, links between Mg/Ca ratios and SST changes in limpets have also been shown to be unreliable [@Graniero2015-zv] and even where they were successful, anomalous patterns in some specimens had to be filtered out, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].
+Problematically, links between Mg/Ca ratios and SST changes in many marine mollusc shells, including limpets, have shown to be unreliable [@Surge2008-ri; @Wanamaker2008-zl; @Schone2010-yl; @Freitas2012-tx; @Graniero2015-zv; @Poulain2015-dg; @Vihtakari2017-wd].
+This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as temperature proxy.
+Confusingly, multiple different temperature equations have been found for the same species (see [@Freitas2012-tx; @Vihtakari2017-wd] and references therein).
+Coeval specimens sharing one locality can also show differences in their relation to SST [@Hausmann2019-fi].
+Where the use of Mg/Ca as a palaeotemperature proxy was successful, anomalous patterns in some specimens had still to be filtered out by hand, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].
 
@@
 in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].
-There remains unclarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].
+There remains a lack of clarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].
 
@@
 
-Relying and adding onto published datasets, allowed us to simultaneously avoid costs for new high-resolution δ^18^O-data, use real-world examples, and also provide pilot-data for areas of existing research interest.
-Generally establishing the usefulness and reliabilty of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.
+By relying and adding onto published datasets, we were able to simultaneously avoid costs for new high-resolution δ^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
+Generally establishing the usefulness and reliability of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.
 
@@
 In total, we studied 16 archaeological specimens.
-Of those, seven come from Quoygrew on Westray dating to 900 – 1200 CE.
+Of those, seven *P.
+vulgata* come from Quoygrew on Westray dating to 900 – 1200 CE.
 These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba].
@@
 
-Nine additional shells have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp.
+Nine additional shells of *N.
+magellanica* (n=6) and *N*.
+*deaureata* (n=3) have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp.
 shells.
 These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells.
+They derive from the site of Heshkaia 35
 

### 2024-06-24 Manuscript.tex
added elemental background

@@
 
-Problematically, links between Mg/Ca ratios and SST changes in limpets
-have also been shown to be unreliable \citep{Graniero2015-zv} and even
-where they were successful, anomalous patterns in some specimens had to
-be filtered out, reducing the overall robustness of the results of those
-successful studies \citep{Ferguson2011-zl}.
+Problematically, links between Mg/Ca ratios and SST changes in many
+marine mollusc shells, including limpets, have shown to be unreliable
+\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2015-zv, Poulain2015-dg, Vihtakari2017-wd}.
+This is particularly the case where there is little available additional
+information on metabolic processes, organic components of the shell
+matrix, intra-increment and intra-shell variability, and growth rates,
+which can independently and unpredictably affect Mg/Ca ratios and
+confound their interpretation as temperature proxy. Confoundingly,
+multiple different temperature equations have been found for the same
+species (see {[}\citet{Freitas2012-tx}; @Vihtakari2017-wd{]} and
+references therein). Coeval specimens sharing one locality can also show
+differences in their relation to SST \citep{Hausmann2019-fi}. Where the
+use of Mg/Ca as a palaeotemperature proxy was successful, anomalous
+patterns in some specimens had still to be filtered out by hand,
+reducing the overall robustness of the results of those successful
+studies \citep{Ferguson2011-zl}.
 

### 2024-06-24 Manuscript.qmd
added iso

@@
 bibliography: bibliography.bib
-format: html
-
-  # elsevier-pdf:
-  #   keep-tex: true
-  #   journal:
-  #     name: Palaeo^3^
-  #     formatting: preprint
-  #     model: 3p
-  #     cite-style: authoryear
+format:
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Palaeo^3^
+      formatting: preprint
+      model: 3p
+      cite-style: authoryear
 output:
@@
 D18O <- read_csv(here("data","D18O.csv")) %>% 
-  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern"))
+  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern")) %>% 
+  mutate(species = case_when(
+      startsWith(sampleID, "ND") ~ "Nacella deaureata",
+      startsWith(sampleID, "NM") ~ "Nacella magellanica",
+      TRUE ~ "Patella vulgata"
+    ))
 
@@
 These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells.
-They derive from the site of Heshkaia 35
+They derive from the site of Heshkaia 35 situated within Moat Bay, around 30 km east of where the modern specimens in our study were collected [@Zangrando2014-qi; @Zangrando2021-qg].
+With radiocarbon dates putting the shells into the Late Holocene (1300 – 1450 CE) they are of similar age — if somewhat younger — as the assemblage from Quoygrew.
 
 ## Methods {#methods}
+
+### Mg/Ca ratios
+
+Mg/Ca elemental imaging was carried out using Laser Induced Breakdown Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany) using previously published methods [@Hausmann2023-ih].
+These involve the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume.
+This plume emits light, which is measured using a synchronised spectrometer.
+The resulting light spectrum quantifies the emission lines of magnesium (MgII; 279.553 nm) and calcium (CaII; 315.887 nm) to determine their intensity ratio.
+While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g.
+[@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca changes within the shell carbonate.
 
@@
 deaureata* and *N.
-magellanica* [@Nicastro2020-ih], where detailed descriptions of the stable isotope analysis methodology can be found.
+magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found.
 In short, all shells were sectioned along the main growth axis to expose the internal growth structure.
-From the exposed section carbonate powder samples were acquired using a micromilling approach with resolutions of 100–200 µm and targeting the calcitic M+2 layer of the shells.
+From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells.
 Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.
 
-### Mg/Ca ratios
+@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential δ^18^O-values of limpet shells by species.
+Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp.
+shells living 2–3 years and *Patella v*.
+regularly over 3 years.
+In those instances where longer lifetimes are encountered (i.e.
+specimen QG1-7246-1, @fig-Pat_iso b).
+Interestingly, the ranges in δ^18^O-values are not similar between the coeval *Nacella* sp.
+specimens with δ^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 3.0‰ and 2.0‰.
+*P.
+vulgata* has similarly variable δ^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.
 
-Mg/Ca elemental imaging was carried out using Laser Induced Breakdown Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany) using previously published methods [@Hausmann2023-ih].
-These involve the ablation of the shell section using an infrared (1064 nm) laser pulse (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume.
-This plume emits light, which is measured using a synchronised spectrometer.
-The resulting light spectrum quantifies the emission lines of magnesium () and calcium () to determine their intensity ratio.
-While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g.
-[@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa].
-
-# Results {#Results}
-
-### Modern Specimens
-
-```{r Modern iso ranges}
-# Ranges <- D18O %>% 
-#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
-# group_by(sampleID,period) %>% 
-#   mutate(range=max(d18O)-min(d18O),
-#          range=round(range,2)) %>% 
-#   select(period, sampleID,range) %>% unique()
-# 
-# Modern_Ranges <- Ranges %>% 
-#   filter(period=="Modern") %>% 
-#   arrange(range)
-
-#Interestingly, the overall annual range in δ18O-values on Orkney (~r Modern_Ranges$range[3]‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp.
-shells (r min(Modern_Ranges$range)‰ to r max(Modern_Ranges$range)‰) despite the different SST ranges.
-
-```
-
-All modern shells show multi-year quasi-sinusoidal records of SST change (@fig-Mod_iso), with *Nacella* sp.
-shells living 2–3 years and *Patella v*.
-over 5 years.
-The δ^18^O values of the Orcadian shell have not been directly compared to the local SST, but accurately reflect the seasonal changes expected barring one extreme, anomalous value in the second summer of recorded growth.
-
-```{r Modern Isotope plot}
-#| fig-cap: Isotope values of modern limpet shells from Cambaceres Bay (*a-f*) and Westray (*g*).
+```{r Nacella_Isotope_plot}
+#| fig-cap: Isotope values of Nacella sp. limpet shells from Cambaceres Bay (*a-f*).
 #| fig-height: 10
 #| fig-width: 10
-#| label: fig-Mod_iso
+#| label: fig-Nac_iso
 # Filter data for Modern period
-modern_data <- D18O %>% filter(period == "Modern")
+Nacella_iso <- D18O %>% filter(grepl("Nacella", species))
 
 # Get unique sample IDs
-sample_ids <- unique(modern_data$sampleID)
+sample_ids <- unique(Nacella_iso$sampleID)
 
@@
   sample_id <- sample_ids[i]
-  plot <- modern_data %>%
+  plot <- Nacella_iso %>%
     filter(sampleID == sample_id) %>%
@@
     geom_path() +
-    geom_point() +
-    xlim(0,15)+
-    labs(
+    geom_point() +    labs(
       x = "Distance from margin in mm",
@@
 # Combine plots using patchwork and add letters
-wrap_plots(plots,nrow=4) + plot_annotation(tag_levels = 'a')
+wrap_plots(plots,nrow=3) + plot_annotation(tag_levels = 'a')
 
@@
 
-### Archaeological Specimens
-
-```{r Archaeological Isotope plot}
-#| fig-cap: Isotope values of Archaeological limpet shells from Quoygrew
+```{r Patella_Isotope_plot}
+#| fig-cap: Isotope values of Patella vulagata limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
 #| fig-height: 10
 #| fig-width: 10
-#| label: fig-Arch_iso
+#| label: fig-Pat_iso
 # Filter data for Modern period
-arch_data <- D18O %>% filter(period == "Archaeological")
+Patella_iso <- D18O %>% filter(grepl("Patella", species))
 
 # Get unique sample IDs
-sample_ids <- unique(arch_data$sampleID)
+sample_ids <- unique(Patella_iso$sampleID)
 
@@
   sample_id <- sample_ids[i]
-  plot <- arch_data %>%
+  plot <- Patella_iso %>%
     filter(sampleID == sample_id) %>%
@@
 
-## Patella vulgata
+# Results {#Results}
 
-## Nacella deaureata
+## *Patella vulgata*
 
-## Nacella magellanica
+## *Nacella deaureata*
+
+## *Nacella magellanica*
 
@@
 <!--# Maybe get the actual data from the image itself. -->
+
+# References
+```{r Modern iso ranges}
+# Ranges <- D18O %>%
+#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
+# group_by(sampleID,period) %>%
+#   mutate(max=max(d18O),
+#          min=min(d18O)) %>%
+#   select(min, sampleID,max) %>% unique()
+# 
+# Modern_Ranges <- Ranges %>%
+#   filter(period=="Modern") %>%
+#   arrange(range)
+
+# Interestingly, the overall annual range in δ^18^O-values on Orkney (\~`r Modern_Ranges$range[3]`‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp.
+shells (`r min(Modern_Ranges$range)`‰ to `r max(Modern_Ranges$range)`‰) despite the different SST ranges.
+
+```

### 2024-06-24 Manuscript.tex
added iso

@@
 which can independently and unpredictably affect Mg/Ca ratios and
-confound their interpretation as temperature proxy. Confoundingly,
+confound their interpretation as temperature proxy. Confusingly,
 multiple different temperature equations have been found for the same
-species (see {[}\citet{Freitas2012-tx}; @Vihtakari2017-wd{]} and
-references therein). Coeval specimens sharing one locality can also show
+species (see \citep{Freitas2012-tx, Vihtakari2017-wd} and references
+therein). Coeval specimens sharing one locality can also show
 differences in their relation to SST \citep{Hausmann2019-fi}. Where the
@@
 \citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
-There remains unclarity for Atlantic limpet species, particularly since
-it was last shown here, that they are not reliable recorders of
-palaeotemperature \citep{Graniero2015-zv}.
+There remains a lack of clarity for Atlantic limpet species,
+particularly since it was last shown here, that they are not reliable
+recorders of palaeotemperature \citep{Graniero2015-zv}.
 
@@
 
-Relying and adding onto published datasets, allowed us to simultaneously
-avoid costs for new high-resolution δ\textsuperscript{18}O-data, use
-real-world examples, and also provide pilot-data for areas of existing
-research interest. Generally establishing the usefulness and reliabilty
-of Mg/Ca as SST proxy in limpets should help to provide a platform for
-future research and an important stepping stone to better understand
-elemental ratios in other marine mollusc shells.
+By relying and adding onto published datasets, we were able to
+simultaneously avoid costs for new high-resolution
+δ\textsuperscript{18}O-data, to use real-world examples, and to also
+provide pilot-data for areas of existing research interest. Generally
+establishing the usefulness and reliability of Mg/Ca as SST proxy in
+limpets should help to provide a platform for future research and an
+important stepping stone to better understand elemental ratios in other
+marine mollusc shells.
 
@@
 
-In total, we studied 16 archaeological specimens. Of those, seven come
-from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
-previously been studied to characterise seasonal temperature change
-during the Medieval Climate Anomaly and the Little Ice Age
-\citep{Surge2012-ba}. Together they show a range of lifespans, with some
-comparatively long for their size (QG1-7246-1, 12 years within 16 mm of
-growth) making them particularly susceptible to time-averaging. That
-said, the majority are under 5 years old. These shells are an ideal
+In total, we studied 16 archaeological specimens. Of those, seven
+\emph{P. vulgata} come from Quoygrew on Westray dating to 900 -- 1200
+CE. These shells have previously been studied to characterise seasonal
+temperature change during the Medieval Climate Anomaly and the Little
+Ice Age \citep{Surge2012-ba}. Together they show a range of lifespans,
+with some comparatively long for their size (QG1-7246-1, 12 years within
+16 mm of growth) making them particularly susceptible to time-averaging.
+That said, the majority are under 5 years old. These shells are an ideal
 comparative dataset from an archaeological assemblage to compare to the
@@
 
-Nine additional shells have been selected for LIBS analysis only, to
+Nine additional shells of \emph{N. magellanica} (n=6) and \emph{N}.
+\emph{deaureata} (n=3) have been selected for LIBS analysis only, to
 provide a comparative dataset to the modern \emph{Nacella} sp. shells.
@@
 ratio distributions are also preserved and apparent in the
-archaeological shells.
+archaeological shells. They derive from the site of Heshkaia 35 situated
+within Moat Bay, around 30 km east of where the modern specimens in our
+study were collected \citep{Zangrando2014-qi, Zangrando2021-qg}. With
+radiocarbon dates putting the shells into the Late Holocene (1300 --
+1450 CE) they are of similar age --- if somewhat younger --- as the
+assemblage from Quoygrew.
 
 \subsection{Methods}\label{methods}
-
-\subsubsection{Oxygen isotopes}\label{oxygen-isotopes}
-
-High-quality oxygen isotope values (δ\textsuperscript{18}O) were
-acquired from previous publications on \emph{P. vulgata}
-\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
-magellanica} \citep{Nicastro2020-ih}, where detailed descriptions of the
-stable isotope analysis methodology can be found. In short, all shells
-were sectioned along the main growth axis to expose the internal growth
-structure. From the exposed section carbonate powder samples were
-acquired using a micromilling approach with resolutions of 100--200 µm
-and targeting the calcitic M+2 layer of the shells. Analytical precision
-of the isotope values was consistently at 0.05--0.10‰ and values are
-reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee
-Belemnite) standard.
 
@@
 involve the ablation of the shell section using an infrared (1064 nm)
-laser pulse (1--2 mJ, 100 Hz) onto a surface area of 20--30 µm to
-generate a plasma plume. This plume emits light, which is measured using
-a synchronised spectrometer. The resulting light spectrum quantifies the
-emission lines of magnesium () and calcium () to determine their
-intensity ratio. While these two peaks alone do not represent the molar
-concentrations of both elements (as opposed to calibration free LIBS;
-e.g. \citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
-correlated to the molar concentration \citep{Hausmann2017-oa}.
+laser (1--2 mJ, 100 Hz) onto a surface area of 20--30 µm to generate a
+plasma plume. This plume emits light, which is measured using a
+synchronised spectrometer. The resulting light spectrum quantifies the
+emission lines of magnesium (MgII; 279.553 nm) and calcium (CaII;
+315.887 nm) to determine their intensity ratio. While these two peaks
+alone do not represent the molar concentrations of both elements (as
+opposed to calibration free LIBS; e.g.
+\citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
+correlated to the molar concentration \citep{Hausmann2017-oa} and can be
+used as a reliable indicator of Mg/Ca changes within the shell
+carbonate.
 
-\section{Results}\label{Results}
+\subsubsection{Oxygen isotopes}\label{oxygen-isotopes}
 
-\subsubsection{Modern Specimens}\label{modern-specimens-1}
+High-quality oxygen isotope values (δ\textsuperscript{18}O) were
+acquired from previous publications on \emph{P. vulgata}
+\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
+magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
+the stable isotope analysis methodology can be found. In short, all
+shells were sectioned along the main growth axis to expose the internal
+growth structure. From the exposed section carbonate powder samples were
+acquired using a micromill with resolutions of 100--200 µm and by
+targeting the calcitic M+2 layer of the shells. Analytical precision of
+the isotope values was consistently at 0.05--0.10‰ and values are
+reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee
+Belemnite) standard.
 
-All modern shells show multi-year quasi-sinusoidal records of SST change
-(Figure~\ref{fig-Mod_iso}), with \emph{Nacella} sp. shells living 2--3
-years and \emph{Patella v}. over 5 years. The δ\textsuperscript{18}O
-values of the Orcadian shell have not been directly compared to the
-local SST, but accurately reflect the seasonal changes expected barring
-one extreme, anomalous value in the second summer of recorded growth.
+Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
+previously acquired sequential δ\textsuperscript{18}O-values of limpet
+shells by species. Due to the high sampling resolution, all shells show
+seasonally resolved and multi-year quasi-sinusoidal records of SST
+change , with \emph{Nacella} sp. shells living 2--3 years and
+\emph{Patella v}. regularly over 3 years. In those instances where
+longer lifetimes are encountered (i.e.~specimen QG1-7246-1,
+Figure~\ref{fig-Pat_iso} b). Interestingly, the ranges in
+δ\textsuperscript{18}O-values are not similar between the coeval
+\emph{Nacella} sp. specimens with δ\textsuperscript{18}O values for the
+previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for
+winter values between 3.0‰ and 2.0‰. \emph{P. vulgata} has similarly
+variable δ\textsuperscript{18}O-values for summer and winter seasons,
+alas the lack of more that one modern specimens, means that we cannot
+quantify the variability of recorded annual ranges between specimens.
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Mod_iso-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.pdf}
 
@@
 
-\caption{\label{fig-Mod_iso}Isotope values of modern limpet shells from
-Cambaceres Bay (\emph{a-f}) and Westray (\emph{g}).}
+\caption{\label{fig-Nac_iso}Isotope values of Nacella sp. limpet shells
+from Cambaceres Bay (\emph{a-f}).}
 
 \end{figure}%
-
-\subsubsection{Archaeological
-Specimens}\label{archaeological-specimens-1}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Arch_iso-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_iso-1.pdf}
 
@@
 
-\caption{\label{fig-Arch_iso}Isotope values of Archaeological limpet
-shells from Quoygrew}
+\caption{\label{fig-Pat_iso}Isotope values of Patella vulagata limpet
+shells from Quoygrew (QG, \emph{a-g}) and Rack Wick Bay (ORK, \emph{h})
+on Westray, Orkney.}
 
@@
 
-\subsection{Patella vulgata}\label{patella-vulgata}
+\section{Results}\label{Results}
 
-\subsection{Nacella deaureata}\label{nacella-deaureata}
+\subsection{\texorpdfstring{\emph{Patella
+vulgata}}{Patella vulgata}}\label{patella-vulgata}
 
-\subsection{Nacella magellanica}\label{nacella-magellanica}
+\subsection{\texorpdfstring{\emph{Nacella
+deaureata}}{Nacella deaureata}}\label{nacella-deaureata}
+
+\subsection{\texorpdfstring{\emph{Nacella
+magellanica}}{Nacella magellanica}}\label{nacella-magellanica}
 
@@
 
+\renewcommand\refname{References}
   \bibliography{bibliography.bib}

### 2024-06-24 Manuscript.qmd
added separate graphs

@@
 
+D18O <-   D18O %>% filter(sampleID %in% MgCa$sampleID)
+
 ```
@@
 
-```{r SST Data}
+```{r SST_Data}
 
@@
 
-```{r SST plot}
+```{r SST_plot}
 #| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
@@
 
+Using this system we carried out elemental imaging of the shell specimens at a resolution of 50 µm distance between sampling spots.
+Each spot was irradiated 10 times with the 3 first spectra used for cleaning steps and the remaining summed to get an average Mg/Ca intensity ratio for each sample spot.
+Subsequently, we re-sampled the section using a line scan at 10 µm resolution.
+This leads to an overlap between sample spots, but also allows for a continuous record without gaps.
+Intensity ratios were filtered for cases with high relative standard deviation (i.e.
+more than 10%).
+These occurred in places where the previous sampling procedures for carbonate powder as part of the oxygen isotope analysis left an uneven sample surfaces introducing variability in the plasma generation and thus uncertainty in the data of these locations.
+
 ### Oxygen isotopes
@@
 specimen QG1-7246-1, @fig-Pat_iso b).
-Interestingly, the ranges in δ^18^O-values are not similar between the coeval *Nacella* sp.
-specimens with δ^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 3.0‰ and 2.0‰.
+Interestingly, the annual ranges in δ^18^O-values are not similar between the coeval *Nacella* sp.
+specimens with δ^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 2.7‰ and 2.0‰.
 *P.
@@
 ```{r Nacella_Isotope_plot}
-#| fig-cap: Isotope values of Nacella sp. limpet shells from Cambaceres Bay (*a-f*).
+#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay (*a-f*).
 #| fig-height: 10
@@
 ```{r Patella_Isotope_plot}
-#| fig-cap: Isotope values of Patella vulagata limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
+#| fig-cap: Isotope values of *Patella vulgata* limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
 #| fig-height: 10
@@
 
-# Results {#Results}
+## Dynamic Time Warping
 
-## *Patella vulgata*
-
-## *Nacella deaureata*
-
-## *Nacella magellanica*
+Because the exact location used in the analysis of stable oxygen isotopes is not possible to reconstruct for each carbonate powder sample, we used dynamic time warping (DTW) to align the time series of δ^18^O-values and Mg/Ca ratios.
+DTW is an algorithm that measures similarity between two proxy-sequences, which may vary in sampling resolution or interval.
+By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences.
+This allows us to compare the proxy data sets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates.
+We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data.
+This involved selecting appropriate distance measures and constraints to ensure meaningful alignment.
+The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.
+The code used for this analysis as well as the data required to run it can be found in this **OSF repository**.
 
@@
 
-#
+# Results {#Results}
 
-```{r Correlation Graphs }
-#| fig-cap: Correlation graphs for all specimens
+## *Patella vulgata*
+
+-   Show one or a few of the LIBS maps as well as the line scans.
+-   Show one the line scans together with isotope data.
+-   Show correlation graphs.
+
+```{r Patella_Correlation_Graphs }
+#| fig-cap: Correlation graphs for *Patella vulgata* specimens
 #| fig-height: 20
-#| fig-width: 10
+#| fig-width: 15
 
@@
   pivot_wider(values_from = "value",names_from = "proxy") %>% 
-  mutate(region = substr(sampleID, 1, 1)) %>% 
+  # mutate(region = substr(sampleID, 1, 1)) %>% 
+  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
+  filter(species=="Patella vulgata") %>% 
+  ggplot()+
+  aes(d18o,mg_ca,col=sampleID)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
+  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
+  # stat_regline_equation(label.x = 2, #label.y = 32
+  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+  #   formula = y~x
+  # ) +
+  scale_colour_manual(values = c("#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063")) +
+scale_fill_manual(values = c( "#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063"))#+
+  # facet_wrap(~region,ncol=2,scales = "free")
+
+# Expanded colour palette with 6 additional colours
+# Updated colour palette with 4 new replacements for the last colours
+# Refined colour palette with 3 additional colours
+# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
+
+P_individual <- P_all+ylim(0,0.8)+
+  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
+    formula = y~x
+  )+
+  theme_cowplot(20)+
+  theme(
+    strip.background = element_rect(fill = "white", colour = NA),
+    strip.text = element_text(colour = "black"),
+    panel.border = element_rect(colour="black")
+  )
+
+P_all/P_individual+plot_layout(heights = c(0.5,0.5))
+
+```
+
+## *Nacella* sp.
+
+```{r Nacella_Correlation_Graphs}
+#| fig-cap: Correlation graphs for *Nacella* sp. specimens
+#| fig-height: 15
+#| fig-width: 15
+
+P_all <- Corr_data %>% 
+  pivot_wider(values_from = "value",names_from = "proxy") %>% 
+  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
+  filter(grepl("Nacella", species)) %>% 
   ggplot()+
@@
 P_individual <- P_all+ylim(0,0.8)+
-  facet_wrap(~sampleID,ncol=3)+#ylim(0.15,NA)+
-  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 0.65,
-    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
+  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
     formula = y~x
   )+
+  theme_cowplot(20)+
   theme(
@@
 
-P_all/P_individual+plot_layout(heights = c(0.5,0.5))
+P_all/P_individual+plot_layout(heights = c(0.8,0.2))
 
@@
 # References
+
 ```{r Modern iso ranges}
@@
 ```
+
+```{r Correlation Graphs, eval=FALSE }
+#| fig-cap: Correlation graphs for all specimens
+#| fig-height: 20
+#| fig-width: 10
+
+P_all <- Corr_data %>% 
+  pivot_wider(values_from = "value",names_from = "proxy") %>% 
+  mutate(region = substr(sampleID, 1, 1)) %>% 
+  ggplot()+
+  aes(d18o,mg_ca,col=sampleID)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
+  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
+  # stat_regline_equation(label.x = 2, #label.y = 32
+  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+  #   formula = y~x
+  # ) +
+  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
+scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
+  # facet_wrap(~region,ncol=2,scales = "free")
+
+# Expanded colour palette with 6 additional colours
+# Updated colour palette with 4 new replacements for the last colours
+# Refined colour palette with 3 additional colours
+# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
+
+P_individual <- P_all+ylim(0,0.8)+
+  facet_wrap(~sampleID,ncol=3)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 0.65,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
+    formula = y~x
+  )+
+  theme(
+    strip.background = element_rect(fill = "white", colour = NA),
+    strip.text = element_text(colour = "black"),
+    panel.border = element_rect(colour="black")
+  )
+
+P_all/P_individual+plot_layout(heights = c(0.5,0.5))
+
+```

### 2024-06-24 Manuscript.tex
added separate graphs

@@
 
+Using this system we carried out elemental imaging of the shell
+specimens at a resolution of 50 µm distance between sampling spots. Each
+spot was irradiated 10 times with the 3 first spectra used for cleaning
+steps and the remaining summed to get an average Mg/Ca intensity ratio
+for each sample spot. Subsequently, we re-sampled the section using a
+line scan at 10 µm resolution. This leads to an overlap between sample
+spots, but also allows for a continuous record without gaps. Intensity
+ratios were filtered for cases with high relative standard deviation
+(i.e.~more than 10\%). These occurred in places where the previous
+sampling procedures for carbonate powder as part of the oxygen isotope
+analysis left an uneven sample surfaces introducing variability in the
+plasma generation and thus uncertainty in the data of these locations.
+
 \subsubsection{Oxygen isotopes}\label{oxygen-isotopes}
@@
 longer lifetimes are encountered (i.e.~specimen QG1-7246-1,
-Figure~\ref{fig-Pat_iso} b). Interestingly, the ranges in
+Figure~\ref{fig-Pat_iso} b). Interestingly, the annual ranges in
 δ\textsuperscript{18}O-values are not similar between the coeval
@@
 previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for
-winter values between 3.0‰ and 2.0‰. \emph{P. vulgata} has similarly
+winter values between 2.7‰ and 2.0‰. \emph{P. vulgata} has similarly
 variable δ\textsuperscript{18}O-values for summer and winter seasons,
@@
 
-\caption{\label{fig-Nac_iso}Isotope values of Nacella sp. limpet shells
-from Cambaceres Bay (\emph{a-f}).}
+\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} sp. limpet
+shells from Cambaceres Bay (\emph{a-f}).}
 
@@
 
-\caption{\label{fig-Pat_iso}Isotope values of Patella vulagata limpet
-shells from Quoygrew (QG, \emph{a-g}) and Rack Wick Bay (ORK, \emph{h})
-on Westray, Orkney.}
+\caption{\label{fig-Pat_iso}Isotope values of \emph{Patella vulgata}
+limpet shells from Quoygrew (QG, \emph{a-g}) and Rack Wick Bay (ORK,
+\emph{h}) on Westray, Orkney.}
 
 \end{figure}%
+
+\subsection{Dynamic Time Warping}\label{dynamic-time-warping}
+
+Because the exact location used in the analysis of stable oxygen
+isotopes is not possible to reconstruct for each carbonate powder
+sample, we used dynamic time warping (DTW) to align the time series of
+δ\textsuperscript{18}O-values and Mg/Ca ratios. DTW is an algorithm that
+measures similarity between two proxy-sequences, which may vary in
+sampling resolution or interval. By stretching or compressing sections
+of the series, DTW finds the probable alignment between the two
+sequences. This allows us to compare the proxy data sets more
+effectively, ensuring that the temporal dynamics of each shell are
+accurately matched despite possible discrepancies in sampling intervals
+or rates. We applied the DTW algorithm using the \texttt{dtw} package in
+R \citep{Giorgino2009-sj, R_Core_Team2020-mk}, which provides a robust
+framework for aligning time series data. This involved selecting
+appropriate distance measures and constraints to ensure meaningful
+alignment. The process involved iterative adjustments to minimise the
+overall distance between corresponding points in the two data sets. The
+code used for this analysis as well as the data required to run it can
+be found in this \textbf{OSF repository}.
 
@@
 
-\subsection{\texorpdfstring{\emph{Nacella
-deaureata}}{Nacella deaureata}}\label{nacella-deaureata}
-
-\subsection{\texorpdfstring{\emph{Nacella
-magellanica}}{Nacella magellanica}}\label{nacella-magellanica}
-
-\section{}\label{section}
+\begin{itemize}
+\tightlist
+\item
+  Show one or a few of the LIBS maps as well as the line scans.
+\item
+  Show one the line scans together with isotope data.
+\item
+  Show correlation graphs.
+\end{itemize}
 
@@
 
-{\centering \includegraphics{Manuscript_files/figure-pdf/Correlation Graphs-1.pdf}
+{\centering \includegraphics{Manuscript_files/figure-pdf/Patella_Correlation_Graphs-1.pdf}
 
@@
 
-\caption{Correlation graphs for all specimens}
+\caption{Correlation graphs for \emph{Patella vulgata} specimens}
+
+\end{figure}%
+
+\subsection{\texorpdfstring{\emph{Nacella}
+sp.}{Nacella sp.}}\label{nacella-sp.}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{Manuscript_files/figure-pdf/Nacella_Correlation_Graphs-1.pdf}
+
+}
+
+\caption{Correlation graphs for \emph{Nacella} sp. specimens}
 

### 2024-06-26 Manuscript.qmd
added results

@@
 
--   Show one or a few of the LIBS maps as well as the line scans.
--   Show one the line scans together with isotope data.
--   Show correlation graphs.
+The elemental imaging of *P.
+vulgata* shells consistently showed repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca intensity ratios indicating and low ratios indicating low temperature periods.
+As examples, @fig-Pat_LIBS shows the 2D-distribution of elemental ratios across the entire section of ORK-LT5 and the preserved anterior side of QG1-7188-1 (additional maps can be found in the supplementaries).
+The repeating patterns are found across the calcitic layers that are exterior to the myostracum (layers m+2 and m+3) as well as the layers that are interior (m-2) [@Fenger2007-gf].
+The layer m-1, which is also interior, consists of aragonite and is seen in @fig-Pat_LIBS in grey, as its Mg/Ca intensity ratio consistently falls below the range of interest (0.2 and above).
+Compared to other patelloid shells, whose interior is almost entirely made of aragonite, this layer is very thin.
+
+```{r Patella_LIBS}
+#| fig-cap: Example LIBS maps and line scans of analysed specimens.
+(*a+b*) ORK-LT and (*c*+*d*) QG1-7188-1.
+Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast.
+Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
+#| fig-height: 10
+#| fig-width: 15
+#| label: fig-Pat_LIBS
+
+#ORK_LT5
+ORK_LT5_map <-
+  read_csv(here("data", "map_data", "ORK_LT_5_A100_All_analyzed_data.csv"))
+
+ORK_LT5_plot <- 
+  ORK_LT5_map %>%
+  filter(std < 0.15) %>%
+  filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
+  ggplot() +
+  aes(x, y, fill = mg_ca) +
+  geom_tile(data = ORK_LT5_map, fill = "grey") +
+  geom_tile(show.legend=FALSE) +
+  geom_path(data=MgCa %>% filter(sampleID=="ORK_LT5"),aes(x,y),col="white",linetype="longdash")+
+  theme_cowplot(20)+
+  scale_fill_viridis_c(limits = c(0.3, 0.7),
+                       breaks = seq(0.3, 0.7, by = 0.1)) +    
+  theme(
+    legend.position = c(0.45, 0.4),
+    legend.title = element_text(hjust = 0.5)
+    )+ # Center the legend title) +
+    coord_fixed() +
+    labs(x = "mm", y = "mm", fill = "Mg/Ca ")
+
+ORK_LT5_line <- 
+  MgCa %>% 
+  filter(sampleID=="ORK_LT5") %>% 
+  filter(relstd < 0.1) %>%
+  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
+  ggplot()+
+  aes(dist,mg_ca,col=mg_ca)+
+  geom_path(show.legend = FALSE)+
+  scale_colour_viridis_c(limits = c(0.3, 0.6),
+                       breaks = seq(0.3, 0.6, by = 0.1)) +    
+  theme_cowplot(20)+
+  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
+
+###########   
+#QG1_7188_1
+###########
+QG1_7188_1_map <-
+  read_csv(here("data", "map_data", "QG1-7188-1-AB50_analyzed_data.txt"))
+
+QG1_7188_1_plot <- 
+  QG1_7188_1_map %>%
+  filter(relstd < 0.1) %>%
+  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
+  ggplot() +
+  aes(x, y, col = mg_ca) +
+  geom_point(data = QG1_7188_1_map, col = "grey") +
+  geom_point(size=1,show.legend=FALSE) +
+  geom_path(data=MgCa %>% filter(sampleID=="QG1-7188-1"),aes(x,y),col="white",linetype="longdash")+
+  scale_colour_viridis_c(limits = c(0.3, 0.6),
+                       breaks = seq(0.3, 0.6, by = 0.1)) +    
+  theme_cowplot(20)+
+  theme(
+    legend.position = c(0.05, 0.78),
+   legend.title = element_text(hjust = 0.5)
+    )+ # Center the legend title) +
+    coord_fixed() +
+    labs(x = "mm", y = "mm", colour = "Mg/Ca")
+
+QG1_7188_1_line <- 
+  MgCa %>% 
+  filter(sampleID=="QG1-7188-1") %>% 
+  filter(relstd < 0.1) %>%
+  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
+  ggplot()+
+  aes(dist,mg_ca,col=mg_ca)+
+  geom_path(show.legend = FALSE)+
+  scale_colour_viridis_c(limits = c(0.3, 0.6),
+                       breaks = seq(0.3, 0.6, by = 0.1)) +    
+  theme_cowplot(20)+
+  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
+
+################
+####FIGURE######
+################
+  (ORK_LT5_plot+ORK_LT5_line)/(QG1_7188_1_plot+QG1_7188_1_line)+plot_annotation(tag_levels = "a")
+```
+
+Both specimens in @fig-Pat_LIBS show increased Mg/Ca intensity ratios towards the outside of the shell and some degree of intra-increment variability.
+This is particularly visible in @fig-Pat_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in the first year of growth (9–12 mm on the y-axis).
+A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0–10 mm on the x-axis).
+
+Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental-imaging.
+That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (\@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D-image, showing the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].
+
+@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its δ^18^O-values (similar graphs for other specimen can be found in the supplementaries).
+The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have several different distances.
+Depending on where one measures the distance and a line scan towards the interior would be shorter than a line scan towards the exterior.
+That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.
+
+```{r Patella_Compare}
+#| fig-cap: Comparison of LIBS line-scan and sequential δ^18^O-values of specimen ORK-LT5 
+#| fig-height: 10
+#| fig-width: 20
+#| label: fig-Pat_Comp
+
+P1 <-
+  MgCa %>% 
+  filter(sampleID=="ORK_LT5") %>% 
+  filter(relstd < 0.1) %>%
+  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
+ggplot()+
+  aes(dist,mg_ca)+
+  geom_path()+
+  theme_cowplot(20)+
+  labs(x = "", y = "Mg/Ca intensity ratio")
+
+P2 <-
+  D18O %>% 
+  filter(sampleID=="ORK_LT5") %>% 
+  ggplot()+
+  aes(dist,d18O)+
+  geom_path()+
+  theme_cowplot(20)+
+  scale_y_reverse()+
+  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))
+
+P1/P2+plot_annotation(tag_levels = "a")
+
+```
+
+Aligning the records programmatically using dynamic time warping, allowed us to compare them directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr).
+The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters (δ^18^Ο = 0.6 – 0.065 \* Mg/Ca).
+The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2).
+These specimens are also the oldest and youngest, respectively, specimens, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies.
+The mean R^2^ value is 0.89, suggesting a generally good fit between the two proxies.
 
 ```{r Patella_Correlation_Graphs }
-#| fig-cap: Correlation graphs for *Patella vulgata* specimens
-#| fig-height: 20
+#| fig-cap: Correlation graphs for *Patella vulgata* specimens.
+#| fig-height: 18
 #| fig-width: 15
+#| label: fig-Pat_Corr
 
@@
   geom_point(show.legend = FALSE)+
+  theme_cowplot(20)+
   labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
@@
 
+```{r Nacella_LIBS}
+#| fig-cap: Example LIBS maps and line scans of analysed specimens.
+(*a+b*) ND-1016-3 and (*c*+*d*) NM-1016-3.
+Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast.
+Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
+#| fig-height: 10
+#| fig-width: 15
+#| label: fig-Nac_LIBS
+
+#ORK_LT5
+ND_1016_map <-
+  read_csv(here("data", "map_data", "ND-106-3-A100B_analyzed_data.csv"))
+
+ND_Line <- read_csv(here("data", "map_data", "ND-106-3-A10-LineScan_analyzed_data.csv")) %>% 
+  mutate(dist=row_number()*0.01)
+
+ND_1016_plot <- 
+  ND_1016_map %>%
+ # filter(std < 0.15) %>%
+  #filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
+  ggplot() +
+  aes(x, y, fill = mg_ca) +
+  geom_tile(data = ND_1016_map, fill = "grey") +
+  geom_tile(show.legend=FALSE) +
+  geom_path(data=ND_Line,aes(x,y),col="white",linetype="longdash")+
+  theme_cowplot(20)+
+  scale_fill_viridis_c(limits = c(0, 1),
+                       breaks = seq(0.3, 0.7, by = 0.1)) +    
+  theme(
+    legend.position = c(0.45, 0.4),
+    legend.title = element_text(hjust = 0.5)
+    )+ # Center the legend title) +
+    coord_fixed() +
+    labs(x = "mm", y = "mm", fill = "Mg/Ca ")
+
+ND_1016_line <- 
+  ND_Line %>% 
+  ggplot()+
+  aes(dist,mg_ca,col=mg_ca)+
+  geom_path(show.legend = FALSE)+
+  scale_colour_viridis_c() +    
+  theme_cowplot(20)+
+  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
+
+###########
+#NM-1016-3
+###########
+NM_1016_map100 <-
+  read_csv(here("data", "map_data", "NM-1016-3-A100_analyzed_data.csv"))
+NM_1016_map50 <-
+  read_csv(here("data", "map_data", "NM-1016-3-A50Edge_analyzed_data.csv"))
+
+NM_Line <- read_csv(here("data", "map_data", "NM-1016-3-A10LineScan2_analyzed_data.csv")) %>% 
+  mutate(dist=row_number()*0.01)
+
+NM_1016_plot100 <- 
+  NM_1016_map100 %>%
+  filter(std < 0.15) %>%
+  ggplot() +
+  aes(x, y, fill = mg_ca) +
+  geom_tile(show.legend=FALSE,size=0.5) +
+  geom_path(data=NM_Line,aes(x,y),col="white",linetype="longdash")+
+  theme_cowplot(20)+
+  scale_fill_viridis_c(limits = c(0, 0.5),
+                       breaks = seq(0, 0.4, by = 0.1)) +    
+  theme(
+    legend.position = c(0.45, 0.4),
+    legend.title = element_text(hjust = 0.5)
+    )+ # Center the legend title) +
+    coord_fixed() +
+   annotate("rect", 
+           xmin = min(NM_1016_map50$x)+15, 
+           xmax = max(NM_1016_map50$x)+15, 
+           ymin = min(NM_1016_map50$y)-2.3, 
+           ymax = max(NM_1016_map50$y)-2, 
+           alpha = 0, linetype="longdash",col = "black")+
+    labs(x = "mm", y = "mm", fill = "Mg/Ca ")
+
+NM_1016_plot50 <- 
+  NM_1016_map50 %>%
+ # filter(std < 0.15) %>%
+  #filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
+  ggplot() +
+  aes(x, y, fill = mg_ca) +
+  geom_tile(show.legend=FALSE,size=0.5) +
+  theme_cowplot(20)+
+  scale_fill_viridis_c(limits = c(0, 0.4),
+                       breaks = seq(0, 0.4, by = 0.1)) +    
+theme(
+    axis.title = element_blank(),
+    axis.text = element_blank(),
+    axis.ticks = element_blank()
+  )+ 
+    coord_fixed() +
+    labs(x = "mm", y = "mm", colour = "Mg/Ca ")
+
+NM_1016_line <- 
+  NM_Line %>% 
+  filter(relstd<0.1) %>% 
+  ggplot()+
+  aes(dist,mg_ca,col=mg_ca)+
+  geom_path(show.legend = FALSE)+
+  scale_colour_viridis_c() +    
+  theme_cowplot(20)+
+  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
+
+NM_1016_plot <- NM_1016_plot100+inset_element(NM_1016_plot50+theme(plot.tag = element_blank()), 0, 0, 0.7, 0.7)
+
+################
+####FIGURE######
+################
+  (ND_1016_plot+ND_1016_line)/(NM_1016_plot+NM_1016_line)+plot_annotation(tag_levels = "a")
+```
+
+Both specimens in @fig-Pat_LIBS show increased Mg/Ca intensity ratios towards the outside of the shell and some degree of intra-increment variability.
+This is particularly visible in @fig-Pat_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in the first year of growth (9–12 mm on the y-axis).
+A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0–10 mm on the x-axis).
+
+Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental-imaging.
+That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (\@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D-image, showing the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].
+
+@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its δ^18^O-values (similar graphs for other specimen can be found in the supplementaries).
+The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have several different distances.
+Depending on where one measures the distance and a line scan towards the interior would be shorter than a line scan towards the exterior.
+That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.
+
+```{r Nacella_Compare}
+#| fig-cap: Comparison of LIBS line-scan and sequential δ^18^O-values of specimen ORK-LT5 
+#| fig-height: 10
+#| fig-width: 20
+#| label: fig-Nac_Comp
+
+P1 <-
+  MgCa %>% 
+  filter(sampleID=="ORK_LT5") %>% 
+  filter(relstd < 0.1) %>%
+  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
+ggplot()+
+  aes(dist,mg_ca)+
+  geom_path()+
+  theme_cowplot(20)+
+  labs(x = "", y = "Mg/Ca intensity ratio")
+
+P2 <-
+  D18O %>% 
+  filter(sampleID=="ORK_LT5") %>% 
+  ggplot()+
+  aes(dist,d18O)+
+  geom_path()+
+  theme_cowplot(20)+
+  scale_y_reverse()+
+  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))
+
+P1/P2+plot_annotation(tag_levels = "a")
+
+```
 ```{r Nacella_Correlation_Graphs}

### 2024-06-26 Manuscript.tex
added results

@@
 
-\begin{itemize}
-\tightlist
-\item
-  Show one or a few of the LIBS maps as well as the line scans.
-\item
-  Show one the line scans together with isotope data.
-\item
-  Show correlation graphs.
-\end{itemize}
+The elemental imaging of \emph{P. vulgata} shells consistently showed
+repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
+intensity ratios indicating and low ratios indicating low temperature
+periods. As examples, Figure~\ref{fig-Pat_LIBS} shows the
+2D-distribution of elemental ratios across the entire section of ORK-LT5
+and the preserved anterior side of QG1-7188-1 (additional maps can be
+found in the supplementaries). The repeating patterns are found across
+the calcitic layers that are exterior to the myostracum (layers m+2 and
+m+3) as well as the layers that are interior (m-2)
+\citep{Fenger2007-gf}. The layer m-1, which is also interior, consists
+of aragonite and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its
+Mg/Ca intensity ratio consistently falls below the range of interest
+(0.2 and above). Compared to other patelloid shells, whose interior is
+almost entirely made of aragonite, this layer is very thin.
 
-\begin{figure}[H]
+\begin{figure}
 
-{\centering \includegraphics{Manuscript_files/figure-pdf/Patella_Correlation_Graphs-1.pdf}
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_LIBS-1.pdf}
 
@@
 
-\caption{Correlation graphs for \emph{Patella vulgata} specimens}
+\caption{\label{fig-Pat_LIBS}Example LIBS maps and line scans of
+analysed specimens. (\emph{a+b}) ORK-LT and (\emph{c}+\emph{d})
+QG1-7188-1. Note that the ranges of Mg/Ca intensity ratios are not
+identical but are chosen arbitrarily to increase contrast. Areas in grey
+consist of low-Mg aragonite and fall outside the chosen colour-range.}
+
+\end{figure}%
+
+Both specimens in Figure~\ref{fig-Pat_LIBS} show increased Mg/Ca
+intensity ratios towards the outside of the shell and some degree of
+intra-increment variability. This is particularly visible in
+Figure~\ref{fig-Pat_LIBS} c, where Mg/Ca intensity ratios range from
+\textasciitilde0.5--0.7 in the first year of growth (9--12 mm on the
+y-axis). A similar pattern is visible in the interior m-2 layer of
+ORK-LT5 (Figure~\ref{fig-Pat_LIBS} a), which increase towards the
+anterior of the shell (0--10 mm on the x-axis).
+
+Line scans also indicate well the quasi-sinusoidal change of Mg/Ca
+intensity ratios expected based on the stable isotope data and
+elemental-imaging. That said, some of the variability of Mg/Ca intensity
+ratios within one season such as the summer period of QG1-7188-1 between
+0.5 and 2.0 mm distance to the shell edge (@fig-Pat\_LIBS d), is visible
+in a more dramatic manner than it appears in the 2D-image, showing the
+downside of line-scanning as opposed to the broader milling approach
+used for the analysis of carbonate powder \citep{Ferguson2011-zl}.
+
+Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
+to its δ\textsuperscript{18}O-values (similar graphs for other specimen
+can be found in the supplementaries). The measurements of the distance
+to the shell edge are not entirely identical, most likely because one
+growth increment can have several different distances. Depending on
+where one measures the distance and a line scan towards the interior
+would be shorter than a line scan towards the exterior. That said,
+increases and decreases of both proxies seem to mirror each other and to
+be well aligned.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Comp-1.pdf}
+
+}
+
+\caption{\label{fig-Pat_Comp}Comparison of LIBS line-scan and sequential
+δ\textsuperscript{18}O-values of specimen ORK-LT5}
+
+\end{figure}%
+
+Aligning the records programmatically using dynamic time warping,
+allowed us to compare them directly to determine specimen-specific
+equations and quantify the correlation of both proxies
+(Figure~\ref{fig-Pat_Corr}). The equations for all specimens are
+different but the majority (barring one: QG1-7246-1) do seem to cluster
+around shared parameters (δ\textsuperscript{18}Ο = 0.6 -- 0.065 *
+Mg/Ca). The various coefficients of determination (R\textsuperscript{2})
+range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens
+are also the oldest and youngest, respectively, specimens, suggesting
+that lower growth rates and time averaging had a negative effect on the
+correlation of both proxies. The mean R\textsuperscript{2} value is
+0.89, suggesting a generally good fit between the two proxies.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Corr-1.pdf}
+
+}
+
+\caption{\label{fig-Pat_Corr}Correlation graphs for \emph{Patella
+vulgata} specimens.}
 
@@
 sp.}{Nacella sp.}}\label{nacella-sp.}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_LIBS-1.pdf}
+
+}
+
+\caption{\label{fig-Nac_LIBS}Example LIBS maps and line scans of
+analysed specimens. (\emph{a+b}) ND-1016-3 and (\emph{c}+\emph{d})
+NM-1016-3. Note that the ranges of Mg/Ca intensity ratios are not
+identical but are chosen arbitrarily to increase contrast. Areas in grey
+consist of low-Mg aragonite and fall outside the chosen colour-range.}
+
+\end{figure}%
+
+Both specimens in Figure~\ref{fig-Pat_LIBS} show increased Mg/Ca
+intensity ratios towards the outside of the shell and some degree of
+intra-increment variability. This is particularly visible in
+Figure~\ref{fig-Pat_LIBS} c, where Mg/Ca intensity ratios range from
+\textasciitilde0.5--0.7 in the first year of growth (9--12 mm on the
+y-axis). A similar pattern is visible in the interior m-2 layer of
+ORK-LT5 (Figure~\ref{fig-Pat_LIBS} a), which increase towards the
+anterior of the shell (0--10 mm on the x-axis).
+
+Line scans also indicate well the quasi-sinusoidal change of Mg/Ca
+intensity ratios expected based on the stable isotope data and
+elemental-imaging. That said, some of the variability of Mg/Ca intensity
+ratios within one season such as the summer period of QG1-7188-1 between
+0.5 and 2.0 mm distance to the shell edge (@fig-Pat\_LIBS d), is visible
+in a more dramatic manner than it appears in the 2D-image, showing the
+downside of line-scanning as opposed to the broader milling approach
+used for the analysis of carbonate powder \citep{Ferguson2011-zl}.
+
+Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
+to its δ\textsuperscript{18}O-values (similar graphs for other specimen
+can be found in the supplementaries). The measurements of the distance
+to the shell edge are not entirely identical, most likely because one
+growth increment can have several different distances. Depending on
+where one measures the distance and a line scan towards the interior
+would be shorter than a line scan towards the exterior. That said,
+increases and decreases of both proxies seem to mirror each other and to
+be well aligned.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Comp-1.pdf}
+
+}
+
+\caption{\label{fig-Nac_Comp}Comparison of LIBS line-scan and sequential
+δ\textsuperscript{18}O-values of specimen ORK-LT5}
+
+\end{figure}%
 

### 2024-06-26 Manuscript.qmd
added Nacella

@@
 
+The LIBS data for *Nacella* sp.
+specimens is less straightforward with shells being much thinner and thus patterns being more difficult to make out.
+Both specimens (ND-1016-3 and NM-1016-3) shown as example in @fig-Nac_LIBS (see additional graphs in supplementaries) also experience increases of Mg/Ca intensity ratio towards the exterior of the shell, similar to the *P.
+vulgata* specimens from Orkney.
+This intra-increment heterogeneity combined with the thinness of the shell, further complicates their analysis compared to *P.
+vulgata* or other *Patella* species [@Hausmann2019-fi].
+Nevertheless, repeating patterns were visible in the anterior section of the shell, which were also mirrored in the inner layers (m-2) of specimens NM-1016-3 (@fig-Nac_LIBS c).
+
 ```{r Nacella_LIBS}
@@
 Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
+Note the inset in (*c*) where the shell edge has been resampled using a higher resolution (50 µm instead of 100 µm) to better understand the Mg/Ca ratios at the shell edge.
 #| fig-height: 10
@@
   theme_cowplot(20)+
-  scale_fill_viridis_c(limits = c(0, 1),
-                       breaks = seq(0.3, 0.7, by = 0.1)) +    
+  scale_fill_viridis_c(limits = c(0, 0.5),
+                       breaks = seq(0, 0.5, by = 0.1)) +    
   theme(
@@
 
-Both specimens in @fig-Pat_LIBS show increased Mg/Ca intensity ratios towards the outside of the shell and some degree of intra-increment variability.
-This is particularly visible in @fig-Pat_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in the first year of growth (9–12 mm on the y-axis).
-A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0–10 mm on the x-axis).
-
-Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental-imaging.
-That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (\@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D-image, showing the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].
+Interestingly, the Mg/Ca intensity ratios in the *Nacella* sp.
+shells were much lower (0.05–0.30) than those seen in other patelloid species (e.g.
+0.3–0.7 in *Patella vulgata* above, or 0.5–1.5 in ) using the same emission lines.
+This is particularly visible in @fig-Nac_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in [@Hausmann2023-ih]).
+While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere, or seen only in aragonitic parts of e.g.
+*Patella* shells.
+
+Line scans reflect well the changes previously indicated by δ^18^O values and also indicate about two years of recorded growth.
+That said, the 2D imaging suggests that in both specimens a third summer can be added to the total record, which is not captured by the line scans.
+The variability seen in the *P.
+vulgata* specimens above is not as strong, which might also be related to the fast and consistent growth rate of the shells.
+
+\##########
 
@@
 ```
+
 ```{r Nacella_Correlation_Graphs}

### 2024-06-26 Manuscript.tex
added Nacella

@@
 
+The LIBS data for \emph{Nacella} sp. specimens is less straightforward
+with shells being much thinner and thus patterns being more difficult to
+make out. Both specimens (ND-1016-3 and NM-1016-3) shown as example in
+Figure~\ref{fig-Nac_LIBS} (see additional graphs in supplementaries)
+also experience increases of Mg/Ca intensity ratio towards the exterior
+of the shell, similar to the \emph{P. vulgata} specimens from Orkney.
+This intra-increment heterogeneity combined with the thinness of the
+shell, further complicates their analysis compared to \emph{P. vulgata}
+or other \emph{Patella} species \citep{Hausmann2019-fi}. Nevertheless,
+repeating patterns were visible in the anterior section of the shell,
+which were also mirrored in the inner layers (m-2) of specimens
+NM-1016-3 (Figure~\ref{fig-Nac_LIBS} c).
+
 \begin{figure}
@@
 identical but are chosen arbitrarily to increase contrast. Areas in grey
-consist of low-Mg aragonite and fall outside the chosen colour-range.}
+consist of low-Mg aragonite and fall outside the chosen colour-range.
+Note the inset in (\emph{c}) where the shell edge has been resampled
+using a higher resolution (50 µm instead of 100 µm) to better understand
+the Mg/Ca ratios at the shell edge.}
 
@@
 
-Both specimens in Figure~\ref{fig-Pat_LIBS} show increased Mg/Ca
-intensity ratios towards the outside of the shell and some degree of
-intra-increment variability. This is particularly visible in
-Figure~\ref{fig-Pat_LIBS} c, where Mg/Ca intensity ratios range from
-\textasciitilde0.5--0.7 in the first year of growth (9--12 mm on the
-y-axis). A similar pattern is visible in the interior m-2 layer of
-ORK-LT5 (Figure~\ref{fig-Pat_LIBS} a), which increase towards the
-anterior of the shell (0--10 mm on the x-axis).
+Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} sp.
+shells were much lower (0.05--0.30) than those seen in other patelloid
+species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in )
+using the same emission lines. This is particularly visible in
+Figure~\ref{fig-Nac_LIBS} c, where Mg/Ca intensity ratios range from
+\textasciitilde0.5--0.7 in \citep{Hausmann2023-ih}). While the intensity
+ratio depends chiefly on the chosen emission lines to calculate the
+proportion of the chosen Magnesium to the Calcium peak, the low ratios
+still indicate lower concentrations than seen elsewhere, or seen only in
+aragonitic parts of e.g.~\emph{Patella} shells.
 
-Line scans also indicate well the quasi-sinusoidal change of Mg/Ca
-intensity ratios expected based on the stable isotope data and
-elemental-imaging. That said, some of the variability of Mg/Ca intensity
-ratios within one season such as the summer period of QG1-7188-1 between
-0.5 and 2.0 mm distance to the shell edge (@fig-Pat\_LIBS d), is visible
-in a more dramatic manner than it appears in the 2D-image, showing the
-downside of line-scanning as opposed to the broader milling approach
-used for the analysis of carbonate powder \citep{Ferguson2011-zl}.
+Line scans reflect well the changes previously indicated by
+δ\textsuperscript{18}O values and Mg/Ca intensity ratios
+
+. That said, some of the variability of Mg/Ca intensity ratios within
+one season such as the summer period of QG1-7188-1 between 0.5 and 2.0
+mm distance to the shell edge (@fig-Pat\_LIBS d), is visible in a more
+dramatic manner than it appears in the 2D-image, showing the downside of
+line-scanning as opposed to the broader milling approach used for the
+analysis of carbonate powder \citep{Ferguson2011-zl}.
 

### 2024-06-26 Manuscript.qmd
added nacella text

@@
 bibliography: bibliography.bib
-format: 
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Palaeo^3^
-      formatting: preprint
-      model: 3p
-      cite-style: authoryear
+format: docx
+#  elsevier-pdf:
+#    keep-tex: true
+#    journal:
+#      name: Palaeo^3^
+#      formatting: preprint
+#      model: 3p
+#      cite-style: authoryear
 output:
@@
 
-P_all/P_individual+plot_layout(heights = c(0.5,0.5))
+P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")
 
@@
 
-\##########
-
-@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its δ^18^O-values (similar graphs for other specimen can be found in the supplementaries).
-The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have several different distances.
-Depending on where one measures the distance and a line scan towards the interior would be shorter than a line scan towards the exterior.
-That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.
+@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their δ^18^O-values (similar graphs for the other *Nacella* sp.
+specimen can be found in the supplementaries).
+Compared to the *P.
+vulgata* records, these mirror each other much more clearly, due to the brief period recorded in the shells and its simple make-up.
 
@@
 P1 <-
-  MgCa %>% 
-  filter(sampleID=="ORK_LT5") %>% 
+  ND_Line %>% 
   filter(relstd < 0.1) %>%
-  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
+  filter(mg_ca < 0.6) %>%
+  filter(dist<10) %>% 
 ggplot()+
@@
   D18O %>% 
-  filter(sampleID=="ORK_LT5") %>% 
+  filter(sampleID=="ND-1016-3") %>% 
   ggplot()+
@@
 
-P1/P2+plot_annotation(tag_levels = "a")
-
-```
+P3 <-
+  NM_Line %>% 
+  filter(relstd < 0.1) %>%
+  filter(dist<9.5) %>%
+ggplot()+
+  aes(dist,mg_ca)+
+  geom_path()+
+  theme_cowplot(20)+
+  labs(x = "", y = "Mg/Ca intensity ratio")
+
+P4 <-
+  D18O %>% 
+  filter(sampleID=="NM-1016-3") %>% 
+  ggplot()+
+  aes(dist,d18O)+
+  geom_path()+
+  theme_cowplot(20)+
+  scale_y_reverse()+
+  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))
+
+ ((P1 / P2) | (P3 / P4)) + plot_annotation(tag_levels = 'a')
+
+```
+
+The *Nacella* sp.
+specimens showed similarly clustered correlations, which interestingly were not grouped by species.
+In fact, specimens ND-1016-3 (*Nacella deaureata)* and NM-1016-3 (*Nacella magellanica*) are almost identical in their fitted equations (@fig-Nac_Corr).
+Both also share a high coefficient of determination (R^2^) of 0.95 and 0.92 respectively, with the remaining specimens being somewhat lower (R^2^ = 0.76 for ND-1016-4 and R^2^=0.84 for NM-1016-3).
+The mean R^2^ value for each *Nacella* species is 0.86 for *Nacella deaureata* and 0.88 for *Nacella magellanica*.
+Why two specimens of different species are so well aligned is not entirely certain and randomness cannot for now be ruled out.
+That said, since they both have a high R^2^ value, it might just be that they were both minimally affected by factors other than SST, including the sampling location of our line scans.
 
@@
 #| fig-cap: Correlation graphs for *Nacella* sp. specimens
-#| fig-height: 15
-#| fig-width: 15
+#| fig-height: 18
+#| fig-width: 10
+#| label: fig-Nac_Corr
 
@@
   stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
+  theme_cowplot(20)+
   # stat_regline_equation(label.x = 2, #label.y = 32
@@
   # ) +
-  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
-scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
+ # ylim(0.05,NA)+
+  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12")) +
+scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12"))#+
   # facet_wrap(~region,ncol=2,scales = "free")
@@
 
-P_individual <- P_all+ylim(0,0.8)+
-  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
-  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
+P_individual <- P_all+ylim(0,0.4)+
+  facet_wrap(~sampleID,ncol=2)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.35,
     aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
@@
 
-P_all/P_individual+plot_layout(heights = c(0.8,0.2))
+P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")
 

### 2024-06-26 Manuscript.tex
added nacella text

@@
 Line scans reflect well the changes previously indicated by
-δ\textsuperscript{18}O values and Mg/Ca intensity ratios
+δ\textsuperscript{18}O values and also indicate about two years of
+recorded growth. That said, the 2D imaging suggests that in both
+specimens a third summer can be added to the total record, which is not
+captured by the line scans. The variability seen in the \emph{P.
+vulgata} specimens above is not as strong, which might also be related
+to the fast and consistent growth rate of the shells.
 
-. That said, some of the variability of Mg/Ca intensity ratios within
-one season such as the summer period of QG1-7188-1 between 0.5 and 2.0
-mm distance to the shell edge (@fig-Pat\_LIBS d), is visible in a more
-dramatic manner than it appears in the 2D-image, showing the downside of
-line-scanning as opposed to the broader milling approach used for the
-analysis of carbonate powder \citep{Ferguson2011-zl}.
-
-Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
-to its δ\textsuperscript{18}O-values (similar graphs for other specimen
-can be found in the supplementaries). The measurements of the distance
-to the shell edge are not entirely identical, most likely because one
-growth increment can have several different distances. Depending on
-where one measures the distance and a line scan towards the interior
-would be shorter than a line scan towards the exterior. That said,
-increases and decreases of both proxies seem to mirror each other and to
-be well aligned.
+Figure~\ref{fig-Nac_Comp} shows the line scans of ND-1016-3 and
+NM-1016-3 in comparison to their δ\textsuperscript{18}O-values (similar
+graphs for the other \emph{Nacella} sp. specimen can be found in the
+supplementaries). Compared to the \emph{P. vulgata} records, these
+mirror each other much more clearly, due to the brief period recorded in
+the shells and its simple make-up.
 
@@
 
-\begin{figure}[H]
+The \emph{Nacella} sp. specimens showed similarly clustered
+correlations, which interestingly were not grouped by species. In fact,
+specimens ND-1016-3 (\emph{Nacella deaureata)} and NM-1016-3
+(\emph{Nacella magellanica}) are almost identical in their fitted
+equations (Figure~\ref{fig-Nac_Corr}). Both also share a high
+coefficient of determination (R\textsuperscript{2}) of 0.95 and 0.92
+respectively, with the remaining specimens being somewhat lower
+(R\textsuperscript{2} = 0.76 for ND-1016-4 and R\textsuperscript{2}=0.84
+for NM-1016-3). The mean R\textsuperscript{2} value for each
+\emph{Nacella} species is 0.86 for \emph{Nacella deaureata} and 0.88 for
+\emph{Nacella magellanica}. Why two specimens of different species are
+so well aligned is not entirely certain and randomness cannot for now be
+ruled out. That said, since they both have a high R\textsuperscript{2}
+value, it might just be that they were both minimally affected by
+factors other than SST, including the sampling location of our line
+scans.
 
-{\centering \includegraphics{Manuscript_files/figure-pdf/Nacella_Correlation_Graphs-1.pdf}
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Corr-1.pdf}
 
@@
 
-\caption{Correlation graphs for \emph{Nacella} sp. specimens}
+\caption{\label{fig-Nac_Corr}Correlation graphs for \emph{Nacella} sp.
+specimens}
 

### 2024-06-27 Manuscript_with_Heshkaia.qmd
first draft

--- 2024-06-26 Manuscript.qmd
@@
         postal-code: 27599-3315
-  - name: Francisco Zangrando
-    affiliations:
-      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
-        address: Avenida Maipú 305
-        city: Ushuaia
-        country: Argentina
-        postal-code: V9410BJA
-  - name: Angelica Tivoli
-    affiliations:
-      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
-        address: Avenida Maipú 305
-        city: Ushuaia
-        country: Argentina
-        postal-code: V9410BJA
   - name: Ivan Briz-Godino
@@
 *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results.
-In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens.
+In this study, we present elemental maps of various such species together with stable oxygen isotope values.
 Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios.
-By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P.
+By reassessing these previous specimens and including a wider range of modern and archaeological samples (n=12) from three patelloid species (*P.
 vulgata*, *N.
@@
 bibliography: bibliography.bib
-format: docx
-#  elsevier-pdf:
-#    keep-tex: true
-#    journal:
-#      name: Palaeo^3^
-#      formatting: preprint
-#      model: 3p
-#      cite-style: authoryear
+format:
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Journal Name
+      formatting: preprint
+      model: 3p
+      cite-style: authoryear
 output:
@@
 ---
-
-# Introduction {#Introduction}
 
@@
     cowplot,ggx,ggpubr,
-    patchwork,
+    patchwork,kableExtra,
     RColorBrewer)
@@
 
+# Introduction {#Introduction}
+
 Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source.
@@
 
-### Modern specimens
-
-The analysed specimens, their origin and respective studies with research background information can be found in [Table 1](#Table_1).
+The analysed specimens, their origin and respective studies with research background information can be found in Table[1](#Table_1).
 Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io].
 While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.
+
+```{r SST_Data}
+
+# Read the CSV files, skipping the comments
+Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
+TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")
+
+# Combine the datasets with new names and add year and month columns
+SST_df <- bind_rows(Westray_SST %>% mutate(dataset = "Westray - Orkney"),
+                         TdF_SST %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
+  mutate(analysed_sst = analysed_sst - 273.15,
+         time = as_datetime(time),
+         year = year(time),
+         month = month(time, label = TRUE)) %>%
+  group_by(dataset, year, month) %>%
+  summarise(mean_sst = mean(analysed_sst, na.rm = TRUE),
+            sd_sst = sd(analysed_sst, na.rm = TRUE))
+
+WestrayMin <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
+  pull(mean_sst) %>% min() %>% round(1)
+WestrayMax <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
+  pull(mean_sst) %>% max() %>% round(1)
+
+CamBayMin <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
+  pull(mean_sst) %>% min() %>% round(1)
+CamBayMax <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
+  pull(mean_sst) %>% max() %>% round(1)
+
+```
+
+There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK).
+Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016.
+The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih].
+SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.
+
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009.
+Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
+The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).
 
@@
 |                |                    |                  |                                        | QG1-7246-1 |                                                             |
-|                | This study         | *N. magellanica* | Heshkaia 35 in Moat Bay - Tierra del Fuego (AR) | Heshkaia_35-1 | *none*                                                      |
-|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-2 |                                                             |
-|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-3 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-4 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-5 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-6 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-7 |                                                             |
-|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-8 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-9 |                                                             |
 
@@
 
-```{r SST_Data}
-
-# Read the CSV files, skipping the comments
-Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
-TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")
-
-# Combine the datasets with new names and add year and month columns
-SST_df <- bind_rows(Westray_SST %>% mutate(dataset = "Westray - Orkney"),
-                         TdF_SST %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
-  mutate(analysed_sst = analysed_sst - 273.15,
-         time = as_datetime(time),
-         year = year(time),
-         month = month(time, label = TRUE)) %>%
-  group_by(dataset, year, month) %>%
-  summarise(mean_sst = mean(analysed_sst, na.rm = TRUE),
-            sd_sst = sd(analysed_sst, na.rm = TRUE))
-
-WestrayMin <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
-  pull(mean_sst) %>% min() %>% round(1)
-WestrayMax <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
-  pull(mean_sst) %>% max() %>% round(1)
-
-CamBayMin <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
-  pull(mean_sst) %>% min() %>% round(1)
-CamBayMax <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
-  pull(mean_sst) %>% max() %>% round(1)
-
-```
-
-There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK).
-Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016.
-The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih].
-SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.
-
-On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009.
-Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
-The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).
-
-```{r SST_plot}
-#| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
-#| fig-height: 5
+```{r SST_plot,fig.pos='H'}
+#| fig-cap: Monthly Sea Surface Temperature values averaged from  years. SST values were sourced from @Good2020-nl .
+#| fig-height: 3
 #| fig-width: 5
@@
 scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
-  labs(title = "Monthly Sea Surface Temperature",
+  labs(title = "",
        x = "Month",
-       y = "Mean Analysed SST (°C)") +
+       y = "Mean Analysed SST (°C)",colour="",fill="") +
     theme(axis.text.x = element_text(angle = 90, hjust = 1),
     legend.position = c(0.1, 0.9))+  
-  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) +
-  labs(colour="",fill="")
-
-```
-
-### Archaeological specimens
-
-In total, we studied 16 archaeological specimens.
-Of those, seven *P.
-vulgata* come from Quoygrew on Westray dating to 900 – 1200 CE.
+  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15))
+
+```
+
+In total, we studied 7 archaeological specimens of *P.
+vulgata* coming from Quoygrew on Westray dating to 900 – 1200 CE.
 These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba].
@@
 These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].
-
-Nine additional shells of *N.
-magellanica* (n=6) and *N*.
-*deaureata* (n=3) have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp.
-shells.
-These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells.
-They derive from the site of Heshkaia 35 situated within Moat Bay, around 30 km east of where the modern specimens in our study were collected [@Zangrando2014-qi; @Zangrando2021-qg].
-With radiocarbon dates putting the shells into the Late Holocene (1300 – 1450 CE) they are of similar age — if somewhat younger — as the assemblage from Quoygrew.
 
@@
 #| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay (*a-f*).
-#| fig-height: 10
-#| fig-width: 10
+#| fig-height: 7
+#| fig-width: 8
 #| label: fig-Nac_iso
@@
 *Patella* shells.
+Similar observations have been made before [@Graniero2017-io].
 
@@
 
-# Discussion
-
-## Other correlations
-
-```{r Ferguson Data}
-#| fig-cap: Correlation graphs for Ferguson et al. specimens
-#| fig-height: 20
-#| fig-width: 10
-
-Ferg_all <-
-  Ferguson_data %>% 
-  ggplot()+
-  aes(d18o,mg_ca,col=shell)+
-  geom_point(show.legend = FALSE)+
-  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
-  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
-  # stat_regline_equation(label.x = 2, #label.y = 32
-  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
-  #   formula = y~x
-  # ) +
-  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
-scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))
-
-# Expanded colour palette with 6 additional colours
-# Updated colour palette with 4 new replacements for the last colours
-# Refined colour palette with 3 additional colours
-# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
-
-Ferg_individual <- Ferg_all+
-  facet_wrap(~shell,ncol=2)+#ylim(0.15,NA)+
-  ggpubr::stat_regline_equation(label.x = 1, label.y = 26,
-    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
-    formula = y~x
-  )+
-  theme(
-    strip.background = element_rect(fill = "white", colour = NA),
-    strip.text = element_text(colour = "black"),
-    panel.border = element_rect(colour="black")
-  )
-
-Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
-
-```
-
-| Species       | Locality       | Specimen | Correlation R^2^        | Study                             |
-|---------------|---------------|---------------|---------------|---------------|
+# Discussion {#Discussion}
+
+The Mg/Ca ratios from the 2D elemental imaging present positive results on SST being the main factor behind Mg/Ca concentration in three Atlantic limpet species.
+Annually repeating patterns of increasing (spring/summer) and decreasing (autum/winter) SST were also visible in the line scans, albeit less straightforward than in the 2D-data.
+Nevertheless, the resulting R^2^-values for each species were high (*P.
+vulgata mean R^2^*=0.89, *N.
+deaureata mean R^2^*=0.86*, N.
+magellanica mean R^2^*=0.88).
+These results are similar to previously published data on *Patella* sp.
+shells (Table [2](#Table_2)).
+R^2^ values of *P.
+depressa* lie between 0.78 and 0.87 [@Garcia-Escarzaga2021-ij], for *P.
+caerulea* from the Mediterranean between 0.78 and 0.96 — although with one outlier of 0.33, which had increased Mg/Ca ratios in increments with rich organic contents [@Hausmann2019-fi].
+
+*P.
+rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.8, respectively [@Ferguson2011-zl] (@fig-Ferg).
+Together with the data from two *P.
+caerulea* shells (JM00 and JM30 with R^2^-values of 0.69 and 0.83), the data from JL2 was used determine a general Mg/Ca-SST relationship with an R^2^-value of 0.79.
+JL1 was disregarded in that calculation, as well as the second year of growth in JM00, which produced anomalous data.
+While no additional information about those specimens is available, it seems likely that the Mg/Ca ratios from JM00, which were interpreted as an ontogenetic trend, were potentially influenced by the same intra-increment heterogeneity we see in some of our shell specimens as well.
+Since these shells were analysed by homogenising the carbonate powder of the m+2 layer and the more exterior m+3 layer, these heterogeneities could potentially increased the Mg/Ca ratio unpredictably, leading to what might look like an ontogenetic trend in a linear dataset.
+Whether these anomalies affected JL1 in the same way, albeit stronger, is not clear but it might well be possible.
+
+::: {#Table_2}
+| Species       | Locality       | Specimen | R^2^                    | Study                             |
+|--------------|--------------|--------------|--------------|-----------------|
 | *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
@@
 |               |                | ISTPC2   | 0.84                    |                                   |
-|               | Crete          | AF1911A  | 0.91[^1]                |                                   |
-|               |                | AF3003A  | 0.92[^2]                |                                   |
+|               | Crete          | AF1911A  | 0.91                    |                                   |
+|               |                | AF3003A  | 0.92                    |                                   |
 |               | Israel         | AKKPC2   | 0.96                    |                                   |
@@
 | *P. rustica*  | Gibraltar      | JL1      | 0.02                    | [@Ferguson2011-zl]                |
-|               |                | JL2      | 0.8 (0.79)              |                                   |
-| *P. caerulea* | Gibraltar      | JM00     | 0.69 (0.79)             |                                   |
-|               |                | JM30     | 0.83 (0.79)             |                                   |
+|               |                | JL2      | 0.8\*\*                 |                                   |
+| *P. caerulea* | Gibraltar      | JM00     | 0.69\*\*                |                                   |
+|               |                | JM30     | 0.83\*\*                |                                   |
 | *P. vulgata*  | Orkney         | ORK-LT5  | not reported, here 0.88 | [@Graniero2017-io] and this study |
-|               |                |          |                         |                                   |
-|               |                |          |                         |                                   |
-|               |                |          |                         |                                   |
-
-: Table 2: Overview of comparative correlations {#tab:correlations}
-
-[^1]: SST only, no other geochemical data available
-
-[^2]: SST only, no geochemical data available
-
-## Comparison of ORK-LT5
-
-```{r}
-
-P1 <-
-  MgCa %>% 
-  filter(sampleID=="ORK_LT5") %>% 
-  # filter(std<0.1) %>% 
-  # filter(mg_ca<0.7) %>% 
-  # filter(dist<20) %>% 
-ggplot()+
-  aes(dist,mg_ca)+
-  geom_path(col="firebrick")+
-  # xlim(0,10)+
-  facet_wrap(~sampleID,scales = "free",nrow=1)
-
-P2 <-
-  D18O %>% 
-  filter(sampleID=="ORK_LT5") %>% 
-  ggplot()+
-  aes(dist,d18O)+
-  geom_path(col="cornflowerblue")+
-  scale_y_reverse()+
-  facet_wrap(~sampleID,scales = "free",nrow=1)
-
-plot_compare <- P1/P2
-
-plot_compare
-
-```
-
-```{r subsample}
+
+: Table showing correlations of Mg/Ca and δ^18^O found in other studies.
+\* indicates that only SST data and no other geochemical data was used.
+\*\* these samples were used to determine a shared R^2^-value of 0.79.
+:::
+
+```{r Ferguson Data}
+#| fig-cap: Correlation graphs for Ferguson et al., 2011 specimens.
+#| fig-height: 20
+#| fig-width: 10
+#| label: fig-Ferg
+
+Ferg_all <-
+  Ferguson_data %>% 
+  ggplot()+
+  aes(d18o,mg_ca,col=shell)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca (mmol/mol)")+
+  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
+  # stat_regline_equation(label.x = 2, #label.y = 32
+  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+  #   formula = y~x
+  # ) +
+  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
+scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))
+
+Ferg_individual <- Ferg_all+
+  facet_wrap(~shell,ncol=2)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 1, label.y = 26,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
+    formula = y~x
+  )+
+  theme(
+    strip.background = element_rect(fill = "white", colour = NA),
+    strip.text = element_text(colour = "black"),
+    panel.border = element_rect(colour="black")
+  )
+
+Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
+```
+
+Another shell whose previous results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with δ^18^O or SST in implication [@Graniero2017-io].
+Different to other specimens of this study, we were able to resample this specimen and getting Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
+These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August).
+Including these missing parts lead to a better alignment of both records and a resulting R^2^-value of 0.88.
+
+```{r ORK_LT5_Subsample}
+#| fig-cap: Comparison of previous and new Mg/Ca data for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca and δ^18^O-values
+#| fig-height: 10
+#| fig-width: 15
+#| label: fig-ORK_sub
 
@@
 
-# Apply the function to your data
-MgCa_filtered <- MgCa %>%
-  filter(sampleID == "ORK_LT5")
-
-MgCa_subsampled_150 <- subsample_data(MgCa_filtered,step_size = 0.15)
-
-MgCa_subsampled_300 <-  subsample_data(MgCa_filtered,step_size = 0.30)
-
-P3.1 <-
-  MgCa_subsampled_150 %>% 
-ggplot()+
-  aes(mean_dist,mean_mg_ca)+
-  geom_path(col="firebrick2")
-
-P3.2 <-
-  MgCa_subsampled_300 %>% 
-ggplot()+
-  aes(mean_dist,mean_mg_ca)+
-  geom_path(col="firebrick4")
-
-P1/P3.1/P3.2
-
-```
-
-```{r}
+#
 # This is data extracted from Fig. 6E of Graniero et al. 2016
@@
 
-```
-
-<!--# Maybe get the actual data from the image itself. -->
+# Apply the function to your data
+MgCa_filtered <- MgCa %>%
+  filter(sampleID == "ORK_LT5")
+
+MgCa_subsampled_150 <- subsample_data(MgCa_filtered,step_size = 0.15)
+
+MgCa_subsampled_300 <-  subsample_data(MgCa_filtered,step_size = 0.30)
+
+P3.1 <-  MgCa %>% 
+  filter(sampleID=="ORK_LT5") %>% 
+  filter(relstd < 0.1) %>%
+  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
+  ggplot()+
+  aes(dist,mg_ca)+
+  geom_path(show.legend = FALSE)+
+    scale_x_continuous(breaks = seq(0,12,by=2),
+                     labels = seq(0,12,by=2))+
+  theme_cowplot(20)+
+    annotate("rect", xmin = 1.5, xmax = 4, ymin = 0.3, ymax = 0.55, alpha = 0.2, fill = "grey50") +
+  annotate("rect", xmin = 0, xmax = 0.5, ymin = 0.3, ymax = 0.55, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
+  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
+
+# P3.2 <-
+#   MgCa_subsampled_300 %>% 
+# ggplot()+
+#   aes(mean_dist,mean_mg_ca)+
+#   geom_path(col="firebrick4")
+
+P3.3 <-
+  GranieroDF %>% 
+ggplot()+
+  aes(dist,mg_ca)+
+  geom_path(col="black")+
+   theme_cowplot(20)+
+  scale_x_continuous(breaks = seq(0,12,by=2),
+                     labels = seq(0,12,by=2))+
+  annotate("rect", xmin = 1.5, xmax = 3, ymin = 10, ymax = 25, alpha = 0.2, fill = "grey50") +
+  annotate("rect", xmin = 0, xmax = 0.08, ymin = 10, ymax = 25, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
+  labs(x = "Distance to edge [mm]", y = "Mg/Ca (mmol/mol)")
+
+P3.1/P3.3
+
+```
+
+While we were not able to resample the *Nacella* sp.
+specimens in Graniero *et al.* [-@Graniero2017-io], we were able to source other modern and archaeological specimens of the same species from Tierra del Fuego, which can serve as an indicator.
+Similar to other patelloid shells, the results of the *Nacella* specimens in our study were successful with high R^2^-values (0.76--0.95) and suggest Mg/Ca as a reliable proxy for SST (@fig-Nac_Comp and supplementaries).
+Why these results are different from the studied *Nacella* sp.
+specimens in Graniero et al.
+[-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and δ^18^O records of their shells (Figure 6 in [@Graniero2017-io] ) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — fell apart and lead to an overall result of an uncorrelated relationship between Mg/Ca and SST.
+
+# Conclusion
+
+This study presents high-resolution and 2D information on Mg/Ca concentrations in modern and archaeological patelloid shells and their relation to sea surface temperature.
+The results for *Patella vulgata*, *Nacella deaureata*, and *Nacella magellanica* were consistently promising with mean R^2^-values of 0.89, 0.86, and 0.88 respectively.
+
+LIBS elemental imaging provided some insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the δ^18^O of the same shells.
+Our 2D approach allowed us to avoid these areas of intra-increment heterogeneity and focus on areas that are predominantly controlled by changes in SST.
+
+These results are particularly promising for future analyses of patelloid shells for the study of season of collection in archaeological sites as well as the reconstruction of sub-annual climatic conditions in the past.
+Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a large part of marine mollusc remains in archaeological sites of those regions.
+Improving their analysis in terms of cost and time efficiency through the benefits of LIBS, which can be used in this context as a screening tool for targeted δ^18^O analysis focusing on annual minima and maxima revealed in the Mg/Ca record, allows researchers to increase sample numbers and produce larger and more in-depth climatic studies than previously possible.
 
 # References
-
-```{r Modern iso ranges}
-# Ranges <- D18O %>%
-#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
-# group_by(sampleID,period) %>%
-#   mutate(max=max(d18O),
-#          min=min(d18O)) %>%
-#   select(min, sampleID,max) %>% unique()
-# 
-# Modern_Ranges <- Ranges %>%
-#   filter(period=="Modern") %>%
-#   arrange(range)
-
-# Interestingly, the overall annual range in δ^18^O-values on Orkney (\~`r Modern_Ranges$range[3]`‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp.
-shells (`r min(Modern_Ranges$range)`‰ to `r max(Modern_Ranges$range)`‰) despite the different SST ranges.
-
-```
-
-```{r Correlation Graphs, eval=FALSE }
-#| fig-cap: Correlation graphs for all specimens
-#| fig-height: 20
-#| fig-width: 10
-
-P_all <- Corr_data %>% 
-  pivot_wider(values_from = "value",names_from = "proxy") %>% 
-  mutate(region = substr(sampleID, 1, 1)) %>% 
-  ggplot()+
-  aes(d18o,mg_ca,col=sampleID)+
-  geom_point(show.legend = FALSE)+
-  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
-  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
-  # stat_regline_equation(label.x = 2, #label.y = 32
-  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
-  #   formula = y~x
-  # ) +
-  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
-scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
-  # facet_wrap(~region,ncol=2,scales = "free")
-
-# Expanded colour palette with 6 additional colours
-# Updated colour palette with 4 new replacements for the last colours
-# Refined colour palette with 3 additional colours
-# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
-
-P_individual <- P_all+ylim(0,0.8)+
-  facet_wrap(~sampleID,ncol=3)+#ylim(0.15,NA)+
-  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 0.65,
-    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
-    formula = y~x
-  )+
-  theme(
-    strip.background = element_rect(fill = "white", colour = NA),
-    strip.text = element_text(colour = "black"),
-    panel.border = element_rect(colour="black")
-  )
-
-P_all/P_individual+plot_layout(heights = c(0.5,0.5))
-
-```

### 2024-06-27 Manuscript_with_Heshkaia.tex
first draft

--- 2024-06-26 Manuscript.tex
@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \makeatletter
@@
 \makeatother
-\journal{Palaeo3}
+\journal{Journal Name}
 \ifLuaTeX
@@
   pdftitle={Interspecies comparisons of Mg/Ca ratios in limpet shells},
-  pdfauthor={Niklas Hausmann; Donna Surge; Francisco Zangrando; Angelica Tivoli; Ivan Briz-Godino},
+  pdfauthor={Niklas Hausmann; Donna Surge; Ivan Briz-Godino},
   pdfkeywords={Sclerochronology, Limpets, Elemental Ratio, Mg/Ca},
@@
 
-\author[3]{Francisco Zangrando%
-%
-}
-
-\author[3]{Angelica Tivoli%
-%
-}
-
-\author[4]{Ivan Briz-Godino%
+\author[3]{Ivan Briz-Godino%
 %
@@
 NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
-\affiliation[3]{organization={CONICET (Consejo Nacional de
-Investigaciones Científicas y Técnicas)},addressline={Avenida Maipú
-305},city={Ushuaia},country={Argentina},countrysep={,},postcode={V9410BJA},postcodesep={}}
-\affiliation[4]{organization={},country={Spain},countrysep={,},postcodesep={}}
+\affiliation[3]{organization={},country={Spain},countrysep={,},postcodesep={}}
 
@@
 In this study, we present elemental maps of various such species
-together with stable oxygen isotope values for some of the specimens.
-Our dataset also includes specimens that were previously unsuccessful in
-providing significant correlations between δ\textsuperscript{18}O and
-Mg/Ca ratios. By reassessing these previous specimens and including a
-wider range of modern and archaeological samples from three patelloid
-species (\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N.
-magellanica}) we further add to the growing set of evidence for the
-reliable use of Mg/Ca ratios to detect palaeotemperature change and
-serve as a means to determine ontogenetic age and season of capture as
-well as to reveal locations of interest within the growth record
-(i.e.~annual temperature minima and maxima) for the targeted analysis
-using δ\textsuperscript{18}O or clumped oxygen isotope analysis.
+together with stable oxygen isotope values. Our dataset also includes
+specimens that were previously unsuccessful in providing significant
+correlations between δ\textsuperscript{18}O and Mg/Ca ratios. By
+reassessing these previous specimens and including a wider range of
+modern and archaeological samples (n=12) from three patelloid species
+(\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N. magellanica}) we
+further add to the growing set of evidence for the reliable use of Mg/Ca
+ratios to detect palaeotemperature change and serve as a means to
+determine ontogenetic age and season of capture as well as to reveal
+locations of interest within the growth record (i.e.~annual temperature
+minima and maxima) for the targeted analysis using
+δ\textsuperscript{18}O or clumped oxygen isotope analysis.
 \end{abstract}
@@
 
-\subsubsection{Modern specimens}\label{modern-specimens}
-
 The analysed specimens, their origin and respective studies with
-research background information can be found in \hyperref[Table_1]{Table
-1}. Here we will briefly summarise their contextual information, which
-can be accessed in more detail at the respective studies
+research background information can be found in
+Table\hyperref[Table_1]{1}. Here we will briefly summarise their
+contextual information, which can be accessed in more detail at the
+respective studies
 \citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While those
@@
 preservation did not all lend them to be re-analysed.
+
+There are two main locations that we sourced specimens from: the Beagle
+Channel in Tierra del Fuego (Argentina) and the island of West Ray in
+Orkney (UK). Four specimens were collected from the Beagle Channel in
+Cambaceres Bay in October 2016. The area receives around 570 mm
+precipitation but the channel waters are predominantly influenced by
+marine currents and in the period of October 2015 to October 2016 stayed
+around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
+August to 10ºC in February.
+
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
+August 2009. Westray lies about 70 km north of the Scottish mainland and
+experiences virtually no freshwater input with salinity values of 34.0
+-- 34.5 psu \citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in
+March to 13.8ºC in August (Figure~\ref{fig-SSTs}).
 
@@
 & & & & QG1-7246-1 & \\
-& This study & \emph{N. magellanica} & Heshkaia 35 in Moat Bay - Tierra
-del Fuego (AR) & Heshkaia\_35-1 & \emph{none} \\
-& & \emph{N. deaureata} & & Heshkaia\_35-2 & \\
-& & \emph{N. deaureata} & & Heshkaia\_35-3 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-4 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-5 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-6 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-7 & \\
-& & \emph{N. deaureata} & & Heshkaia\_35-8 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-9 & \\
 \end{longtable}
 
-There are two main locations that we sourced specimens from: the Beagle
-Channel in Tierra del Fuego (Argentina) and the island of West Ray in
-Orkney (UK). Four specimens were collected from the Beagle Channel in
-Cambaceres Bay in October 2016. The area receives around 570 mm
-precipitation but the channel waters are predominantly influenced by
-marine currents and in the period of October 2015 to October 2016 stayed
-around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
-August to 10ºC in February.
-
-On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
-August 2009. Westray lies about 70 km north of the Scottish mainland and
-experiences virtually no freshwater input with salinity values of 34.0
--- 34.5 psu \citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in
-March to 13.8ºC in August (Figure~\ref{fig-SSTs}).
-
-\begin{figure}
+\begin{figure}[H]
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-SSTs-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-SSTs-1.pdf}
 
@@
 
-\caption{\label{fig-SSTs}SST values averaged from years. SST values were
-sourced from \citet{Good2020-nl}}
+\caption{\label{fig-SSTs}Monthly Sea Surface Temperature values averaged
+from years. SST values were sourced from \citet{Good2020-nl} .}
 
@@
 
-\subsubsection{Archaeological specimens}\label{archaeological-specimens}
-
-In total, we studied 16 archaeological specimens. Of those, seven
-\emph{P. vulgata} come from Quoygrew on Westray dating to 900 -- 1200
-CE. These shells have previously been studied to characterise seasonal
-temperature change during the Medieval Climate Anomaly and the Little
-Ice Age \citep{Surge2012-ba}. Together they show a range of lifespans,
-with some comparatively long for their size (QG1-7246-1, 12 years within
-16 mm of growth) making them particularly susceptible to time-averaging.
-That said, the majority are under 5 years old. These shells are an ideal
+In total, we studied 7 archaeological specimens of \emph{P. vulgata}
+coming from Quoygrew on Westray dating to 900 -- 1200 CE. These shells
+have previously been studied to characterise seasonal temperature change
+during the Medieval Climate Anomaly and the Little Ice Age
+\citep{Surge2012-ba}. Together they show a range of lifespans, with some
+comparatively long for their size (QG1-7246-1, 12 years within 16 mm of
+growth) making them particularly susceptible to time-averaging. That
+said, the majority are under 5 years old. These shells are an ideal
 comparative dataset from an archaeological assemblage to compare to the
 modern shell ORK\_LT5 \citep{Graniero2017-io}.
-
-Nine additional shells of \emph{N. magellanica} (n=6) and \emph{N}.
-\emph{deaureata} (n=3) have been selected for LIBS analysis only, to
-provide a comparative dataset to the modern \emph{Nacella} sp. shells.
-These have not been analysed for δ\textsuperscript{18}O-values, but
-rather as a test to see whether patterns visible in the modern Mg/Ca
-ratio distributions are also preserved and apparent in the
-archaeological shells. They derive from the site of Heshkaia 35 situated
-within Moat Bay, around 30 km east of where the modern specimens in our
-study were collected \citep{Zangrando2014-qi, Zangrando2021-qg}. With
-radiocarbon dates putting the shells into the Late Holocene (1300 --
-1450 CE) they are of similar age --- if somewhat younger --- as the
-assemblage from Quoygrew.
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Nac_iso-1.pdf}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_iso-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Pat_iso-1.pdf}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_LIBS-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Pat_LIBS-1.pdf}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Comp-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Pat_Comp-1.pdf}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Corr-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Pat_Corr-1.pdf}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_LIBS-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Nac_LIBS-1.pdf}
 
@@
 still indicate lower concentrations than seen elsewhere, or seen only in
-aragonitic parts of e.g.~\emph{Patella} shells.
+aragonitic parts of e.g.~\emph{Patella} shells. Similar observations
+have been made before \citep{Graniero2017-io}.
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Comp-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Nac_Comp-1.pdf}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Corr-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Nac_Corr-1.pdf}
 
@@
 
-\section{Discussion}\label{discussion}
+\section{Discussion}\label{Discussion}
 
-\subsection{Other correlations}\label{other-correlations}
+The Mg/Ca ratios from the 2D elemental imaging present positive results
+on SST being the main factor behind Mg/Ca concentration in three
+Atlantic limpet species. Annually repeating patterns of increasing
+(spring/summer) and decreasing (autum/winter) SST were also visible in
+the line scans, albeit less straightforward than in the 2D-data.
+Nevertheless, the resulting R\textsuperscript{2}-values for each species
+were high (\emph{P. vulgata mean R\textsuperscript{2}}=0.89, \emph{N.
+deaureata mean R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
+R\textsuperscript{2}}=0.88). These results are similar to previously
+published data on \emph{Patella} sp. shells (Table
+\hyperref[Table_2]{2}). R\textsuperscript{2} values of \emph{P.
+depressa} lie between 0.78 and 0.87 \citep{Garcia-Escarzaga2021-ij}, for
+\emph{P. caerulea} from the Mediterranean between 0.78 and 0.96 ---
+although with one outlier of 0.33, which had increased Mg/Ca ratios in
+increments with rich organic contents \citep{Hausmann2019-fi}.
 
-\begin{figure}[H]
+\emph{P. rustica} has two studied specimens (JL1 and JL2) with
+R\textsuperscript{2}-values of 0.02 and 0.8, respectively
+\citep{Ferguson2011-zl} (Figure~\ref{fig-Ferg}). Together with the data
+from two \emph{P. caerulea} shells (JM00 and JM30 with
+R\textsuperscript{2}-values of 0.69 and 0.83), the data from JL2 was
+used determine a general Mg/Ca-SST relationship with an
+R\textsuperscript{2}-value of 0.79. JL1 was disregarded in that
+calculation, as well as the second year of growth in JM00, which
+produced anomalous data. While no additional information about those
+specimens is available, it seems likely that the Mg/Ca ratios from JM00,
+which were interpreted as an ontogenetic trend, were potentially
+influenced by the same intra-increment heterogeneity we see in some of
+our shell specimens as well. Since these shells were analysed by
+homogenising the carbonate powder of the m+2 layer and the more exterior
+m+3 layer, these heterogeneities could potentially increased the Mg/Ca
+ratio unpredictably, leading to what might look like an ontogenetic
+trend in a linear dataset. Whether these anomalies affected JL1 in the
+same way, albeit stronger, is not clear but it might well be possible.
 
-{\centering \includegraphics{Manuscript_files/figure-pdf/Ferguson Data-1.pdf}
-
-}
-
-\caption{Correlation graphs for Ferguson et al.~specimens}
-
-\end{figure}%
-
+\phantomsection\label{Table_2}
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
-\caption{Table 2: Overview of comparative correlations
-\{\#tab:correlations\}}\tabularnewline
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1918}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1918}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1918}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1918}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}@{}}
+\caption{Table showing correlations of Mg/Ca and δ\textsuperscript{18}O
+found in other studies. * indicates that only SST data and no other
+geochemical data was used. ** these samples were used to determine a
+shared R\textsuperscript{2}-value of 0.79.}\tabularnewline
 \toprule\noalign{}
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Correlation R\textsuperscript{2}
+R\textsuperscript{2}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
@@
 & & ISTPC2 & 0.84 & \\
-& Crete & AF1911A & 0.91\footnote{SST only, no other geochemical data
-  available} & \\
-& & AF3003A & 0.92\footnote{SST only, no geochemical data available}
-& \\
+& Crete & AF1911A & 0.91 & \\
+& & AF3003A & 0.92 & \\
 & Israel & AKKPC2 & 0.96 & \\
@@
 \emph{P. rustica} & Gibraltar & JL1 & 0.02 & \citep{Ferguson2011-zl} \\
-& & JL2 & 0.8 (0.79) & \\
-\emph{P. caerulea} & Gibraltar & JM00 & 0.69 (0.79) & \\
-& & JM30 & 0.83 (0.79) & \\
+& & JL2 & 0.8** & \\
+\emph{P. caerulea} & Gibraltar & JM00 & 0.69** & \\
+& & JM30 & 0.83** & \\
 \emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
 \citep{Graniero2017-io} and this study \\
-& & & & \\
-& & & & \\
-& & & & \\
 \end{longtable}
 
-\subsection{Comparison of ORK-LT5}\label{comparison-of-ork-lt5}
+\begin{figure}
 
-\includegraphics{Manuscript_files/figure-pdf/unnamed-chunk-1-1.pdf}
+\centering{
 
-\includegraphics{Manuscript_files/figure-pdf/subsample-1.pdf}
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Ferg-1.pdf}
+
+}
+
+\caption{\label{fig-Ferg}Correlation graphs for Ferguson et al., 2011
+specimens.}
+
+\end{figure}%
+
+Another shell whose previous results we aim to explain further is
+ORK-LT5, which has previously been shown to have no substantial
+correlation with δ\textsuperscript{18}O or SST in implication
+\citep{Graniero2017-io}. Different to other specimens of this study, we
+were able to resample this specimen and getting Mg/Ca data on a slightly
+higher resolution (20--30 µm spots in a 100 µm raster compared to 50 µm
+spots spaced at 150--300 µm) and using 2 dimensions, as well as using a
+continuous high-resolution line-scan at (20--30 µm spots spaced at 10
+µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
+previously missed summer peaks at around 2--3 mm distance to the shell
+edge (grey box in Figure~\ref{fig-ORK_sub}) and revealed additional
+growth at the very edge (dashed box in Figure~\ref{fig-ORK_sub})
+mirroring the change towards warmer temperatures at the time of
+collection (August). Including these missing parts lead to a better
+alignment of both records and a resulting R\textsuperscript{2}-value of
+0.88.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-ORK_sub-1.pdf}
+
+}
+
+\caption{\label{fig-ORK_sub}Comparison of previous and new Mg/Ca data
+for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca
+and δ\textsuperscript{18}O-values}
+
+\end{figure}%
+
+While we were not able to resample the \emph{Nacella} sp. specimens in
+Graniero \emph{et al.} \citeyearpar{Graniero2017-io}, we were able to
+source other modern and archaeological specimens of the same species
+from Tierra del Fuego, which can serve as an indicator. Similar to other
+patelloid shells, the results of the \emph{Nacella} specimens in our
+study were successful with high R\textsuperscript{2}-values (0.76--0.95)
+and suggest Mg/Ca as a reliable proxy for SST (Figure~\ref{fig-Nac_Comp}
+and supplementaries). Why these results are different from the studied
+\emph{Nacella} sp. specimens in Graniero et al.
+\citeyearpar{Graniero2017-io} is not obvious, but the occasional
+alignment of the Mg/Ca and δ\textsuperscript{18}O records of their
+shells (Figure 6 in \citep{Graniero2017-io} ) indicate that there is
+some relationship that --- potentially due to the low sampling
+resolution or intra-increment heterogeneity --- fell apart and lead to
+an overall result of an uncorrelated relationship between Mg/Ca and SST.
+
+\section{Conclusion}\label{conclusion}
+
+This study presents high-resolution and 2D information on Mg/Ca
+concentrations in modern and archaeological patelloid shells and their
+relation to sea surface temperature. The results for \emph{Patella
+vulgata}, \emph{Nacella deaureata}, and \emph{Nacella magellanica} were
+consistently promising with mean R\textsuperscript{2}-values of 0.89,
+0.86, and 0.88 respectively.
+
+LIBS elemental imaging provided some insight into elemental anomalies
+found particulary in the external parts (m+3 layer) of the shell, which
+introduce intra-incremental heterogeneity and which cannot be reconciled
+with modern SST changes or SST changes recorded in the
+δ\textsuperscript{18}O of the same shells. Our 2D approach allowed us to
+avoid these areas of intra-increment heterogeneity and focus on areas
+that are predominantly controlled by changes in SST.
+
+These results are particularly promising for future analyses of
+patelloid shells for the study of season of collection in archaeological
+sites as well as the reconstruction of sub-annual climatic conditions in
+the past. Shells of the species studied here are widely available along
+the European Atlantic façade and the Argentinian coastline and make up a
+large part of marine mollusc remains in archaeological sites of those
+regions. Improving their analysis in terms of cost and time efficiency
+through the benefits of LIBS, which can be used in this context as a
+screening tool for targeted δ\textsuperscript{18}O analysis focusing on
+annual minima and maxima revealed in the Mg/Ca record, allows
+researchers to increase sample numbers and produce larger and more
+in-depth climatic studies than previously possible.
 

### 2024-06-27 Manuscript.qmd
cleaned up format

@@
 *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results.
-In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens.
-Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios.
-By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P.
+In this study, we present elemental maps of various such species together with stable oxygen isotope values.
+Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between $\delta$^18^O and Mg/Ca ratios.
+By reassessing these previous specimens and including a wider range of modern and archaeological samples (n=12) from three patelloid species (*P.
 vulgata*, *N.
@@
 magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e.
-annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
+annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O or clumped oxygen isotope analysis.
 keywords: 
@@
 bibliography: bibliography.bib
-format: docx
-#  elsevier-pdf:
-#    keep-tex: true
-#    journal:
-#      name: Palaeo^3^
-#      formatting: preprint
-#      model: 3p
-#      cite-style: authoryear
+format:
+  elsevier-pdf:
+    keep-tex: true
+    latex_engine: xelatex
+    journal:
+      name: Palaeo^3^
+      formatting: preprint
+      model: 3p
+      cite-style: authoryear
 output:
   units: cm
+
 ---
-
-# Introduction {#Introduction}
 
@@
 
-  theme_set(theme_cowplot())
+  theme_set(theme_cowplot(10))
 
@@
 
+# Introduction {#Introduction}
+
 Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source.
 They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
-Determining past sea surface temperature (SST) change largely relies on the measurement of δ^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than δ^18^O-values, which are also affected by changes in salinity.
+Determining past sea surface temperature (SST) change largely relies on the measurement of $\delta$^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than $\delta$^18^O-values, which are also affected by changes in salinity.
 In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].
 
-Problematically, links between Mg/Ca ratios and SST changes in many marine mollusc shells, including limpets, have shown to be unreliable [@Surge2008-ri; @Wanamaker2008-zl; @Schone2010-yl; @Freitas2012-tx; @Graniero2015-zv; @Poulain2015-dg; @Vihtakari2017-wd].
+Problematically, links between Mg/Ca ratios and SST changes in many marine mollusc shells, including limpets, have shown to be unreliable [@Surge2008-ri; @Wanamaker2008-zl; @Schone2010-yl; @Freitas2012-tx; @Graniero2017-io; @Poulain2015-dg; @Vihtakari2017-wd].
 This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as temperature proxy.
@@
 
-By relying and adding onto published datasets, we were able to simultaneously avoid costs for new high-resolution δ^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
+By relying and adding onto published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
 Generally establishing the usefulness and reliability of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.
@@
 
-### Modern specimens
-
-The analysed specimens, their origin and respective studies with research background information can be found in [Table 1](#Table_1).
+The analysed specimens, their origin and respective studies with research background information can be found in Table[1](#Table_1).
 Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io].
 While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.
-
-::: {#Table_1}
-| Context        | Study              | Species          | Location                                        | Sample ID     | Previous analyses                                           |
-|------------|------------|------------|------------|------------|------------|
-| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay - Tierra del Fuego (AR)          | ND-1016-3     | Stable oxygen and carbon isotope analysis                   |
-|                |                    |                  |                                                 | ND-1016-4     |                                                             |
-|                |                    | *N. magellanica* |                                                 | NM-1016-1     |                                                             |
-|                |                    |                  |                                                 | NM-1016-3     |                                                             |
-|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - Westray (UK)                    | ORK-LT5       | Stable oxygen and carbon isotope analysis; Mg, Li, Sr, Ca,. |
-| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - Westray (UK)                    | QG1-7188-1    | Stable oxygen and carbon isotope analysis                   |
-|                |                    |                  |                                                 | QG1-7189-2    |                                                             |
-|                |                    |                  |                                                 | QG2-1061-1    |                                                             |
-|                |                    |                  |                                                 | QG2-1064-1    |                                                             |
-|                |                    |                  |                                                 | QG2-7180-1    |                                                             |
-|                |                    |                  |                                                 | QG2-7180-2    |                                                             |
-|                |                    |                  |                                                 | QG1-7246-1    |                                                             |
-|                | This study         | *N. magellanica* | Heshkaia 35 in Moat Bay - Tierra del Fuego (AR) | Heshkaia_35-1 | *none*                                                      |
-|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-2 |                                                             |
-|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-3 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-4 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-5 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-6 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-7 |                                                             |
-|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-8 |                                                             |
-|                |                    | *N. magellanica* |                                                 | Heshkaia_35-9 |                                                             |
-
-: Overview of the modern and archaeological limpet specimens analysed in this study.
-:::
 
@@
 
+::: {#Table_1}
+\fontsize{8pt}{8pt}\selectfont
+
+| Context        | Study              | Species          | Location        | Sample ID  | Previous analyses |
+|------|------------|-------|--------|------|----------|
+| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay  | ND-1016-3  | Stable oxygen     |
+|                |                    |                  | \- Tierra del   | ND-1016-4  | and carbon        |
+|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  | isotope analysis  |
+|                |                    |                  |                 | NM-1016-3  |                   |
+|                |                    |                  |                 |            |                   |
+|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | Stable oxygen     |
+|                |                    |                  | Westray (UK)    |            | and carbon        |
+|                |                    |                  |                 |            | isotope analysis; |
+|                |                    |                  |                 |            | Mg, Li, Sr, Ca.   |
+|                |                    |                  |                 |            |                   |
+| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | Stable oxygen     |
+|                |                    |                  | Westray (UK)    | QG1-7189-2 | and carbon        |
+|                |                    |                  |                 | QG2-1061-1 | isotope analysis  |
+|                |                    |                  |                 | QG2-1064-1 |                   |
+|                |                    |                  |                 | QG2-7180-1 |                   |
+|                |                    |                  |                 | QG2-7180-2 |                   |
+|                |                    |                  |                 | QG1-7246-1 |                   |
+
+: Overview of the modern and archaeological limpet specimens analysed in this study.
+:::
+
+\normalsize
 There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK).
@@
 
-```{r SST_plot}
-#| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
-#| fig-height: 5
-#| fig-width: 5
+```{r SST_plot,fig.pos='H'}
+#| fig-cap: Monthly Sea Surface Temperature values averaged from  years. SST values were sourced from @Good2020-nl.
+#| fig-height: 3
+#| fig-width: 6
 #| label: fig-SSTs
@@
 scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
-  labs(title = "Monthly Sea Surface Temperature",
+  labs(title = "",
        x = "Month",
-       y = "Mean Analysed SST (°C)") +
+       y = "Mean Analysed SST (°C)",colour="",fill="") +
     theme(axis.text.x = element_text(angle = 90, hjust = 1),
           legend.position = c(0.1, 0.9))+  
-  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) +
-  labs(colour="",fill="")
-
-```
-
-### Archaeological specimens
-
-In total, we studied 16 archaeological specimens.
-Of those, seven *P.
-vulgata* come from Quoygrew on Westray dating to 900 – 1200 CE.
+  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15))
+
+```
+
+In total, we studied 7 archaeological specimens of *P.
+vulgata* coming from Quoygrew on Westray dating to 900 – 1200 CE.
 These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba].
@@
 
-### Oxygen isotopes
-
-High-quality oxygen isotope values (δ^18^O) were acquired from previous publications on *P.
-vulgata* [@Surge2012-ba; @Graniero2017-io], *N.
-deaureata* and *N.
-magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found.
-In short, all shells were sectioned along the main growth axis to expose the internal growth structure.
-From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells.
-Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.
-
-@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential δ^18^O-values of limpet shells by species.
-Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp.
-shells living 2–3 years and *Patella v*.
-regularly over 3 years.
-In those instances where longer lifetimes are encountered (i.e.
-specimen QG1-7246-1, @fig-Pat_iso b).
-Interestingly, the annual ranges in δ^18^O-values are not similar between the coeval *Nacella* sp.
-specimens with δ^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 2.7‰ and 2.0‰.
-*P.
-vulgata* has similarly variable δ^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.
-
-```{r Nacella_Isotope_plot}
-#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay (*a-f*).
-#| fig-height: 10
+```{r Nacella_Isotope_plot,dev='png',dpi=300}
+#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay.
+#| fig-height: 5
 #| fig-width: 10
@@
       x = "Distance from margin in mm",
-      y = expression(delta^18*O~("\u2030")),
+      y = expression(paste(delta^{18},"O (\u2030)")),
       subtitle = sample_id,
@@
 # Combine plots using patchwork and add letters
-wrap_plots(plots,nrow=3) + plot_annotation(tag_levels = 'a')
-
-```
-
-```{r Patella_Isotope_plot}
+wrap_plots(plots,nrow=2) + plot_annotation(tag_levels = 'a')
+
+```
+
+### Oxygen isotopes
+
+High-quality oxygen isotope values ($\delta$^18^O) were acquired from previous publications on *P.
+vulgata* [@Surge2012-ba; @Graniero2017-io], *N.
+deaureata* and *N.
+magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found.
+In short, all shells were sectioned along the main growth axis to expose the internal growth structure.
+From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells.
+Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.
+
+@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species.
+Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp.
+shells living 2–3 years and *Patella v*.
+regularly over 3 years.
+In those instances where longer lifetimes are encountered (i.e.
+specimen QG1-7246-1, @fig-Pat_iso b).
+
+```{r Patella_Isotope_plot,dev='png',dpi=300,fig.pos='H'}
 #| fig-cap: Isotope values of *Patella vulgata* limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
@@
 
+ Interestingly, the annual ranges in $\delta$^18^O-values are not similar between the coeval *Nacella* sp.
+specimens with $\delta$^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 2.7‰ and 2.0‰.
+*P.
+vulgata* has similarly variable $\delta$^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.
+
 ## Dynamic Time Warping
 
-Because the exact location used in the analysis of stable oxygen isotopes is not possible to reconstruct for each carbonate powder sample, we used dynamic time warping (DTW) to align the time series of δ^18^O-values and Mg/Ca ratios.
+Because the exact location used in the analysis of stable oxygen isotopes is not possible to reconstruct for each carbonate powder sample, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O-values and Mg/Ca ratios.
 DTW is an algorithm that measures similarity between two proxy-sequences, which may vary in sampling resolution or interval.
@@
 The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.
-The code used for this analysis as well as the data required to run it can be found in this **OSF repository**.
 
@@
 
-```{r Patella_LIBS}
+```{r Patella_LIBS,fig.pos='H'}
 #| fig-cap: Example LIBS maps and line scans of analysed specimens.
@@
 Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
-#| fig-height: 10
-#| fig-width: 15
+#| fig-height: 5
+#| fig-width: 10
 #| label: fig-Pat_LIBS
@@
   geom_path(data=MgCa %>% filter(sampleID=="ORK_LT5"),aes(x,y),col="white",linetype="longdash")+
-  theme_cowplot(20)+
+  theme_cowplot(8)+
   scale_fill_viridis_c(limits = c(0.3, 0.7),
@@
                        breaks = seq(0.3, 0.6, by = 0.1)) +    
-  theme_cowplot(20)+
+  theme_cowplot(8)+
   labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
@@
   aes(x, y, col = mg_ca) +
-  geom_point(data = QG1_7188_1_map, col = "grey") +
-  geom_point(size=1,show.legend=FALSE) +
+  geom_point(data = QG1_7188_1_map, col = "grey",size=0.2) +
+  geom_point(size=0.2,show.legend=FALSE) +
   geom_path(data=MgCa %>% filter(sampleID=="QG1-7188-1"),aes(x,y),col="white",linetype="longdash")+
@@
                        breaks = seq(0.3, 0.6, by = 0.1)) +    
-  theme_cowplot(20)+
+  theme_cowplot(8)+
   theme(
@@
 Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental-imaging.
-That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (\@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D-image, showing the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].
-
-@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its δ^18^O-values (similar graphs for other specimen can be found in the supplementaries).
+That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D-image, showing the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].
+
+@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O-values (similar graphs for other specimen can be found in the supplementaries).
 The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have several different distances.
@@
 
-```{r Patella_Compare}
-#| fig-cap: Comparison of LIBS line-scan and sequential δ^18^O-values of specimen ORK-LT5 
-#| fig-height: 10
-#| fig-width: 20
+```{r Patella_Compare,dev='png',dpi=300}
+#| fig-cap: Comparison of LIBS line-scan and sequential $\delta$^18^O-values of specimen ORK-LT5 
+#| fig-height: 7
+#| fig-width: 16
 #| label: fig-Pat_Comp
@@
   geom_path()+
-  theme_cowplot(20)+
+  theme_cowplot(17)+
   labs(x = "", y = "Mg/Ca intensity ratio")
@@
   geom_path()+
-  theme_cowplot(20)+
+  theme_cowplot(18)+
   scale_y_reverse()+
-  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))
+  labs(x = "Distance to edge [mm]", y = expression(paste(delta^{18},"O (\u2030)")))
 
@@
 Aligning the records programmatically using dynamic time warping, allowed us to compare them directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr).
-The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters (δ^18^Ο = 0.6 – 0.065 \* Mg/Ca).
+The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 – 0.065 \* Mg/Ca).
 The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2).
@@
 
-```{r Patella_Correlation_Graphs }
+```{r Patella_Correlation_Graphs,dev='png',dpi=300}
 #| fig-cap: Correlation graphs for *Patella vulgata* specimens.
-#| fig-height: 18
-#| fig-width: 15
+#| fig-height: 8
+#| fig-width: 16
 #| label: fig-Pat_Corr
@@
   theme_cowplot(20)+
-  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
+  labs(x=expression(paste(delta^{18},"O (\u2030)")),y="Mg/Ca intensity ratio")+
   stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
@@
 
-P_individual <- P_all+ylim(0,0.8)+
+P_individual <- P_all+ylim(0.2,0.8)+
   facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
@@
 
-P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")
+#P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")
+P_individual
 
@@
 
-```{r Nacella_LIBS}
+```{r Nacella_LIBS,fig.pos='H'}
 #| fig-cap: Example LIBS maps and line scans of analysed specimens.
@@
 Note the inset in (*c*) where the shell edge has been resampled using a higher resolution (50 µm instead of 100 µm) to better understand the Mg/Ca ratios at the shell edge.
-#| fig-height: 10
-#| fig-width: 15
+#| fig-height: 7
+#| fig-width: 16
 #| label: fig-Nac_LIBS
@@
 ################
-  (ND_1016_plot+ND_1016_line)/(NM_1016_plot+NM_1016_line)+plot_annotation(tag_levels = "a")
+Maps <- ND_1016_plot|NM_1016_plot
+Lines <- ND_1016_line/NM_1016_line
+Maps+Lines+ plot_annotation(tag_levels = list(c("a", "b", "","c","d")))
+
+#(ND_1016_plot+ND_1016_line)/(NM_1016_plot+NM_1016_line)+plot_annotation(tag_levels = "a")
+
 ```
@@
 *Patella* shells.
-
-Line scans reflect well the changes previously indicated by δ^18^O values and also indicate about two years of recorded growth.
+Similar observations have been made before [@Graniero2017-io].
+
+Line scans reflect well the changes previously indicated by $\delta$^18^O values and also indicate about two years of recorded growth.
 That said, the 2D imaging suggests that in both specimens a third summer can be added to the total record, which is not captured by the line scans.
@@
 
-@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their δ^18^O-values (similar graphs for the other *Nacella* sp.
+@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their $\delta$^18^O-values (similar graphs for the other *Nacella* sp.
 specimen can be found in the supplementaries).
@@
 
-```{r Nacella_Compare}
-#| fig-cap: Comparison of LIBS line-scan and sequential δ^18^O-values of specimen ORK-LT5 
+```{r Nacella_Compare,dev='png',dpi=300}
+#| fig-cap: Comparison of LIBS line-scan and sequential $\delta$^18^O-values of specimen ORK-LT5 
 #| fig-height: 10
@@
   scale_y_reverse()+
-  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))
+  labs(x = "Distance to edge [mm]", y = expression(paste(delta^{18},"O (\u2030)")))
 
@@
 
-```{r Nacella_Correlation_Graphs}
+```{r Nacella_Correlation_Graphs,dev='png',dpi=300}
 #| fig-cap: Correlation graphs for *Nacella* sp. specimens
-#| fig-height: 18
-#| fig-width: 10
+#| fig-height: 5
+#| fig-width: 16
 #| label: fig-Nac_Corr
@@
   geom_point(show.legend = FALSE)+
-  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
+  labs(x=expression(paste(delta^{18},"O (\u2030)")),y="Mg/Ca intensity ratio")+
   stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
-  theme_cowplot(20)+
+  theme_cowplot(10)+
   # stat_regline_equation(label.x = 2, #label.y = 32
@@
 
+Nacella_Corr_data <- Corr_data %>% 
+  pivot_wider(values_from = "value",names_from = "proxy") %>% 
+  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
+  filter(grepl("Nacella", species))
+
 P_individual <- P_all+ylim(0,0.4)+
-  facet_wrap(~sampleID,ncol=2)+#ylim(0.15,NA)+
+  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
   ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.35,
@@
   theme_cowplot(20)+
+    coord_fixed(ratio = 6) +
   theme(
@@
 
-P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")
-
-```
-
-# Discussion
-
-## Other correlations
-
-```{r Ferguson Data}
-#| fig-cap: Correlation graphs for Ferguson et al. specimens
-#| fig-height: 20
-#| fig-width: 10
-
-Ferg_all <-
-  Ferguson_data %>% 
-  ggplot()+
-  aes(d18o,mg_ca,col=shell)+
-  geom_point(show.legend = FALSE)+
-  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
-  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
-  # stat_regline_equation(label.x = 2, #label.y = 32
-  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
-  #   formula = y~x
-  # ) +
-  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
-scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))
-
-# Expanded colour palette with 6 additional colours
-# Updated colour palette with 4 new replacements for the last colours
-# Refined colour palette with 3 additional colours
-# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
-
-Ferg_individual <- Ferg_all+
-  facet_wrap(~shell,ncol=2)+#ylim(0.15,NA)+
-  ggpubr::stat_regline_equation(label.x = 1, label.y = 26,
-    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
-    formula = y~x
-  )+
-  theme(
-    strip.background = element_rect(fill = "white", colour = NA),
-    strip.text = element_text(colour = "black"),
-    panel.border = element_rect(colour="black")
-  )
-
-Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
-
-```
-
-| Species       | Locality       | Specimen | Correlation R^2^        | Study                             |
-|---------------|---------------|---------------|---------------|---------------|
+P_individual
+#P_all/P_individual+plot_layout(heights = c(0.5,0.5))+plot_annotation(tag_levels = "a")
+
+```
+
+# Discussion {#Discussion}
+
+The Mg/Ca ratios from the 2D elemental imaging present positive results on SST being the main factor behind Mg/Ca concentration in three Atlantic limpet species.
+Annually repeating patterns of increasing (spring/summer) and decreasing (autum/winter) SST were also visible in the line scans, albeit less straightforward than in the 2D-data.
+Nevertheless, the resulting R^2^-values for each species were high (*P.
+vulgata mean R^2^*=0.89, *N.
+deaureata mean R^2^*=0.86*, N.
+magellanica mean R^2^*=0.88).
+These results are similar to previously published data on *Patella* sp.
+shells (Table [2](#Table_2)).
+R^2^ values of *P.
+depressa* lie between 0.78 and 0.87 [@Garcia-Escarzaga2021-ij], for *P.
+caerulea* from the Mediterranean between 0.78 and 0.96 — although with one outlier of 0.33 (specimen MP64A), which had increased Mg/Ca ratios in increments with rich organic contents [@Hausmann2019-fi].
+
+::: {#Table_2}
+\fontsize{8pt}{8pt}\selectfont
+
+| Species       | Locality       | Specimen | R^2^                    | Study                             |
+|----------|---------------|------------|----------|-------------------------|
 | *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
@@
 | *P. rustica*  | Gibraltar      | JL1      | 0.02                    | [@Ferguson2011-zl]                |
-|               |                | JL2      | 0.8 (0.79)              |                                   |
-| *P. caerulea* | Gibraltar      | JM00     | 0.69 (0.79)             |                                   |
-|               |                | JM30     | 0.83 (0.79)             |                                   |
+|               |                | JL2      | 0.8\*\*                 |                                   |
+| *P. caerulea* | Gibraltar      | JM00     | 0.69\*\*                |                                   |
+|               |                | JM30     | 0.83\*\*                |                                   |
 | *P. vulgata*  | Orkney         | ORK-LT5  | not reported, here 0.88 | [@Graniero2017-io] and this study |
-|               |                |          |                         |                                   |
-|               |                |          |                         |                                   |
-|               |                |          |                         |                                   |
-
-: Table 2: Overview of comparative correlations {#tab:correlations}
-
-[^1]: SST only, no other geochemical data available
-
-[^2]: SST only, no geochemical data available
-
-## Comparison of ORK-LT5
-
-```{r}
-
-P1 <-
-  MgCa %>% 
-  filter(sampleID=="ORK_LT5") %>% 
-  # filter(std<0.1) %>% 
-  # filter(mg_ca<0.7) %>% 
-  # filter(dist<20) %>% 
-ggplot()+
-  aes(dist,mg_ca)+
-  geom_path(col="firebrick")+
-  # xlim(0,10)+
-  facet_wrap(~sampleID,scales = "free",nrow=1)
-
-P2 <-
-  D18O %>% 
-  filter(sampleID=="ORK_LT5") %>% 
-  ggplot()+
-  aes(dist,d18O)+
-  geom_path(col="cornflowerblue")+
-  scale_y_reverse()+
-  facet_wrap(~sampleID,scales = "free",nrow=1)
-
-plot_compare <- P1/P2
-
-plot_compare
-
-```
-
-```{r subsample}
+
+: Table showing correlations of Mg/Ca and $\delta$^18^O found in other studies.
+\* indicates that only SST data and no other geochemical data was used.
+\*\* these samples were used to determine a shared R^2^-value of 0.79.
+:::
+\normalsize
+
+*P.
+rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.8, respectively [@Ferguson2011-zl] (@fig-Ferg).
+Together with the data from two *P.
+caerulea* shells (JM00 and JM30 with R^2^-values of 0.69 and 0.83), the data from JL2 was used determine a general Mg/Ca-SST relationship with an R^2^-value of 0.79.
+JL1 was disregarded in that calculation, as well as the second year of growth in JM00, which produced anomalous data.
+While no additional information about those specimens is available, it seems likely that the Mg/Ca ratios from JM00, which were interpreted as an ontogenetic trend, were potentially influenced by the same intra-increment heterogeneity we see in some of our shell specimens as well.
+Since these shells were analysed by homogenising the carbonate powder of the m+2 layer and the more exterior m+3 layer, these heterogeneities could potentially increased the Mg/Ca ratio unpredictably, leading to what might look like an ontogenetic trend in a linear dataset.
+Whether these anomalies affected JL1 in the same way, albeit stronger, is not clear but it might well be possible.
+
+```{r Ferguson Data, dev='png'}
+#| fig-cap: Correlation graphs for Ferguson et al., 2011 specimens.
+#| fig-height: 5
+#| fig-width: 16
+#| label: fig-Ferg
+
+Ferg_all <-
+  Ferguson_data %>% 
+  ggplot()+
+  aes(d18o,mg_ca,col=shell)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression(paste(delta^{18},"O (\u2030)")),y="Mg/Ca (mmol/mol)")+
+  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
+  # stat_regline_equation(label.x = 2, #label.y = 32
+  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+  #   formula = y~x
+  # ) +
+  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
+scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))
+
+Ferg_individual <- Ferg_all+
+  facet_wrap(~shell,ncol=4)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 26,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
+    formula = y~x
+  )+  
+  theme_cowplot(20)+
+  coord_fixed(ratio =0.22) +
+  theme(
+    strip.background = element_rect(fill = "white", colour = NA),
+    strip.text = element_text(colour = "black"),
+    panel.border = element_rect(colour="black")
+  )
+Ferg_individual
+#Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
+```
+
+Another shell whose previous results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with $\delta$^18^O or SST in implication [@Graniero2017-io].
+Different to other specimens of this study, we were able to resample this specimen and getting Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
+These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August).
+Including these missing parts lead to a better alignment of both records and a resulting R^2^-value of 0.88.
+
+```{r ORK_LT5_Subsample}
+#| fig-cap: Comparison of previous and new Mg/Ca data for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca and $\delta$^18^O-values
+#| fig-height: 4.5
+#| fig-width: 15
+#| label: fig-ORK_sub
 
@@
 
-```
-
-<!--# Maybe get the actual data from the image itself. -->
+# Apply the function to your data
+MgCa_filtered <- MgCa %>%
+  filter(sampleID == "ORK_LT5")
+
+MgCa_subsampled_150 <- subsample_data(MgCa_filtered,step_size = 0.15)
+
+MgCa_subsampled_300 <-  subsample_data(MgCa_filtered,step_size = 0.30)
+
+P3.1 <-  MgCa %>% 
+  filter(sampleID=="ORK_LT5") %>% 
+  filter(relstd < 0.1) %>%
+  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
+  ggplot()+
+  aes(dist,mg_ca)+
+  geom_path(show.legend = FALSE)+
+    scale_x_continuous(breaks = seq(0,12,by=2),
+                     labels = seq(0,12,by=2))+
+  theme_cowplot(12)+
+    annotate("rect", xmin = 1.5, xmax = 4, ymin = 0.3, ymax = 0.55, alpha = 0.2, fill = "grey50") +
+  annotate("rect", xmin = 0, xmax = 0.5, ymin = 0.3, ymax = 0.55, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
+  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
+
+# P3.2 <-
+#   MgCa_subsampled_300 %>% 
+# ggplot()+
+#   aes(mean_dist,mean_mg_ca)+
+#   geom_path(col="firebrick4")
+
+P3.3 <-
+  GranieroDF %>% 
+ggplot()+
+  aes(dist,mg_ca)+
+  geom_path(col="black")+
+   theme_cowplot(12)+
+  scale_x_continuous(breaks = seq(0,12,by=2),
+                     labels = seq(0,12,by=2))+
+  annotate("rect", xmin = 1.5, xmax = 3, ymin = 10, ymax = 25, alpha = 0.2, fill = "grey50") +
+  annotate("rect", xmin = 0, xmax = 0.08, ymin = 10, ymax = 25, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
+  labs(x = "Distance to edge [mm]", y = "Mg/Ca (mmol/mol)")
+
+P3.1/P3.3
+
+```
+
+While we were not able to resample the *Nacella* sp.
+specimens in Graniero *et al.* [-@Graniero2017-io], we were able to source other modern and archaeological specimens of the same species from Tierra del Fuego, which can serve as an indicator.
+Similar to other patelloid shells, the results of the *Nacella* specimens in our study were successful with high R^2^-values (0.76--0.95) and suggest Mg/Ca as a reliable proxy for SST (@fig-Nac_Comp and supplementaries).
+Why these results are different from the studied *Nacella* sp.
+specimens in Graniero et al.
+[-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and $\delta$^18^O records of their shells (Figure 6 in [@Graniero2017-io] ) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — fell apart and lead to an overall result of an uncorrelated relationship between Mg/Ca and SST.
+
+# Conclusion
+
+This study presents high-resolution and 2D information on Mg/Ca concentrations in modern and archaeological patelloid shells and their relation to sea surface temperature.
+The results for *Patella vulgata*, *Nacella deaureata*, and *Nacella magellanica* were consistently promising with mean R^2^-values of 0.89, 0.86, and 0.88 respectively.
+
+LIBS elemental imaging provided some insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the $\delta$^18^O of the same shells.
+Our 2D approach allowed us to avoid these areas of intra-increment heterogeneity and focus on areas that are predominantly controlled by changes in SST.
+
+These results are particularly promising for future analyses of patelloid shells for the study of season of collection in archaeological sites as well as the reconstruction of sub-annual climatic conditions in the past.
+Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a large part of marine mollusc remains in archaeological sites of those regions.
+Improving their analysis in terms of cost and time efficiency through the benefits of LIBS, which can be used in this context as a screening tool for targeted $\delta$^18^O analysis focusing on annual minima and maxima revealed in the Mg/Ca record, allows researchers to increase sample numbers and produce larger and more in-depth climatic studies than previously possible.
 
@@
 
-```{r Modern iso ranges}
-# Ranges <- D18O %>%
-#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
-# group_by(sampleID,period) %>%
-#   mutate(max=max(d18O),
-#          min=min(d18O)) %>%
-#   select(min, sampleID,max) %>% unique()
-# 
-# Modern_Ranges <- Ranges %>%
-#   filter(period=="Modern") %>%
-#   arrange(range)
-
-# Interestingly, the overall annual range in δ^18^O-values on Orkney (\~`r Modern_Ranges$range[3]`‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp.
-shells (`r min(Modern_Ranges$range)`‰ to `r max(Modern_Ranges$range)`‰) despite the different SST ranges.
-
-```
-
-```{r Correlation Graphs, eval=FALSE }
-#| fig-cap: Correlation graphs for all specimens
-#| fig-height: 20
-#| fig-width: 10
-
-P_all <- Corr_data %>% 
-  pivot_wider(values_from = "value",names_from = "proxy") %>% 
-  mutate(region = substr(sampleID, 1, 1)) %>% 
-  ggplot()+
-  aes(d18o,mg_ca,col=sampleID)+
-  geom_point(show.legend = FALSE)+
-  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
-  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
-  # stat_regline_equation(label.x = 2, #label.y = 32
-  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
-  #   formula = y~x
-  # ) +
-  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
-scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
-  # facet_wrap(~region,ncol=2,scales = "free")
-
-# Expanded colour palette with 6 additional colours
-# Updated colour palette with 4 new replacements for the last colours
-# Refined colour palette with 3 additional colours
-# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
-
-P_individual <- P_all+ylim(0,0.8)+
-  facet_wrap(~sampleID,ncol=3)+#ylim(0.15,NA)+
-  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 0.65,
-    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
-    formula = y~x
-  )+
-  theme(
-    strip.background = element_rect(fill = "white", colour = NA),
-    strip.text = element_text(colour = "black"),
-    panel.border = element_rect(colour="black")
-  )
-
-P_all/P_individual+plot_layout(heights = c(0.5,0.5))
-
-```
+```{r}
+#header stuff
+#include-in-header:
+#  text: |
+#    \usepackage{graphicx}
+#    \usepackage{booktabs}
+#    \usepackage{amsmath}
+#    \usepackage{unicode-math}
+#    \setmathfont{XITS Math}
+```

### 2024-06-27 Manuscript.tex
cleaned up format

@@
 In this study, we present elemental maps of various such species
-together with stable oxygen isotope values for some of the specimens.
-Our dataset also includes specimens that were previously unsuccessful in
-providing significant correlations between δ\textsuperscript{18}O and
-Mg/Ca ratios. By reassessing these previous specimens and including a
-wider range of modern and archaeological samples from three patelloid
-species (\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N.
-magellanica}) we further add to the growing set of evidence for the
-reliable use of Mg/Ca ratios to detect palaeotemperature change and
-serve as a means to determine ontogenetic age and season of capture as
-well as to reveal locations of interest within the growth record
-(i.e.~annual temperature minima and maxima) for the targeted analysis
-using δ\textsuperscript{18}O or clumped oxygen isotope analysis.
+together with stable oxygen isotope values. Our dataset also includes
+specimens that were previously unsuccessful in providing significant
+correlations between \(\delta\)\textsuperscript{18}O and Mg/Ca ratios.
+By reassessing these previous specimens and including a wider range of
+modern and archaeological samples (n=12) from three patelloid species
+(\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N. magellanica}) we
+further add to the growing set of evidence for the reliable use of Mg/Ca
+ratios to detect palaeotemperature change and serve as a means to
+determine ontogenetic age and season of capture as well as to reveal
+locations of interest within the growth record (i.e.~annual temperature
+minima and maxima) for the targeted analysis using
+\(\delta\)\textsuperscript{18}O or clumped oxygen isotope analysis.
 \end{abstract}
@@
 Determining past sea surface temperature (SST) change largely relies on
-the measurement of δ\textsuperscript{18}O-values within the calcitic
-parts of the shell, but attempts have been made to also use elemental
-ratios, such as magnesium to calcium (Mg/Ca), to have an alternative
-measure, that potentially provides a more accurate SST estimate than
-δ\textsuperscript{18}O-values, which are also affected by changes in
-salinity. In addition, the data acquisition of elemental ratios ---
-either through laser-ablation-isotope-ratio-mass-spectrometry
-(LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much
-faster and cost-effective, increasing the number of specimens that can
-be studied overall \citep{Durham2017-fh, Hausmann2023-ih}.
+the measurement of \(\delta\)\textsuperscript{18}O-values within the
+calcitic parts of the shell, but attempts have been made to also use
+elemental ratios, such as magnesium to calcium (Mg/Ca), to have an
+alternative measure, that potentially provides a more accurate SST
+estimate than \(\delta\)\textsuperscript{18}O-values, which are also
+affected by changes in salinity. In addition, the data acquisition of
+elemental ratios --- either through
+laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser
+induced breakdown spectroscopy (LIBS), can be much faster and
+cost-effective, increasing the number of specimens that can be studied
+overall \citep{Durham2017-fh, Hausmann2023-ih}.
 
@@
 simultaneously avoid costs for new high-resolution
-δ\textsuperscript{18}O-data, to use real-world examples, and to also
-provide pilot-data for areas of existing research interest. Generally
-establishing the usefulness and reliability of Mg/Ca as SST proxy in
-limpets should help to provide a platform for future research and an
-important stepping stone to better understand elemental ratios in other
-marine mollusc shells.
+\(\delta\)\textsuperscript{18}O-data, to use real-world examples, and to
+also provide pilot-data for areas of existing research interest.
+Generally establishing the usefulness and reliability of Mg/Ca as SST
+proxy in limpets should help to provide a platform for future research
+and an important stepping stone to better understand elemental ratios in
+other marine mollusc shells.
 
@@
 \phantomsection\label{Table_1}
+\fontsize{8pt}{8pt}\selectfont
+
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
 \caption{Overview of the modern and archaeological limpet specimens
@@
 Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
-- Tierra del Fuego (AR) & ND-1016-3 & Stable oxygen and carbon isotope
-analysis \\
-& & & & ND-1016-4 & \\
-& & \emph{N. magellanica} & & NM-1016-1 & \\
+& ND-1016-3 & Stable oxygen \\
+& & & - Tierra del & ND-1016-4 & and carbon \\
+& & \emph{N. magellanica} & Fuego (AR) & NM-1016-1 & isotope analysis \\
 & & & & NM-1016-3 & \\
-& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - Westray
-(UK) & ORK-LT5 & Stable oxygen and carbon isotope analysis; Mg, Li, Sr,
-Ca,. \\
+& & & & & \\
+& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - &
+ORK-LT5 & Stable oxygen \\
+& & & Westray (UK) & & and carbon \\
+& & & & & isotope analysis; \\
+& & & & & Mg, Li, Sr, Ca. \\
+& & & & & \\
 Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
-Bay - Westray (UK) & QG1-7188-1 & Stable oxygen and carbon isotope
-analysis \\
-& & & & QG1-7189-2 & \\
-& & & & QG2-1061-1 & \\
+Bay - & QG1-7188-1 & Stable oxygen \\
+& & & Westray (UK) & QG1-7189-2 & and carbon \\
+& & & & QG2-1061-1 & isotope analysis \\
 & & & & QG2-1064-1 & \\
@@
 & & & & QG1-7246-1 & \\
-& This study & \emph{N. magellanica} & Heshkaia 35 in Moat Bay - Tierra
-del Fuego (AR) & Heshkaia\_35-1 & \emph{none} \\
-& & \emph{N. deaureata} & & Heshkaia\_35-2 & \\
-& & \emph{N. deaureata} & & Heshkaia\_35-3 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-4 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-5 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-6 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-7 & \\
-& & \emph{N. deaureata} & & Heshkaia\_35-8 & \\
-& & \emph{N. magellanica} & & Heshkaia\_35-9 & \\
 \end{longtable}
+
+\normalsize
 
@@
 
-\begin{figure}
+\begin{figure}[H]
 
@@
 
-\caption{\label{fig-SSTs}SST values averaged from years. SST values were
-sourced from \citet{Good2020-nl}}
+\caption{\label{fig-SSTs}Monthly Sea Surface Temperature values averaged
+from years. SST values were sourced from \citet{Good2020-nl}.}
 
@@
 
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.png}
+
+}
+
+\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} sp. limpet
+shells from Cambaceres Bay.}
+
+\end{figure}%
+
 \subsubsection{Oxygen isotopes}\label{oxygen-isotopes}
 
-High-quality oxygen isotope values (δ\textsuperscript{18}O) were
-acquired from previous publications on \emph{P. vulgata}
+High-quality oxygen isotope values (\(\delta\)\textsuperscript{18}O)
+were acquired from previous publications on \emph{P. vulgata}
 \citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
@@
 Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
-previously acquired sequential δ\textsuperscript{18}O-values of limpet
-shells by species. Due to the high sampling resolution, all shells show
-seasonally resolved and multi-year quasi-sinusoidal records of SST
-change , with \emph{Nacella} sp. shells living 2--3 years and
+previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
+limpet shells by species. Due to the high sampling resolution, all
+shells show seasonally resolved and multi-year quasi-sinusoidal records
+of SST change , with \emph{Nacella} sp. shells living 2--3 years and
 \emph{Patella v}. regularly over 3 years. In those instances where
 longer lifetimes are encountered (i.e.~specimen QG1-7246-1,
-Figure~\ref{fig-Pat_iso} b). Interestingly, the annual ranges in
-δ\textsuperscript{18}O-values are not similar between the coeval
-\emph{Nacella} sp. specimens with δ\textsuperscript{18}O values for the
-previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for
-winter values between 2.7‰ and 2.0‰. \emph{P. vulgata} has similarly
-variable δ\textsuperscript{18}O-values for summer and winter seasons,
-alas the lack of more that one modern specimens, means that we cannot
-quantify the variability of recorded annual ranges between specimens.
+Figure~\ref{fig-Pat_iso} b).
 
-\begin{figure}
+\begin{figure}[H]
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.pdf}
-
-}
-
-\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} sp. limpet
-shells from Cambaceres Bay (\emph{a-f}).}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_iso-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_iso-1.png}
 
@@
 
+Interestingly, the annual ranges in
+\(\delta\)\textsuperscript{18}O-values are not similar between the
+coeval \emph{Nacella} sp. specimens with \(\delta\)\textsuperscript{18}O
+values for the previous summer temperatures ranging from 0.4‰ to above
+1.0‰, and for winter values between 2.7‰ and 2.0‰. \emph{P. vulgata} has
+similarly variable \(\delta\)\textsuperscript{18}O-values for summer and
+winter seasons, alas the lack of more that one modern specimens, means
+that we cannot quantify the variability of recorded annual ranges
+between specimens.
+
 \subsection{Dynamic Time Warping}\label{dynamic-time-warping}
@@
 sample, we used dynamic time warping (DTW) to align the time series of
-δ\textsuperscript{18}O-values and Mg/Ca ratios. DTW is an algorithm that
-measures similarity between two proxy-sequences, which may vary in
-sampling resolution or interval. By stretching or compressing sections
-of the series, DTW finds the probable alignment between the two
-sequences. This allows us to compare the proxy data sets more
-effectively, ensuring that the temporal dynamics of each shell are
+\(\delta\)\textsuperscript{18}O-values and Mg/Ca ratios. DTW is an
+algorithm that measures similarity between two proxy-sequences, which
+may vary in sampling resolution or interval. By stretching or
+compressing sections of the series, DTW finds the probable alignment
+between the two sequences. This allows us to compare the proxy data sets
+more effectively, ensuring that the temporal dynamics of each shell are
 accurately matched despite possible discrepancies in sampling intervals
@@
 alignment. The process involved iterative adjustments to minimise the
-overall distance between corresponding points in the two data sets. The
-code used for this analysis as well as the data required to run it can
-be found in this \textbf{OSF repository}.
+overall distance between corresponding points in the two data sets.
 
@@
 ratios within one season such as the summer period of QG1-7188-1 between
-0.5 and 2.0 mm distance to the shell edge (@fig-Pat\_LIBS d), is visible
-in a more dramatic manner than it appears in the 2D-image, showing the
-downside of line-scanning as opposed to the broader milling approach
-used for the analysis of carbonate powder \citep{Ferguson2011-zl}.
+0.5 and 2.0 mm distance to the shell edge (Figure~\ref{fig-Pat_LIBS} d),
+is visible in a more dramatic manner than it appears in the 2D-image,
+showing the downside of line-scanning as opposed to the broader milling
+approach used for the analysis of carbonate powder
+\citep{Ferguson2011-zl}.
 
 Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
-to its δ\textsuperscript{18}O-values (similar graphs for other specimen
-can be found in the supplementaries). The measurements of the distance
-to the shell edge are not entirely identical, most likely because one
-growth increment can have several different distances. Depending on
-where one measures the distance and a line scan towards the interior
-would be shorter than a line scan towards the exterior. That said,
-increases and decreases of both proxies seem to mirror each other and to
-be well aligned.
+to its \(\delta\)\textsuperscript{18}O-values (similar graphs for other
+specimen can be found in the supplementaries). The measurements of the
+distance to the shell edge are not entirely identical, most likely
+because one growth increment can have several different distances.
+Depending on where one measures the distance and a line scan towards the
+interior would be shorter than a line scan towards the exterior. That
+said, increases and decreases of both proxies seem to mirror each other
+and to be well aligned.
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Comp-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Comp-1.png}
 
@@
 \caption{\label{fig-Pat_Comp}Comparison of LIBS line-scan and sequential
-δ\textsuperscript{18}O-values of specimen ORK-LT5}
+\(\delta\)\textsuperscript{18}O-values of specimen ORK-LT5}
 
@@
 different but the majority (barring one: QG1-7246-1) do seem to cluster
-around shared parameters (δ\textsuperscript{18}Ο = 0.6 -- 0.065 *
-Mg/Ca). The various coefficients of determination (R\textsuperscript{2})
-range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens
-are also the oldest and youngest, respectively, specimens, suggesting
-that lower growth rates and time averaging had a negative effect on the
-correlation of both proxies. The mean R\textsuperscript{2} value is
-0.89, suggesting a generally good fit between the two proxies.
+around shared parameters (\(\delta\)\textsuperscript{18}Ο = 0.6 -- 0.065
+* Mg/Ca). The various coefficients of determination
+(R\textsuperscript{2}) range between 0.81 (QG1-7246-1) and 0.95
+(QG1-7189-2). These specimens are also the oldest and youngest,
+respectively, specimens, suggesting that lower growth rates and time
+averaging had a negative effect on the correlation of both proxies. The
+mean R\textsuperscript{2} value is 0.89, suggesting a generally good fit
+between the two proxies.
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Corr-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Corr-1.png}
 
@@
 still indicate lower concentrations than seen elsewhere, or seen only in
-aragonitic parts of e.g.~\emph{Patella} shells.
+aragonitic parts of e.g.~\emph{Patella} shells. Similar observations
+have been made before \citep{Graniero2017-io}.
 
 Line scans reflect well the changes previously indicated by
-δ\textsuperscript{18}O values and also indicate about two years of
-recorded growth. That said, the 2D imaging suggests that in both
+\(\delta\)\textsuperscript{18}O values and also indicate about two years
+of recorded growth. That said, the 2D imaging suggests that in both
 specimens a third summer can be added to the total record, which is not
@@
 Figure~\ref{fig-Nac_Comp} shows the line scans of ND-1016-3 and
-NM-1016-3 in comparison to their δ\textsuperscript{18}O-values (similar
-graphs for the other \emph{Nacella} sp. specimen can be found in the
-supplementaries). Compared to the \emph{P. vulgata} records, these
-mirror each other much more clearly, due to the brief period recorded in
-the shells and its simple make-up.
+NM-1016-3 in comparison to their \(\delta\)\textsuperscript{18}O-values
+(similar graphs for the other \emph{Nacella} sp. specimen can be found
+in the supplementaries). Compared to the \emph{P. vulgata} records,
+these mirror each other much more clearly, due to the brief period
+recorded in the shells and its simple make-up.
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Comp-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Comp-1.png}
 
@@
 \caption{\label{fig-Nac_Comp}Comparison of LIBS line-scan and sequential
-δ\textsuperscript{18}O-values of specimen ORK-LT5}
+\(\delta\)\textsuperscript{18}O-values of specimen ORK-LT5}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Corr-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Corr-1.png}
 
@@
 
-\section{Discussion}\label{discussion}
+\section{Discussion}\label{Discussion}
 
-\subsection{Other correlations}\label{other-correlations}
+The Mg/Ca ratios from the 2D elemental imaging present positive results
+on SST being the main factor behind Mg/Ca concentration in three
+Atlantic limpet species. Annually repeating patterns of increasing
+(spring/summer) and decreasing (autum/winter) SST were also visible in
+the line scans, albeit less straightforward than in the 2D-data.
+Nevertheless, the resulting R\textsuperscript{2}-values for each species
+were high (\emph{P. vulgata mean R\textsuperscript{2}}=0.89, \emph{N.
+deaureata mean R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
+R\textsuperscript{2}}=0.88). These results are similar to previously
+published data on \emph{Patella} sp. shells (Table
+\hyperref[Table_2]{2}). R\textsuperscript{2} values of \emph{P.
+depressa} lie between 0.78 and 0.87 \citep{Garcia-Escarzaga2021-ij}, for
+\emph{P. caerulea} from the Mediterranean between 0.78 and 0.96 ---
+although with one outlier of 0.33 (specimen MP64A), which had increased
+Mg/Ca ratios in increments with rich organic contents
+\citep{Hausmann2019-fi}.
 
-\begin{figure}[H]
-
-{\centering \includegraphics{Manuscript_files/figure-pdf/Ferguson Data-1.pdf}
-
-}
-
-\caption{Correlation graphs for Ferguson et al.~specimens}
-
-\end{figure}%
+\phantomsection\label{Table_2}
+\fontsize{8pt}{8pt}\selectfont
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
-\caption{Table 2: Overview of comparative correlations
-\{\#tab:correlations\}}\tabularnewline
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}@{}}
+\caption{Table showing correlations of Mg/Ca and
+\(\delta\)\textsuperscript{18}O found in other studies. * indicates that
+only SST data and no other geochemical data was used. ** these samples
+were used to determine a shared R\textsuperscript{2}-value of
+0.79.}\tabularnewline
 \toprule\noalign{}
@@
 \emph{P. rustica} & Gibraltar & JL1 & 0.02 & \citep{Ferguson2011-zl} \\
-& & JL2 & 0.8 (0.79) & \\
-\emph{P. caerulea} & Gibraltar & JM00 & 0.69 (0.79) & \\
-& & JM30 & 0.83 (0.79) & \\
+& & JL2 & 0.8** & \\
+\emph{P. caerulea} & Gibraltar & JM00 & 0.69** & \\
+& & JM30 & 0.83** & \\
 \emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
 \citep{Graniero2017-io} and this study \\
-& & & & \\
-& & & & \\
-& & & & \\
 \end{longtable}
 
-\subsection{Comparison of ORK-LT5}\label{comparison-of-ork-lt5}
+\normalsize
 
-\includegraphics{Manuscript_files/figure-pdf/unnamed-chunk-1-1.pdf}
+\emph{P. rustica} has two studied specimens (JL1 and JL2) with
+R\textsuperscript{2}-values of 0.02 and 0.8, respectively
+\citep{Ferguson2011-zl} (Figure~\ref{fig-Ferg}). Together with the data
+from two \emph{P. caerulea} shells (JM00 and JM30 with
+R\textsuperscript{2}-values of 0.69 and 0.83), the data from JL2 was
+used determine a general Mg/Ca-SST relationship with an
+R\textsuperscript{2}-value of 0.79. JL1 was disregarded in that
+calculation, as well as the second year of growth in JM00, which
+produced anomalous data. While no additional information about those
+specimens is available, it seems likely that the Mg/Ca ratios from JM00,
+which were interpreted as an ontogenetic trend, were potentially
+influenced by the same intra-increment heterogeneity we see in some of
+our shell specimens as well. Since these shells were analysed by
+homogenising the carbonate powder of the m+2 layer and the more exterior
+m+3 layer, these heterogeneities could potentially increased the Mg/Ca
+ratio unpredictably, leading to what might look like an ontogenetic
+trend in a linear dataset. Whether these anomalies affected JL1 in the
+same way, albeit stronger, is not clear but it might well be possible.
 
-\includegraphics{Manuscript_files/figure-pdf/subsample-1.pdf}
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Ferg-1.png}
+
+}
+
+\caption{\label{fig-Ferg}Correlation graphs for Ferguson et al., 2011
+specimens.}
+
+\end{figure}%
+
+Another shell whose previous results we aim to explain further is
+ORK-LT5, which has previously been shown to have no substantial
+correlation with \(\delta\)\textsuperscript{18}O or SST in implication
+\citep{Graniero2017-io}. Different to other specimens of this study, we
+were able to resample this specimen and getting Mg/Ca data on a slightly
+higher resolution (20--30 µm spots in a 100 µm raster compared to 50 µm
+spots spaced at 150--300 µm) and using 2 dimensions, as well as using a
+continuous high-resolution line-scan at (20--30 µm spots spaced at 10
+µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
+previously missed summer peaks at around 2--3 mm distance to the shell
+edge (grey box in Figure~\ref{fig-ORK_sub}) and revealed additional
+growth at the very edge (dashed box in Figure~\ref{fig-ORK_sub})
+mirroring the change towards warmer temperatures at the time of
+collection (August). Including these missing parts lead to a better
+alignment of both records and a resulting R\textsuperscript{2}-value of
+0.88.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-ORK_sub-1.pdf}
+
+}
+
+\caption{\label{fig-ORK_sub}Comparison of previous and new Mg/Ca data
+for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca
+and \(\delta\)\textsuperscript{18}O-values}
+
+\end{figure}%
+
+While we were not able to resample the \emph{Nacella} sp. specimens in
+Graniero \emph{et al.} \citeyearpar{Graniero2017-io}, we were able to
+source other modern and archaeological specimens of the same species
+from Tierra del Fuego, which can serve as an indicator. Similar to other
+patelloid shells, the results of the \emph{Nacella} specimens in our
+study were successful with high R\textsuperscript{2}-values (0.76--0.95)
+and suggest Mg/Ca as a reliable proxy for SST (Figure~\ref{fig-Nac_Comp}
+and supplementaries). Why these results are different from the studied
+\emph{Nacella} sp. specimens in Graniero et al.
+\citeyearpar{Graniero2017-io} is not obvious, but the occasional
+alignment of the Mg/Ca and \(\delta\)\textsuperscript{18}O records of
+their shells (Figure 6 in \citep{Graniero2017-io} ) indicate that there
+is some relationship that --- potentially due to the low sampling
+resolution or intra-increment heterogeneity --- fell apart and lead to
+an overall result of an uncorrelated relationship between Mg/Ca and SST.
+
+\section{Conclusion}\label{conclusion}
+
+This study presents high-resolution and 2D information on Mg/Ca
+concentrations in modern and archaeological patelloid shells and their
+relation to sea surface temperature. The results for \emph{Patella
+vulgata}, \emph{Nacella deaureata}, and \emph{Nacella magellanica} were
+consistently promising with mean R\textsuperscript{2}-values of 0.89,
+0.86, and 0.88 respectively.
+
+LIBS elemental imaging provided some insight into elemental anomalies
+found particulary in the external parts (m+3 layer) of the shell, which
+introduce intra-incremental heterogeneity and which cannot be reconciled
+with modern SST changes or SST changes recorded in the
+\(\delta\)\textsuperscript{18}O of the same shells. Our 2D approach
+allowed us to avoid these areas of intra-increment heterogeneity and
+focus on areas that are predominantly controlled by changes in SST.
+
+These results are particularly promising for future analyses of
+patelloid shells for the study of season of collection in archaeological
+sites as well as the reconstruction of sub-annual climatic conditions in
+the past. Shells of the species studied here are widely available along
+the European Atlantic façade and the Argentinian coastline and make up a
+large part of marine mollusc remains in archaeological sites of those
+regions. Improving their analysis in terms of cost and time efficiency
+through the benefits of LIBS, which can be used in this context as a
+screening tool for targeted \(\delta\)\textsuperscript{18}O analysis
+focusing on annual minima and maxima revealed in the Mg/Ca record,
+allows researchers to increase sample numbers and produce larger and
+more in-depth climatic studies than previously possible.
 

### 2024-06-27 Manuscript_with Heshkaia.qmd
cleaned up format

--- 2024-06-27 Manuscript_with_Heshkaia.qmd
@@
         postal-code: 27599-3315
+  - name: Francisco Zangrando
+    affiliations:
+      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
+        address: Avenida Maipú 305
+        city: Ushuaia
+        country: Argentina
+        postal-code: V9410BJA
+  - name: Angelica Tivoli
+    affiliations:
+      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
+        address: Avenida Maipú 305
+        city: Ushuaia
+        country: Argentina
+        postal-code: V9410BJA
   - name: Ivan Briz-Godino
@@
 *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results.
-In this study, we present elemental maps of various such species together with stable oxygen isotope values.
+In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens.
 Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios.
-By reassessing these previous specimens and including a wider range of modern and archaeological samples (n=12) from three patelloid species (*P.
+By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P.
 vulgata*, *N.
@@
 
-The analysed specimens, their origin and respective studies with research background information can be found in Table[1](#Table_1).
+### Modern specimens
+
+The analysed specimens, their origin and respective studies with research background information can be found in [Table 1](#Table_1).
 Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io].
 While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.
-
-```{r SST_Data}
-
-# Read the CSV files, skipping the comments
-Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
-TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")
-
-# Combine the datasets with new names and add year and month columns
-SST_df <- bind_rows(Westray_SST %>% mutate(dataset = "Westray - Orkney"),
-                         TdF_SST %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
-  mutate(analysed_sst = analysed_sst - 273.15,
-         time = as_datetime(time),
-         year = year(time),
-         month = month(time, label = TRUE)) %>%
-  group_by(dataset, year, month) %>%
-  summarise(mean_sst = mean(analysed_sst, na.rm = TRUE),
-            sd_sst = sd(analysed_sst, na.rm = TRUE))
-
-WestrayMin <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
-  pull(mean_sst) %>% min() %>% round(1)
-WestrayMax <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
-  pull(mean_sst) %>% max() %>% round(1)
-
-CamBayMin <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
-  pull(mean_sst) %>% min() %>% round(1)
-CamBayMax <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
-  pull(mean_sst) %>% max() %>% round(1)
-
-```
-
-There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK).
-Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016.
-The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih].
-SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.
-
-On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009.
-Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
-The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).
 
@@
 |                |                    |                  |                                                 | QG1-7246-1    |                                                             |
+|                | This study         | *N. magellanica* | Heshkaia 35 in Moat Bay - Tierra del Fuego (AR) | Heshkaia_35-1 | *none*                                                      |
+|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-2 |                                                             |
+|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-3 |                                                             |
+|                |                    | *N. magellanica* |                                                 | Heshkaia_35-4 |                                                             |
+|                |                    | *N. magellanica* |                                                 | Heshkaia_35-5 |                                                             |
+|                |                    | *N. magellanica* |                                                 | Heshkaia_35-6 |                                                             |
+|                |                    | *N. magellanica* |                                                 | Heshkaia_35-7 |                                                             |
+|                |                    | *N. deaureata*   |                                                 | Heshkaia_35-8 |                                                             |
+|                |                    | *N. magellanica* |                                                 | Heshkaia_35-9 |                                                             |
 
@@
 
-```{r SST_plot,fig.pos='H'}
-#| fig-cap: Monthly Sea Surface Temperature values averaged from  years. SST values were sourced from @Good2020-nl .
-#| fig-height: 3
+```{r SST_Data}
+
+# Read the CSV files, skipping the comments
+Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
+TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")
+
+# Combine the datasets with new names and add year and month columns
+SST_df <- bind_rows(Westray_SST %>% mutate(dataset = "Westray - Orkney"),
+                         TdF_SST %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
+  mutate(analysed_sst = analysed_sst - 273.15,
+         time = as_datetime(time),
+         year = year(time),
+         month = month(time, label = TRUE)) %>%
+  group_by(dataset, year, month) %>%
+  summarise(mean_sst = mean(analysed_sst, na.rm = TRUE),
+            sd_sst = sd(analysed_sst, na.rm = TRUE))
+
+WestrayMin <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
+  pull(mean_sst) %>% min() %>% round(1)
+WestrayMax <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
+  pull(mean_sst) %>% max() %>% round(1)
+
+CamBayMin <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
+  pull(mean_sst) %>% min() %>% round(1)
+CamBayMax <- SST_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
+  pull(mean_sst) %>% max() %>% round(1)
+
+```
+
+There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK).
+Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016.
+The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih].
+SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.
+
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009.
+Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
+The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).
+
+```{r SST_plot}
+#| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
+#| fig-height: 5
 #| fig-width: 5
@@
 scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
-  labs(title = "",
+  labs(title = "Monthly Sea Surface Temperature",
        x = "Month",
-       y = "Mean Analysed SST (°C)",colour="",fill="") +
+       y = "Mean Analysed SST (°C)") +
     theme(axis.text.x = element_text(angle = 90, hjust = 1),
     legend.position = c(0.1, 0.9))+  
-  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15))
-
-```
-
-In total, we studied 7 archaeological specimens of *P.
-vulgata* coming from Quoygrew on Westray dating to 900 – 1200 CE.
+  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) +
+  labs(colour="",fill="")
+
+```
+
+### Archaeological specimens
+
+In total, we studied 16 archaeological specimens.
+Of those, seven *P.
+vulgata* come from Quoygrew on Westray dating to 900 – 1200 CE.
 These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba].
@@
 These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].
+
+Nine additional shells of *N.
+magellanica* (n=6) and *N*.
+*deaureata* (n=3) have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp.
+shells.
+These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells.
+They derive from the site of Heshkaia 35 situated within Moat Bay, around 30 km east of where the modern specimens in our study were collected [@Zangrando2014-qi; @Zangrando2021-qg].
+With radiocarbon dates putting the shells into the Late Holocene (1300 – 1450 CE) they are of similar age — if somewhat younger — as the assemblage from Quoygrew.
 
@@
 #| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay (*a-f*).
-#| fig-height: 7
-#| fig-width: 8
+#| fig-height: 10
+#| fig-width: 10
 #| label: fig-Nac_iso
@@
 These results are similar to previously published data on *Patella* sp.
-shells (Table [2](#Table_2)).
+shells (@tbl-other_corrs).
 R^2^ values of *P.
@@
 *P.
-rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.8, respectively [@Ferguson2011-zl] (@fig-Ferg).
+rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.8, respectively [@Ferguson2011-zl] @fig-Ferg.
 Together with the data from two *P.
@@
 
-::: {#Table_2}
+```{r Other_Correlations}
+#| tbl-cap: Table showing correlations of Mg/Ca and δ^18^O found in other studies.
+* indicates that only SST data and no other geochemical data was used.
+** these samples were used to determine a shared R^2^-value of 0.79.
+#| label: tbl-other_corrs
+```
+
 | Species       | Locality       | Specimen | R^2^                    | Study                             |
-|--------------|--------------|--------------|--------------|-----------------|
+|-------------|-------------|-------------|-------------|----------------------|
 | *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
@@
 | *P. vulgata*  | Orkney         | ORK-LT5  | not reported, here 0.88 | [@Graniero2017-io] and this study |
-
-: Table showing correlations of Mg/Ca and δ^18^O found in other studies.
-\* indicates that only SST data and no other geochemical data was used.
-\*\* these samples were used to determine a shared R^2^-value of 0.79.
-:::
 
@@
 # References
+
+```{r Modern iso ranges, eval=FALSE}
+# Ranges <- D18O %>%
+#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
+# group_by(sampleID,period) %>%
+#   mutate(max=max(d18O),
+#          min=min(d18O)) %>%
+#   select(min, sampleID,max) %>% unique()
+# 
+# Modern_Ranges <- Ranges %>%
+#   filter(period=="Modern") %>%
+#   arrange(range)
+
+# Interestingly, the overall annual range in δ^18^O-values on Orkney (\~`r Modern_Ranges$range[3]`‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp.
+shells (`r min(Modern_Ranges$range)`‰ to `r max(Modern_Ranges$range)`‰) despite the different SST ranges.
+
+```
+
+```{r Correlation Graphs, eval=FALSE }
+#| fig-cap: Correlation graphs for all specimens
+#| fig-height: 20
+#| fig-width: 10
+
+P_all <- Corr_data %>% 
+  pivot_wider(values_from = "value",names_from = "proxy") %>% 
+  mutate(region = substr(sampleID, 1, 1)) %>% 
+  ggplot()+
+  aes(d18o,mg_ca,col=sampleID)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression("‰ δ" ^ "18" * "O [VPDB]"),y="Mg/Ca intensity ratio")+
+  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
+  # stat_regline_equation(label.x = 2, #label.y = 32
+  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+  #   formula = y~x
+  # ) +
+  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1")) +
+scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12", "#007f7f", "#1f78b4", "#99d8c9", "#542788","black", "#d73027", "#EC7063","pink1"))#+
+  # facet_wrap(~region,ncol=2,scales = "free")
+
+# Expanded colour palette with 6 additional colours
+# Updated colour palette with 4 new replacements for the last colours
+# Refined colour palette with 3 additional colours
+# c("#007f7f", "#d73027", "#ffc100", "black", "#1a9850", "#542788", "#3288bd", "#99d8c9")
+
+P_individual <- P_all+ylim(0,0.8)+
+  facet_wrap(~sampleID,ncol=3)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 0.65,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,
+    formula = y~x
+  )+
+  theme(
+    strip.background = element_rect(fill = "white", colour = NA),
+    strip.text = element_text(colour = "black"),
+    panel.border = element_rect(colour="black")
+  )
+
+P_all/P_individual+plot_layout(heights = c(0.5,0.5))
+
+```

### 2024-06-28 Extract data.Rmd
edits

--- 2024-05-30 including QG2-7246.Rmd
@@
 ```{r}
-QGShell <- read_csv(file=here("data","QG1-7246-1-A30Centre_analyzed_data.csv"), col_names = c("x", "y", "z", "mg_ca", "std", "relstd")) %>%
-  mutate(sampleID = "QG1-7246-1") %>%
-  filter(relstd<0.1,
-         x>1.3) %>% 
-  mutate(x=round(x,2)) %>% 
-  group_by(y) %>% 
-  mutate(mg_ca=mean(mg_ca),
-         std=mean(std),
-         n=n()) %>% 
-  ungroup() %>% 
-  filter(n>1) %>% 
-  select(sampleID,everything(),-n) %>% 
-  arrange(y) %>% 
-  mutate(dist=row_number()*0.03)
+MgCa <- read_csv(here("data","MgCa.csv"))
 
-QGShell %>% 
+NM_1016_1 <- MgCa %>% 
+  filter(sampleID=="NM-1016-1") #%>% 
+  # select(-sampleID,-dist)
+
+NM_1016_1 %>% 
+  filter(std<0.01) %>% 
   ggplot()+
-  aes(y,mg_ca)+
-  geom_line()
+  aes(x=dist,y=mg_ca)+
+  geom_path()
+
+# write_csv(QG1_7246,file = here("data","export","QG1_7246.csv"))
 
 ```
-
-```{r}
-
-write_csv(bind_rows(MgCa,QGShell),file=here("data","MgCa.csv"))
-
-```

### 2024-06-28 Manuscript.qmd
edits

@@
 ---
-title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
+title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
 author:
@@
     affiliations:
-      - name: 
-        address: 
-        city: 
+      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
+        address: Avinguda Diagonal, 643
+        city: Barcelona
         country: Spain
-        postal-code: 
+        postal-code: 08028 
 abstract: |
-  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures.
-While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species.
-*Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results.
-In this study, we present elemental maps of various such species together with stable oxygen isotope values.
-Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between $\delta$^18^O and Mg/Ca ratios.
-By reassessing these previous specimens and including a wider range of modern and archaeological samples (n=12) from three patelloid species (*P.
+  This study provides a reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST).
+While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this is not the case for other patelloid species.
+*Patella vulgata*, *Nacella deaureata*, and *Nacella magellanica* have been studied using Mg/Ca with mixed or contrary results.
+In this study, we present elemental maps of various such species together with stable oxygen isotope values that confirm a good relationship with SST.
+Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O and Mg/Ca ratios.
+By reassessing these species and including modern and archaeological samples (n=12) from three patelloid species (*P.
 vulgata*, *N.
 deaureata*, and *N.
-magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e.
+magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change.
+As a result, these species can in the future serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e.
 annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O or clumped oxygen isotope analysis.
@@
   - Sclerochronology
-  - Limpets
-  - Elemental Ratio
+  - Patella
+  - Nacella
+  - LIBS
   - Mg/Ca
@@
   units: cm
-
 ---
@@
 
-Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source.
+Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Robson2023-oz].
 They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
-Determining past sea surface temperature (SST) change largely relies on the measurement of $\delta$^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than $\delta$^18^O-values, which are also affected by changes in salinity.
+Determining past sea surface temperature (SST) change largely relies on the measurement of $\delta$^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure, that potentially provides a more accurate SST estimate than $\delta$^18^O-values, which are also affected by changes in salinity.
 In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].
@@
 This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as temperature proxy.
-Confusingly, multiple different temperature equations have been found for the same species (see [@Freitas2012-tx; @Vihtakari2017-wd] and references therein).
+Confusingly, in some cases multiple different temperature equations have been found for the same species (see Freitas *et al*.
+[-@Freitas2012-tx] and Vihtakari *et al*.
+[-@Vihtakari2017-wd] and references therein).
 Coeval specimens sharing one locality can also show differences in their relation to SST [@Hausmann2019-fi].
@@
 in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].
-There remains a lack of clarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2015-zv].
+There remains a lack of clarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2017-io].
 
@@
 LIBS allows us to carry out 2D imaging of entire shell sections, which helps us to navigate the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.
-
-By relying and adding onto published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
-Generally establishing the usefulness and reliability of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.
+By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
+Generally establishing the usefulness and reliability of Mg/Ca as SST proxy in limpets should help to provide a platform for such future research and an important stepping stone to a better understanding of elemental ratios in other marine mollusc shells.
 
@@
 
-The analysed specimens, their origin and respective studies with research background information can be found in Table[1](#Table_1).
+The analysed specimens, their origin and respective studies with research background information can be found in Table [1](#Table_1).
 Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io].
-While those studies also included other specimens, their accessibility or state of preservation did not all lend them to be re-analysed.
+While those studies also included other specimens, their accessibility or state of preservation did not allow them to be re-analysed.
 
@@
 \normalsize
+
 There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK).
-Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016.
+Four modern specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016.
 The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih].
-SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.
-
-On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009.
+SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February (@fig-SSTs - red).
+
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has previously been analysed using stable carbon and oxygen isotopes, as well as LA-ICP-MS [@Graniero2017-io].
 Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
-The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs).
+The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs - blue).
 
@@
 
-In total, we studied 7 archaeological specimens of *P.
+We also studied 7 archaeological specimens of *P.
 vulgata* coming from Quoygrew on Westray dating to 900 – 1200 CE.
@@
 That said, the majority are under 5 years old.
-These shells are an ideal comparative dataset from an archaeological assemblage to compare to the modern shell ORK_LT5 [@Graniero2017-io].
 
@@
 These involve the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume.
-This plume emits light, which is measured using a synchronised spectrometer.
-The resulting light spectrum quantifies the emission lines of magnesium (MgII; 279.553 nm) and calcium (CaII; 315.887 nm) to determine their intensity ratio.
+This plume emits light, which is measured using a synchronised spectrometer (delay time = 0.5 µs, integration time = 1 µs).
+The resulting light spectrum quantifies the emission lines of Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine their intensity ratio.
 While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g.
 [@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca changes within the shell carbonate.
-
-Using this system we carried out elemental imaging of the shell specimens at a resolution of 50 µm distance between sampling spots.
-Each spot was irradiated 10 times with the 3 first spectra used for cleaning steps and the remaining summed to get an average Mg/Ca intensity ratio for each sample spot.
+All values reported here at in arbitrary units based on the ratio of intensities of the peaks above.
+
+Using this system we carried out elemental imaging of the shell specimens at a resolution of 50--100 µm distance between sampling spots.
+Each spot was irradiated 10 times with the 3 first spectra discarded as a cleaning step and the remaining spectra summed to get an average Mg/Ca intensity ratio for each sample spot.
 Subsequently, we re-sampled the section using a line scan at 10 µm resolution.
@@
 more than 10%).
-These occurred in places where the previous sampling procedures for carbonate powder as part of the oxygen isotope analysis left an uneven sample surfaces introducing variability in the plasma generation and thus uncertainty in the data of these locations.
+These occurred in places where the previous sampling procedures for carbonate powder as part of the oxygen isotope analysis left an uneven sample surface introducing variability in the plasma generation and thus uncertainty in the data of these locations.
+
+### Oxygen isotopes
+
+High-quality oxygen isotope values ($\delta$^18^O) were acquired from previous publications on *P.
+vulgata* [@Surge2012-ba; @Graniero2017-io], *N.
+deaureata* and *N.
+magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found.
+In short, all shells were sectioned along the main growth axis to expose the internal growth structure.
+From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells.
+Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.
 
@@
 
-### Oxygen isotopes
-
-High-quality oxygen isotope values ($\delta$^18^O) were acquired from previous publications on *P.
-vulgata* [@Surge2012-ba; @Graniero2017-io], *N.
-deaureata* and *N.
-magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found.
-In short, all shells were sectioned along the main growth axis to expose the internal growth structure.
-From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells.
-Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.
-
-@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species.
-Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp.
-shells living 2–3 years and *Patella v*.
-regularly over 3 years.
+@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* sp., and *Patella vulgata*, respectively.
+Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* sp.
+shells living 2–3 years and *P.
+vulgata* living regularly over 3 years.
 In those instances where longer lifetimes are encountered (i.e.
-specimen QG1-7246-1, @fig-Pat_iso b).
+specimen QG1-7246-1, @fig-Pat_iso b) the temporal-resolution is reduced from more than a dozen of samples per year, to only around 7.
 
@@
 
- Interestingly, the annual ranges in $\delta$^18^O-values are not similar between the coeval *Nacella* sp.
-specimens with $\delta$^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 2.7‰ and 2.0‰.
-*P.
-vulgata* has similarly variable $\delta$^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.
-
 ## Dynamic Time Warping
 
-Because the exact location used in the analysis of stable oxygen isotopes is not possible to reconstruct for each carbonate powder sample, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O-values and Mg/Ca ratios.
+Because it is not possible to directly associate the exact sample locations used in the analysis of stable oxygen isotopes with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O-values and Mg/Ca ratios.
 DTW is an algorithm that measures similarity between two proxy-sequences, which may vary in sampling resolution or interval.
@@
 The layer m-1, which is also interior, consists of aragonite and is seen in @fig-Pat_LIBS in grey, as its Mg/Ca intensity ratio consistently falls below the range of interest (0.2 and above).
-Compared to other patelloid shells, whose interior is almost entirely made of aragonite, this layer is very thin.
+Compared to other patelloid shells, some of whose interior is almost entirely made of aragonite, this layer is very thin.
 
@@
 @fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O-values (similar graphs for other specimen can be found in the supplementaries).
-The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have several different distances.
-Depending on where one measures the distance and a line scan towards the interior would be shorter than a line scan towards the exterior.
+The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior.
 That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.
@@
 The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2).
-These specimens are also the oldest and youngest, respectively, specimens, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies.
-The mean R^2^ value is 0.89, suggesting a generally good fit between the two proxies.
+These specimens are also the oldest and youngest specimens, respectively, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies.
+The mean R^2^ value of the individual *P.
+vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies.
 
@@
 vulgata* or other *Patella* species [@Hausmann2019-fi].
-Nevertheless, repeating patterns were visible in the anterior section of the shell, which were also mirrored in the inner layers (m-2) of specimens NM-1016-3 (@fig-Nac_LIBS c).
+Nevertheless, repeating patterns were visible in the anterior section of the shell, which were also mirrored in the inner layers (m-2) of specimens NM-1016-3 (@fig-Nac_LIBS b).
 
 ```{r Nacella_LIBS,fig.pos='H'}
-#| fig-cap: Example LIBS maps and line scans of analysed specimens.
-(*a+b*) ND-1016-3 and (*c*+*d*) NM-1016-3.
+#| fig-cap: Example LIBS maps and line scans of analysed *Nacella* sp.
+specimens.
+(*a+c*) ND-1016-3 and (*b*+*d*) NM-1016-3.
 Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast.
@@
   scale_colour_viridis_c() +    
-  theme_cowplot(20)+
+  theme_cowplot(15)+
   labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
@@
 shells were much lower (0.05–0.30) than those seen in other patelloid species (e.g.
-0.3–0.7 in *Patella vulgata* above, or 0.5–1.5 in ) using the same emission lines.
-This is particularly visible in @fig-Nac_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in [@Hausmann2023-ih]).
-While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere, or seen only in aragonitic parts of e.g.
+0.3–0.7 in *Patella vulgata* above, or 0.5–1.5 in Hausmann *et al*.
+[-@Hausmann2023-ih]) using the same emission lines.
+While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the chosen Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere, or seen only in aragonitic parts of e.g.
 *Patella* shells.
@@
 | Species       | Locality       | Specimen | R^2^                    | Study                             |
-|----------|---------------|------------|----------|-------------------------|
+|-------------|-------------|-------------|-------------|--------------------|
 | *P. depressa* | Northern Spain | LAN541   | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
@@
 |               | Libya          | MO31A    | 0.83                    |                                   |
-|               |                | MP64A    | 0.33                    |                                   |
+|               |                | **MP64A**    | 0.33                    |                                   |
 |               |                | MP67A    | 0.96                    |                                   |
@@
 |               |                | KIZPC2   | 0.86                    |                                   |
-| *P. rustica*  | Gibraltar      | JL1      | 0.02                    | [@Ferguson2011-zl]                |
+| *P. rustica*  | Gibraltar      | **JL1**      | 0.02                    | [@Ferguson2011-zl]                |
 |               |                | JL2      | 0.8\*\*                 |                                   |
@@
 :::
+
 \normalsize

### 2024-06-28 Manuscript.tex
edits

@@
 \hypersetup{
-  pdftitle={Interspecies comparisons of Mg/Ca ratios in limpet shells},
+  pdftitle={Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells},
   pdfauthor={Niklas Hausmann; Donna Surge; Ivan Briz-Godino},
-  pdfkeywords={Sclerochronology, Limpets, Elemental Ratio, Mg/Ca},
+  pdfkeywords={Sclerochronology, Patella, Nacella, LIBS, Mg/Ca},
   colorlinks=true,
@@
 \begin{frontmatter}
-\title{Interspecies comparisons of Mg/Ca ratios in limpet shells}
+\title{Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic
+limpet shells}
 \author[1]{Niklas Hausmann%
@@
 NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
-\affiliation[3]{organization={},country={Spain},countrysep={,},postcodesep={}}
+\affiliation[3]{organization={Department of Evolutionary Biology,
+Ecology and Environmental Science, Biodiversity Research Institute
+(IRBio), University of Barcelona},addressline={Avinguda Diagonal,
+643},city={Barcelona},country={Spain},countrysep={,},postcode={8028},postcodesep={}}
 
@@
 \begin{abstract}
-This study provides a short reassessment of the use of Magnesium to
-Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea
-surface temperatures. While \emph{Patella vulgata} along the Spanish
-shoreline has since then repeatedly produced reliable correlations
-between sea surface temperature and Mg/Ca ratios, this is not the case
-for other patelloid species. \emph{Patella vulgata} and \emph{Nacella
-deaureata} have been studied using Mg/Ca with mixed or contrary results.
-In this study, we present elemental maps of various such species
-together with stable oxygen isotope values. Our dataset also includes
-specimens that were previously unsuccessful in providing significant
-correlations between \(\delta\)\textsuperscript{18}O and Mg/Ca ratios.
-By reassessing these previous specimens and including a wider range of
-modern and archaeological samples (n=12) from three patelloid species
-(\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N. magellanica}) we
-further add to the growing set of evidence for the reliable use of Mg/Ca
-ratios to detect palaeotemperature change and serve as a means to
-determine ontogenetic age and season of capture as well as to reveal
-locations of interest within the growth record (i.e.~annual temperature
-minima and maxima) for the targeted analysis using
-\(\delta\)\textsuperscript{18}O or clumped oxygen isotope analysis.
+This study provides a reassessment of the use of Magnesium to Calcium
+(Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface
+temperatures (SST). While \emph{Patella depressa} along the Spanish
+shoreline has repeatedly produced reliable correlations between SST and
+Mg/Ca ratios, this is not the case for other patelloid species.
+\emph{Patella vulgata}, \emph{Nacella deaureata}, and \emph{Nacella
+magellanica} have been studied using Mg/Ca with mixed or contrary
+results. In this study, we present elemental maps of various such
+species together with stable oxygen isotope values that confirm a good
+relationship with SST. Our dataset also reassesses a specimen which was
+previously unsuccessful in providing significant correlations between
+\(\delta\)\textsuperscript{18}O and Mg/Ca ratios. By reassessing these
+species and including modern and archaeological samples (n=12) from
+three patelloid species (\emph{P. vulgata}, \emph{N. deaureata}, and
+\emph{N. magellanica}) we further add to the growing set of evidence for
+the reliable use of Mg/Ca ratios to detect palaeotemperature change. As
+a result, these species can in the future serve as a means to determine
+ontogenetic age and season of capture as well as to reveal locations of
+interest within the growth record (i.e.~annual temperature minima and
+maxima) for the targeted analysis using \(\delta\)\textsuperscript{18}O
+or clumped oxygen isotope analysis.
 \end{abstract}
@@
 \begin{keyword}
-    Sclerochronology \sep Limpets \sep Elemental Ratio \sep 
+    Sclerochronology \sep Patella \sep Nacella \sep LIBS \sep 
     Mg/Ca
@@
 past shorelines due to their robust carbonate structure and the
-long-term use of limpets as a marine food source. They have successfully
-been studied in the past in the context of coastal subsistence
-economies, site occupation on seasonal
+long-term use of limpets as a marine food source \citep{Robson2023-oz}.
+They have successfully been studied in the past in the context of
+coastal subsistence economies, site occupation on seasonal
 \citep{Shackleton1973-ij, Parker2018-wf, Bosch2018-ud} or long-term
@@
 calcitic parts of the shell, but attempts have been made to also use
-elemental ratios, such as magnesium to calcium (Mg/Ca), to have an
-alternative measure, that potentially provides a more accurate SST
-estimate than \(\delta\)\textsuperscript{18}O-values, which are also
-affected by changes in salinity. In addition, the data acquisition of
-elemental ratios --- either through
-laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser
-induced breakdown spectroscopy (LIBS), can be much faster and
-cost-effective, increasing the number of specimens that can be studied
-overall \citep{Durham2017-fh, Hausmann2023-ih}.
+elemental ratios, such as magnesium to calcium (Mg/Ca)
+\citep{Graniero2017-io, Ferguson2011-zl}, to have an alternative
+measure, that potentially provides a more accurate SST estimate than
+\(\delta\)\textsuperscript{18}O-values, which are also affected by
+changes in salinity. In addition, the data acquisition of elemental
+ratios --- either through laser-ablation-isotope-ratio-mass-spectrometry
+(LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much
+faster and cost-effective, increasing the number of specimens that can
+be studied overall \citep{Durham2017-fh, Hausmann2023-ih}.
 
@@
 marine mollusc shells, including limpets, have shown to be unreliable
-\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2015-zv, Poulain2015-dg, Vihtakari2017-wd}.
+\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2017-io, Poulain2015-dg, Vihtakari2017-wd}.
 This is particularly the case where there is little available additional
@@
 which can independently and unpredictably affect Mg/Ca ratios and
-confound their interpretation as temperature proxy. Confusingly,
-multiple different temperature equations have been found for the same
-species (see \citep{Freitas2012-tx, Vihtakari2017-wd} and references
+confound their interpretation as temperature proxy. Confusingly, in some
+cases multiple different temperature equations have been found for the
+same species (see Freitas \emph{et al}. \citeyearpar{Freitas2012-tx} and
+Vihtakari \emph{et al}. \citeyearpar{Vihtakari2017-wd} and references
 therein). Coeval specimens sharing one locality can also show
@@
 particularly since it was last shown here, that they are not reliable
-recorders of palaeotemperature \citep{Graniero2015-zv}.
+recorders of palaeotemperature \citep{Graniero2017-io}.
 
@@
 temperature-related changes from the internal and less understood
-factors that influence the Mg/Ca ratio.
-
-By relying and adding onto published datasets, we were able to
-simultaneously avoid costs for new high-resolution
-\(\delta\)\textsuperscript{18}O-data, to use real-world examples, and to
-also provide pilot-data for areas of existing research interest.
-Generally establishing the usefulness and reliability of Mg/Ca as SST
-proxy in limpets should help to provide a platform for future research
-and an important stepping stone to better understand elemental ratios in
-other marine mollusc shells.
+factors that influence the Mg/Ca ratio. By working with already
+published datasets, we were able to simultaneously avoid costs for new
+high-resolution \(\delta\)\textsuperscript{18}O-data, to use real-world
+examples, and to also provide pilot-data for areas of existing research
+interest. Generally establishing the usefulness and reliability of Mg/Ca
+as SST proxy in limpets should help to provide a platform for such
+future research and an important stepping stone to a better
+understanding of elemental ratios in other marine mollusc shells.
 
@@
 The analysed specimens, their origin and respective studies with
-research background information can be found in
-Table\hyperref[Table_1]{1}. Here we will briefly summarise their
-contextual information, which can be accessed in more detail at the
-respective studies
-\citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While those
-studies also included other specimens, their accessibility or state of
-preservation did not all lend them to be re-analysed.
+research background information can be found in Table
+\hyperref[Table_1]{1}. Here we will briefly summarise their contextual
+information, which can be accessed in more detail at the respective
+studies \citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While
+those studies also included other specimens, their accessibility or
+state of preservation did not allow them to be re-analysed.
 
@@
 Channel in Tierra del Fuego (Argentina) and the island of West Ray in
-Orkney (UK). Four specimens were collected from the Beagle Channel in
-Cambaceres Bay in October 2016. The area receives around 570 mm
-precipitation but the channel waters are predominantly influenced by
+Orkney (UK). Four modern specimens were collected from the Beagle
+Channel in Cambaceres Bay in October 2016. The area receives around 570
+mm precipitation but the channel waters are predominantly influenced by
 marine currents and in the period of October 2015 to October 2016 stayed
 around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
-August to 10ºC in February.
+August to 10ºC in February (Figure~\ref{fig-SSTs} - red).
 
 On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
-August 2009. Westray lies about 70 km north of the Scottish mainland and
-experiences virtually no freshwater input with salinity values of 34.0
--- 34.5 psu \citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in
-March to 13.8ºC in August (Figure~\ref{fig-SSTs}).
+August 2009 and has previously been analysed using stable carbon and
+oxygen isotopes, as well as LA-ICP-MS \citep{Graniero2017-io}. Westray
+lies about 70 km north of the Scottish mainland and experiences
+virtually no freshwater input with salinity values of 34.0 -- 34.5 psu
+\citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in March to
+13.8ºC in August (Figure~\ref{fig-SSTs} - blue).
 
@@
 
-In total, we studied 7 archaeological specimens of \emph{P. vulgata}
-coming from Quoygrew on Westray dating to 900 -- 1200 CE. These shells
-have previously been studied to characterise seasonal temperature change
+We also studied 7 archaeological specimens of \emph{P. vulgata} coming
+from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
+previously been studied to characterise seasonal temperature change
 during the Medieval Climate Anomaly and the Little Ice Age
@@
 growth) making them particularly susceptible to time-averaging. That
-said, the majority are under 5 years old. These shells are an ideal
-comparative dataset from an archaeological assemblage to compare to the
-modern shell ORK\_LT5 \citep{Graniero2017-io}.
+said, the majority are under 5 years old.
 
@@
 plasma plume. This plume emits light, which is measured using a
-synchronised spectrometer. The resulting light spectrum quantifies the
-emission lines of magnesium (MgII; 279.553 nm) and calcium (CaII;
-315.887 nm) to determine their intensity ratio. While these two peaks
-alone do not represent the molar concentrations of both elements (as
-opposed to calibration free LIBS; e.g.
-\citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
-correlated to the molar concentration \citep{Hausmann2017-oa} and can be
-used as a reliable indicator of Mg/Ca changes within the shell
-carbonate.
+synchronised spectrometer (delay time = 0.5 µs, integration time = 1
+µs). The resulting light spectrum quantifies the emission lines of
+Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine
+their intensity ratio. While these two peaks alone do not represent the
+molar concentrations of both elements (as opposed to calibration free
+LIBS; e.g. \citep{Martinez-Minchero2022-jz}), the intensity ratio is
+linearly correlated to the molar concentration \citep{Hausmann2017-oa}
+and can be used as a reliable indicator of Mg/Ca changes within the
+shell carbonate. All values reported here at in arbitrary units based on
+the ratio of intensities of the peaks above.
 
 Using this system we carried out elemental imaging of the shell
-specimens at a resolution of 50 µm distance between sampling spots. Each
-spot was irradiated 10 times with the 3 first spectra used for cleaning
-steps and the remaining summed to get an average Mg/Ca intensity ratio
-for each sample spot. Subsequently, we re-sampled the section using a
-line scan at 10 µm resolution. This leads to an overlap between sample
-spots, but also allows for a continuous record without gaps. Intensity
-ratios were filtered for cases with high relative standard deviation
-(i.e.~more than 10\%). These occurred in places where the previous
-sampling procedures for carbonate powder as part of the oxygen isotope
-analysis left an uneven sample surfaces introducing variability in the
-plasma generation and thus uncertainty in the data of these locations.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.png}
-
-}
-
-\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} sp. limpet
-shells from Cambaceres Bay.}
-
-\end{figure}%
+specimens at a resolution of 50--100 µm distance between sampling spots.
+Each spot was irradiated 10 times with the 3 first spectra discarded as
+a cleaning step and the remaining spectra summed to get an average Mg/Ca
+intensity ratio for each sample spot. Subsequently, we re-sampled the
+section using a line scan at 10 µm resolution. This leads to an overlap
+between sample spots, but also allows for a continuous record without
+gaps. Intensity ratios were filtered for cases with high relative
+standard deviation (i.e.~more than 10\%). These occurred in places where
+the previous sampling procedures for carbonate powder as part of the
+oxygen isotope analysis left an uneven sample surface introducing
+variability in the plasma generation and thus uncertainty in the data of
+these locations.
 
@@
 
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.png}
+
+}
+
+\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} sp. limpet
+shells from Cambaceres Bay.}
+
+\end{figure}%
+
 Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
 previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
-limpet shells by species. Due to the high sampling resolution, all
-shells show seasonally resolved and multi-year quasi-sinusoidal records
-of SST change , with \emph{Nacella} sp. shells living 2--3 years and
-\emph{Patella v}. regularly over 3 years. In those instances where
-longer lifetimes are encountered (i.e.~specimen QG1-7246-1,
-Figure~\ref{fig-Pat_iso} b).
+limpet shells by species (\emph{Nacella} sp., and \emph{Patella
+vulgata}, respectively. Due to the high sampling resolution, all shells
+show seasonally resolved and multi-year quasi-sinusoidal records of SST
+change, with \emph{Nacella} sp. shells living 2--3 years and \emph{P.
+vulgata} living regularly over 3 years. In those instances where longer
+lifetimes are encountered (i.e.~specimen QG1-7246-1,
+Figure~\ref{fig-Pat_iso} b) the temporal-resolution is reduced from more
+than a dozen of samples per year, to only around 7.
 
@@
 
-Interestingly, the annual ranges in
-\(\delta\)\textsuperscript{18}O-values are not similar between the
-coeval \emph{Nacella} sp. specimens with \(\delta\)\textsuperscript{18}O
-values for the previous summer temperatures ranging from 0.4‰ to above
-1.0‰, and for winter values between 2.7‰ and 2.0‰. \emph{P. vulgata} has
-similarly variable \(\delta\)\textsuperscript{18}O-values for summer and
-winter seasons, alas the lack of more that one modern specimens, means
-that we cannot quantify the variability of recorded annual ranges
-between specimens.
-
 \subsection{Dynamic Time Warping}\label{dynamic-time-warping}
 
-Because the exact location used in the analysis of stable oxygen
-isotopes is not possible to reconstruct for each carbonate powder
-sample, we used dynamic time warping (DTW) to align the time series of
+Because it is not possible to directly associate the exact sample
+locations used in the analysis of stable oxygen isotopes with the LIBS
+data, we used dynamic time warping (DTW) to align the time series of
 \(\delta\)\textsuperscript{18}O-values and Mg/Ca ratios. DTW is an
@@
 Mg/Ca intensity ratio consistently falls below the range of interest
-(0.2 and above). Compared to other patelloid shells, whose interior is
-almost entirely made of aragonite, this layer is very thin.
+(0.2 and above). Compared to other patelloid shells, some of whose
+interior is almost entirely made of aragonite, this layer is very thin.
 
@@
 distance to the shell edge are not entirely identical, most likely
-because one growth increment can have several different distances.
-Depending on where one measures the distance and a line scan towards the
-interior would be shorter than a line scan towards the exterior. That
-said, increases and decreases of both proxies seem to mirror each other
-and to be well aligned.
+because one growth increment can have a range of distances to the edge,
+depending on where one measures, and a line scan along the interior
+would be shorter than a line scan along the exterior. That said,
+increases and decreases of both proxies seem to mirror each other and to
+be well aligned.
 
@@
 (R\textsuperscript{2}) range between 0.81 (QG1-7246-1) and 0.95
-(QG1-7189-2). These specimens are also the oldest and youngest,
-respectively, specimens, suggesting that lower growth rates and time
+(QG1-7189-2). These specimens are also the oldest and youngest
+specimens, respectively, suggesting that lower growth rates and time
 averaging had a negative effect on the correlation of both proxies. The
-mean R\textsuperscript{2} value is 0.89, suggesting a generally good fit
-between the two proxies.
+mean R\textsuperscript{2} value of the individual \emph{P. vulgata}
+specimens is 0.89, suggesting a generally good fit between the two
+proxies.
 
@@
 which were also mirrored in the inner layers (m-2) of specimens
-NM-1016-3 (Figure~\ref{fig-Nac_LIBS} c).
+NM-1016-3 (Figure~\ref{fig-Nac_LIBS} b).
 
@@
 \caption{\label{fig-Nac_LIBS}Example LIBS maps and line scans of
-analysed specimens. (\emph{a+b}) ND-1016-3 and (\emph{c}+\emph{d})
-NM-1016-3. Note that the ranges of Mg/Ca intensity ratios are not
-identical but are chosen arbitrarily to increase contrast. Areas in grey
-consist of low-Mg aragonite and fall outside the chosen colour-range.
-Note the inset in (\emph{c}) where the shell edge has been resampled
-using a higher resolution (50 µm instead of 100 µm) to better understand
-the Mg/Ca ratios at the shell edge.}
+analysed \emph{Nacella} sp. specimens. (\emph{a+c}) ND-1016-3 and
+(\emph{b}+\emph{d}) NM-1016-3. Note that the ranges of Mg/Ca intensity
+ratios are not identical but are chosen arbitrarily to increase
+contrast. Areas in grey consist of low-Mg aragonite and fall outside the
+chosen colour-range. Note the inset in (\emph{c}) where the shell edge
+has been resampled using a higher resolution (50 µm instead of 100 µm)
+to better understand the Mg/Ca ratios at the shell edge.}
 
@@
 shells were much lower (0.05--0.30) than those seen in other patelloid
-species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in )
-using the same emission lines. This is particularly visible in
-Figure~\ref{fig-Nac_LIBS} c, where Mg/Ca intensity ratios range from
-\textasciitilde0.5--0.7 in \citep{Hausmann2023-ih}). While the intensity
-ratio depends chiefly on the chosen emission lines to calculate the
-proportion of the chosen Magnesium to the Calcium peak, the low ratios
-still indicate lower concentrations than seen elsewhere, or seen only in
-aragonitic parts of e.g.~\emph{Patella} shells. Similar observations
-have been made before \citep{Graniero2017-io}.
+species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in
+Hausmann \emph{et al}. \citeyearpar{Hausmann2023-ih}) using the same
+emission lines. While the intensity ratio depends chiefly on the chosen
+emission lines to calculate the proportion of the chosen Magnesium to
+the chosen Calcium peak, the low ratios still indicate lower
+concentrations than seen elsewhere, or seen only in aragonitic parts of
+e.g.~\emph{Patella} shells. Similar observations have been made before
+\citep{Graniero2017-io}.
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3472}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}@{}}
 \caption{Table showing correlations of Mg/Ca and
@@
 & Libya & MO31A & 0.83 & \\
-& & MP64A & 0.33 & \\
+& & \textbf{MP64A} & 0.33 & \\
 & & MP67A & 0.96 & \\
@@
 & & KIZPC2 & 0.86 & \\
-\emph{P. rustica} & Gibraltar & JL1 & 0.02 & \citep{Ferguson2011-zl} \\
+\emph{P. rustica} & Gibraltar & \textbf{JL1} & 0.02 &
+\citep{Ferguson2011-zl} \\
 & & JL2 & 0.8** & \\

### 2024-06-28 Supplementary Files.qmd
edits

---
title: "Supplementary Files for “*Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells*”"
format: docx
---

**Authors: Niklas Hausmann, Donna Surge, Ivan Briz-Godino**

```{r Settings}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)
```

# Overview

The supplementary files for this study include a comprehensive set of elemental maps and linescans of various [*Patella vulgata*](#Patella) and [*Nacella* sp.](#Nacella) species, which were analysed to reassess the use of Magnesium to Calcium (Mg/Ca) ratios as palaeothermometers in Atlantic limpet shells. These supplementary files are organised by species and specimen identifiers, with both maps and linescans included for each sample.

For each species, the images are grouped to provide a clear comparison between different specimens and sampling techniques. This dataset supports the main findings of the study by offering additional evidence for the reliable use of Mg/Ca ratios in detecting palaeotemperature changes.

# *Nacella* sp. {#Nacella}

### ND-1016-3

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Nacella/Map/ND-1016-3.png")
knitr::include_graphics("Supplementary Files/Nacella/Linescan/ND-1016-3.png")
```

### ND-1016-4

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Nacella/Map/ND-1016-4.png")
knitr::include_graphics("Supplementary Files/Nacella/Linescan/ND-1016-4.png")
```

### NM-1016-1

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Nacella/Map/NM-1016-1.png")
knitr::include_graphics("Supplementary Files/Nacella/Linescan/NM-1016-1.png")
```

### NM-1016-3

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Nacella/Map/NM-1016-3-Edge.png")
knitr::include_graphics("Supplementary Files/Nacella/Map/NM-1016-3.png")
knitr::include_graphics("Supplementary Files/Nacella/Linescan/NM-1016-3.png")
```

# *Patella vulgata* {#Patella}

### ORK-LT5

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Patella/Map/ORK-LT5.png")
knitr::include_graphics("Supplementary Files/Patella/Linescan/ORK-LT5.png")
```

### QG1-7189-2

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Patella/Map/QG1-7189-2.png")
knitr::include_graphics("Supplementary Files/Patella/Linescan/QG1-7189-2.png")
```

### QG1-7246-1

This shell had a very long lifetime, which made the analysis not as clear as in other shells due to increased time-averaging. Interestingly, the record was also visible in the centre part of the shell, which is shown here.

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Patella/Map/QG1-7246-1-centre30.png")
knitr::include_graphics("Supplementary Files/Patella/Map/QG1-7246-1-centre200.png")
knitr::include_graphics("Supplementary Files/Patella/Map/QG1-7246-1.png")
knitr::include_graphics("Supplementary Files/Patella/Linescan/QG1-7246-1-centre.png")
knitr::include_graphics("Supplementary Files/Patella/Linescan/QG1-7246-1.png")
```

### QG2-1061-1

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Patella/Map/QG2-1061-1.png")
knitr::include_graphics("Supplementary Files/Patella/Linescan/QG2-1061-1.png")
```

### QG2-7180-1

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Patella/Map/QG2-7180-1.png")
knitr::include_graphics("Supplementary Files/Patella/Linescan/QG2-7180-1.png")
```

### QG2-7188-1

```{r, fig.align= 'center', out.width = "49%"}
knitr::include_graphics("Supplementary Files/Patella/Map/QG2-7188-1.png")
knitr::include_graphics("Supplementary Files/Patella/Linescan/QG2-7188-1.png")
```

### 2024-06-28 Manuscript.qmd
more edits

@@
     coord_fixed() +
-    labs(x = "mm", y = "mm", fill = "Mg/Ca ")
+    labs(x = "mm", y = "mm", fill = "Mg/Ca ", subtitle="ORK-LT5")
 
@@
   theme_cowplot(8)+
-  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
+  labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio")
 
@@
     coord_fixed() +
-    labs(x = "mm", y = "mm", colour = "Mg/Ca")
+    labs(x = "mm", y = "mm", colour = "Mg/Ca", subtitle="QG1-7188-1")
 
@@
   theme_cowplot(17)+
-  labs(x = "", y = "Mg/Ca intensity ratio")
+  labs(x = "", y = "Mg/Ca intensity ratio",subtitle="ORK-LT5 ")
 
@@
   scale_y_reverse()+
-  labs(x = "Distance to edge [mm]", y = expression(paste(delta^{18},"O (\u2030)")))
+  labs(x = "Distance to edge (mm)", y = expression(paste(delta^{18},"O (\u2030)")))
 
@@
   theme_cowplot(20)+
-  scale_fill_viridis_c(limits = c(0, 0.5),
+  scale_fill_viridis_c(limits = c(0.05, 0.45),
                        breaks = seq(0, 0.5, by = 0.1)) +    
@@
     coord_fixed() +
-    labs(x = "mm", y = "mm", fill = "Mg/Ca ")
+    labs(x = "mm", y = "mm", fill = "Mg/Ca ",subtitle="ND-1016-3")
 
@@
   theme_cowplot(15)+
-  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
+  labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio",subtitle="ND-1016-3")
 
@@
   theme_cowplot(20)+
-  scale_fill_viridis_c(limits = c(0, 0.5),
+  scale_fill_viridis_c(limits = c(0.1, 0.4),
                        breaks = seq(0, 0.4, by = 0.1)) +    
@@
            alpha = 0, linetype="longdash",col = "black")+
-    labs(x = "mm", y = "mm", fill = "Mg/Ca ")
+    labs(x = "mm", y = "mm", fill = "Mg/Ca ",subtitle="NM-1016-3")
 
@@
   theme_cowplot(20)+
-  scale_fill_viridis_c(limits = c(0, 0.4),
+  scale_fill_viridis_c(limits = c(0.05, 0.4),
                        breaks = seq(0, 0.4, by = 0.1)) +    
@@
   theme_cowplot(15)+
-  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
+  labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio",subtitle="NM-1016-3")
 
@@
 Similar observations have been made before [@Graniero2017-io].
+Strong intra-increment heterogeneities are visible along the exterior of the shells (m+3 layer), where Mg/Ca intensity ratios are higher, particularly in specimen ND-1016-3 (@fig-Nac_LIBS).
+The step from lower to higher along the increment is not gradual, as is the case in NM-1016-3 (see inset of @fig-Nac_LIBS b), but rather step-wise.
+Sampling across this boundary in a linear scan would most likely lead to erroneous results.
+Also sampling across gradual changes as in NM-1016-3 can influence the overall correlation graph for a specimen.
 
@@
 That said, the 2D imaging suggests that in both specimens a third summer can be added to the total record, which is not captured by the line scans.
-The variability seen in the *P.
-vulgata* specimens above is not as strong, which might also be related to the fast and consistent growth rate of the shells.
 
@@
 ```{r Nacella_Compare,dev='png',dpi=300}
-#| fig-cap: Comparison of LIBS line-scan and sequential $\delta$^18^O-values of specimen ORK-LT5 
-#| fig-height: 10
-#| fig-width: 20
+#| fig-cap: Comparison of LIBS line-scan and sequential $\delta$^18^O-values of *Nacella* sp. specimens. 
+#| fig-height: 7
+#| fig-width: 16
 #| label: fig-Nac_Comp
@@
   geom_path()+
-  theme_cowplot(20)+
-  labs(x = "", y = "Mg/Ca intensity ratio")
+  theme_cowplot(15)+
+  labs(x = "", y = "Mg/Ca intensity ratio", subtitle="ND-1016-3")
 
@@
   geom_path()+
-  theme_cowplot(20)+
+  theme_cowplot(15)+
   scale_y_reverse()+
-  labs(x = "Distance to edge [mm]", y = expression(paste(delta^{18},"O (\u2030)")))
+  labs(x = "Distance to edge (mm)", y = expression(paste(delta^{18},"O (\u2030)")))
 
@@
   geom_path()+
-  theme_cowplot(20)+
-  labs(x = "", y = "Mg/Ca intensity ratio")
+  theme_cowplot(15)+
+  labs(x = "", y = "Mg/Ca intensity ratio", subtitle="NM-1016-3")
 
@@
 These results are similar to previously published data on *Patella* sp.
-shells (Table [2](#Table_2)).
-R^2^ values of *P.
+shells (Table [2](#Table_2)): R^2^ values of *P.
 depressa* lie between 0.78 and 0.87 [@Garcia-Escarzaga2021-ij], for *P.
-caerulea* from the Mediterranean between 0.78 and 0.96 — although with one outlier of 0.33 (specimen MP64A), which had increased Mg/Ca ratios in increments with rich organic contents [@Hausmann2019-fi].
+caerulea* from the Mediterranean between 0.78 and 0.96 — barring one outlier of 0.33 (specimen MP64A), which had increased Mg/Ca ratios in increments with rich organic contents [@Hausmann2019-fi].
 
@@
 | Species       | Locality       | Specimen  | R^2^                    | Study                             |
-|-------------|-------------|-------------|-------------|--------------------|
+|--------|----------|-------|-------------|--------------------------|
 | *P. depressa* | Northern Spain | LAN541    | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
@@
 |               | Libya          | MO31A     | 0.83                    |                                   |
-|               |                | **MP64A**    | 0.33                    |                                   |
+|               |                | **MP64A** | **0.33**                |                                   |
 |               |                | MP67A     | 0.96                    |                                   |
@@
 |               |                | KIZPC2    | 0.86                    |                                   |
-| *P. rustica*  | Gibraltar      | **JL1**      | 0.02                    | [@Ferguson2011-zl]                |
+| *P. rustica*  | Gibraltar      | **JL1**   | **0.02**                | [@Ferguson2011-zl]                |
 |               |                | JL2       | 0.8\*\*                 |                                   |
@@
 
-: Table showing correlations of Mg/Ca and $\delta$^18^O found in other studies.
+: Correlations of Mg/Ca with $\delta$^18^O and SST found in other studies.
+Bold R^2^ values are interpreted as outliers.
 \* indicates that only SST data and no other geochemical data was used.
@@
 JL1 was disregarded in that calculation, as well as the second year of growth in JM00, which produced anomalous data.
-While no additional information about those specimens is available, it seems likely that the Mg/Ca ratios from JM00, which were interpreted as an ontogenetic trend, were potentially influenced by the same intra-increment heterogeneity we see in some of our shell specimens as well.
+While no additional information about those specimens is available, it seems likely that discarded the Mg/Ca ratios from JM00, which were interpreted as an ontogenetic trend, were potentially influenced by the same intra-increment heterogeneity we see in some of our shell specimens as well.
 Since these shells were analysed by homogenising the carbonate powder of the m+2 layer and the more exterior m+3 layer, these heterogeneities could potentially increased the Mg/Ca ratio unpredictably, leading to what might look like an ontogenetic trend in a linear dataset.
@@
 ```{r Ferguson Data, dev='png'}
-#| fig-cap: Correlation graphs for Ferguson et al., 2011 specimens.
+#| fig-cap: Correlation graphs for Ferguson et al., (2011) specimens.
 #| fig-height: 5
@@
   facet_wrap(~shell,ncol=4)+#ylim(0.15,NA)+
-  ggpubr::stat_regline_equation(label.x = 0.5, label.y = 26,
+  ggpubr::stat_regline_equation(label.x = 0.3, label.y = 26,
     aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
@@
 
-Another shell whose previous results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with $\delta$^18^O or SST in implication [@Graniero2017-io].
-Different to other specimens of this study, we were able to resample this specimen and getting Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
+Another shell whose results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with $\delta$^18^O or SST in implication [@Graniero2017-io].
+Different to other specimens of this study, we were able to resample this specimen and to acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
 These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August).
-Including these missing parts lead to a better alignment of both records and a resulting R^2^-value of 0.88.
-
-```{r ORK_LT5_Subsample}
+Including these missing parts lead to a better alignment of both records and a resulting R^2^-value of 0.88 (@fig-ORK_sub).
+
+```{r ORK_LT5}
 #| fig-cap: Comparison of previous and new Mg/Ca data for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca and $\delta$^18^O-values
@@
 #| label: fig-ORK_sub
-
-# Function to create the desired subsampled tibble
-subsample_data <- function(data, window_size = 0.05, step_size = 0.15) {
-  max_dist <- max(data$dist)
-  subsample_intervals <- seq(0, max_dist, by = step_size)
-
-  subsampled_data <- subsample_intervals %>%
-    purrr::map_df(~ {
-      window_start <- .x
-      window_end <- .x + window_size
-
-      data_in_window <- data %>%
-        filter(dist >= window_start & dist < window_end)
-
-      if (nrow(data_in_window) > 0) {
-        summarised_data <- data_in_window %>%
-          summarise(across(c(x, y, z, mg_ca, std, relstd, dist), mean, .names = "mean_{.col}"))
-        summarised_data <- summarised_data %>%
-          mutate(window_start = window_start, window_end = window_end)
-        return(summarised_data)
-      } else {
-        return(tibble())
-      }
-    })
-
-  return(subsampled_data)
-}
 
@@
 
-# Apply the function to your data
-MgCa_filtered <- MgCa %>%
-  filter(sampleID == "ORK_LT5")
-
-MgCa_subsampled_150 <- subsample_data(MgCa_filtered,step_size = 0.15)
-
-MgCa_subsampled_300 <-  subsample_data(MgCa_filtered,step_size = 0.30)
-
 P3.1 <-  MgCa %>% 
@@
   annotate("rect", xmin = 0, xmax = 0.5, ymin = 0.3, ymax = 0.55, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
-  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")
-
-# P3.2 <-
-#   MgCa_subsampled_300 %>% 
-# ggplot()+
-#   aes(mean_dist,mean_mg_ca)+
-#   geom_path(col="firebrick4")
+  labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio")
 
@@
   annotate("rect", xmin = 0, xmax = 0.08, ymin = 10, ymax = 25, alpha = 0, col="grey60", linetype="longdash", linewidth=1) +
-  labs(x = "Distance to edge [mm]", y = "Mg/Ca (mmol/mol)")
+  labs(x = "Distance to edge (mm)", y = "Mg/Ca (mmol/mol)")
 
@@
 
-While we were not able to resample the *Nacella* sp.
-specimens in Graniero *et al.* [-@Graniero2017-io], we were able to source other modern and archaeological specimens of the same species from Tierra del Fuego, which can serve as an indicator.
+We were not able to resample the *Nacella* sp.
+specimens in Graniero *et al.* [-@Graniero2017-io], which had similarly unrelated records of $\delta$^18^O and Mg/Ca and pointed to no relationship between elemental ratios and SST at all.
+However, we were able to source other modern specimens of the same species from Tierra del Fuego, which shall serve here as an indicator.
 Similar to other patelloid shells, the results of the *Nacella* specimens in our study were successful with high R^2^-values (0.76--0.95) and suggest Mg/Ca as a reliable proxy for SST (@fig-Nac_Comp and supplementaries).
 Why these results are different from the studied *Nacella* sp.
-specimens in Graniero et al.
-[-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and $\delta$^18^O records of their shells (Figure 6 in [@Graniero2017-io] ) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — fell apart and lead to an overall result of an uncorrelated relationship between Mg/Ca and SST.
+specimens in Graniero *et al*.
+[-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and $\delta$^18^O records of their shells (Figure 6 in Graniero *et al*.
+[-@Graniero2017-io]) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — is not perfectly visible, leading to an overall result of an uncorrelated relationship between Mg/Ca and SST.
 
@@
 
-This study presents high-resolution and 2D information on Mg/Ca concentrations in modern and archaeological patelloid shells and their relation to sea surface temperature.
+This study presents high-resolution and 2D information on Mg/Ca concentrations in modern and archaeological patelloid shells and their relationship to sea surface temperature.
 The results for *Patella vulgata*, *Nacella deaureata*, and *Nacella magellanica* were consistently promising with mean R^2^-values of 0.89, 0.86, and 0.88 respectively.
@@
 These results are particularly promising for future analyses of patelloid shells for the study of season of collection in archaeological sites as well as the reconstruction of sub-annual climatic conditions in the past.
-Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a large part of marine mollusc remains in archaeological sites of those regions.
-Improving their analysis in terms of cost and time efficiency through the benefits of LIBS, which can be used in this context as a screening tool for targeted $\delta$^18^O analysis focusing on annual minima and maxima revealed in the Mg/Ca record, allows researchers to increase sample numbers and produce larger and more in-depth climatic studies than previously possible.
+Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a significant part of marine mollusc remains in archaeological sites of those regions [@Colonese2011-ab; @Villagran2011-ld; @Zangrando2016-yl].
+Improving their analysis in terms of cost and time efficiency through the benefits of LIBS, which can be used in this context as a screening tool for targeted $\delta$^18^O analysis focusing on annual minima and maxima revealed in the Mg/Ca record, will allow researchers to increase sample numbers and produce larger and more in-depth climatic studies than previously possible.
 

### 2024-06-28 Manuscript.tex
more edits

@@
 e.g.~\emph{Patella} shells. Similar observations have been made before
-\citep{Graniero2017-io}.
+\citep{Graniero2017-io}. Strong intra-increment heterogeneities are
+visible along the exterior of the shells (m+3 layer), where Mg/Ca
+intensity ratios are higher, particularly in specimen ND-1016-3
+(Figure~\ref{fig-Nac_LIBS}). The step from lower to higher along the
+increment is not gradual, as is the case in NM-1016-3 (see inset of
+Figure~\ref{fig-Nac_LIBS} b), but rather step-wise. Sampling across this
+boundary in a linear scan would most likely lead to erroneous results.
+Also sampling across gradual changes as in NM-1016-3 can influence the
+overall correlation graph for a specimen.
 
@@
 specimens a third summer can be added to the total record, which is not
-captured by the line scans. The variability seen in the \emph{P.
-vulgata} specimens above is not as strong, which might also be related
-to the fast and consistent growth rate of the shells.
+captured by the line scans.
 
@@
 \caption{\label{fig-Nac_Comp}Comparison of LIBS line-scan and sequential
-\(\delta\)\textsuperscript{18}O-values of specimen ORK-LT5}
+\(\delta\)\textsuperscript{18}O-values of \emph{Nacella} sp. specimens.}
 
@@
 published data on \emph{Patella} sp. shells (Table
-\hyperref[Table_2]{2}). R\textsuperscript{2} values of \emph{P.
+\hyperref[Table_2]{2}): R\textsuperscript{2} values of \emph{P.
 depressa} lie between 0.78 and 0.87 \citep{Garcia-Escarzaga2021-ij}, for
 \emph{P. caerulea} from the Mediterranean between 0.78 and 0.96 ---
-although with one outlier of 0.33 (specimen MP64A), which had increased
-Mg/Ca ratios in increments with rich organic contents
-\citep{Hausmann2019-fi}.
+barring one outlier of 0.33 (specimen MP64A), which had increased Mg/Ca
+ratios in increments with rich organic contents \citep{Hausmann2019-fi}.
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}@{}}
-\caption{Table showing correlations of Mg/Ca and
-\(\delta\)\textsuperscript{18}O found in other studies. * indicates that
-only SST data and no other geochemical data was used. ** these samples
-were used to determine a shared R\textsuperscript{2}-value of
-0.79.}\tabularnewline
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1562}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1094}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2031}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4062}}@{}}
+\caption{Correlations of Mg/Ca with \(\delta\)\textsuperscript{18}O and
+SST found in other studies. Bold R\textsuperscript{2} values are
+interpreted as outliers. * indicates that only SST data and no other
+geochemical data was used. ** these samples were used to determine a
+shared R\textsuperscript{2}-value of 0.79.}\tabularnewline
 \toprule\noalign{}
@@
 & Libya & MO31A & 0.83 & \\
-& & \textbf{MP64A} & 0.33 & \\
+& & \textbf{MP64A} & \textbf{0.33} & \\
 & & MP67A & 0.96 & \\
@@
 & & KIZPC2 & 0.86 & \\
-\emph{P. rustica} & Gibraltar & \textbf{JL1} & 0.02 &
+\emph{P. rustica} & Gibraltar & \textbf{JL1} & \textbf{0.02} &
 \citep{Ferguson2011-zl} \\
@@
 produced anomalous data. While no additional information about those
-specimens is available, it seems likely that the Mg/Ca ratios from JM00,
-which were interpreted as an ontogenetic trend, were potentially
-influenced by the same intra-increment heterogeneity we see in some of
-our shell specimens as well. Since these shells were analysed by
-homogenising the carbonate powder of the m+2 layer and the more exterior
-m+3 layer, these heterogeneities could potentially increased the Mg/Ca
-ratio unpredictably, leading to what might look like an ontogenetic
-trend in a linear dataset. Whether these anomalies affected JL1 in the
-same way, albeit stronger, is not clear but it might well be possible.
+specimens is available, it seems likely that discarded the Mg/Ca ratios
+from JM00, which were interpreted as an ontogenetic trend, were
+potentially influenced by the same intra-increment heterogeneity we see
+in some of our shell specimens as well. Since these shells were analysed
+by homogenising the carbonate powder of the m+2 layer and the more
+exterior m+3 layer, these heterogeneities could potentially increased
+the Mg/Ca ratio unpredictably, leading to what might look like an
+ontogenetic trend in a linear dataset. Whether these anomalies affected
+JL1 in the same way, albeit stronger, is not clear but it might well be
+possible.
 
@@
 
-\caption{\label{fig-Ferg}Correlation graphs for Ferguson et al., 2011
+\caption{\label{fig-Ferg}Correlation graphs for Ferguson et al., (2011)
 specimens.}
@@
 
-Another shell whose previous results we aim to explain further is
-ORK-LT5, which has previously been shown to have no substantial
-correlation with \(\delta\)\textsuperscript{18}O or SST in implication
+Another shell whose results we aim to explain further is ORK-LT5, which
+has previously been shown to have no substantial correlation with
+\(\delta\)\textsuperscript{18}O or SST in implication
 \citep{Graniero2017-io}. Different to other specimens of this study, we
-were able to resample this specimen and getting Mg/Ca data on a slightly
-higher resolution (20--30 µm spots in a 100 µm raster compared to 50 µm
-spots spaced at 150--300 µm) and using 2 dimensions, as well as using a
-continuous high-resolution line-scan at (20--30 µm spots spaced at 10
-µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
+were able to resample this specimen and to acquire Mg/Ca data on a
+slightly higher resolution (20--30 µm spots in a 100 µm raster compared
+to 50 µm spots spaced at 150--300 µm) and using 2 dimensions, as well as
+using a continuous high-resolution line-scan at (20--30 µm spots spaced
+at 10 µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
 previously missed summer peaks at around 2--3 mm distance to the shell
@@
 alignment of both records and a resulting R\textsuperscript{2}-value of
-0.88.
+0.88 (Figure~\ref{fig-ORK_sub}).
 
@@
 
-While we were not able to resample the \emph{Nacella} sp. specimens in
-Graniero \emph{et al.} \citeyearpar{Graniero2017-io}, we were able to
-source other modern and archaeological specimens of the same species
-from Tierra del Fuego, which can serve as an indicator. Similar to other
-patelloid shells, the results of the \emph{Nacella} specimens in our
-study were successful with high R\textsuperscript{2}-values (0.76--0.95)
-and suggest Mg/Ca as a reliable proxy for SST (Figure~\ref{fig-Nac_Comp}
-and supplementaries). Why these results are different from the studied
-\emph{Nacella} sp. specimens in Graniero et al.
-\citeyearpar{Graniero2017-io} is not obvious, but the occasional
-alignment of the Mg/Ca and \(\delta\)\textsuperscript{18}O records of
-their shells (Figure 6 in \citep{Graniero2017-io} ) indicate that there
-is some relationship that --- potentially due to the low sampling
-resolution or intra-increment heterogeneity --- fell apart and lead to
-an overall result of an uncorrelated relationship between Mg/Ca and SST.
+We were not able to resample the \emph{Nacella} sp. specimens in
+Graniero \emph{et al.} \citeyearpar{Graniero2017-io}, which had
+similarly unrelated records of \(\delta\)\textsuperscript{18}O and Mg/Ca
+and pointed to no relationship between elemental ratios and SST at all.
+However, we were able to source other modern specimens of the same
+species from Tierra del Fuego, which shall serve here as an indicator.
+Similar to other patelloid shells, the results of the \emph{Nacella}
+specimens in our study were successful with high
+R\textsuperscript{2}-values (0.76--0.95) and suggest Mg/Ca as a reliable
+proxy for SST (Figure~\ref{fig-Nac_Comp} and supplementaries). Why these
+results are different from the studied \emph{Nacella} sp. specimens in
+Graniero \emph{et al}. \citeyearpar{Graniero2017-io} is not obvious, but
+the occasional alignment of the Mg/Ca and
+\(\delta\)\textsuperscript{18}O records of their shells (Figure 6 in
+Graniero \emph{et al}. \citeyearpar{Graniero2017-io}) indicate that
+there is some relationship that --- potentially due to the low sampling
+resolution or intra-increment heterogeneity --- is not perfectly
+visible, leading to an overall result of an uncorrelated relationship
+between Mg/Ca and SST.
 
@@
 concentrations in modern and archaeological patelloid shells and their
-relation to sea surface temperature. The results for \emph{Patella
+relationship to sea surface temperature. The results for \emph{Patella
 vulgata}, \emph{Nacella deaureata}, and \emph{Nacella magellanica} were
@@
 the European Atlantic façade and the Argentinian coastline and make up a
-large part of marine mollusc remains in archaeological sites of those
-regions. Improving their analysis in terms of cost and time efficiency
-through the benefits of LIBS, which can be used in this context as a
-screening tool for targeted \(\delta\)\textsuperscript{18}O analysis
-focusing on annual minima and maxima revealed in the Mg/Ca record,
-allows researchers to increase sample numbers and produce larger and
-more in-depth climatic studies than previously possible.
+significant part of marine mollusc remains in archaeological sites of
+those regions
+\citep{Colonese2011-ab, Villagran2011-ld, Zangrando2016-yl}. Improving
+their analysis in terms of cost and time efficiency through the benefits
+of LIBS, which can be used in this context as a screening tool for
+targeted \(\delta\)\textsuperscript{18}O analysis focusing on annual
+minima and maxima revealed in the Mg/Ca record, will allow researchers
+to increase sample numbers and produce larger and more in-depth climatic
+studies than previously possible.
 

### 2024-06-28 Supplementary files 2 - isotope graphs.qmd
more edits

---
title: "Comparison of geochemical datasets for “*Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells*”"
format: docx
---

# Overview
These supplementary files include the linescan information for Mg/Ca and $\delta$^18^O records of all analysed [*Patella vulgata*](#Patella) and [*Nacella* sp.](#Nacella) shells, which were analysed to reassess the use of Magnesium to Calcium (Mg/Ca) ratios as palaeothermometers in Atlantic limpet shells. These supplementary files are organised by species and specimen identifiers.

```{r Library and datasets}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)
# Packages
  pacman::p_load(
    here,
    janitor,
    tidyverse,
    cowplot,ggx,ggpubr,
    patchwork,
    RColorBrewer)

  theme_set(theme_cowplot(10))

# Scripts
# source("DTW_Script.R")
source("DTW_Script-update.R")

# Files

D18O <- read_csv(here("data","D18O.csv")) %>% 
  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern")) %>% 
  mutate(species = case_when(
      startsWith(sampleID, "ND") ~ "Nacella deaureata",
      startsWith(sampleID, "NM") ~ "Nacella magellanica",
      TRUE ~ "Patella vulgata"
    ))

MgCa <- read_csv(here("data","MgCa.csv"))

Ferguson_data <- read_csv(here("data","Ferguson-data.csv")) %>% 
  na.omit()

# # Check overlaps
# # Find common sample IDs
# intersect(MgCa$sampleID, D18O$sampleID)
# 
# # Find sample IDs only in MgCa
# setdiff(MgCa$sampleID, D18O$sampleID)
# 
# # Find sample IDs only in D18O
# setdiff(D18O$sampleID, MgCa$sampleID)

D18O <-   D18O %>% filter(sampleID %in% MgCa$sampleID)

```

```{r}
#| fig-height: 7
#| fig-width: 16

# Calculate the 2.5th and 97.5th percentiles for mg_ca within each sampleID
percentiles <- MgCa %>%
  group_by(sampleID) %>%
  summarise(lower = quantile(mg_ca, 0.025, na.rm = TRUE),
            upper = quantile(mg_ca, 0.975, na.rm = TRUE))

# Function to create the plots
create_plots <- function(sample) {
  lower <- percentiles %>% filter(sampleID == sample) %>% pull(lower)
  upper <- percentiles %>% filter(sampleID == sample) %>% pull(upper)

  P1 <-
    MgCa %>%
    filter(sampleID == sample) %>%
    filter(relstd < 0.1) %>%
    filter(mg_ca >= lower & mg_ca <= upper) %>%
    ggplot() +
    aes(dist, mg_ca) +
    geom_path() +
    theme_cowplot(17) +
    labs(x = "", y = "Mg/Ca intensity ratio", title = sample)

  P2 <-
    D18O %>%
    filter(sampleID == sample) %>%
    ggplot() +
    aes(dist, d18O) +
    geom_path() +
    theme_cowplot(18) +
    scale_y_reverse() +
    labs(x = "Distance to edge (mm)", y = expression(paste(delta^{18}, "O (\u2030)")))

  P1 / P2 + plot_annotation(tag_levels = "a")
}

# Loop through each sampleID and create the plots
unique_sampleIDs <- unique(MgCa$sampleID)

for(sample in unique_sampleIDs) {
  plot_combined <- create_plots(sample)
  print(plot_combined)
}

```

### 2024-06-29 Manuscript_docx.qmd
added docx file for easier export

--- 2024-06-28 Manuscript.qmd
@@
 bibliography: bibliography.bib
-format:
-  elsevier-pdf:
-    keep-tex: true
-    latex_engine: xelatex
-    journal:
-      name: Palaeo^3^
-      formatting: preprint
-      model: 3p
-      cite-style: authoryear
+format: docx
+    # citation-style: palaeogeography-palaeoclimatology-palaeoecology.csl
 output:

### 2024-07-04 Manuscript.qmd
corrections 1

@@
 abstract: |
-  This study provides a reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST).
-While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this is not the case for other patelloid species.
-*Patella vulgata*, *Nacella deaureata*, and *Nacella magellanica* have been studied using Mg/Ca with mixed or contrary results.
-In this study, we present elemental maps of various such species together with stable oxygen isotope values that confirm a good relationship with SST.
-Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O and Mg/Ca ratios.
-By reassessing these species and including modern and archaeological samples (n=12) from three patelloid species (*P.
+  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST).
+While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species.
+*Patella vulgata*, *Nacella deaureata*, and *N.
+magellanica* have been studied using Mg/Ca ratios with mixed or contrary results.
+In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta&^18^O) that confirm a good relationship with SST.
+Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios.
+By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P.
 vulgata*, *N.
@@
 magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change.
-As a result, these species can in the future serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e.
-annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O or clumped oxygen isotope analysis.
+As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e.
+annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
 keywords: 
@@
 
-Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Robson2023-oz].
-They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
-Determining past sea surface temperature (SST) change largely relies on the measurement of $\delta$^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure, that potentially provides a more accurate SST estimate than $\delta$^18^O-values, which are also affected by changes in salinity.
+Limpet shells are commonly found within archaeological sites and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Robson2023-oz].
+They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
+Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of $\delta$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than $\delta$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater.
 In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].
@@
 Problematically, links between Mg/Ca ratios and SST changes in many marine mollusc shells, including limpets, have shown to be unreliable [@Surge2008-ri; @Wanamaker2008-zl; @Schone2010-yl; @Freitas2012-tx; @Graniero2017-io; @Poulain2015-dg; @Vihtakari2017-wd].
-This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as temperature proxy.
+This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as a temperature proxy.
 Confusingly, in some cases multiple different temperature equations have been found for the same species (see Freitas *et al*.
@@
 Coeval specimens sharing one locality can also show differences in their relation to SST [@Hausmann2019-fi].
-Where the use of Mg/Ca as a palaeotemperature proxy was successful, anomalous patterns in some specimens had still to be filtered out by hand, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].
-
-While recent research particularly of *Patella* sp.
-in the Mediterranean and Southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].
-There remains a lack of clarity for Atlantic limpet species, particularly since it was last shown here, that they are not reliable recorders of palaeotemperature [@Graniero2017-io].
+Where the use of Mg/Ca ratios as a palaeotemperature proxy was successful, anomalous patterns in some specimens had to be filtered out by hand, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].
+
+Recent research particularly of *Patella* spp.
+in the Mediterranean and southwest Europe have provided promising results [@Hausmann2019-fi; @Garcia-Escarzaga2015-jc; @Garcia-Escarzaga2018-nf].
+However, there remains a lack of clarity for Atlantic limpet species, particularly since it was shown, that the Mg/Ca ratio in their shells are not reliable recorders of palaeotemperature [@Graniero2017-io].
 
@@
 To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS.
-LIBS allows us to carry out 2D imaging of entire shell sections, which helps us to navigate the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.
+LIBS allows us to carry out 2D imaging of entire shell sections, which allows navigation of the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.
 By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
-Generally establishing the usefulness and reliability of Mg/Ca as SST proxy in limpets should help to provide a platform for such future research and an important stepping stone to a better understanding of elemental ratios in other marine mollusc shells.
+Generally establishing the usefulness and reliability of Mg/Ca ratios as SST proxy in limpets provides a platform for future research and an important steppingstone to a better understanding of elemental ratios in other marine mollusc shells.
 
@@
 |------|------------|-------|--------|------|----------|
-| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay  | ND-1016-3  | Stable oxygen     |
-|                |                    |                  | \- Tierra del   | ND-1016-4  | and carbon        |
-|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  | isotope analysis  |
+| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C     |
+|                |                    |                  | \- Tierra del   | ND-1016-4  | analysis          |
+|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  |                   |
 |                |                    |                  |                 | NM-1016-3  |                   |
 |                |                    |                  |                 |            |                   |
-|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | Stable oxygen     |
-|                |                    |                  | Westray (UK)    |            | and carbon        |
-|                |                    |                  |                 |            | isotope analysis; |
+|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | $\delta$^18^O and $\delta$^13^C     |
+|                |                    |                  | Westray (UK)    |            | analysis;         |
 |                |                    |                  |                 |            | Mg, Li, Sr, Ca.   |
+|                |                    |                  |                 |            |                   |
 |                |                    |                  |                 |            |                   |
@@
 
-There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK).
+Specimens were sourced from two main locations: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in the Orkney archipelago of Scotland (UK).
 Four modern specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016.
-The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih].
-SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February (@fig-SSTs - red).
-
-On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has previously been analysed using stable carbon and oxygen isotopes, as well as LA-ICP-MS [@Graniero2017-io].
+The area receives around 570 mm of precipitation, but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units) [@Nicastro2020-ih].
+SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February (@fig-SSTs -- red line).
+
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for $\delta$^18^O and $\delta$^13^C values, as well as LA-ICP-MS [@Graniero2017-io].
 Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
-The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs - blue).
+The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs -- blue line).
 
@@
 We also studied 7 archaeological specimens of *P.
-vulgata* coming from Quoygrew on Westray dating to 900 – 1200 CE.
-These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly and the Little Ice Age [@Surge2012-ba].
-Together they show a range of lifespans, with some comparatively long for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to time-averaging.
-That said, the majority are under 5 years old.
+vulgata* from the Quoygrew site on Westray dating to 900 – 1200 CE.
+These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly (810 -- 1229 CE) [@Surge2012-ba].
+Together they show a range of lifespans, with some comparatively long lived for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to within-shell time-averaging because of slow growth rates.
+That said, the majority are under 5 years old yielding sufficiently fast growth to nearly capture seasonal amplitudes of SST.
 
@@
 
-Mg/Ca elemental imaging was carried out using Laser Induced Breakdown Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany) using previously published methods [@Hausmann2023-ih].
-These involve the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume.
+Mg/Ca elemental imaging was carried out using LIBS at the Leibniz Zentrum für Archäologie (Mainz-Germany) following previously published methods [@Hausmann2023-ih].
+The imaging involves the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume.
 This plume emits light, which is measured using a synchronised spectrometer (delay time = 0.5 µs, integration time = 1 µs).
@@
 While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g.
-[@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca changes within the shell carbonate.
-All values reported here at in arbitrary units based on the ratio of intensities of the peaks above.
+[@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca variation within the shell carbonate.
+All values reported here are in arbitrary units based on the ratio of intensities of the peaks above.
 
 Using this system we carried out elemental imaging of the shell specimens at a resolution of 50--100 µm distance between sampling spots.
-Each spot was irradiated 10 times with the 3 first spectra discarded as a cleaning step and the remaining spectra summed to get an average Mg/Ca intensity ratio for each sample spot.
+Each spot was irradiated 10 times with the first 3 spectra discarded as a cleaning step and the remaining spectra summed to get an average Mg/Ca intensity ratio for each sample spot.
 Subsequently, we re-sampled the section using a line scan at 10 µm resolution.
 This leads to an overlap between sample spots, but also allows for a continuous record without gaps.
-Intensity ratios were filtered for cases with high relative standard deviation (i.e.
-more than 10%).
+Intensity ratios were filtered for cases with high relative standard deviation (i.e., more than 10%).
 These occurred in places where the previous sampling procedures for carbonate powder as part of the oxygen isotope analysis left an uneven sample surface introducing variability in the plasma generation and thus uncertainty in the data of these locations.
 
-### Oxygen isotopes
-
-High-quality oxygen isotope values ($\delta$^18^O) were acquired from previous publications on *P.
+### Oxygen isotope ratios
+
+High-quality $\delta$^18^O values were acquired from previous publications on *P.
 vulgata* [@Surge2012-ba; @Graniero2017-io], *N.
 deaureata* and *N.
-magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found.
-In short, all shells were sectioned along the main growth axis to expose the internal growth structure.
+magellanica* [@Nicastro2020-ih], in which detailed descriptions of the methodology for the stable isotope analyses can be found.
+In short, all shells were sectioned along the maximum growth axis to expose the internal growth structure.
 From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells.
-Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.
+Analytical precision of the isotope ratios was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.
 
@@
 
-@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* sp., and *Patella vulgata*, respectively.
-Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* sp.
+@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* sp., and *Patella vulgata*, respectively).
+Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* sp.
 shells living 2–3 years and *P.
 vulgata* living regularly over 3 years.
-In those instances where longer lifetimes are encountered (i.e.
-specimen QG1-7246-1, @fig-Pat_iso b) the temporal-resolution is reduced from more than a dozen of samples per year, to only around 7.
+In those instances where longer lifespans resulted in slow growth rates below 1 mm/yr (i.e.
+specimen QG1-7246-1, @fig-Pat_iso b), the temporal-resolution is reduced from more than a dozen of samples per year, to only around 7.
 
@@
 
-Because it is not possible to directly associate the exact sample locations used in the analysis of stable oxygen isotopes with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O-values and Mg/Ca ratios.
-DTW is an algorithm that measures similarity between two proxy-sequences, which may vary in sampling resolution or interval.
+Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O values and Mg/Ca ratios.
+DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval.
 By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences.
-This allows us to compare the proxy data sets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates.
+This allows us to compare the proxy datasets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates.
 We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data.
-This involved selecting appropriate distance measures and constraints to ensure meaningful alignment.
+This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in OSF Repository).
 The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.
@@
 Compared to other patelloid shells, some of whose interior is almost entirely made of aragonite, this layer is very thin.
-
+\newline
 ```{r Patella_LIBS,fig.pos='H'}
 #| fig-cap: Example LIBS maps and line scans of analysed specimens.
-(*a+b*) ORK-LT and (*c*+*d*) QG1-7188-1.
+(*a+c*) ORK-LT and (*b*+*d*) QG1-7188-1.
 Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast.
 Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
-#| fig-height: 5
-#| fig-width: 10
+#| fig-height: 6
+#| fig-width: 6
 #| label: fig-Pat_LIBS
@@
   filter(std < 0.15) %>%
+ #filter(x>0) %>% 
   filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
@@
   theme_cowplot(8)+
+    theme(plot.tag.position = c(-0.01, 1))+ # Adjust horizontal and vertical position
   labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio")
@@
   theme_cowplot(8)+
+  theme(plot.tag.position = c(-0.01, 1))+ # Adjust horizontal and vertical position
   labs(x = "Distance to edge (mm)", y = "Mg/Ca intensity ratio")
@@
 ################
-  (ORK_LT5_plot+ORK_LT5_line)/(QG1_7188_1_plot+QG1_7188_1_line)+plot_annotation(tag_levels = "a")
+#(ORK_LT5_plot+ORK_LT5_line)/(QG1_7188_1_plot+QG1_7188_1_line)+plot_annotation(tag_levels = "a")
+
+#QG1_7188_1_line/ORK_LT5_line/
+maps <-     (ORK_LT5_plot+labs(tag="a"))+(QG1_7188_1_plot+labs(tag = "b"))+plot_layout(widths = c(0.4,0.2))#heights=c(0.55,1.2))
+
+lines <-   (ORK_LT5_line+labs(tag="c"))/(QG1_7188_1_line+labs(tag = "d"))+plot_layout(heights=c(0.5,0.5))
+
+maps/lines+plot_layout(heights=c(0.6,0.5))
+
 ```

### 2024-07-04 Manuscript.tex
corrections 1

@@
 \begin{abstract}
-This study provides a reassessment of the use of Magnesium to Calcium
-(Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface
+This study provides a reassessment of Magnesium to Calcium (Mg/Ca)
+ratios in Atlantic limpet shells to determine past sea surface
 temperatures (SST). While \emph{Patella depressa} along the Spanish
 shoreline has repeatedly produced reliable correlations between SST and
-Mg/Ca ratios, this is not the case for other patelloid species.
-\emph{Patella vulgata}, \emph{Nacella deaureata}, and \emph{Nacella
-magellanica} have been studied using Mg/Ca with mixed or contrary
-results. In this study, we present elemental maps of various such
-species together with stable oxygen isotope values that confirm a good
-relationship with SST. Our dataset also reassesses a specimen which was
-previously unsuccessful in providing significant correlations between
-\(\delta\)\textsuperscript{18}O and Mg/Ca ratios. By reassessing these
-species and including modern and archaeological samples (n=12) from
-three patelloid species (\emph{P. vulgata}, \emph{N. deaureata}, and
-\emph{N. magellanica}) we further add to the growing set of evidence for
-the reliable use of Mg/Ca ratios to detect palaeotemperature change. As
-a result, these species can in the future serve as a means to determine
-ontogenetic age and season of capture as well as to reveal locations of
-interest within the growth record (i.e.~annual temperature minima and
-maxima) for the targeted analysis using \(\delta\)\textsuperscript{18}O
-or clumped oxygen isotope analysis.
+Mg/Ca ratios, this relationship is not the case for other patelloid
+species. \emph{Patella vulgata}, \emph{Nacella deaureata}, and \emph{N.
+magellanica} have been studied using Mg/Ca ratios with mixed or contrary
+results. In this study, we present elemental maps of these three species
+together with oxygen isotope ratios (\$\delta\&\textsuperscript{18}O)
+that confirm a good relationship with SST. Our dataset also reassesses a
+specimen which was previously unsuccessful in providing significant
+correlations between \(\delta\)\textsuperscript{18}O values and Mg/Ca
+ratios. By reassessing these species and including modern and
+archaeological specimens (n=12) from three patelloid species (\emph{P.
+vulgata}, \emph{N. deaureata}, and \emph{N. magellanica}) we further add
+to the growing set of evidence for the reliable use of Mg/Ca ratios to
+detect palaeotemperature change. As a result, these species can in the
+future serve to determine ontogenetic age and season of capture as well
+as to reveal locations of interest within the growth record (i.e.~annual
+temperature minima and maxima) for the targeted analysis using
+\(\delta\)\textsuperscript{18}O and clumped isotope analysis.
 \end{abstract}
@@
 
-Limpet shells are commonly found within archaeological sites and and
-past shorelines due to their robust carbonate structure and the
-long-term use of limpets as a marine food source \citep{Robson2023-oz}.
-They have successfully been studied in the past in the context of
-coastal subsistence economies, site occupation on seasonal
+Limpet shells are commonly found within archaeological sites and past
+shorelines due to their robust carbonate structure and the long-term use
+of limpets as a marine food source \citep{Robson2023-oz}. They have been
+successfully studied in the context of coastal subsistence economies and
+site occupation on seasonal
 \citep{Shackleton1973-ij, Parker2018-wf, Bosch2018-ud} or long-term
-scale \citep{Ortiz2015-mr}, as well as palaeotemperature
+scales \citep{Ortiz2015-mr}, as well as archives of palaeotemperature
 \citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
-Determining past sea surface temperature (SST) change largely relies on
-the measurement of \(\delta\)\textsuperscript{18}O-values within the
-calcitic parts of the shell, but attempts have been made to also use
-elemental ratios, such as magnesium to calcium (Mg/Ca)
-\citep{Graniero2017-io, Ferguson2011-zl}, to have an alternative
-measure, that potentially provides a more accurate SST estimate than
-\(\delta\)\textsuperscript{18}O-values, which are also affected by
-changes in salinity. In addition, the data acquisition of elemental
-ratios --- either through laser-ablation-isotope-ratio-mass-spectrometry
-(LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much
-faster and cost-effective, increasing the number of specimens that can
-be studied overall \citep{Durham2017-fh, Hausmann2023-ih}.
+Determining past sea surface temperature (SST) change recorded in
+patelloid limpet shells largely relies on the measurement of
+\(\delta\)\textsuperscript{18}O values within the calcitic parts of the
+shell, but attempts have been made to also use elemental ratios, such as
+magnesium to calcium (Mg/Ca) ratios
+\citep{Graniero2017-io, Ferguson2011-zl}, to have an alternative measure
+that potentially provides a more accurate SST estimate than
+\(\delta\)\textsuperscript{18}O-values,as the latter proxy is affected
+by changes in the mixing of fresh- and saltwater. In addition, the data
+acquisition of elemental ratios --- either through
+laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser
+induced breakdown spectroscopy (LIBS), can be much faster and
+cost-effective, increasing the number of specimens that can be studied
+overall \citep{Durham2017-fh, Hausmann2023-ih}.
 
@@
 which can independently and unpredictably affect Mg/Ca ratios and
-confound their interpretation as temperature proxy. Confusingly, in some
-cases multiple different temperature equations have been found for the
-same species (see Freitas \emph{et al}. \citeyearpar{Freitas2012-tx} and
-Vihtakari \emph{et al}. \citeyearpar{Vihtakari2017-wd} and references
-therein). Coeval specimens sharing one locality can also show
+confound their interpretation as a temperature proxy. Confusingly, in
+some cases multiple different temperature equations have been found for
+the same species (see Freitas \emph{et al}. \citeyearpar{Freitas2012-tx}
+and Vihtakari \emph{et al}. \citeyearpar{Vihtakari2017-wd} and
+references therein). Coeval specimens sharing one locality can also show
 differences in their relation to SST \citep{Hausmann2019-fi}. Where the
-use of Mg/Ca as a palaeotemperature proxy was successful, anomalous
-patterns in some specimens had still to be filtered out by hand,
+use of Mg/Ca ratios as a palaeotemperature proxy was successful,
+anomalous patterns in some specimens had to be filtered out by hand,
 reducing the overall robustness of the results of those successful
@@
 
-While recent research particularly of \emph{Patella} sp. in the
-Mediterranean and Southwest Europe have provided promising results
+Recent research particularly of \emph{Patella} spp. in the Mediterranean
+and southwest Europe have provided promising results
 \citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
-There remains a lack of clarity for Atlantic limpet species,
-particularly since it was last shown here, that they are not reliable
-recorders of palaeotemperature \citep{Graniero2017-io}.
+However, there remains a lack of clarity for Atlantic limpet species,
+particularly since it was shown, that the Mg/Ca ratio in their shells
+are not reliable recorders of palaeotemperature \citep{Graniero2017-io}.
 
@@
 contexts using LIBS. LIBS allows us to carry out 2D imaging of entire
-shell sections, which helps us to navigate the complex elemental
+shell sections, which allows navigation of the complex elemental
 structure of the shell and better separate the external,
@@
 interest. Generally establishing the usefulness and reliability of Mg/Ca
-as SST proxy in limpets should help to provide a platform for such
-future research and an important stepping stone to a better
-understanding of elemental ratios in other marine mollusc shells.
+ratios as SST proxy in limpets provides a platform for future research
+and an important steppingstone to a better understanding of elemental
+ratios in other marine mollusc shells.
 
@@
 Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
-& ND-1016-3 & Stable oxygen \\
-& & & - Tierra del & ND-1016-4 & and carbon \\
-& & \emph{N. magellanica} & Fuego (AR) & NM-1016-1 & isotope analysis \\
+& ND-1016-3 & \(\delta\)\textsuperscript{18}O and
+\(\delta\)\textsuperscript{13}C \\
+& & & - Tierra del & ND-1016-4 & analysis \\
+& & \emph{N. magellanica} & Fuego (AR) & NM-1016-1 & \\
 & & & & NM-1016-3 & \\
@@
 & \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - &
-ORK-LT5 & Stable oxygen \\
-& & & Westray (UK) & & and carbon \\
-& & & & & isotope analysis; \\
+ORK-LT5 & \(\delta\)\textsuperscript{18}O and
+\(\delta\)\textsuperscript{13}C \\
+& & & Westray (UK) & & analysis; \\
 & & & & & Mg, Li, Sr, Ca. \\
+& & & & & \\
 & & & & & \\
@@
 
-There are two main locations that we sourced specimens from: the Beagle
-Channel in Tierra del Fuego (Argentina) and the island of West Ray in
-Orkney (UK). Four modern specimens were collected from the Beagle
-Channel in Cambaceres Bay in October 2016. The area receives around 570
-mm precipitation but the channel waters are predominantly influenced by
-marine currents and in the period of October 2015 to October 2016 stayed
-around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
-August to 10ºC in February (Figure~\ref{fig-SSTs} - red).
+Specimens were sourced from two main locations: the Beagle Channel in
+Tierra del Fuego (Argentina) and the island of West Ray in the Orkney
+archipelago of Scotland (UK). Four modern specimens were collected from
+the Beagle Channel in Cambaceres Bay in October 2016. The area receives
+around 570 mm of precipitation, but the channel waters are predominantly
+influenced by marine currents and in the period of October 2015 to
+October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units)
+\citep{Nicastro2020-ih}. SSTs range from 4.7ºC in August to 10ºC in
+February (Figure~\ref{fig-SSTs} -- red line).
 
 On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
-August 2009 and has previously been analysed using stable carbon and
-oxygen isotopes, as well as LA-ICP-MS \citep{Graniero2017-io}. Westray
-lies about 70 km north of the Scottish mainland and experiences
-virtually no freshwater input with salinity values of 34.0 -- 34.5 psu
+August 2009 and has been previously analysed for
+\(\delta\)\textsuperscript{18}O and \(\delta\)\textsuperscript{13}C
+values, as well as LA-ICP-MS \citep{Graniero2017-io}. Westray lies about
+70 km north of the Scottish mainland and experiences virtually no
+freshwater input with salinity values of 34.0 -- 34.5 psu
 \citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in March to
-13.8ºC in August (Figure~\ref{fig-SSTs} - blue).
+13.8ºC in August (Figure~\ref{fig-SSTs} -- blue line).
 
@@
 
-We also studied 7 archaeological specimens of \emph{P. vulgata} coming
-from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
+We also studied 7 archaeological specimens of \emph{P. vulgata} from the
+Quoygrew site on Westray dating to 900 -- 1200 CE. These shells have
 previously been studied to characterise seasonal temperature change
-during the Medieval Climate Anomaly and the Little Ice Age
+during the Medieval Climate Anomaly (810 -- 1229 CE)
 \citep{Surge2012-ba}. Together they show a range of lifespans, with some
-comparatively long for their size (QG1-7246-1, 12 years within 16 mm of
-growth) making them particularly susceptible to time-averaging. That
-said, the majority are under 5 years old.
+comparatively long lived for their size (QG1-7246-1, 12 years within 16
+mm of growth) making them particularly susceptible to within-shell
+time-averaging because of slow growth rates. That said, the majority are
+under 5 years old yielding sufficiently fast growth to nearly capture
+seasonal amplitudes of SST.
 
@@
 
-Mg/Ca elemental imaging was carried out using Laser Induced Breakdown
-Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany)
-using previously published methods \citep{Hausmann2023-ih}. These
-involve the ablation of the shell section using an infrared (1064 nm)
-laser (1--2 mJ, 100 Hz) onto a surface area of 20--30 µm to generate a
-plasma plume. This plume emits light, which is measured using a
-synchronised spectrometer (delay time = 0.5 µs, integration time = 1
-µs). The resulting light spectrum quantifies the emission lines of
-Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine
-their intensity ratio. While these two peaks alone do not represent the
-molar concentrations of both elements (as opposed to calibration free
-LIBS; e.g. \citep{Martinez-Minchero2022-jz}), the intensity ratio is
-linearly correlated to the molar concentration \citep{Hausmann2017-oa}
-and can be used as a reliable indicator of Mg/Ca changes within the
-shell carbonate. All values reported here at in arbitrary units based on
-the ratio of intensities of the peaks above.
+Mg/Ca elemental imaging was carried out using LIBS at the Leibniz
+Zentrum für Archäologie (Mainz-Germany) following previously published
+methods \citep{Hausmann2023-ih}. The imaging involves the ablation of
+the shell section using an infrared (1064 nm) laser (1--2 mJ, 100 Hz)
+onto a surface area of 20--30 µm to generate a plasma plume. This plume
+emits light, which is measured using a synchronised spectrometer (delay
+time = 0.5 µs, integration time = 1 µs). The resulting light spectrum
+quantifies the emission lines of Magnesium (MgII; 279.553 nm) and
+Calcium (CaII; 315.887 nm) to determine their intensity ratio. While
+these two peaks alone do not represent the molar concentrations of both
+elements (as opposed to calibration free LIBS; e.g.
+\citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
+correlated to the molar concentration \citep{Hausmann2017-oa} and can be
+used as a reliable indicator of Mg/Ca variation within the shell
+carbonate. All values reported here are in arbitrary units based on the
+ratio of intensities of the peaks above.
 
@@
 
-\subsubsection{Oxygen isotopes}\label{oxygen-isotopes}
+\subsubsection{Oxygen isotope ratios}\label{oxygen-isotope-ratios}
 
-High-quality oxygen isotope values (\(\delta\)\textsuperscript{18}O)
-were acquired from previous publications on \emph{P. vulgata}
+High-quality \(\delta\)\textsuperscript{18}O values were acquired from
+previous publications on \emph{P. vulgata}
 \citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
@@
 interior is almost entirely made of aragonite, this layer is very thin.
+\newline
 
@@
 \caption{\label{fig-Pat_LIBS}Example LIBS maps and line scans of
-analysed specimens. (\emph{a+b}) ORK-LT and (\emph{c}+\emph{d})
+analysed specimens. (\emph{a+c}) ORK-LT and (\emph{b}+\emph{d})
 QG1-7188-1. Note that the ranges of Mg/Ca intensity ratios are not

### 2024-07-04 Manuscript.qmd
Donna comments

@@
 | Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
-|------|------------|-------|--------|------|----------|
+|------------|------------|------------|------------|------------|------------|
 | Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C |
@@
 |                |                    |                  |                 |            |                                       |
-| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | Stable oxygen     |
-|                |                    |                  | Westray (UK)    | QG1-7189-2 | and carbon        |
-|                |                    |                  |                 | QG2-1061-1 | isotope analysis  |
+| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O                         |
+|                |                    |                  | Westray (UK)    | QG1-7189-2 | and $\delta$^13^C                            |
+|                |                    |                  |                 | QG2-1061-1 | analysis                      |
 |                |                    |                  |                 | QG2-1064-1 |                                       |
@@
 ```{r Nacella_Isotope_plot,dev='png',dpi=300}
-#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay.
+#| fig-cap: Isotope values of *Nacella* spp. limpet shells from Cambaceres Bay.
 #| fig-height: 5
@@
 
-@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* sp., and *Patella vulgata*, respectively).
-Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* sp.
+@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively).
+Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* spp.
 shells living 2–3 years and *P.
@@
 We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data.
-This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in OSF Repository).
+This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in [@Hausmann_2024OSF]).
 The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.
@@
 The elemental imaging of *P.
-vulgata* shells consistently showed repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca intensity ratios indicating and low ratios indicating low temperature periods.
+vulgata* shells consistently showed repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca intensity ratios indicating high temperature periods and low ratios indicating low temperature periods.
 As examples, @fig-Pat_LIBS shows the 2D-distribution of elemental ratios across the entire section of ORK-LT5 and the preserved anterior side of QG1-7188-1 (additional maps can be found in the supplementaries).
-The repeating patterns are found across the calcitic layers that are exterior to the myostracum (layers m+2 and m+3) as well as the layers that are interior (m-2) [@Fenger2007-gf].
+The repeating patterns are found across the calcitic layers that are exterior to the myostracum (layers m+2 and m+3) as well as the layers that are interior (m-2)(see Fenger et al.
+[-@Fenger2007-gf] for a detailed explanations of microstructural layers in *P.
+vulgaata*).
 The layer m-1, which is also interior, consists of aragonite and is seen in @fig-Pat_LIBS in grey, as its Mg/Ca intensity ratio consistently falls below the range of interest (0.2 and above).
@@
 \newline
+
 ```{r Patella_LIBS,fig.pos='H'}
@@
 
-Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental-imaging.
-That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D-image, showing the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].
-
-@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O-values (similar graphs for other specimen can be found in the supplementaries).
+Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental imaging.
+That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D image.
+This shows the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].
+
+@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O values (similar graphs for other specimen can be found in the supplementaries).
 The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior.
@@
 
-Aligning the records programmatically using dynamic time warping, allowed us to compare them directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr).
+Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr).
 The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 – 0.065 \* Mg/Ca).
@@
 
-## *Nacella* sp.
-
-The LIBS data for *Nacella* sp.
-specimens is less straightforward with shells being much thinner and thus patterns being more difficult to make out.
-Both specimens (ND-1016-3 and NM-1016-3) shown as example in @fig-Nac_LIBS (see additional graphs in supplementaries) also experience increases of Mg/Ca intensity ratio towards the exterior of the shell, similar to the *P.
+## *Nacella* spp.
+
+The LIBS data for *Nacella* spp.
+specimens is less straightforward with shells being much thinner than *P.
+vulgata* and thus patterns being more difficult to discern.
+Both specimens (ND-1016-3 and NM-1016-3) shown as examples in @fig-Nac_LIBS (see additional graphs in supplementaries) also experience increases of Mg/Ca intensity ratio towards the exterior of the shell, similar to the *P.
 vulgata* specimens from Orkney.
@@
 ```{r Nacella_LIBS,fig.pos='H'}
-#| fig-cap: Example LIBS maps and line scans of analysed *Nacella* sp.
+#| fig-cap: Example LIBS maps and line scans of analysed *Nacella* spp.
 specimens.
@@
 
-Interestingly, the Mg/Ca intensity ratios in the *Nacella* sp.
+Interestingly, the Mg/Ca intensity ratios in the *Nacella* spp.
 shells were much lower (0.05–0.30) than those seen in other patelloid species (e.g.
-0.3–0.7 in *Patella vulgata* above, or 0.5–1.5 in Hausmann *et al*.
+0.3–0.7 in *P.
+vulgata* above, or 0.5–1.5 in Hausmann *et al*.
 [-@Hausmann2023-ih]) using the same emission lines.
-While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the chosen Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere, or seen only in aragonitic parts of e.g.
+While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the chosen Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere or seen only in aragonitic parts of e.g.
 *Patella* shells.
@@
 Strong intra-increment heterogeneities are visible along the exterior of the shells (m+3 layer), where Mg/Ca intensity ratios are higher, particularly in specimen ND-1016-3 (@fig-Nac_LIBS).
-The step from lower to higher along the increment is not gradual, as is the case in NM-1016-3 (see inset of @fig-Nac_LIBS b), but rather step-wise.
+The step from lower to higher along the increment is not gradual, as is the case in NM-1016-3 (see inset of @fig-Nac_LIBS b), but rather stepwise.
 Sampling across this boundary in a linear scan would most likely lead to erroneous results.
@@
 
-@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their $\delta$^18^O-values (similar graphs for the other *Nacella* sp.
-specimen can be found in the supplementaries).
+@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their $\delta$^18^O values (similar graphs for the other *Nacella* spp.
+specimens can be found in the supplementaries).
 Compared to the *P.
-vulgata* records, these mirror each other much more clearly, due to the brief period recorded in the shells and its simple make-up.
+vulgata* records, these mirror each other much more clearly, due to the brief period recorded in the shells and thus a lower chance of misaligning annual extremes.
 
 ```{r Nacella_Compare,dev='png',dpi=300}
-#| fig-cap: Comparison of LIBS line-scan and sequential $\delta$^18^O-values of *Nacella* sp. specimens. 
+#| fig-cap: Comparison of LIBS line-scan and sequential $\delta$^18^O-values of *Nacella* spp. specimens. 
 #| fig-height: 7
@@
 
-The *Nacella* sp.
+The *Nacella* spp.
 specimens showed similarly clustered correlations, which interestingly were not grouped by species.
-In fact, specimens ND-1016-3 (*Nacella deaureata)* and NM-1016-3 (*Nacella magellanica*) are almost identical in their fitted equations (@fig-Nac_Corr).
+In fact, specimens ND-1016-3 (*Nacella deaureata)* and NM-1016-3 (*N.
+magellanica*) are almost identical in their fitted equations (@fig-Nac_Corr).
 Both also share a high coefficient of determination (R^2^) of 0.95 and 0.92 respectively, with the remaining specimens being somewhat lower (R^2^ = 0.76 for ND-1016-4 and R^2^=0.84 for NM-1016-3).
-The mean R^2^ value for each *Nacella* species is 0.86 for *Nacella deaureata* and 0.88 for *Nacella magellanica*.
+The mean R^2^ value for each *N.* species is 0.86 for *N.
+deaureata* and 0.88 for *N.
+magellanica*.
 Why two specimens of different species are so well aligned is not entirely certain and randomness cannot for now be ruled out.
@@
 ```{r Nacella_Correlation_Graphs,dev='png',dpi=300}
-#| fig-cap: Correlation graphs for *Nacella* sp. specimens
+#| fig-cap: Correlation graphs for *Nacella* spp. specimens
 #| fig-height: 5
@@
 
-The Mg/Ca ratios from the 2D elemental imaging present positive results on SST being the main factor behind Mg/Ca concentration in three Atlantic limpet species.
+The Mg/Ca ratios from the 2D elemental imaging present positive results suggesting that SST is the main factor influencing variations in Mg/Ca ratios in three Atlantic limpet species.
 Annually repeating patterns of increasing (spring/summer) and decreasing (autum/winter) SST were also visible in the line scans, albeit less straightforward than in the 2D-data.
@@
 magellanica mean R^2^*=0.88).
-These results are similar to previously published data on *Patella* sp.
-shells (Table [2](#Table_2)): R^2^ values of *P.
+These results are similar to previously published data on other *Patella* species shells (Table [2](#Table_2)): R^2^ values of *P.
 depressa* lie between 0.78 and 0.87 [@Garcia-Escarzaga2021-ij], for *P.
@@
 | Species       | Locality       | Specimen  | R^2^                    | Study                             |
-|--------|----------|-------|-------------|--------------------------|
+|------|------------|-------|--------|------|----------|
 | *P. depressa* | Northern Spain | LAN541    | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
@@
 
-: Correlations of Mg/Ca with $\delta$^18^O and SST found in other studies.
+: Correlations of Mg/Ca ratios with $\delta$^18^O values and SST found in other studies.
 Bold R^2^ values are interpreted as outliers.
@@
 *P.
-rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.8, respectively [@Ferguson2011-zl] (@fig-Ferg).
+rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.80, respectively [@Ferguson2011-zl] (@fig-Ferg).
 Together with the data from two *P.
-caerulea* shells (JM00 and JM30 with R^2^-values of 0.69 and 0.83), the data from JL2 was used determine a general Mg/Ca-SST relationship with an R^2^-value of 0.79.
+caerulea* shells (JM00 and JM30 with R^2^-values of 0.69 and 0.83), the data from JL2 was used to determine a general Mg/Ca-SST relationship with an R^2^ value of 0.79.
 JL1 was disregarded in that calculation, as well as the second year of growth in JM00, which produced anomalous data.
-While no additional information about those specimens is available, it seems likely that discarded the Mg/Ca ratios from JM00, which were interpreted as an ontogenetic trend, were potentially influenced by the same intra-increment heterogeneity we see in some of our shell specimens as well.
-Since these shells were analysed by homogenising the carbonate powder of the m+2 layer and the more exterior m+3 layer, these heterogeneities could potentially increased the Mg/Ca ratio unpredictably, leading to what might look like an ontogenetic trend in a linear dataset.
+While no additional information about those specimens is available, it seems likely that the discarded Mg/Ca ratios from JM00, which were interpreted as an ontogenetic trend, were potentially influenced by the same intra-increment heterogeneity we see in some of our shell specimens as well.
+Since these shells were analysed by homogenising the carbonate powder of the m+2 layer and the more exterior m+3 layer, these heterogeneities could have potentially increased the Mg/Ca ratio unpredictably, leading to what might look like an ontogenetic trend in a linear dataset.
 Whether these anomalies affected JL1 in the same way, albeit stronger, is not clear but it might well be possible.
@@
 
-Another shell whose results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with $\delta$^18^O or SST in implication [@Graniero2017-io].
-Different to other specimens of this study, we were able to resample this specimen and to acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
+Another shell whose results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with $\delta$^18^O values or SST by implication [@Graniero2017-io].
+In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
 These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August).
@@
 
-We were not able to resample the *Nacella* sp.
-specimens in Graniero *et al.* [-@Graniero2017-io], which had similarly unrelated records of $\delta$^18^O and Mg/Ca and pointed to no relationship between elemental ratios and SST at all.
+We were not able to resample the *Nacella* spp.
+specimens in Graniero *et al.* [-@Graniero2017-io], which had similarly unrelated records of $\delta$^18^O values and Mg/Ca ratios and pointed to no relationship between elemental ratios and SST at all.
 However, we were able to source other modern specimens of the same species from Tierra del Fuego, which shall serve here as an indicator.
-Similar to other patelloid shells, the results of the *Nacella* specimens in our study were successful with high R^2^-values (0.76--0.95) and suggest Mg/Ca as a reliable proxy for SST (@fig-Nac_Comp and supplementaries).
-Why these results are different from the studied *Nacella* sp.
+Similar to other patelloid shells, the results of the *Nacella* specimens in our study were successful with high R^2^ values (0.76--0.95) and suggest Mg/Ca as a reliable proxy for SST (@fig-Nac_Comp and supplementaries).
+Why these results are different from the studied *Nacella* spp.
 specimens in Graniero *et al*.
 [-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and $\delta$^18^O records of their shells (Figure 6 in Graniero *et al*.
-[-@Graniero2017-io]) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — is not perfectly visible, leading to an overall result of an uncorrelated relationship between Mg/Ca and SST.
+[-@Graniero2017-io]) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — is not perfectly visible, leading to an overall result of an uncorrelated relationship between Mg/Ca ratios and SST.
 
@@
 
-This study presents high-resolution and 2D information on Mg/Ca concentrations in modern and archaeological patelloid shells and their relationship to sea surface temperature.
-The results for *Patella vulgata*, *Nacella deaureata*, and *Nacella magellanica* were consistently promising with mean R^2^-values of 0.89, 0.86, and 0.88 respectively.
-
-LIBS elemental imaging provided some insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the $\delta$^18^O of the same shells.
+This study uses LIBS as a novel analytical tool to present high-resolution and 2D information on Mg/Ca concentrations in modern and archaeological patelloid shells and their relationship to sea surface temperature.
+The results for *P.
+vulgata*, *N.
+deaureata*, and *N.
+magellanica* were consistently promising with mean R^2^ values of 0.89, 0.86, and 0.88 respectively.
+
+LIBS elemental imaging provided insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the $\delta$^18^O values of the same shells.
 Our 2D approach allowed us to avoid these areas of intra-increment heterogeneity and focus on areas that are predominantly controlled by changes in SST.
@@
 Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a significant part of marine mollusc remains in archaeological sites of those regions [@Colonese2011-ab; @Villagran2011-ld; @Zangrando2016-yl].
-Improving their analysis in terms of cost and time efficiency through the benefits of LIBS, which can be used in this context as a screening tool for targeted $\delta$^18^O analysis focusing on annual minima and maxima revealed in the Mg/Ca record, will allow researchers to increase sample numbers and produce larger and more in-depth climatic studies than previously possible.
+By using LIBS as a screening tool for targeted \delta^{18}O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency.
+This novel method will enable them to increase sample numbers and produce larger, more in-depth climatic studies than previously possible.
 

### 2024-07-04 Manuscript.tex
Donna comments

@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
 \caption{Overview of the modern and archaeological limpet specimens
@@
 Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
-Bay - & QG1-7188-1 & Stable oxygen \\
-& & & Westray (UK) & QG1-7189-2 & and carbon \\
-& & & & QG2-1061-1 & isotope analysis \\
+Bay - & QG1-7188-1 & \(\delta\)\textsuperscript{18}O \\
+& & & Westray (UK) & QG1-7189-2 & and \(\delta\)\textsuperscript{13}C \\
+& & & & QG2-1061-1 & analysis \\
 & & & & QG2-1064-1 & \\
@@
 specimens at a resolution of 50--100 µm distance between sampling spots.
-Each spot was irradiated 10 times with the 3 first spectra discarded as
+Each spot was irradiated 10 times with the first 3 spectra discarded as
 a cleaning step and the remaining spectra summed to get an average Mg/Ca
@@
 gaps. Intensity ratios were filtered for cases with high relative
-standard deviation (i.e.~more than 10\%). These occurred in places where
-the previous sampling procedures for carbonate powder as part of the
-oxygen isotope analysis left an uneven sample surface introducing
+standard deviation (i.e., more than 10\%). These occurred in places
+where the previous sampling procedures for carbonate powder as part of
+the oxygen isotope analysis left an uneven sample surface introducing
 variability in the plasma generation and thus uncertainty in the data of
@@
 magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
-the stable isotope analysis methodology can be found. In short, all
-shells were sectioned along the main growth axis to expose the internal
-growth structure. From the exposed section carbonate powder samples were
-acquired using a micromill with resolutions of 100--200 µm and by
-targeting the calcitic M+2 layer of the shells. Analytical precision of
-the isotope values was consistently at 0.05--0.10‰ and values are
-reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee
-Belemnite) standard.
+the methodology for the stable isotope analyses can be found. In short,
+all shells were sectioned along the maximum growth axis to expose the
+internal growth structure. From the exposed section carbonate powder
+samples were acquired using a micromill with resolutions of 100--200 µm
+and by targeting the calcitic M+2 layer of the shells. Analytical
+precision of the isotope ratios was consistently at 0.05--0.10‰ and
+values are reported in per mil units (‰) relative to the VPDB (Vienna
+Pee Dee Belemnite) standard.
 
@@
 
-\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} sp. limpet
+\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} spp. limpet
 shells from Cambaceres Bay.}
@@
 previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
-limpet shells by species (\emph{Nacella} sp., and \emph{Patella
-vulgata}, respectively. Due to the high sampling resolution, all shells
+limpet shells by species (\emph{Nacella} spp., and \emph{Patella
+vulgata}, respectively). Due to the high-resolution sampling, all shells
 show seasonally resolved and multi-year quasi-sinusoidal records of SST
-change, with \emph{Nacella} sp. shells living 2--3 years and \emph{P.
+change, with \emph{Nacella} spp. shells living 2--3 years and \emph{P.
 vulgata} living regularly over 3 years. In those instances where longer
-lifetimes are encountered (i.e.~specimen QG1-7246-1,
-Figure~\ref{fig-Pat_iso} b) the temporal-resolution is reduced from more
-than a dozen of samples per year, to only around 7.
+lifespans resulted in slow growth rates below 1 mm/yr (i.e.~specimen
+QG1-7246-1, Figure~\ref{fig-Pat_iso} b), the temporal-resolution is
+reduced from more than a dozen of samples per year, to only around 7.
 
@@
 Because it is not possible to directly associate the exact sample
-locations used in the analysis of stable oxygen isotopes with the LIBS
+locations used in the analysis of oxygen isotope ratios with the LIBS
 data, we used dynamic time warping (DTW) to align the time series of
-\(\delta\)\textsuperscript{18}O-values and Mg/Ca ratios. DTW is an
-algorithm that measures similarity between two proxy-sequences, which
+\(\delta\)\textsuperscript{18}O values and Mg/Ca ratios. DTW is an
+algorithm that measures similarity between two proxy sequences, which
 may vary in sampling resolution or interval. By stretching or
 compressing sections of the series, DTW finds the probable alignment
-between the two sequences. This allows us to compare the proxy data sets
+between the two sequences. This allows us to compare the proxy datasets
 more effectively, ensuring that the temporal dynamics of each shell are
@@
 appropriate distance measures and constraints to ensure meaningful
-alignment. The process involved iterative adjustments to minimise the
-overall distance between corresponding points in the two data sets.
+alignment (see code in \citep{Hausmann_2024OSF}). The process involved
+iterative adjustments to minimise the overall distance between
+corresponding points in the two data sets.
 
@@
 repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
-intensity ratios indicating and low ratios indicating low temperature
-periods. As examples, Figure~\ref{fig-Pat_LIBS} shows the
-2D-distribution of elemental ratios across the entire section of ORK-LT5
-and the preserved anterior side of QG1-7188-1 (additional maps can be
-found in the supplementaries). The repeating patterns are found across
-the calcitic layers that are exterior to the myostracum (layers m+2 and
-m+3) as well as the layers that are interior (m-2)
-\citep{Fenger2007-gf}. The layer m-1, which is also interior, consists
-of aragonite and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its
-Mg/Ca intensity ratio consistently falls below the range of interest
-(0.2 and above). Compared to other patelloid shells, some of whose
-interior is almost entirely made of aragonite, this layer is very thin.
-\newline
+intensity ratios indicating high temperature periods and low ratios
+indicating low temperature periods. As examples,
+Figure~\ref{fig-Pat_LIBS} shows the 2D-distribution of elemental ratios
+across the entire section of ORK-LT5 and the preserved anterior side of
+QG1-7188-1 (additional maps can be found in the supplementaries). The
+repeating patterns are found across the calcitic layers that are
+exterior to the myostracum (layers m+2 and m+3) as well as the layers
+that are interior (m-2)(see Fenger et al. \citeyearpar{Fenger2007-gf}
+for a detailed explanations of microstructural layers in \emph{P.
+vulgaata}). The layer m-1, which is also interior, consists of aragonite
+and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its Mg/Ca intensity
+ratio consistently falls below the range of interest (0.2 and above).
+Compared to other patelloid shells, some of whose interior is almost
+entirely made of aragonite, this layer is very thin. \newline
 
@@
 Line scans also indicate well the quasi-sinusoidal change of Mg/Ca
-intensity ratios expected based on the stable isotope data and
-elemental-imaging. That said, some of the variability of Mg/Ca intensity
-ratios within one season such as the summer period of QG1-7188-1 between
-0.5 and 2.0 mm distance to the shell edge (Figure~\ref{fig-Pat_LIBS} d),
-is visible in a more dramatic manner than it appears in the 2D-image,
-showing the downside of line-scanning as opposed to the broader milling
+intensity ratios expected based on the stable isotope data and elemental
+imaging. That said, some of the variability of Mg/Ca intensity ratios
+within one season such as the summer period of QG1-7188-1 between 0.5
+and 2.0 mm distance to the shell edge (Figure~\ref{fig-Pat_LIBS} d), is
+visible in a more dramatic manner than it appears in the 2D image. This
+shows the downside of line-scanning as opposed to the broader milling
 approach used for the analysis of carbonate powder
@@
 Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
-to its \(\delta\)\textsuperscript{18}O-values (similar graphs for other
+to its \(\delta\)\textsuperscript{18}O values (similar graphs for other
 specimen can be found in the supplementaries). The measurements of the
@@
 Aligning the records programmatically using dynamic time warping,
-allowed us to compare them directly to determine specimen-specific
-equations and quantify the correlation of both proxies
-(Figure~\ref{fig-Pat_Corr}). The equations for all specimens are
-different but the majority (barring one: QG1-7246-1) do seem to cluster
-around shared parameters (\(\delta\)\textsuperscript{18}Ο = 0.6 -- 0.065
-* Mg/Ca). The various coefficients of determination
-(R\textsuperscript{2}) range between 0.81 (QG1-7246-1) and 0.95
-(QG1-7189-2). These specimens are also the oldest and youngest
-specimens, respectively, suggesting that lower growth rates and time
-averaging had a negative effect on the correlation of both proxies. The
-mean R\textsuperscript{2} value of the individual \emph{P. vulgata}
-specimens is 0.89, suggesting a generally good fit between the two
-proxies.
+allowed us to compare both proxies for each specimen directly to
+determine specimen-specific equations and quantify the correlation of
+both proxies (Figure~\ref{fig-Pat_Corr}). The equations for all
+specimens are different but the majority (barring one: QG1-7246-1) do
+seem to cluster around shared parameters
+(\(\delta\)\textsuperscript{18}Ο = 0.6 -- 0.065 * Mg/Ca). The various
+coefficients of determination (R\textsuperscript{2}) range between 0.81
+(QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest
+and youngest specimens, respectively, suggesting that lower growth rates
+and time averaging had a negative effect on the correlation of both
+proxies. The mean R\textsuperscript{2} value of the individual \emph{P.
+vulgata} specimens is 0.89, suggesting a generally good fit between the
+two proxies.
 
@@
 \subsection{\texorpdfstring{\emph{Nacella}
-sp.}{Nacella sp.}}\label{nacella-sp.}
+spp.}{Nacella spp.}}\label{nacella-spp.}
 
-The LIBS data for \emph{Nacella} sp. specimens is less straightforward
-with shells being much thinner and thus patterns being more difficult to
-make out. Both specimens (ND-1016-3 and NM-1016-3) shown as example in
-Figure~\ref{fig-Nac_LIBS} (see additional graphs in supplementaries)
-also experience increases of Mg/Ca intensity ratio towards the exterior
-of the shell, similar to the \emph{P. vulgata} specimens from Orkney.
-This intra-increment heterogeneity combined with the thinness of the
-shell, further complicates their analysis compared to \emph{P. vulgata}
-or other \emph{Patella} species \citep{Hausmann2019-fi}. Nevertheless,
-repeating patterns were visible in the anterior section of the shell,
-which were also mirrored in the inner layers (m-2) of specimens
-NM-1016-3 (Figure~\ref{fig-Nac_LIBS} b).
+The LIBS data for \emph{Nacella} spp. specimens is less straightforward
+with shells being much thinner than \emph{P. vulgata} and thus patterns
+being more difficult to discern. Both specimens (ND-1016-3 and
+NM-1016-3) shown as examples in Figure~\ref{fig-Nac_LIBS} (see
+additional graphs in supplementaries) also experience increases of Mg/Ca
+intensity ratio towards the exterior of the shell, similar to the
+\emph{P. vulgata} specimens from Orkney. This intra-increment
+heterogeneity combined with the thinness of the shell, further
+complicates their analysis compared to \emph{P. vulgata} or other
+\emph{Patella} species \citep{Hausmann2019-fi}. Nevertheless, repeating
+patterns were visible in the anterior section of the shell, which were
+also mirrored in the inner layers (m-2) of specimens NM-1016-3
+(Figure~\ref{fig-Nac_LIBS} b).
 
@@
 \caption{\label{fig-Nac_LIBS}Example LIBS maps and line scans of
-analysed \emph{Nacella} sp. specimens. (\emph{a+c}) ND-1016-3 and
+analysed \emph{Nacella} spp. specimens. (\emph{a+c}) ND-1016-3 and
 (\emph{b}+\emph{d}) NM-1016-3. Note that the ranges of Mg/Ca intensity
@@
 
-Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} sp.
+Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} spp.
 shells were much lower (0.05--0.30) than those seen in other patelloid
-species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in
+species (e.g.~0.3--0.7 in \emph{P. vulgata} above, or 0.5--1.5 in
 Hausmann \emph{et al}. \citeyearpar{Hausmann2023-ih}) using the same
@@
 the chosen Calcium peak, the low ratios still indicate lower
-concentrations than seen elsewhere, or seen only in aragonitic parts of
+concentrations than seen elsewhere or seen only in aragonitic parts of
 e.g.~\emph{Patella} shells. Similar observations have been made before
@@
 increment is not gradual, as is the case in NM-1016-3 (see inset of
-Figure~\ref{fig-Nac_LIBS} b), but rather step-wise. Sampling across this
+Figure~\ref{fig-Nac_LIBS} b), but rather stepwise. Sampling across this
 boundary in a linear scan would most likely lead to erroneous results.
@@
 Figure~\ref{fig-Nac_Comp} shows the line scans of ND-1016-3 and
-NM-1016-3 in comparison to their \(\delta\)\textsuperscript{18}O-values
-(similar graphs for the other \emph{Nacella} sp. specimen can be found
+NM-1016-3 in comparison to their \(\delta\)\textsuperscript{18}O values
+(similar graphs for the other \emph{Nacella} spp. specimens can be found
 in the supplementaries). Compared to the \emph{P. vulgata} records,
 these mirror each other much more clearly, due to the brief period
-recorded in the shells and its simple make-up.
+recorded in the shells and thus a lower chance of misaligning annual
+extremes.
 
@@
 \caption{\label{fig-Nac_Comp}Comparison of LIBS line-scan and sequential
-\(\delta\)\textsuperscript{18}O-values of \emph{Nacella} sp. specimens.}
+\(\delta\)\textsuperscript{18}O-values of \emph{Nacella} spp.
+specimens.}
 
@@
 
-The \emph{Nacella} sp. specimens showed similarly clustered
+The \emph{Nacella} spp. specimens showed similarly clustered
 correlations, which interestingly were not grouped by species. In fact,
-specimens ND-1016-3 (\emph{Nacella deaureata)} and NM-1016-3
-(\emph{Nacella magellanica}) are almost identical in their fitted
-equations (Figure~\ref{fig-Nac_Corr}). Both also share a high
-coefficient of determination (R\textsuperscript{2}) of 0.95 and 0.92
-respectively, with the remaining specimens being somewhat lower
-(R\textsuperscript{2} = 0.76 for ND-1016-4 and R\textsuperscript{2}=0.84
-for NM-1016-3). The mean R\textsuperscript{2} value for each
-\emph{Nacella} species is 0.86 for \emph{Nacella deaureata} and 0.88 for
-\emph{Nacella magellanica}. Why two specimens of different species are
-so well aligned is not entirely certain and randomness cannot for now be
-ruled out. That said, since they both have a high R\textsuperscript{2}
-value, it might just be that they were both minimally affected by
-factors other than SST, including the sampling location of our line
-scans.
+specimens ND-1016-3 (\emph{Nacella deaureata)} and NM-1016-3 (\emph{N.
+magellanica}) are almost identical in their fitted equations
+(Figure~\ref{fig-Nac_Corr}). Both also share a high coefficient of
+determination (R\textsuperscript{2}) of 0.95 and 0.92 respectively, with
+the remaining specimens being somewhat lower (R\textsuperscript{2} =
+0.76 for ND-1016-4 and R\textsuperscript{2}=0.84 for NM-1016-3). The
+mean R\textsuperscript{2} value for each \emph{N.} species is 0.86 for
+\emph{N. deaureata} and 0.88 for \emph{N. magellanica}. Why two
+specimens of different species are so well aligned is not entirely
+certain and randomness cannot for now be ruled out. That said, since
+they both have a high R\textsuperscript{2} value, it might just be that
+they were both minimally affected by factors other than SST, including
+the sampling location of our line scans.
 
@@
 
-\caption{\label{fig-Nac_Corr}Correlation graphs for \emph{Nacella} sp.
+\caption{\label{fig-Nac_Corr}Correlation graphs for \emph{Nacella} spp.
 specimens}
@@
 The Mg/Ca ratios from the 2D elemental imaging present positive results
-on SST being the main factor behind Mg/Ca concentration in three
-Atlantic limpet species. Annually repeating patterns of increasing
-(spring/summer) and decreasing (autum/winter) SST were also visible in
-the line scans, albeit less straightforward than in the 2D-data.
-Nevertheless, the resulting R\textsuperscript{2}-values for each species
-were high (\emph{P. vulgata mean R\textsuperscript{2}}=0.89, \emph{N.
-deaureata mean R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
+suggesting that SST is the main factor influencing variations in Mg/Ca
+ratios in three Atlantic limpet species. Annually repeating patterns of
+increasing (spring/summer) and decreasing (autum/winter) SST were also
+visible in the line scans, albeit less straightforward than in the
+2D-data. Nevertheless, the resulting R\textsuperscript{2}-values for
+each species were high (\emph{P. vulgata mean
+R\textsuperscript{2}}=0.89, \emph{N. deaureata mean
+R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
 R\textsuperscript{2}}=0.88). These results are similar to previously
-published data on \emph{Patella} sp. shells (Table
+published data on other \emph{Patella} species shells (Table
 \hyperref[Table_2]{2}): R\textsuperscript{2} values of \emph{P.
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1562}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1094}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2031}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4062}}@{}}
-\caption{Correlations of Mg/Ca with \(\delta\)\textsuperscript{18}O and
-SST found in other studies. Bold R\textsuperscript{2} values are
-interpreted as outliers. * indicates that only SST data and no other
-geochemical data was used. ** these samples were used to determine a
-shared R\textsuperscript{2}-value of 0.79.}\tabularnewline
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3067}}@{}}
+\caption{Correlations of Mg/Ca ratios with
+\(\delta\)\textsuperscript{18}O values and SST found in other studies.
+Bold R\textsuperscript{2} values are interpreted as outliers. *
+indicates that only SST data and no other geochemical data was used. **
+these samples were used to determine a shared R\textsuperscript{2}-value
+of 0.79.}\tabularnewline
 \toprule\noalign{}
@@
 \emph{P. rustica} has two studied specimens (JL1 and JL2) with
-R\textsuperscript{2}-values of 0.02 and 0.8, respectively
+R\textsuperscript{2}-values of 0.02 and 0.80, respectively
 \citep{Ferguson2011-zl} (Figure~\ref{fig-Ferg}). Together with the data
@@
 R\textsuperscript{2}-values of 0.69 and 0.83), the data from JL2 was
-used determine a general Mg/Ca-SST relationship with an
-R\textsuperscript{2}-value of 0.79. JL1 was disregarded in that
+used to determine a general Mg/Ca-SST relationship with an
+R\textsuperscript{2} value of 0.79. JL1 was disregarded in that
 calculation, as well as the second year of growth in JM00, which
 produced anomalous data. While no additional information about those
-specimens is available, it seems likely that discarded the Mg/Ca ratios
+specimens is available, it seems likely that the discarded Mg/Ca ratios
 from JM00, which were interpreted as an ontogenetic trend, were
@@
 by homogenising the carbonate powder of the m+2 layer and the more
-exterior m+3 layer, these heterogeneities could potentially increased
-the Mg/Ca ratio unpredictably, leading to what might look like an
-ontogenetic trend in a linear dataset. Whether these anomalies affected
-JL1 in the same way, albeit stronger, is not clear but it might well be
-possible.
+exterior m+3 layer, these heterogeneities could have potentially
+increased the Mg/Ca ratio unpredictably, leading to what might look like
+an ontogenetic trend in a linear dataset. Whether these anomalies
+affected JL1 in the same way, albeit stronger, is not clear but it might
+well be possible.
 
@@
 has previously been shown to have no substantial correlation with
-\(\delta\)\textsuperscript{18}O or SST in implication
-\citep{Graniero2017-io}. Different to other specimens of this study, we
-were able to resample this specimen and to acquire Mg/Ca data on a
+\(\delta\)\textsuperscript{18}O values or SST by implication
+\citep{Graniero2017-io}. In contrast to other specimens in this study,
+we were able to resample this specimen and acquire Mg/Ca data on a
 slightly higher resolution (20--30 µm spots in a 100 µm raster compared
@@
 
-We were not able to resample the \emph{Nacella} sp. specimens in
+We were not able to resample the \emph{Nacella} spp. specimens in
 Graniero \emph{et al.} \citeyearpar{Graniero2017-io}, which had
-similarly unrelated records of \(\delta\)\textsuperscript{18}O and Mg/Ca
-and pointed to no relationship between elemental ratios and SST at all.
-However, we were able to source other modern specimens of the same
-species from Tierra del Fuego, which shall serve here as an indicator.
-Similar to other patelloid shells, the results of the \emph{Nacella}
-specimens in our study were successful with high
-R\textsuperscript{2}-values (0.76--0.95) and suggest Mg/Ca as a reliable
+similarly unrelated records of \(\delta\)\textsuperscript{18}O values
+and Mg/Ca ratios and pointed to no relationship between elemental ratios
+and SST at all. However, we were able to source other modern specimens
+of the same species from Tierra del Fuego, which shall serve here as an
+indicator. Similar to other patelloid shells, the results of the
+\emph{Nacella} specimens in our study were successful with high
+R\textsuperscript{2} values (0.76--0.95) and suggest Mg/Ca as a reliable
 proxy for SST (Figure~\ref{fig-Nac_Comp} and supplementaries). Why these
-results are different from the studied \emph{Nacella} sp. specimens in
+results are different from the studied \emph{Nacella} spp. specimens in
 Graniero \emph{et al}. \citeyearpar{Graniero2017-io} is not obvious, but
@@
 visible, leading to an overall result of an uncorrelated relationship
-between Mg/Ca and SST.
+between Mg/Ca ratios and SST.
 
@@
 
-This study presents high-resolution and 2D information on Mg/Ca
-concentrations in modern and archaeological patelloid shells and their
-relationship to sea surface temperature. The results for \emph{Patella
-vulgata}, \emph{Nacella deaureata}, and \emph{Nacella magellanica} were
-consistently promising with mean R\textsuperscript{2}-values of 0.89,
-0.86, and 0.88 respectively.
+This study uses LIBS as a novel analytical tool to present
+high-resolution and 2D information on Mg/Ca concentrations in modern and
+archaeological patelloid shells and their relationship to sea surface
+temperature. The results for \emph{P. vulgata}, \emph{N. deaureata}, and
+\emph{N. magellanica} were consistently promising with mean
+R\textsuperscript{2} values of 0.89, 0.86, and 0.88 respectively.
 
-LIBS elemental imaging provided some insight into elemental anomalies
-found particulary in the external parts (m+3 layer) of the shell, which
+LIBS elemental imaging provided insight into elemental anomalies found
+particulary in the external parts (m+3 layer) of the shell, which
 introduce intra-incremental heterogeneity and which cannot be reconciled
 with modern SST changes or SST changes recorded in the
-\(\delta\)\textsuperscript{18}O of the same shells. Our 2D approach
-allowed us to avoid these areas of intra-increment heterogeneity and
-focus on areas that are predominantly controlled by changes in SST.
+\(\delta\)\textsuperscript{18}O values of the same shells. Our 2D
+approach allowed us to avoid these areas of intra-increment
+heterogeneity and focus on areas that are predominantly controlled by
+changes in SST.
 
@@
 those regions
-\citep{Colonese2011-ab, Villagran2011-ld, Zangrando2016-yl}. Improving
-their analysis in terms of cost and time efficiency through the benefits
-of LIBS, which can be used in this context as a screening tool for
-targeted \(\delta\)\textsuperscript{18}O analysis focusing on annual
-minima and maxima revealed in the Mg/Ca record, will allow researchers
-to increase sample numbers and produce larger and more in-depth climatic
-studies than previously possible.
+\citep{Colonese2011-ab, Villagran2011-ld, Zangrando2016-yl}. By using
+LIBS as a screening tool for targeted \delta\^{}\{18\}O analysis,
+focusing on annual minima and maxima revealed in the Mg/Ca record,
+researchers can improve cost and time efficiency. This novel method will
+enable them to increase sample numbers and produce larger, more in-depth
+climatic studies than previously possible.
 

### 2024-07-04 Manuscript.qmd
1st submission

@@
         postal-code: 27599-3315
-  - name: Ivan Briz-Godino
+  - name: Ivan Briz i Godino
     affiliations:
@@
 While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species.
-*Patella vulgata*, *Nacella deaureata*, and *N.
+*Patella vulgata*, *Nacella deaurata*, and *N.
 magellanica* have been studied using Mg/Ca ratios with mixed or contrary results.
-In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta&^18^O) that confirm a good relationship with SST.
+In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST.
 Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios.
@@
 vulgata*, *N.
-deaureata*, and *N.
+deaurata*, and *N.
 magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change.
@@
 #| warning: FALSE
-
+#knitr::opts_chunk$set(fig.path = "")
 knitr::opts_chunk$set(echo=FALSE,message = FALSE,warning = FALSE)
@@
   mutate(species = case_when(
-      startsWith(sampleID, "ND") ~ "Nacella deaureata",
+      startsWith(sampleID, "ND") ~ "Nacella deaurata",
       startsWith(sampleID, "NM") ~ "Nacella magellanica",
@@
 
-Limpet shells are commonly found within archaeological sites and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Robson2023-oz].
+Limpet shells are commonly found within archaeological sites and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Alvarez2011-rm, @Robson2023-oz].
 They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
@@
 | Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
-|------------|------------|------------|------------|------------|------------|
-| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C |
+|------|------------|-------|--------|------|----------|
+| Modern         | [@Nicastro2020-ih] | *N. deaurata*   | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C |
 |                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
@@
 vulgata* [@Surge2012-ba; @Graniero2017-io], *N.
-deaureata* and *N.
+deaurata* and *N.
 magellanica* [@Nicastro2020-ih], in which detailed descriptions of the methodology for the stable isotope analyses can be found.
@@
 [-@Fenger2007-gf] for a detailed explanations of microstructural layers in *P.
-vulgaata*).
+vulgata*).
 The layer m-1, which is also interior, consists of aragonite and is seen in @fig-Pat_LIBS in grey, as its Mg/Ca intensity ratio consistently falls below the range of interest (0.2 and above).
@@
 Both specimens in @fig-Pat_LIBS show increased Mg/Ca intensity ratios towards the outside of the shell and some degree of intra-increment variability.
-This is particularly visible in @fig-Pat_LIBS c, where Mg/Ca intensity ratios range from \~0.5–0.7 in the first year of growth (9–12 mm on the y-axis).
+This is particularly visible in @fig-Pat_LIBS b, where Mg/Ca intensity ratios range from \~0.5–0.7 in the first year of growth (9–12 mm on the y-axis).
 A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0–10 mm on the x-axis).
@@
 specimens showed similarly clustered correlations, which interestingly were not grouped by species.
-In fact, specimens ND-1016-3 (*Nacella deaureata)* and NM-1016-3 (*N.
+In fact, specimens ND-1016-3 (*Nacella deaurata)* and NM-1016-3 (*N.
 magellanica*) are almost identical in their fitted equations (@fig-Nac_Corr).
@@
 The mean R^2^ value for each *N.* species is 0.86 for *N.
-deaureata* and 0.88 for *N.
+deaurata* and 0.88 for *N.
 magellanica*.
@@
 vulgata mean R^2^*=0.89, *N.
-deaureata mean R^2^*=0.86*, N.
+deaurata mean R^2^*=0.86*, N.
 magellanica mean R^2^*=0.88).
@@
 vulgata*, *N.
-deaureata*, and *N.
+deaurata*, and *N.
 magellanica* were consistently promising with mean R^2^ values of 0.89, 0.86, and 0.88 respectively.
@@
 This novel method will enable them to increase sample numbers and produce larger, more in-depth climatic studies than previously possible.
+
+# Acknowledgements
 

### 2024-07-04 Manuscript.tex
1st submission

@@
   pdftitle={Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells},
-  pdfauthor={Niklas Hausmann; Donna Surge; Ivan Briz-Godino},
+  pdfauthor={Niklas Hausmann; Donna Surge; Ivan Briz i Godino},
   pdfkeywords={Sclerochronology, Patella, Nacella, LIBS, Mg/Ca},
@@
 
-\author[3]{Ivan Briz-Godino%
+\author[3]{Ivan Briz i Godino%
 %
@@
 Mg/Ca ratios, this relationship is not the case for other patelloid
-species. \emph{Patella vulgata}, \emph{Nacella deaureata}, and \emph{N.
+species. \emph{Patella vulgata}, \emph{Nacella deaurata}, and \emph{N.
 magellanica} have been studied using Mg/Ca ratios with mixed or contrary
 results. In this study, we present elemental maps of these three species
-together with oxygen isotope ratios (\$\delta\&\textsuperscript{18}O)
+together with oxygen isotope ratios (\(\delta\)\textsuperscript{18}O)
 that confirm a good relationship with SST. Our dataset also reassesses a
@@
 archaeological specimens (n=12) from three patelloid species (\emph{P.
-vulgata}, \emph{N. deaureata}, and \emph{N. magellanica}) we further add
+vulgata}, \emph{N. deaurata}, and \emph{N. magellanica}) we further add
 to the growing set of evidence for the reliable use of Mg/Ca ratios to
@@
 shorelines due to their robust carbonate structure and the long-term use
-of limpets as a marine food source \citep{Robson2023-oz}. They have been
+of limpets as a marine food source
+\citep[\citet{Robson2023-oz}]{Alvarez2011-rm}. They have been
 successfully studied in the context of coastal subsistence economies and
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
 \caption{Overview of the modern and archaeological limpet specimens
@@
 \endlastfoot
-Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
-& ND-1016-3 & \(\delta\)\textsuperscript{18}O and
+Modern & \citep{Nicastro2020-ih} & \emph{N. deaurata} & Cambaceres Bay &
+ND-1016-3 & \(\delta\)\textsuperscript{18}O and
 \(\delta\)\textsuperscript{13}C \\
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-SSTs-1.pdf}
+\includegraphics{fig-SSTs-1.pdf}
 
@@
 previous publications on \emph{P. vulgata}
-\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
+\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaurata} and \emph{N.
 magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.png}
+\includegraphics{fig-Nac_iso-1.png}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_iso-1.png}
+\includegraphics{fig-Pat_iso-1.png}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_LIBS-1.pdf}
+\includegraphics{fig-Pat_LIBS-1.pdf}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Comp-1.png}
+\includegraphics{fig-Pat_Comp-1.png}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Corr-1.png}
+\includegraphics{fig-Pat_Corr-1.png}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_LIBS-1.pdf}
+\includegraphics{fig-Nac_LIBS-1.pdf}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Comp-1.png}
+\includegraphics{fig-Nac_Comp-1.png}
 
@@
 correlations, which interestingly were not grouped by species. In fact,
-specimens ND-1016-3 (\emph{Nacella deaureata)} and NM-1016-3 (\emph{N.
+specimens ND-1016-3 (\emph{Nacella deaurata)} and NM-1016-3 (\emph{N.
 magellanica}) are almost identical in their fitted equations
@@
 mean R\textsuperscript{2} value for each \emph{N.} species is 0.86 for
-\emph{N. deaureata} and 0.88 for \emph{N. magellanica}. Why two
-specimens of different species are so well aligned is not entirely
-certain and randomness cannot for now be ruled out. That said, since
-they both have a high R\textsuperscript{2} value, it might just be that
-they were both minimally affected by factors other than SST, including
-the sampling location of our line scans.
+\emph{N. deaurata} and 0.88 for \emph{N. magellanica}. Why two specimens
+of different species are so well aligned is not entirely certain and
+randomness cannot for now be ruled out. That said, since they both have
+a high R\textsuperscript{2} value, it might just be that they were both
+minimally affected by factors other than SST, including the sampling
+location of our line scans.
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Corr-1.png}
+\includegraphics{fig-Nac_Corr-1.png}
 
@@
 each species were high (\emph{P. vulgata mean
-R\textsuperscript{2}}=0.89, \emph{N. deaureata mean
+R\textsuperscript{2}}=0.89, \emph{N. deaurata mean
 R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1733}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3067}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
 \caption{Correlations of Mg/Ca ratios with
@@
 Study
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
 \end{minipage} \\
@@
 \emph{P. depressa} & Northern Spain & LAN541 & 0.87 &
-\citep{Garcia-Escarzaga2021-ij} \\
-& & LAN545 & 0.86 & \\
-& & LAN554 & 0.78 & \\
-& & LAN559 & 0.82 & \\
-\emph{P. caerulea} & Croatia & ISTPC1 & 0.9 & \citep{Hausmann2019-fi} \\
-& & ISTPC2 & 0.84 & \\
-& Crete & AF1911A & 0.91 & \\
-& & AF3003A & 0.92 & \\
-& Israel & AKKPC2 & 0.96 & \\
-& & AKKPC3 & 0.89 & \\
-& & FRMPC1 & 0.84 & \\
-& & FRMPC2 & 0.96 & \\
-& Libya & MO31A & 0.83 & \\
-& & \textbf{MP64A} & \textbf{0.33} & \\
-& & MP67A & 0.96 & \\
-& & MP68A & 0.81 & \\
-& Malta & MA10 & 0.82 & \\
-& Tunisia & TUNPC1 & 0.81 & \\
-& & TUNPC2 & 0.78 & \\
-& Turkey & ANTPC1 & 0.95 & \\
-& & ANTPC2 & 0.93 & \\
-& & KIZPC1 & 0.94 & \\
-& & KIZPC2 & 0.86 & \\
+\citep{Garcia-Escarzaga2021-ij} & \\
+& & LAN545 & 0.86 & & \\
+& & LAN554 & 0.78 & & \\
+& & LAN559 & 0.82 & & \\
+\emph{P. caerulea} & Croatia & ISTPC1 & 0.9 & \citep{Hausmann2019-fi}
+& \\
+& & ISTPC2 & 0.84 & & \\
+& Crete & AF1911A & 0.91 & & \\
+& & AF3003A & 0.92 & & \\
+& Israel & AKKPC2 & 0.96 & & \\
+& & AKKPC3 & 0.89 & & \\
+& & FRMPC1 & 0.84 & & \\
+& & FRMPC2 & 0.96 & & \\
+& Libya & MO31A & 0.83 & & \\
+& & \textbf{MP64A} & \textbf{0.33} & & \\
+& & MP67A & 0.96 & & \\
+& & MP68A & 0.81 & & \\
+& Malta & MA10 & 0.82 & & \\
+& Tunisia & TUNPC1 & 0.81 & & \\
+& & TUNPC2 & 0.78 & & \\
+& Turkey & ANTPC1 & 0.95 & & \\
+& & ANTPC2 & 0.93 & & \\
+& & KIZPC1 & 0.94 & & \\
+& & KIZPC2 & 0.86 & & \\
 \emph{P. rustica} & Gibraltar & \textbf{JL1} & \textbf{0.02} &
-\citep{Ferguson2011-zl} \\
-& & JL2 & 0.8** & \\
-\emph{P. caerulea} & Gibraltar & JM00 & 0.69** & \\
-& & JM30 & 0.83** & \\
+\citep{Ferguson2011-zl} & \\
+& & JL2 & 0.8** & & \\
+\emph{P. caerulea} & Gibraltar & JM00 & 0.69** & & \\
+& & JM30 & 0.83** & & \\
 \emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
-\citep{Graniero2017-io} and this study \\
+\citep{Graniero2017-io} and this study & \\
 \end{longtable}
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-Ferg-1.png}
+\includegraphics{fig-Ferg-1.png}
 
@@
 
-\includegraphics{Manuscript_files/figure-pdf/fig-ORK_sub-1.pdf}
+\includegraphics{fig-ORK_sub-1.pdf}
 
@@
 archaeological patelloid shells and their relationship to sea surface
-temperature. The results for \emph{P. vulgata}, \emph{N. deaureata}, and
+temperature. The results for \emph{P. vulgata}, \emph{N. deaurata}, and
 \emph{N. magellanica} were consistently promising with mean
@@
 \renewcommand\refname{References}
-  \bibliography{bibliography.bib}
+ \bibliographystyle{spmpsci}
+\bibliography{bibliography}
 

### 2024-07-04 Upload/main.tex
1st submission

--- 2024-07-04 Manuscript.tex
@@
 results. In this study, we present elemental maps of these three species
-together with oxygen isotope ratios (\(\delta\)\textsuperscript{18}O)
+together with oxygen isotope ratios ($\delta$\textsuperscript{18}O)
 that confirm a good relationship with SST. Our dataset also reassesses a
 specimen which was previously unsuccessful in providing significant
-correlations between \(\delta\)\textsuperscript{18}O values and Mg/Ca
+correlations between $\delta$\textsuperscript{18}O values and Mg/Ca
 ratios. By reassessing these species and including modern and
@@
 temperature minima and maxima) for the targeted analysis using
-\(\delta\)\textsuperscript{18}O and clumped isotope analysis.
+$\delta$\textsuperscript{18}O and clumped isotope analysis.
 \end{abstract}
@@
 patelloid limpet shells largely relies on the measurement of
-\(\delta\)\textsuperscript{18}O values within the calcitic parts of the
+$\delta$\textsuperscript{18}O values within the calcitic parts of the
 shell, but attempts have been made to also use elemental ratios, such as
@@
 that potentially provides a more accurate SST estimate than
-\(\delta\)\textsuperscript{18}O-values,as the latter proxy is affected
+$\delta$\textsuperscript{18}O-values,as the latter proxy is affected
 by changes in the mixing of fresh- and saltwater. In addition, the data
@@
 published datasets, we were able to simultaneously avoid costs for new
-high-resolution \(\delta\)\textsuperscript{18}O-data, to use real-world
+high-resolution $\delta$\textsuperscript{18}O-data, to use real-world
 examples, and to also provide pilot-data for areas of existing research
@@
 Modern & \citep{Nicastro2020-ih} & \emph{N. deaurata} & Cambaceres Bay &
-ND-1016-3 & \(\delta\)\textsuperscript{18}O and
-\(\delta\)\textsuperscript{13}C \\
+ND-1016-3 & $\delta$\textsuperscript{18}O and
+$\delta$\textsuperscript{13}C \\
 & & & - Tierra del & ND-1016-4 & analysis \\
@@
 & \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - &
-ORK-LT5 & \(\delta\)\textsuperscript{18}O and
-\(\delta\)\textsuperscript{13}C \\
+ORK-LT5 & $\delta$\textsuperscript{18}O and
+$\delta$\textsuperscript{13}C \\
 & & & Westray (UK) & & analysis; \\
@@
 Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
-Bay - & QG1-7188-1 & \(\delta\)\textsuperscript{18}O \\
-& & & Westray (UK) & QG1-7189-2 & and \(\delta\)\textsuperscript{13}C \\
-& & & & QG2-1061-1 & analysis \\
+Bay - & QG1-7188-1 & $\delta$\textsuperscript{18}O and \(\delta\)\textsuperscript{13}C \\
+& & & Westray (UK) & QG1-7189-2 & analysis \\
+& & & & QG2-1061-1 &  \\
 & & & & QG2-1064-1 & \\
@@
 August 2009 and has been previously analysed for
-\(\delta\)\textsuperscript{18}O and \(\delta\)\textsuperscript{13}C
+$\delta$\textsuperscript{18}O and \(\delta\)\textsuperscript{13}C
 values, as well as LA-ICP-MS \citep{Graniero2017-io}. Westray lies about
@@
 
-High-quality \(\delta\)\textsuperscript{18}O values were acquired from
+High-quality $\delta$\textsuperscript{18}O values were acquired from
 previous publications on \emph{P. vulgata}
@@
 Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
-previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
+previously acquired sequential $\delta$\textsuperscript{18}O-values of
 limpet shells by species (\emph{Nacella} spp., and \emph{Patella
@@
 data, we used dynamic time warping (DTW) to align the time series of
-\(\delta\)\textsuperscript{18}O values and Mg/Ca ratios. DTW is an
+$\delta$\textsuperscript{18}O values and Mg/Ca ratios. DTW is an
 algorithm that measures similarity between two proxy sequences, which
@@
 for a detailed explanations of microstructural layers in \emph{P.
-vulgaata}). The layer m-1, which is also interior, consists of aragonite
+vulgata}). The layer m-1, which is also interior, consists of aragonite
 and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its Mg/Ca intensity
@@
 intra-increment variability. This is particularly visible in
-Figure~\ref{fig-Pat_LIBS} c, where Mg/Ca intensity ratios range from
+Figure~\ref{fig-Pat_LIBS} b, where Mg/Ca intensity ratios range from
 \textasciitilde0.5--0.7 in the first year of growth (9--12 mm on the
@@
 Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
-to its \(\delta\)\textsuperscript{18}O values (similar graphs for other
+to its $\delta$\textsuperscript{18}O values (similar graphs for other
 specimen can be found in the supplementaries). The measurements of the
@@
 \caption{\label{fig-Pat_Comp}Comparison of LIBS line-scan and sequential
-\(\delta\)\textsuperscript{18}O-values of specimen ORK-LT5}
+$\delta$\textsuperscript{18}O-values of specimen ORK-LT5}
 
@@
 seem to cluster around shared parameters
-(\(\delta\)\textsuperscript{18}Ο = 0.6 -- 0.065 * Mg/Ca). The various
+($\delta$\textsuperscript{18}O = 0.6 -- 0.065 * Mg/Ca). The various
 coefficients of determination (R\textsuperscript{2}) range between 0.81
@@
 Line scans reflect well the changes previously indicated by
-\(\delta\)\textsuperscript{18}O values and also indicate about two years
+$\delta$\textsuperscript{18}O values and also indicate about two years
 of recorded growth. That said, the 2D imaging suggests that in both
@@
 Figure~\ref{fig-Nac_Comp} shows the line scans of ND-1016-3 and
-NM-1016-3 in comparison to their \(\delta\)\textsuperscript{18}O values
+NM-1016-3 in comparison to their $\delta$\textsuperscript{18}O values
 (similar graphs for the other \emph{Nacella} spp. specimens can be found
@@
 \caption{\label{fig-Nac_Comp}Comparison of LIBS line-scan and sequential
-\(\delta\)\textsuperscript{18}O-values of \emph{Nacella} spp.
+$\delta$\textsuperscript{18}O-values of \emph{Nacella} spp.
 specimens.}
@@
 \caption{Correlations of Mg/Ca ratios with
-\(\delta\)\textsuperscript{18}O values and SST found in other studies.
+$\delta$\textsuperscript{18}O values and SST found in other studies.
 Bold R\textsuperscript{2} values are interpreted as outliers. *
@@
 has previously been shown to have no substantial correlation with
-\(\delta\)\textsuperscript{18}O values or SST by implication
+$\delta$\textsuperscript{18}O values or SST by implication
 \citep{Graniero2017-io}. In contrast to other specimens in this study,
@@
 for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca
-and \(\delta\)\textsuperscript{18}O-values}
+and $\delta$\textsuperscript{18}O}
 
@@
 Graniero \emph{et al.} \citeyearpar{Graniero2017-io}, which had
-similarly unrelated records of \(\delta\)\textsuperscript{18}O values
+similarly unrelated records of $\delta$\textsuperscript{18}O values
 and Mg/Ca ratios and pointed to no relationship between elemental ratios
@@
 the occasional alignment of the Mg/Ca and
-\(\delta\)\textsuperscript{18}O records of their shells (Figure 6 in
+$\delta$\textsuperscript{18}O records of their shells (Figure 6 in
 Graniero \emph{et al}. \citeyearpar{Graniero2017-io}) indicate that
@@
 with modern SST changes or SST changes recorded in the
-\(\delta\)\textsuperscript{18}O values of the same shells. Our 2D
+$\delta$\textsuperscript{18}O values of the same shells. Our 2D
 approach allowed us to avoid these areas of intra-increment
@@
 \citep{Colonese2011-ab, Villagran2011-ld, Zangrando2016-yl}. By using
-LIBS as a screening tool for targeted \delta\^{}\{18\}O analysis,
+LIBS as a screening tool for targeted $\delta$\textsuperscript{18}O analysis,
 focusing on annual minima and maxima revealed in the Mg/Ca record,
@@
 
+\section{Acknowledgements}\label{acknowledgements}
+Niklas Hausmann was supported by the German Research Foundation 
+under the Emmy Noether Program (project number: 439799406). The authors 
+also thank the Applied Photonics Lab of the Institute of Electronic Structure and 
+Laser of the Foundation for Research and Technology - Hellas for close support 
+with the LIBS system.
+
 \renewcommand\refname{References}
- \bibliographystyle{spmpsci}
 \bibliography{bibliography}

### 2024-07-05 Manuscript.qmd
1st submission

@@
   elsevier-pdf:
-    keep-tex: true
+    keep-tex: FALSE
     latex_engine: xelatex
@@
 We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data.
-This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in [@Hausmann_2024OSF]).
+This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in [@Hausmann_Surge_Godino_2024]).
 The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.
@@
 
-Another shell whose results we aim to explain further is ORK-LT5, which has previously been shown to have no substantial correlation with $\delta$^18^O values or SST by implication [@Graniero2017-io].
+Another shell whose results we aim to explain further is ORK-LT5 (@fig-Pat_LIBS), which has previously been shown to have no substantial correlation with $\delta$^18^O values or SST by implication [@Graniero2017-io].
 In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
@@
 Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a significant part of marine mollusc remains in archaeological sites of those regions [@Colonese2011-ab; @Villagran2011-ld; @Zangrando2016-yl].
-By using LIBS as a screening tool for targeted \delta^{18}O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency.
+By using LIBS as a screening tool for targeted $\delta$^18^O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency.
 This novel method will enable them to increase sample numbers and produce larger, more in-depth climatic studies than previously possible.

### 2024-07-05 DTW_Script.R

@@ -1,90 +1,0 @@
-shell_dtw_data <- function(D18o, MgCa, show.dtw.plot = FALSE, plot_alignment = FALSE, return_data = TRUE) {
-  require(tidyverse)
-  require(dtw)
-  require(ggplotify)
-  
-  if (is.null(show.dtw.plot)) 
-    show.dtw.plot <- FALSE
-  
-  plot_alignment_local <- plot_alignment
-  data_to_return <- NULL
-  
-  suppressMessages({
-    suppressWarnings({
-      D18o_warp <- D18o %>% mutate(d18o = scale(d18o)) %>%
-        pull(d18o)
-      MgCa_warp <- MgCa %>% mutate(mg_ca = scale(mg_ca)) %>%
-        pull(mg_ca)
-      alignment <- dtw(D18o_warp * -1,
-                       MgCa_warp,
-                       keep = TRUE,
-                       step.pattern = symmetric1)
-      Aligned_d18o <- D18o %>% mutate(x = 1:length(D18o$d18o),
-                                      proxy = "d18o") %>% rename(value = d18o) %>%
-        select(x, value, proxy)
-      Aligned_mg_ca <-
-        MgCa %>% pull(mg_ca) %>% .[alignment$index2] %>%
-        as.data.frame() %>% clean_names() %>% rename(value = x) %>%
-        cbind(x = alignment$index1) %>% mutate(proxy = "mg_ca")
-      
-      # Code for plotting alignment data
-      if (plot_alignment) {
-        plot(alignment,
-             type = "twoway",
-             offset = 10)
-      }
-      
-      Data_aligned <- rbind(Aligned_d18o, Aligned_mg_ca) %>%
-        group_by(x, proxy) %>% summarise(value = mean(value)) %>%
-        ungroup() %>% mutate(value = round(value + 1e-05 *
-                                             row_number(), 5)) %>% group_by(proxy, value) %>%
-        summarise(x = mean(x))
-      lmp <- function(modelobject) {
-        if (class(modelobject) != "lm")
-          stop("Not an object of class 'lm' ")
-        f <- summary(modelobject)$fstatistic
-        p <- pf(f[1], f[2], f[3], lower.tail = F)
-        attributes(p) <- NULL
-        return(p)
-      }
-      corr_data <-
-        Data_aligned %>% pivot_wider(names_from = "proxy",
-                                     values_from = "value") %>% select(d18o, mg_ca)
[…]

### 2024-07-06 Manuscript.qmd
added linenumbers

@@
   units: cm
+header-includes:
+  - \usepackage{lineno}
+  - \linenumbers
 ---
@@
 |                |                    |                  |                 |            |                                       |
-| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O                         |
-|                |                    |                  | Westray (UK)    | QG1-7189-2 | and $\delta$^13^C                            |
-|                |                    |                  |                 | QG2-1061-1 | analysis                      |
+| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O and $\delta$^13^C       |
+|                |                    |                  | Westray (UK)    | QG1-7189-2 |       analysis                        |
+|                |                    |                  |                 | QG2-1061-1 |                                       |
 |                |                    |                  |                 | QG2-1064-1 |                                       |

### 2024-08-26 Manuscript.qmd
added spaces

@@
 They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
-Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of $\delta$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than $\delta$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater.
+Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of \$\\delta\$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than \$\\delta\$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater.
 In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].
@@
 | Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
-|------|------------|-------|--------|------|----------|
-| Modern         | [@Nicastro2020-ih] | *N. deaurata*   | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C        |
+|------------|------------|------------|------------|------------|------------|
+| Modern         | [@Nicastro2020-ih] | *N. deaurata*    | Cambaceres Bay  | ND-1016-3  | \$\\delta\$^18^O and \$\\delta\$^13^C |
 |                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
@@
 |                |                    |                  |                 |            |                                       |
-|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | $\delta$^18^O and $\delta$^13^C       |
+|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | \$\\delta\$^18^O and \$\\delta\$^13^C |
 |                |                    |                  | Westray (UK)    |            | analysis;                             |
@@
 |                |                    |                  |                 |            |                                       |
-| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O and $\delta$^13^C       |
+| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | \$\\delta\$^18^O and \$\\delta\$^13^C |
 |                |                    |                  | Westray (UK)    | QG1-7189-2 | analysis                              |
@@
 
-| Species       | Locality       | Specimen  | R^2^                    | Study                             |
-|------|------------|-------|--------|------|----------|
-| *P. depressa* | Northern Spain | LAN541    | 0.87                    | [@Garcia-Escarzaga2021-ij]        |
-|               |                | LAN545    | 0.86                    |                                   |
-|               |                | LAN554    | 0.78                    |                                   |
-|               |                | LAN559    | 0.82                    |                                   |
-| *P. caerulea* | Croatia        | ISTPC1    | 0.9                     | [@Hausmann2019-fi]                |
-|               |                | ISTPC2    | 0.84                    |                                   |
-|               | Crete          | AF1911A   | 0.91                    |                                   |
-|               |                | AF3003A   | 0.92                    |                                   |
-|               | Israel         | AKKPC2    | 0.96                    |                                   |
-|               |                | AKKPC3    | 0.89                    |                                   |
-|               |                | FRMPC1    | 0.84                    |                                   |
-|               |                | FRMPC2    | 0.96                    |                                   |
-|               | Libya          | MO31A     | 0.83                    |                                   |
-|               |                | **MP64A** | **0.33**                |                                   |
-|               |                | MP67A     | 0.96                    |                                   |
-|               |                | MP68A     | 0.81                    |                                   |
-|               | Malta          | MA10      | 0.82                    |                                   |
-|               | Tunisia        | TUNPC1    | 0.81                    |                                   |
-|               |                | TUNPC2    | 0.78                    |                                   |
-|               | Turkey         | ANTPC1    | 0.95                    |                                   |
-|               |                | ANTPC2    | 0.93                    |                                   |
-|               |                | KIZPC1    | 0.94                    |                                   |
-|               |                | KIZPC2    | 0.86                    |                                   |
-| *P. rustica*  | Gibraltar      | **JL1**   | **0.02**                | [@Ferguson2011-zl]                |
-|               |                | JL2       | 0.8\*\*                 |                                   |
-| *P. caerulea* | Gibraltar      | JM00      | 0.69\*\*                |                                   |
-|               |                | JM30      | 0.83\*\*                |                                   |
-| *P. vulgata*  | Orkney         | ORK-LT5   | not reported, here 0.88 | [@Graniero2017-io] and this study |
+| Species       | Locality       | Specimen  | R^2^                    | Study                             |     |
+|------------|------------|------------|------------|------------|------------|
+| *P. depressa* | Northern Spain | LAN541    | 0.87                    | [@Garcia-Escarzaga2021-ij]        |     |
+|               |                | LAN545    | 0.86                    |                                   |     |
+|               |                | LAN554    | 0.78                    |                                   |     |
+|               |                | LAN559    | 0.82                    |                                   |     |
+| *P. caerulea* | Croatia        | ISTPC1    | 0.9                     | [@Hausmann2019-fi]                |     |
+|               |                | ISTPC2    | 0.84                    |                                   |     |
+|               | Crete          | AF1911A   | 0.91                    |                                   |     |
+|               |                | AF3003A   | 0.92                    |                                   |     |
+|               | Israel         | AKKPC2    | 0.96                    |                                   |     |
+|               |                | AKKPC3    | 0.89                    |                                   |     |
+|               |                | FRMPC1    | 0.84                    |                                   |     |
+|               |                | FRMPC2    | 0.96                    |                                   |     |
+|               | Libya          | MO31A     | 0.83                    |                                   |     |
+|               |                | **MP64A** | **0.33**                |                                   |     |
+|               |                | MP67A     | 0.96                    |                                   |     |
+|               |                | MP68A     | 0.81                    |                                   |     |
+|               | Malta          | MA10      | 0.82                    |                                   |     |
+|               | Tunisia        | TUNPC1    | 0.81                    |                                   |     |
+|               |                | TUNPC2    | 0.78                    |                                   |     |
+|               | Turkey         | ANTPC1    | 0.95                    |                                   |     |
+|               |                | ANTPC2    | 0.93                    |                                   |     |
+|               |                | KIZPC1    | 0.94                    |                                   |     |
+|               |                | KIZPC2    | 0.86                    |                                   |     |
+| *P. rustica*  | Gibraltar      | **JL1**   | **0.02**                | [@Ferguson2011-zl]                |     |
+|               |                | JL2       | 0.8\*\*                 |                                   |     |
+| *P. caerulea* | Gibraltar      | JM00      | 0.69\*\*                |                                   |     |
+|               |                | JM30      | 0.83\*\*                |                                   |     |
+| *P. vulgata*  | Orkney         | ORK-LT5   | not reported, here 0.88 | [@Graniero2017-io] and this study |     |
 

### 2024-09-12 Manuscript.qmd
Added time warped graphs into supplementary and changed dtw function to provide standard deviation for warped series

@@
   This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST).
-While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species.
-*Patella vulgata*, *Nacella deaurata*, and *N.
-magellanica* have been studied using Mg/Ca ratios with mixed or contrary results.
-In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST.
+While *Patella depressa* along the Spanish shoreline and *Patella caerulea* in the Mediterranean have repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species.
+*Patella vulgata*, *Nacella deaurata* have been studied using Mg/Ca ratios with mixed or contrary results.
+In this study, we present elemental maps of these two species as well as *Nacella magellanica* together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST.
 Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios.
@@
 
+source(here("data","IsoError.R"))
+IsoError <- IsoError %>% 
+  arrange(sampleID)
+
 MgCa <- read_csv(here("data","MgCa.csv"))
@@
 However, there remains a lack of clarity for Atlantic limpet species, particularly since it was shown, that the Mg/Ca ratio in their shells are not reliable recorders of palaeotemperature [@Graniero2017-io].
-
-In this study we will repeat and expand the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies.
+Using modern reference material Graniero *et al*.
+[-@Granieo2017-io] tested multiple elemental ratios (Mg/Ca, Sr/Ca, Li/Ca, Li/Mg, and Sr/Li) acquired via LA-ICP-MS measurements with no ratios showing correlations with corresponding SST changes nor any sinusoidal patterns that would indicate a seasonal environmental control of elemental ratios.
+This study included two species also used in our study (*Patella vulgata*, *Nacella deaurata*), with one *P.
+vulgata* specimen (i.e.
+specimen ORK-LT5) being analysed by both.
+We also added specimens of the species *Nacella magellanica,* which — similar to the other two species — is also an established palaeothermometer using \$\\delta\$^18^O-values [@Colonese2012-ct; @gutiérrez-zugasti2017; @Nicastro2020-ih].
+
+In this study, we repeat and expand upon the analysis of Atlantic limpets to determine the reliability of Mg/Ca ratios as palaeotemperature proxies.
 To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS.
-LIBS allows us to carry out 2D imaging of entire shell sections, which allows navigation of the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.
-By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
+LIBS allows us to carry out 2D imaging of entire shell sections, which facilitates navigation of the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.
+By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution \$\\delta\$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
 Generally establishing the usefulness and reliability of Mg/Ca ratios as SST proxy in limpets provides a platform for future research and an important steppingstone to a better understanding of elemental ratios in other marine mollusc shells.
@@
 The analysed specimens, their origin and respective studies with research background information can be found in Table [1](#Table_1).
-Here we will briefly summarise their contextual information, which can be accessed in more detail at the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io].
+Here we will briefly summarise their contextual information, which can be accessed in more detail from the respective studies [@Nicastro2020-ih; @Surge2012-ba; @Graniero2017-io].
 While those studies also included other specimens, their accessibility or state of preservation did not allow them to be re-analysed.
 
-```{r SST_Data}
+```{r SS_Data}
 
@@
 Westray_SST <- read_csv(here("data","Westray-SST.csv"), comment = "#")
+Westray_SSS <- read_csv(here("data","Westray-SSS.csv"), comment = "#")
 TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")
+TdF_SSS <- read_csv(here("data","Tierra_del_Fuego-SSS.csv"), comment = "#")
 
@@
 
+# Combine the datasets with new names and add year and month columns
+SSS_df <- bind_rows(Westray_SSS %>% mutate(dataset = "Westray - Orkney"),
+                         TdF_SSS %>% mutate(dataset = "Cambaceres Bay - Tierra del Fuego")) %>%
+  mutate(analysed_sss = so,
+         time = as_datetime(time),
+         year = year(time),
+         month = month(time, label = TRUE)) %>%
+  group_by(dataset, year, month) %>%
+  summarise(mean_sss = mean(analysed_sss, na.rm = TRUE),
+            sd_sss = sd(analysed_sss, na.rm = TRUE))
+
 WestrayMin <- SST_df %>% filter(dataset=="Westray - Orkney") %>% 
@@
   pull(mean_sst) %>% max() %>% round(1)
+
+WestrayMin <- SSS_df %>% filter(dataset=="Westray - Orkney") %>% 
+  pull(mean_sss) %>% min() %>% round(1)
+WestrayMax <- SSS_df %>% filter(dataset=="Westray - Orkney") %>% 
+  pull(mean_sss) %>% max() %>% round(1)
+
+CamBayMin <- SSS_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
+  pull(mean_sss) %>% min() %>% round(1)
+CamBayMax <- SSS_df %>% filter(dataset=="Cambaceres Bay - Tierra del Fuego") %>% 
+  pull(mean_sss) %>% max() %>% round(1)
 
@@
 
-On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for $\delta$^18^O and $\delta$^13^C values, as well as LA-ICP-MS [@Graniero2017-io].
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for \$\\delta\$^18^O and \$\\delta\$^13^C values, as well as via LA-ICP-MS [@Graniero2017-io].
 Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
@@
 ```{r SST_plot,fig.pos='H'}
-#| fig-cap: Monthly Sea Surface Temperature values averaged from  years. SST values were sourced from @Good2020-nl.
+#| fig-cap: Monthly sea surface temperature values averaged from 2007 to 2023, monthly sea surface salinity values from 1993 to 2021. Values were sourced from @Good2020-nl.
 #| fig-height: 3
@@
 # Plot the data
-SST_df %>% 
+SST_plot <- SST_df %>% 
     group_by(dataset, month) %>%
@@
 
+# Plot the data
+SSS_plot <- SSS_df %>% 
+    group_by(dataset, month) %>%
+  summarise(sd_sss = sd(mean_sss, na.rm = TRUE),
+            mean_sss = mean(mean_sss, na.rm = TRUE)) %>%
+ggplot(aes(x = month, y = mean_sss, colour = dataset,group=dataset)) +
+  geom_ribbon(aes(ymin = mean_sss - sd_sss, ymax = mean_sss + sd_sss,fill=dataset), alpha = 0.2,col="white") +
+  geom_path() +
+  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
+scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
+  labs(title = "",
+       x = "Month",
+       y = "Mean Analysed SSS (so)",colour="",fill="") +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1),
+          legend.position = "none")+  
+  scale_y_continuous(breaks = c(32,33, 34,35,36),limits = c(32,36))
+
+SST_plot +SSS_plot
+
 ```
@@
 These shells have previously been studied to characterise seasonal temperature change during the Medieval Climate Anomaly (810 -- 1229 CE) [@Surge2012-ba].
-Together they show a range of lifespans, with some comparatively long lived for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to within-shell time-averaging because of slow growth rates.
-That said, the majority are under 5 years old yielding sufficiently fast growth to nearly capture seasonal amplitudes of SST.
 
@@
 
-### Mg/Ca ratios
-
-Mg/Ca elemental imaging was carried out using LIBS at the Leibniz Zentrum für Archäologie (Mainz-Germany) following previously published methods [@Hausmann2023-ih].
-The imaging involves the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume.
-This plume emits light, which is measured using a synchronised spectrometer (delay time = 0.5 µs, integration time = 1 µs).
-The resulting light spectrum quantifies the emission lines of Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine their intensity ratio.
-While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g.
-[@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca variation within the shell carbonate.
-All values reported here are in arbitrary units based on the ratio of intensities of the peaks above.
-
-Using this system we carried out elemental imaging of the shell specimens at a resolution of 50--100 µm distance between sampling spots.
-Each spot was irradiated 10 times with the first 3 spectra discarded as a cleaning step and the remaining spectra summed to get an average Mg/Ca intensity ratio for each sample spot.
-Subsequently, we re-sampled the section using a line scan at 10 µm resolution.
-This leads to an overlap between sample spots, but also allows for a continuous record without gaps.
-Intensity ratios were filtered for cases with high relative standard deviation (i.e., more than 10%).
-These occurred in places where the previous sampling procedures for carbonate powder as part of the oxygen isotope analysis left an uneven sample surface introducing variability in the plasma generation and thus uncertainty in the data of these locations.
-
 ### Oxygen isotope ratios
 
-High-quality $\delta$^18^O values were acquired from previous publications on *P.
+High-quality \$\\delta\$^18^O values were acquired from previous publications on *P.
 vulgata* [@Surge2012-ba; @Graniero2017-io], *N.
@@
 In short, all shells were sectioned along the maximum growth axis to expose the internal growth structure.
-From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells.
+From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic layers of the shells.
+The exact location of the sampled carbonate powder is not available apart from the distance to the growth edge, where reported.
 Analytical precision of the isotope ratios was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.
@@
     geom_path() +
+    geom_errorbar(aes(ymin = d18O - 0.1, ymax = d18O + 0.1), width = 0.2) +
     geom_point() +    labs(
@@
 
-@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively).
+@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential \$\\delta\$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively).
 Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* spp.
@@
 specimen QG1-7246-1, @fig-Pat_iso b), the temporal-resolution is reduced from more than a dozen of samples per year, to only around 7.
+Together they show a range of lifespans, with some comparatively long lived for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to within-shell time-averaging because of slow growth rates.
+That said, the majority are under 5 years old yielding sufficiently fast growth to nearly capture seasonal amplitudes of SST.
 
@@
 # Filter data for Modern period
-Patella_iso <- D18O %>% filter(grepl("Patella", species))
+Patella_iso <- D18O %>% filter(grepl("Patella", species)) #%>%
 
@@
 sample_ids <- unique(Patella_iso$sampleID)
+
+Patella_iso <- Patella_iso %>% 
+  arrange(sampleID) %>%
+  bind_cols(d18Osd=IsoError$d18Osd)
 
@@
     geom_path() +
+    geom_errorbar(aes(ymin = d18O - d18Osd, ymax = d18O + d18Osd), width = 0.2) +
     geom_point() +
@@
 
+### Mg/Ca ratios
+
+Mg/Ca elemental imaging was carried out using LIBS at the Leibniz Zentrum für Archäologie (Mainz-Germany) following previously published methods [@Hausmann2023-ih].
+The imaging involves the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume.
+This plume emits light, which is measured using a synchronised spectrometer (delay time = 0.5 µs, integration time = 1 µs).
+The resulting light spectrum quantifies the emission lines of Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine their intensity ratio.
+While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g.
+[@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca variation within the shell carbonate.
+All values reported here are in arbitrary units based on the ratio of intensities of the peaks above.
+
+Using this system we carried out elemental imaging of the shell specimens at a resolution of 50--100 µm distance between sampling spots.
+Each spot was irradiated 10 times with the first 3 spectra discarded as a cleaning step and the remaining spectra summed to get an average Mg/Ca intensity ratio for each sample spot.
+Subsequently, we re-sampled the section using a line scan at 10 µm resolution.
+This leads to an overlap between sample spots, but also allows for a continuous record without gaps.
+Intensity ratios were filtered for cases with high relative standard deviation (i.e., more than 10%).
+These occurred in places where the previous sampling procedures for carbonate powder as part of the oxygen isotope analysis left an uneven sample surface introducing variability in the plasma generation and thus uncertainty in the data of these locations.
+
+###
+
 ## Dynamic Time Warping
 
-Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O values and Mg/Ca ratios.
+Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of \$\\delta\$^18^O values and Mg/Ca ratios.
 DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval.
@@
 We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data.
-This involved selecting appropriate distance measures and constraints to ensure meaningful alignment (see code in [@Hausmann_Surge_Godino_2024]).
-The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.
+This can involve selecting appropriate distance measures and constraints to ensure meaningful alignment.
+We used the default settings for symmetric step patterns and euclidean distance measures (see code in [@Hausmann_Surge_Godino_2024]).
 
@@
   # Append temp_data to Corr_data
-  Corr_data <- bind_rows(Corr_data, temp_data)
+  Corr_data <- bind_rows(Corr_data, temp_data) %>% 
+    arrange(x)
 }
+
+write_csv(Corr_data,here("data","Corr_data.csv"))
 
@@
 
-@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O values (similar graphs for other specimen can be found in the supplementaries).
+@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its \$\\delta\$^18^O values (similar graphs for other specimen can be found in the supplementaries).
 The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior.
@@
 Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr).
-The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 – 0.065 \* Mg/Ca).
+The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters (\$\\delta\$^18^Ο = 0.6 – 0.065 \* Mg/Ca).
 The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2).
@@
     formula = y~x
+  )+
+  stat_cor(
+    aes(label = paste("P-value: ", ..p..)),
+    label.x = 0.2, label.y = 0.6, 
+    size = 6
   )+
@@
 
-Line scans reflect well the changes previously indicated by $\delta$^18^O values and also indicate about two years of recorded growth.
+Line scans reflect well the changes previously indicated by \$\\delta\$^18^O values and also indicate about two years of recorded growth.
 That said, the 2D imaging suggests that in both specimens a third summer can be added to the total record, which is not captured by the line scans.
 
-@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their $\delta$^18^O values (similar graphs for the other *Nacella* spp.
+@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their \$\\delta\$^18^O values (similar graphs for the other *Nacella* spp.
 specimens can be found in the supplementaries).
@@
 | *P. vulgata*     | Orkney           | ORK-LT5    | not reported, here 0.88 | [@Graniero2017-io] and this study |     |
-
-: Correlations of Mg/Ca ratios with $\delta$^18^O values and SST found in other studies.
+|                  |                  | QG1-7188-1 | 0.88                    | this study                        |     |
+|                  |                  | QG1-7189-2 | 0.95                    |                                   |     |
+|                  |                  | QG1-7246-1 | 0.81                    |                                   |     |
+|                  |                  | QG2-1061-1 | 0.90                    |                                   |     |
+|                  |                  | QG2-1064-1 | 0.90                    |                                   |     |
+|                  |                  | QG2-7180-1 | 0.87                    |                                   |     |
+|                  |                  | QG2-7180-2 | 0.90                    |                                   |     |
+| *N. deaurata*    | Tierra del Fuego | ND-1016-3  | 0.95                    |                                   |     |
+|                  |                  | ND-1016-4  | 0.76                    |                                   |     |
+| *N. magellanica* | Tierra del Fuego | NM-1016-1  | 0.92                    |                                   |     |
+|                  |                  | NM-1016-3  | 0.84                    |                                   |     |
+
+: Correlations of Mg/Ca ratios with \$\\delta\$^18^O values and SST found in other studies.
 Bold R^2^ values are interpreted as outliers.
@@
 Whether these anomalies affected JL1 in the same way, albeit stronger, is not clear but it might well be possible.
+Previous studies of *P.
+caerulea* found similar patterns of anomalous Mg/Ca values in the m+3 layer [@Hausmann2019-fi].
+These areas can be avoided using a 2-dimensional approach and if the growth increment is sufficiently wide.
+These heterogeneities within growth increments are little understood but some research points towards a zonation of the mollusc's mantle and along the growth edge [@Lazareth2013-jf].
+This zonation could differentially influence the transport of Mg ions to the extrapallial fluid and into the shell.
 
@@
 
-Another shell whose results we aim to explain further is ORK-LT5 (@fig-Pat_LIBS), which has previously been shown to have no substantial correlation with $\delta$^18^O values or SST by implication [@Graniero2017-io].
+Another shell whose results we aim to explain further is ORK-LT5 (@fig-Pat_LIBS), which has previously been shown to have no substantial correlation with \$\\delta\$^18^O values or SST by implication [@Graniero2017-io].
 In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
@@
 We were not able to resample the *Nacella* spp.
-specimens in Graniero *et al.* [-@Graniero2017-io], which had similarly unrelated records of $\delta$^18^O values and Mg/Ca ratios and pointed to no relationship between elemental ratios and SST at all.
+specimens in Graniero *et al.* [-@Graniero2017-io], which had similarly unrelated records of \$\\delta\$^18^O values and Mg/Ca ratios and pointed to no relationship between elemental ratios and SST at all.
 However, we were able to source other modern specimens of the same species from Tierra del Fuego, which shall serve here as an indicator.
@@
 specimens in Graniero *et al*.
-[-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and $\delta$^18^O records of their shells (Figure 6 in Graniero *et al*.
+[-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and \$\\delta\$^18^O records of their shells (Figure 6 in Graniero *et al*.
 [-@Graniero2017-io]) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — is not perfectly visible, leading to an overall result of an uncorrelated relationship between Mg/Ca ratios and SST.
@@
 
-LIBS elemental imaging provided insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the $\delta$^18^O values of the same shells.
+LIBS elemental imaging provided insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the \$\\delta\$^18^O values of the same shells.
 Our 2D approach allowed us to avoid these areas of intra-increment heterogeneity and focus on areas that are predominantly controlled by changes in SST.
@@
 Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a significant part of marine mollusc remains in archaeological sites of those regions [@Colonese2011-ab; @Villagran2011-ld; @Zangrando2016-yl].
-By using LIBS as a screening tool for targeted $\delta$^18^O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency.
+By using LIBS as a screening tool for targeted \$\\delta\$^18^O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency.
 This novel method will enable them to increase sample numbers and produce larger, more in-depth climatic studies than previously possible.

### 2024-09-12 Supplementary files 2 - isotope graphs.qmd
Added time warped graphs into supplementary and changed dtw function to provide standard deviation for warped series

@@
 
+Corr_data <- read.csv(here("data","Corr_data.csv"))
+
+#This below caused problems
+#%>% 
+  #   pivot_wider(values_from = "value",names_from = "proxy") %>%
+  # left_join(D18O %>% select(sampleID) %>% unique(),by="sampleID") %>%
+  # pivot_longer(cols=c(d18o,mg_ca),values_to = "value",names_to = "proxy") %>% 
+  # arrange(x)
+
 # # Check overlaps
@@
 
-  P1 <-
+  P1.1 <-
     MgCa %>%
@@
 
-  P2 <-
+  P1.2 <-
     D18O %>%
@@
 
-  P1 / P2 + plot_annotation(tag_levels = "a")
+   P2.1 <-
+    Corr_data %>%
+    filter(sampleID == sample) %>%
+    filter(proxy == "mg_ca") %>%
+    ggplot() +
+    aes(x, value) +
+    # geom_errorbar(aes(ymin = value - sd, ymax = value + sd, width = 1)) +
+    geom_ribbon(aes(ymin = value - sd, ymax = value + sd), alpha = 0.5) +
+    geom_path() +
+    theme_cowplot(17) +
+    labs(x = "", y = "")
+
+  P2.2 <-
+    Corr_data %>%
+    filter(sampleID == sample) %>%
+    filter(proxy == "d18o") %>%
+          group_by(x) %>% 
+      mutate(sd=sd(value)) %>% 
+      ungroup() %>% 
+    ggplot() +
+    aes(x, value) +
+    # geom_errorbar(aes(ymin = value - sd, ymax = value + sd, width = 0.2)) +
+    geom_ribbon(aes(ymin = value - sd, ymax = value + sd), alpha = 0.5) +
+    geom_path() +
+    theme_cowplot(18) +
+    scale_y_reverse() +
+    labs(x = "Warped time index", y = "")
+
+    P1.1 / P1.2 |  P2.1 / P2.2 + plot_annotation(tag_levels = "a")
+
 }

### 2024-09-12 DTW_Script-update.R

@@ -19,13 +19,22 @@ shell_dtw_data <- function(D18o, MgCa, show.dtw.plot = FALSE, plot_alignment = F
                        MgCa_warp,
                        keep = TRUE,
                        step.pattern = symmetric2)
+      
       Aligned_d18o <- D18o %>% mutate(x = 1:length(D18o$d18o),
                                       proxy = "d18o") %>% rename(value = d18o) %>%
-        select(x, value, proxy)
+        select(x, value, proxy) %>% 
+        group_by(x) %>% 
+        mutate(sd=sd(value,na.rm=TRUE)) %>% 
+        ungroup()
+      
+      
       Aligned_mg_ca <-
         MgCa %>% pull(mg_ca) %>% .[alignment$index2] %>%
         as.data.frame() %>% clean_names() %>% rename(value = x) %>%
-        cbind(x = alignment$index1) %>% mutate(proxy = "mg_ca")
+        cbind(x = alignment$index1) %>% mutate(proxy = "mg_ca") %>% 
+        group_by(x) %>% 
+        mutate(sd=sd(value,na.rm=TRUE)) %>% 
+        ungroup()
       
       # Code for plotting alignment data
       if (plot_alignment) {
@@ -35,9 +44,9 @@ shell_dtw_data <- function(D18o, MgCa, show.dtw.plot = FALSE, plot_alignment = F
       }
       
       Data_aligned <- rbind(Aligned_d18o, Aligned_mg_ca) %>%
-        group_by(x, proxy) %>% summarise(value = mean(value)) %>%
+        group_by(x, proxy,sd) %>% summarise(value = mean(value)) %>%
         ungroup() %>% mutate(value = round(value + 1e-05 *
-                                             row_number(), 5)) %>% group_by(proxy, value) %>%
+                                             row_number(), 5)) %>% group_by(proxy, value,sd) %>%
         summarise(x = mean(x))
       lmp <- function(modelobject) {
         if (class(modelobject) != "lm")

### 2024-09-12 data/IsoError.R

@@ -1,0 +1,635 @@
+IsoError <- tibble::tribble(
+           ~sampleID,   ~d18Osd,
+"QG1-7188-1", 0.005,
+"QG1-7188-1", 0.024,
+"QG1-7188-1",  0.02,
+"QG1-7188-1", 0.008,
+"QG1-7188-1", 0.062,
+"QG1-7188-1", 0.045,
+"QG1-7188-1", 0.071,
+"QG1-7188-1", 0.064,
+"QG1-7188-1", 0.086,
+"QG1-7188-1", 0.002,
+"QG1-7188-1", 0.011,
+"QG1-7188-1",   0.1,
+"QG1-7188-1", 0.047,
+"QG1-7188-1", 0.051,
+"QG1-7188-1", 0.021,
+"QG1-7188-1", 0.079,
+"QG1-7188-1", 0.043,
+"QG1-7188-1",  0.07,
+"QG1-7188-1", 0.074,
+"QG1-7188-1", 0.012,
+"QG1-7188-1", 0.014,
+"QG1-7188-1", 0.106,
+"QG1-7188-1", 0.327,
+"QG1-7188-1", 0.049,
+"QG1-7188-1", 0.092,
+"QG1-7188-1", 0.021,
+"QG1-7188-1", 0.066,
+"QG1-7188-1", 0.014,
+"QG1-7188-1", 0.029,
+"QG1-7188-1", 0.043,
+"QG1-7188-1", 0.072,
+"QG1-7188-1", 0.018,
+"QG1-7188-1", 0.009,
+"QG1-7188-1", 0.016,
+"QG1-7188-1", 0.028,
+"QG1-7188-1", 0.028,
+"QG1-7188-1", 0.057,
+"QG1-7188-1", 0.054,
+"QG1-7188-1", 0.083,
+"QG1-7188-1", 0.044,
+"QG1-7188-1",  0.05,
+"QG1-7188-1", 0.044,
+"QG1-7188-1", 0.018,
+"QG1-7188-1", 0.023,
+"QG1-7188-1", 0.027,
+"QG1-7188-1",  0.11,
+"QG1-7188-1", 0.081,
+"QG1-7188-1", 0.093,
+"QG1-7188-1", 0.044,
+"QG1-7188-1", 0.049,
+  "QG1-7189-2",  0.09,
+  "QG1-7189-2",   0.1,
+  "QG1-7189-2", 0.082,
+  "QG1-7189-2", 0.071,
+  "QG1-7189-2", 0.028,
[…]

### 2024-09-13 Manuscript.qmd
Corrections

@@
 Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
+White dashed lines indicate location of line scans.
 #| fig-height: 6
@@
 vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies.
+P-values are overwhelmingly significant with the highest being p=1e^-23^.
 
@@
     label.x = 0.2, label.y = 0.6, 
-    size = 6
+    size = 6,show.legend = FALSE
   )+
@@
     strip.text = element_text(colour = "black"),
-    panel.border = element_rect(colour="black")
+    panel.border = element_rect(colour="black"),
+    legend.position = "none"
   )
@@
 Note the inset in (*c*) where the shell edge has been resampled using a higher resolution (50 µm instead of 100 µm) to better understand the Mg/Ca ratios at the shell edge.
+White dashed lines indicate location of line scans.
 #| fig-height: 7
@@
 deaurata* and 0.88 for *N.
-magellanica*.
+magellanica* with p-values for all speciemns being far below 0.0001.
 Why two specimens of different species are so well aligned is not entirely certain and randomness cannot for now be ruled out.
@@
     coord_fixed(ratio = 6) +
+  stat_cor(
+  aes(label = paste("P-value: ", round(..p.., 10))),
+    label.x = 0.2, label.y = 0.3, 
+    size = 6,show.legend = FALSE
+  )+
   theme(
@@
 
-The Mg/Ca ratios from the 2D elemental imaging present positive results suggesting that SST is the main factor influencing variations in Mg/Ca ratios in three Atlantic limpet species.
+The Mg/Ca ratios from the 2D elemental imaging present results suggesting that SST is the main factor influencing variations in Mg/Ca ratios in three Atlantic limpet species.
 Annually repeating patterns of increasing (spring/summer) and decreasing (autum/winter) SST were also visible in the line scans, albeit less straightforward than in the 2D-data.
-Nevertheless, the resulting R^2^-values for each species were high (*P.
-vulgata mean R^2^*=0.89, *N.
-deaurata mean R^2^*=0.86*, N.
-magellanica mean R^2^*=0.88).
+Nevertheless, the resulting mean R^2^-values comparin Mg/Ca and δ^18^O values for each species were high (*P.
+vulgata R^2^*=0.89, *N.
+deaurata R^2^*=0.86*, N.
+magellanica R^2^*=0.88).
 These results are similar to previously published data on other *Patella* species shells (Table [2](#Table_2)): R^2^ values of *P.
@@
 | *P. rustica*     | Gibraltar        | **JL1**    | **0.02**                | [@Ferguson2011-zl]                |     |
-|                  |                  | JL2        | 0.8\*\*                 |                                   |     |
-| *P. caerulea*    | Gibraltar        | JM00       | 0.69\*\*                |                                   |     |
-|                  |                  | JM30       | 0.83\*\*                |                                   |     |
+|                  |                  | JL2        | 0.8\*                   |                                   |     |
+| *P. caerulea*    | Gibraltar        | JM00       | 0.69\*                  |                                   |     |
+|                  |                  | JM30       | 0.83\*                  |                                   |     |
 | *P. vulgata*     | Orkney           | ORK-LT5    | not reported, here 0.88 | [@Graniero2017-io] and this study |     |
@@
 
-: Correlations of Mg/Ca ratios with \$\\delta\$^18^O values and SST found in other studies.
+: Correlations of Mg/Ca ratios with \$\\delta\$^18^O values found in other studies.
 Bold R^2^ values are interpreted as outliers.
-\* indicates that only SST data and no other geochemical data was used.
-\*\* these samples were used to determine a shared R^2^-value of 0.79.
+\* indicates samples that were used to determine a shared R^2^-value of 0.79.
 :::
@@
 *P.
-rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.80, respectively [@Ferguson2011-zl] (@fig-Ferg).
+rustica* has two studied specimens (JL1 and JL2) with R^2^-values of 0.02 and 0.80, respectively [@Ferguson2011-zl] (see also Supplementaries).
 Together with the data from two *P.
@@
 
-```{r Ferguson Data, dev='png'}
-#| fig-cap: Correlation graphs for Ferguson et al., (2011) specimens.
-#| fig-height: 5
-#| fig-width: 16
-#| label: fig-Ferg
-
-Ferg_all <-
-  Ferguson_data %>% 
-  ggplot()+
-  aes(d18o,mg_ca,col=shell)+
-  geom_point(show.legend = FALSE)+
-  labs(x=expression(paste(delta^{18},"O (\u2030)")),y="Mg/Ca (mmol/mol)")+
-  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
-  # stat_regline_equation(label.x = 2, #label.y = 32
-  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
-  #   formula = y~x
-  # ) +
-  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
-scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))
-
-Ferg_individual <- Ferg_all+
-  facet_wrap(~shell,ncol=4)+#ylim(0.15,NA)+
-  ggpubr::stat_regline_equation(label.x = 0.3, label.y = 26,
-    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
-    formula = y~x
-  )+  
-  theme_cowplot(20)+
-  coord_fixed(ratio =0.22) +
-  theme(
-    strip.background = element_rect(fill = "white", colour = NA),
-    strip.text = element_text(colour = "black"),
-    panel.border = element_rect(colour="black")
-  )
-Ferg_individual
-#Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
-```
-
-Another shell whose results we aim to explain further is ORK-LT5 (@fig-Pat_LIBS), which has previously been shown to have no substantial correlation with \$\\delta\$^18^O values or SST by implication [@Graniero2017-io].
+Another shell whose results we aim to explain further is ORK-LT5 (see Supplementary Information), which has previously been shown to have no substantial correlation with \$\\delta\$^18^O values or SST by implication [@Graniero2017-io].
 In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
@@
 
+## Temperature correlation
+
+We further related the Mg/Ca ratios of the modern shells to actual SST data from the respective localities (@fig-SSTs).
+The data for ORK-LT5 from Orkney were cut to the time period between the start of our available SST data (2007) and the time of collection (August 2009).
+For the *Nacella* spp.
+from Cambaceres Bay we chose an overlapping time period, to make the results more comparable between the specimens, which is the lifespan of the youngest specimen NM-1016-3 from January 2015 to October 2016, the time of collection for all 4 specimens.
+
+As indicated by the stable isotope values, each shell individually has a good correlation with SST (R^2^ values between 0.93 to 0.96 with p-values consistently below 0.001) (@fig-SST_Corr), which are even better than the correlation between elemental ratios and stable isotope values.
+This improvement is most likely due to the simple and sinusoidal SST curve, which is easier to fit to other data.
+
+```{r SST-Corr-data}
+reference <- 
+  TdF_SST %>% 
+  filter(time<"2016-10-01") %>%
+  filter(time>"2015-01-01") %>%
+  mutate(analysed_sst = analysed_sst - 273.15)
+
+query <- MgCa %>% 
+  filter(grepl("1016",sampleID)) %>% 
+  filter(!(sampleID=="ND-1016-4"&dist>2.3)) %>% 
+  filter(!(sampleID=="NM-1016-1"&dist>7.2)) %>% 
+  filter(!(sampleID=="ND-1016-3"&dist>6)) %>% 
+  arrange(-dist)
+
+TdF_plot <- reference %>% 
+  ggplot()+
+  aes(time,analysed_sst)+
+  geom_path()+
+labs(x="",y="SST in ºC")
+
+Nacella_MgCa_plot <- query %>% 
+  ggplot()+
+  aes(dist,mg_ca)+
+  geom_path()+
+  # scale_x_reverse()+
+  labs(x="Distance to shell edge in mm", y="Mg/Ca intensity ratio")+
+  facet_wrap(~sampleID,ncol=1,scales="free")
+
+#SST_P1 <- TdF_plot/Nacella_MgCa_plot+plot_layout(heights=c(1,5))
+
+Samples_SST <- unique(query$sampleID)
+
+Corr_data_SST <- tibble()
+
+for (sample in Samples_SST) {
+
+  temp_reference <- reference %>% 
+    select(analysed_sst) %>% 
+    mutate(sample_seq = row_number())
+
+  temp_MgCa <- query %>% 
+    filter(sampleID == sample) %>% 
+    select(mg_ca) %>% 
+    mutate(sample_seq = row_number())
+
+   Ref_warp <- temp_reference %>% mutate(analysed_sst = scale(analysed_sst)) %>%
+        pull(analysed_sst)
+   MgCa_warp <- temp_MgCa %>% mutate(mg_ca = scale(mg_ca)) %>%
+        pull(mg_ca)
+
+      alignment <- dtw(Ref_warp,
+                       MgCa_warp,
+                       keep = TRUE,
+                       step.pattern = symmetric2)
+
+      Aligned_ref <- temp_reference %>% mutate(x = 1:length(temp_reference$analysed_sst),
+                                      proxy = "sst") %>% rename(value = analysed_sst) %>%
+        select(x, value, proxy) %>% 
+        group_by(x) %>% 
+        mutate(sd=sd(value,na.rm=TRUE)) %>% 
+        ungroup()
+
+      Aligned_mg_ca <-
+        temp_MgCa %>% pull(mg_ca) %>% .[alignment$index2] %>%
+        as.data.frame() %>% clean_names() %>% rename(value = x) %>%
+        cbind(x = alignment$index1) %>% mutate(proxy = "mg_ca") %>% 
+        group_by(x) %>% 
+        mutate(sd=sd(value,na.rm=TRUE)) %>% 
+        ungroup()
+
+      # # Code for plotting alignment data
+        # plot(alignment,
+        #      type = "twoway",
+        #      offset = 10)
+
+      Data_aligned <- rbind(Aligned_ref, Aligned_mg_ca) %>%
+        group_by(x, proxy,sd) %>% summarise(value = mean(value)) %>%
+        ungroup() %>% mutate(value = round(value + 1e-05 *
+                                             row_number(),5)) %>% group_by(proxy, value,sd) %>%
+        summarise(x = mean(x))
+
+      lmp <- function(modelobject) {
+        if (class(modelobject) != "lm")
+          stop("Not an object of class 'lm' ")
+        f <- summary(modelobject)$fstatistic
+        p <- pf(f[1], f[2], f[3], lower.tail = F)
+        attributes(p) <- NULL
+        return(p)
+      }
+      # corr_data <-
+      #   Data_aligned %>% pivot_wider(names_from = "proxy",
+      #                                values_from = "value",values_fn = mean) %>% select(sst, mg_ca)
+
+      # LM <- lm(sst ~ mg_ca, data = corr_data,)
+      # R2 <- round(summary(LM)[[8]], 2)
+      # P_value <-
+      #   ifelse(lmp(LM) < 0.001,
+      #          "p<0.001",
+      #          ifelse(
+      #            lmp(LM) <
+      #              0.01,
+      #            "p<0.01",
+      #            ifelse(lmp(LM) < 0.05, "p<0.05",
+      #                   "ns!!!")
+      #          ))
+      # Stats <- paste0("R²=", R2, ", ", P_value)
+
+  # Add the sampleID column to temp_data
+  temp_data <- Data_aligned %>% 
+    mutate(sampleID = sample) %>% filter(!(sampleID=="ND-1016-4"&proxy=="mg_ca"&value>0.25))
+
+  # Append temp_data to Corr_data
+  Corr_data_SST <- bind_rows(Corr_data_SST, temp_data) %>% 
+    arrange(x)
+
+}
+
+##################### ADDING ORK-5
+
+reference <- 
+  Westray_SST %>% 
+  filter(time<"2009-08-15") %>%
+  mutate(analysed_sst = analysed_sst - 273.15)
+
+sample <- "ORK_LT5"
+
+query <- MgCa %>% 
+  filter(sampleID==sample) %>% 
+    filter(relstd < 0.1) %>%
+  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
+  filter(dist<4) %>% 
+  arrange(-dist)
+
+  temp_reference <- reference %>% 
+    select(analysed_sst) %>% 
+    mutate(sample_seq = row_number())
+
+  temp_MgCa <- query %>% 
+    filter(sampleID == sample) %>% 
+    select(mg_ca) %>% 
+    mutate(sample_seq = row_number())
+
+   Ref_warp <- temp_reference %>% mutate(analysed_sst = scale(analysed_sst)) %>%
+        pull(analysed_sst)
+
+   MgCa_warp <- temp_MgCa %>% mutate(mg_ca = scale(mg_ca)) %>%
+        pull(mg_ca)
+
+      alignment <- dtw(Ref_warp,
+                       MgCa_warp,
+                       keep = TRUE,
+                       step.pattern = symmetric2)
+
+      Aligned_ref <- temp_reference %>% mutate(x = 1:length(temp_reference$analysed_sst),
+                                      proxy = "sst") %>% rename(value = analysed_sst) %>%
+        select(x, value, proxy) %>% 
+        group_by(x) %>% 
+        mutate(sd=sd(value,na.rm=TRUE)) %>% 
+        ungroup()
+
+      Aligned_mg_ca <-
+        temp_MgCa %>% pull(mg_ca) %>% .[alignment$index2] %>%
+        as.data.frame() %>% clean_names() %>% rename(value = x) %>%
+        cbind(x = alignment$index1) %>% mutate(proxy = "mg_ca") %>% 
+        group_by(x) %>% 
+        mutate(sd=sd(value,na.rm=TRUE)) %>% 
+        ungroup()
+
+      # # Code for plotting alignment data
+        # plot(alignment,
+        #      type = "twoway",
+        #      offset = 10)
+
+      Data_aligned <- rbind(Aligned_ref, Aligned_mg_ca) %>%
+        group_by(x, proxy,sd) %>% summarise(value = mean(value)) %>%
+        ungroup() %>% mutate(value = round(value + 1e-05 *
+                                             row_number(),5)) %>% group_by(proxy, value,sd) %>%
+        summarise(x = mean(x))
+
+      # lmp <- function(modelobject) {
+      #   if (class(modelobject) != "lm")
+      #     stop("Not an object of class 'lm' ")
+      #   f <- summary(modelobject)$fstatistic
+      #   p <- pf(f[1], f[2], f[3], lower.tail = F)
+      #   attributes(p) <- NULL
+      #   return(p)
+      # }
+      # corr_data <-
+      #   Data_aligned %>% pivot_wider(names_from = "proxy",
+      #                                values_from = "value",values_fn = mean) %>% select(sst, mg_ca)
+      # 
+      # 
+      # 
+      # LM <- lm(sst ~ mg_ca, data = corr_data,)
+      # R2 <- round(summary(LM)[[8]], 2)
+      # P_value <-
+      #   ifelse(lmp(LM) < 0.001,
+      #          "p<0.001",
+      #          ifelse(
+      #            lmp(LM) <
+      #              0.01,
+      #            "p<0.01",
+      #            ifelse(lmp(LM) < 0.05, "p<0.05",
+      #                   "ns!!!")
+      #          ))
+      # Stats <- paste0("R²=", R2, ", ", P_value)
+
+  # Add the sampleID column to temp_data
+  temp_data <- Data_aligned %>% 
+    mutate(sampleID = sample)
+
+  # Append temp_data to Corr_data
+  Corr_data_SST <- bind_rows(Corr_data_SST, temp_data) %>% 
+    arrange(x)
+```
+
+```{r SST-Corr}
+#| fig-cap: Comparison of Mg/Ca and SST in all modern shells.
+#| fig-height: 10
+#| fig-width: 16
+#| label: fig-SST_Corr
+
+Corr_data_SST %>% 
+  pivot_wider(values_from = "value",names_from = "proxy",values_fn = mean) %>% 
+  ggplot()+
+  aes(sst,mg_ca,col=sampleID)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression("SST in ºC"),y="Mg/Ca intensity ratio")+
+  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
+  theme_cowplot(20)+
+  stat_regline_equation(label.x = 2, label.y = 0.55,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+    formula = y~x,size=5
+  ) +
+  theme(legend.position = "right")+
+  labs(col="Specimen")+
+  facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
+  scale_colour_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12","#542788")) +
+scale_fill_manual(values = c( "#33a02c","#2ecc71", "#ffc100", "#f39c12","#542788"))+
+  theme_cowplot(20)+
+  coord_fixed(ratio = 20) +
+  stat_cor(
+  aes(label = paste("P-value: ", round(..p.., 10))),
+     label.x = 2, label.y = 0.5,
+  show.legend = FALSE,size=5
+  )+
+  ylim(0,0.6)
+   theme(
+    strip.background = element_rect(fill = "white", colour = NA),
+    strip.text = element_text(colour = "black"),
+    panel.border = element_rect(colour="black"),
+    legend.position="none"
+  )
+
+# Corr_Samples <- Corr_data_SST %>% pull(sampleID) %>% unique()
+# 
+# Stats <- tibble()
+# 
+# for (sample in Corr_Samples){
+#   
+#   
+# LM <- lm(sst ~ mg_ca, data = Corr_data_SST %>% 
+#     filter(sampleID==sample) %>% pivot_wider(values_from = "value",names_from = "proxy",values_fn = mean)) 
+#   
+# R2 <- round(summary(LM)[[8]], 2)
+# P_value <-        ifelse(lmp(LM) < 0.001,
+#                "p<0.001",
+#                ifelse(
+#                  lmp(LM) <
+#                    0.01,
+#                  "p<0.01",
+#                  ifelse(lmp(LM) < 0.05, "p<0.05",
+#                         "ns!!!")
+#                ))
+#       Tempstats <- tibble(sample, R2, P_value)
+# 
+# Stats <- rbind(Stats,Tempstats)
+# 
+# }
+# 
+# print(Stats)
+# 
+# 
+#SST_P1 <- 
+```
+
 # Conclusion

### 2024-09-13 Supplementary files 2 - isotope graphs.qmd
Corrections

@@
 These supplementary files are organised by species and specimen identifiers.
+
+We also show correlation graphs of
 
@@
 
+## Geochemical records of shells in this study
 ```{r}
@@
 ```
+
+## Geochemical records of modern shells in other studies mentioned within the main text
+
+``` {r Ferguson Data, dev='png'}
+#| fig-cap: Correlation graphs for Ferguson et al., (2011) specimens.
+#| fig-height: 5
+#| fig-width: 16
+#| label: fig-Ferg
+
+Ferguson_data <- read_csv(here("data","Ferguson-data.csv")) %>% 
+  na.omit()
+
+Ferg_all <-
+  Ferguson_data %>% 
+  ggplot()+
+  aes(d18o,mg_ca,col=shell)+
+  geom_point(show.legend = FALSE)+
+  labs(x=expression(paste(delta^{18},"O (\u2030)")),y="Mg/Ca (mmol/mol)")+
+  stat_smooth(aes(fill = shell,col=shell), method = "lm", formula = y~x,show.legend = FALSE) +
+  # stat_regline_equation(label.x = 2, #label.y = 32
+  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
+  #   formula = y~x
+  # ) +
+  scale_colour_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063")) +
+scale_fill_manual(values = c(  "#EC7063", "#007f7f", "#1f78b4", "#542788","black", "#d73027", "#EC7063"))
+
+Ferg_individual <- Ferg_all+
+  facet_wrap(~shell,ncol=4)+#ylim(0.15,NA)+
+  ggpubr::stat_regline_equation(label.x = 0.3, label.y = 26,
+    aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
+    formula = y~x
+  )+  
+  theme_cowplot(20)+
+  coord_fixed(ratio =0.22) +
+  theme(
+    strip.background = element_rect(fill = "white", colour = NA),
+    strip.text = element_text(colour = "black"),
+    panel.border = element_rect(colour="black")
+  )
+Ferg_individual
+#Ferg_all/Ferg_individual+plot_layout(heights = c(0.5,0.5))
+
+```
+
+## References
+* Ferguson, J.E., Henderson, G.M., Fa, D.A., Finlayson, J.C., Charnley, N.R., 2011. Increased seasonality in the western
+mediterranean during the last glacial from limpet shell geochemistry. *Earth Planet. Sci. Lett.* 308, 325–333. doi:10.1016/j.
+epsl.2011.05.054.

### 2024-09-13 Manuscript.qmd
more corrections

@@
 They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
-Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of \$\\delta\$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than \$\\delta\$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater.
+Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of $\delta$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than $\delta$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater.
 In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].
@@
 specimen ORK-LT5) being analysed by both.
-We also added specimens of the species *Nacella magellanica,* which — similar to the other two species — is also an established palaeothermometer using \$\\delta\$^18^O-values [@Colonese2012-ct; @gutiérrez-zugasti2017; @Nicastro2020-ih].
+We also added specimens of the species *Nacella magellanica,* which — similar to the other two species — is also an established palaeothermometer using $\delta$^18^O-values [@Colonese2012-ct; @gutiérrez-zugasti2017; @Nicastro2020-ih].
 
@@
 LIBS allows us to carry out 2D imaging of entire shell sections, which facilitates navigation of the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.
-By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution \$\\delta\$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
+By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
 Generally establishing the usefulness and reliability of Mg/Ca ratios as SST proxy in limpets provides a platform for future research and an important steppingstone to a better understanding of elemental ratios in other marine mollusc shells.
@@
 TdF_SST <- read_csv(here("data","Tierra_del_Fuego-SST.csv"), comment = "#")
-TdF_SSS <- read_csv(here("data","Tierra_del_Fuego-SSS.csv"), comment = "#")
+TdF_SSS <- read_csv(here("data","Tierra_del_Fuego-SSS.csv"), comment = "#") %>% 
+  mutate(so=so-2)
 
@@
 |------------|------------|------------|------------|------------|------------|
-| Modern         | [@Nicastro2020-ih] | *N. deaurata*    | Cambaceres Bay  | ND-1016-3  | \$\\delta\$^18^O and \$\\delta\$^13^C |
+| Modern         | [@Nicastro2020-ih] | *N. deaurata*    | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C |
 |                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
@@
 |                |                    |                  |                 |            |                                       |
-|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | \$\\delta\$^18^O and \$\\delta\$^13^C |
+|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | $\delta$^18^O and $\delta$^13^C |
 |                |                    |                  | Westray (UK)    |            | analysis;                             |
@@
 |                |                    |                  |                 |            |                                       |
-| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | \$\\delta\$^18^O and \$\\delta\$^13^C |
+| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O and $\delta$^13^C |
 |                |                    |                  | Westray (UK)    | QG1-7189-2 | analysis                              |
@@
 
-On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for \$\\delta\$^18^O and \$\\delta\$^13^C values, as well as via LA-ICP-MS [@Graniero2017-io].
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for $\delta$^18^O and $\delta$$^13^C values, as well as via LA-ICP-MS [@Graniero2017-io].
 Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
@@
 ```{r SST_plot,fig.pos='H'}
-#| fig-cap: Monthly sea surface temperature values averaged from 2007 to 2023, monthly sea surface salinity values from 1993 to 2021. Values were sourced from @Good2020-nl.
+#| fig-cap: Monthly sea surface temperature values averaged from 2007 to 2023. Values were sourced from @Good2020-nl.
 #| fig-height: 3
@@
        x = "Month",
-       y = "Mean Analysed SSS (so)",colour="",fill="") +
+       y = "Mean Analysed SSS (psu)",colour="",fill="") +
     theme(axis.text.x = element_text(angle = 90, hjust = 1),
           legend.position = "none")+  
-  scale_y_continuous(breaks = c(32,33, 34,35,36),limits = c(32,36))
-
-SST_plot +SSS_plot
+  scale_y_continuous(breaks = c(30,31,32,33, 34,35,36),limits = c(30,36))
+
+SST_plot# +SSS_plot
 
@@
 
-High-quality \$\\delta\$^18^O values were acquired from previous publications on *P.
+High-quality $\delta$^18^O values were acquired from previous publications on *P.
 vulgata* [@Surge2012-ba; @Graniero2017-io], *N.
@@
 
-@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential \$\\delta\$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively).
+@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species (*Nacella* spp., and *Patella vulgata*, respectively).
 Due to the high-resolution sampling, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change, with *Nacella* spp.
@@
 
-###
-
 ## Dynamic Time Warping
 
-Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of \$\\delta\$^18^O values and Mg/Ca ratios.
+Because it is not possible to directly associate the exact sample locations used in the analysis of oxygen isotope ratios with the LIBS data, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O values and Mg/Ca ratios.
 DTW is an algorithm that measures similarity between two proxy sequences, which may vary in sampling resolution or interval.
@@
 
-@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its \$\\delta\$^18^O values (similar graphs for other specimen can be found in the supplementaries).
+@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O values (similar graphs for other specimen can be found in the supplementaries).
 The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior.
@@
 Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr).
-The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters (\$\\delta\$^18^Ο = 0.6 – 0.065 \* Mg/Ca).
+The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 – 0.065 \* Mg/Ca).
 The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2).
@@
 vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies.
-P-values are overwhelmingly significant with the highest being p=1e^-23^.
+P-values are overwhelmingly significant and consistently below p<1e^-6^.
 
@@
 
-P_individual <- P_all+ylim(0.2,0.8)+
+P_individual <- P_all+ylim(0.2,0.72)+
   facet_wrap(~sampleID,ncol=4)+#ylim(0.15,NA)+
-  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.65,
+  ggpubr::stat_regline_equation(label.x = 0.2, label.y = 0.7,
     aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),show.legend = FALSE,size=6,
@@
   )+
-  stat_cor(
-    aes(label = paste("P-value: ", ..p..)),
-    label.x = 0.2, label.y = 0.6, 
-    size = 6,show.legend = FALSE
-  )+
+geom_text(aes(x = 0.2, y = 0.65, label = "p-value < 0.0001", col = sampleID),size=6,hjust=0) +
+
+  # THIS BIT BELOW IS COMMENTED OUT BECAUSE THE P VALUES ARE SO SMALL THAT THEY SHOW AS 0. I ADDED THE LINE ABOVE INSTEAD.
+  # stat_cor(
+  # aes(label = paste("P-value: ", round(..p.., 5))),
+  #   label.x = 0.2, label.y = 0.6, 
+  #   size = 6,show.legend = FALSE
+  # )+
   theme_cowplot(20)+
@@
 
-Line scans reflect well the changes previously indicated by \$\\delta\$^18^O values and also indicate about two years of recorded growth.
+Line scans reflect well the changes previously indicated by $\delta$^18^O values and also indicate about two years of recorded growth.
 That said, the 2D imaging suggests that in both specimens a third summer can be added to the total record, which is not captured by the line scans.
 
-@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their \$\\delta\$^18^O values (similar graphs for the other *Nacella* spp.
+@fig-Nac_Comp shows the line scans of ND-1016-3 and NM-1016-3 in comparison to their $\delta$^18^O values (similar graphs for the other *Nacella* spp.
 specimens can be found in the supplementaries).
@@
     coord_fixed(ratio = 6) +
-  stat_cor(
-  aes(label = paste("P-value: ", round(..p.., 10))),
-    label.x = 0.2, label.y = 0.3, 
-    size = 6,show.legend = FALSE
-  )+
+  geom_text(aes(x = 0.2, y = 0.3, label = "p-value < 0.0001", col = sampleID),size=6,hjust=0) +
+
+  # THIS BIT BELOW IS COMMENTED OUT BECAUSE THE P VALUES ARE SO SMALL THAT THEY SHOW AS 0. I ADDED THE LINE ABOVE INSTEAD.
+  # stat_cor(
+  # aes(label = paste("P-value: ", round(..p.., 5))),
+  #   label.x = 0.2, label.y = 0.3, 
+  #   size = 6,show.legend = FALSE
+  # )+
   theme(
@@
 
-: Correlations of Mg/Ca ratios with \$\\delta\$^18^O values found in other studies.
+: Correlations of Mg/Ca ratios with $\delta$^18^O values found in other studies.
 Bold R^2^ values are interpreted as outliers.
@@
 
-Another shell whose results we aim to explain further is ORK-LT5 (see Supplementary Information), which has previously been shown to have no substantial correlation with \$\\delta\$^18^O values or SST by implication [@Graniero2017-io].
+Another shell whose results we aim to explain further is ORK-LT5 (see Supplementary Information), which has previously been shown to have no substantial correlation with $\delta$^18^O values or SST by implication [@Graniero2017-io].
 In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
@@
 We were not able to resample the *Nacella* spp.
-specimens in Graniero *et al.* [-@Graniero2017-io], which had similarly unrelated records of \$\\delta\$^18^O values and Mg/Ca ratios and pointed to no relationship between elemental ratios and SST at all.
+specimens in Graniero *et al.* [-@Graniero2017-io], which had similarly unrelated records of $\delta$^18^O values and Mg/Ca ratios and pointed to no relationship between elemental ratios and SST at all.
 However, we were able to source other modern specimens of the same species from Tierra del Fuego, which shall serve here as an indicator.
@@
 specimens in Graniero *et al*.
-[-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and \$\\delta\$^18^O records of their shells (Figure 6 in Graniero *et al*.
+[-@Graniero2017-io] is not obvious, but the occasional alignment of the Mg/Ca and $\delta$^18^O records of their shells (Figure 6 in Graniero *et al*.
 [-@Graniero2017-io]) indicate that there is some relationship that — potentially due to the low sampling resolution or intra-increment heterogeneity — is not perfectly visible, leading to an overall result of an uncorrelated relationship between Mg/Ca ratios and SST.
@@
 
-```{r SST-Corr}
+```{r SST-Corr,dev='png',dpi=300}
 #| fig-cap: Comparison of Mg/Ca and SST in all modern shells.
@@
   theme_cowplot(20)+
-  stat_regline_equation(label.x = 2, label.y = 0.55,
+  stat_regline_equation(label.x = 5, label.y = 0.55,
     aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
-    formula = y~x,size=5
+    formula = y~x,size=6
   ) +
@@
   theme_cowplot(20)+
-  coord_fixed(ratio = 20) +
-  stat_cor(
-  aes(label = paste("P-value: ", round(..p.., 10))),
-     label.x = 2, label.y = 0.5,
-  show.legend = FALSE,size=5
-  )+
-  ylim(0,0.6)
+  coord_fixed(ratio = 15) +
+  geom_text(aes(x = 5, y = 0.5, label = "p-value < 0.0001", col = sampleID),size=6,hjust=0) +
+
+  # THIS BIT BELOW IS COMMENTED OUT BECAUSE THE P VALUES ARE SO SMALL THAT THEY SHOW AS 0. I ADDED THE LINE ABOVE INSTEAD.
+  # stat_cor(
+  # aes(label = paste("P-value: ", round(..p.., 10))),
+  #    label.x = 5, label.y = 0.5,
+  # show.legend = FALSE,size=5
+  # )+
+  ylim(0,0.6)+
    theme(
@@
 
-# Corr_Samples <- Corr_data_SST %>% pull(sampleID) %>% unique()
-# 
-# Stats <- tibble()
-# 
-# for (sample in Corr_Samples){
-#   
-#   
-# LM <- lm(sst ~ mg_ca, data = Corr_data_SST %>% 
-#     filter(sampleID==sample) %>% pivot_wider(values_from = "value",names_from = "proxy",values_fn = mean)) 
-#   
-# R2 <- round(summary(LM)[[8]], 2)
-# P_value <-        ifelse(lmp(LM) < 0.001,
-#                "p<0.001",
-#                ifelse(
-#                  lmp(LM) <
-#                    0.01,
-#                  "p<0.01",
-#                  ifelse(lmp(LM) < 0.05, "p<0.05",
-#                         "ns!!!")
-#                ))
-#       Tempstats <- tibble(sample, R2, P_value)
-# 
-# Stats <- rbind(Stats,Tempstats)
-# 
-# }
-# 
-# print(Stats)
-# 
-# 
-#SST_P1 <- 
 ```
@@
 
-LIBS elemental imaging provided insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the \$\\delta\$^18^O values of the same shells.
+LIBS elemental imaging provided insight into elemental anomalies found particulary in the external parts (m+3 layer) of the shell, which introduce intra-incremental heterogeneity and which cannot be reconciled with modern SST changes or SST changes recorded in the $\delta$^18^O values of the same shells.
 Our 2D approach allowed us to avoid these areas of intra-increment heterogeneity and focus on areas that are predominantly controlled by changes in SST.
@@
 Shells of the species studied here are widely available along the European Atlantic façade and the Argentinian coastline and make up a significant part of marine mollusc remains in archaeological sites of those regions [@Colonese2011-ab; @Villagran2011-ld; @Zangrando2016-yl].
-By using LIBS as a screening tool for targeted \$\\delta\$^18^O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency.
+By using LIBS as a screening tool for targeted $\delta$^18^O analysis, focusing on annual minima and maxima revealed in the Mg/Ca record, researchers can improve cost and time efficiency.
 This novel method will enable them to increase sample numbers and produce larger, more in-depth climatic studies than previously possible.

### 2024-10-14 Manuscript.qmd
Published

@@
     affiliations:
-      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
-        address: Avinguda Diagonal, 643
-        city: Barcelona
-        country: Spain
-        postal-code: 08028 
+      - name: CONICET-CADIC
+        address: AV. B. Houssay, 200.
+        city: Ushuaia 
+        country: Argentina
+        postal-code: 9410 
 abstract: |
-  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST).
+  This study provides a reassessment of magnesium to calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST).
 While *Patella depressa* along the Spanish shoreline and *Patella caerulea* in the Mediterranean have repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species.
-*Patella vulgata*, *Nacella deaurata* have been studied using Mg/Ca ratios with mixed or contrary results.
+Particularly, *Patella vulgata* but also *Nacella deaurata* have been studied using Mg/Ca ratios with mixed or contrary results.
 In this study, we present elemental maps of these two species as well as *Nacella magellanica* together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST.
@@
 As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e.
-annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
+annual temperature minima and maxima) for targeted $\delta$^18^O and clumped isotope analysis.
 keywords: 
@@
   elsevier-pdf:
-    keep-tex: FALSE
+    keep-tex: TRUE
     latex_engine: xelatex
@@
 Limpet shells are commonly found within archaeological sites and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source [@Alvarez2011-rm, @Robson2023-oz].
-They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
+They have been successfully studied in the context of coastal subsistence economies and site occupation on seasonal [@Shackleton1973-ij; @Parker2020-oy; @Bosch2018-ud] or long-term scales [@Ortiz2015-mr], as well as archives of palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl].
 Determining past sea surface temperature (SST) change recorded in patelloid limpet shells largely relies on the measurement of $\delta$^18^O values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca) ratios [@Graniero2017-io; @Ferguson2011-zl], to have an alternative measure that potentially provides a more accurate SST estimate than $\delta$^18^O-values,as the latter proxy is affected by changes in the mixing of fresh- and saltwater.
@@
 Using modern reference material Graniero *et al*.
-[-@Granieo2017-io] tested multiple elemental ratios (Mg/Ca, Sr/Ca, Li/Ca, Li/Mg, and Sr/Li) acquired via LA-ICP-MS measurements with no ratios showing correlations with corresponding SST changes nor any sinusoidal patterns that would indicate a seasonal environmental control of elemental ratios.
+[-@Graniero2017-io] tested multiple elemental ratios (Mg/Ca, Sr/Ca, Li/Ca, Li/Mg, and Sr/Li) acquired via LA-ICP-MS measurements with no ratios showing correlations with corresponding SST changes nor any sinusoidal patterns that would indicate a seasonal environmental control of elemental ratios.
 This study included two species also used in our study (*Patella vulgata*, *Nacella deaurata*), with one *P.
@@
 To do this, we sampled a set of previously published and unpublished limpet specimens dating to modern and archaeological contexts using LIBS.
-LIBS allows us to carry out 2D imaging of entire shell sections, which facilitates navigation of the complex elemental structure of the shell and better separate the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.
+LIBS allows us to carry out 2D imaging of entire shell sections, which facilitates navigation of the complex elemental structure of the shell and a better separation of the external, temperature-related changes from the internal and less understood factors that influence the Mg/Ca ratio.
 By working with already published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest.
@@
 | Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
-|------------|------------|------------|------------|------------|------------|
+|----------|-----------------|------------|------------|---------|------------|
 | Modern         | [@Nicastro2020-ih] | *N. deaurata*    | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C |
@@
 |                |                    |                  | Westray (UK)    |            | analysis;                             |
-|                |                    |                  |                 |            | Mg, Li, Sr, Ca.                       |
+|                |                    |                  |                 |            | Mg, Li, Sr, Ca                       |
 |                |                    |                  |                 |            |                                       |
@@
 
-Specimens were sourced from two main locations: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in the Orkney archipelago of Scotland (UK).
+Specimens were sourced from two main locations: the Beagle Channel in Tierra del Fuego (Argentina) and the island of Westray in the Orkney archipelago of Scotland (UK).
 Four modern specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016.
-The area receives around 570 mm of precipitation, but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units) [@Nicastro2020-ih].
+The area receives around 570 mm of annual precipitation, but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units) [@Nicastro2020-ih].
 SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February (@fig-SSTs -- red line).
 
-On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for $\delta$^18^O and $\delta$$^13^C values, as well as via LA-ICP-MS [@Graniero2017-io].
+On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in August 2009 and has been previously analysed for $\delta$^18^O and $\delta$^13^C values, as well as for elemental ratios via LA-ICP-MS [@Graniero2017-io].
 Westray lies about 70 km north of the Scottish mainland and experiences virtually no freshwater input with salinity values of 34.0 – 34.5 psu [@Inall2009-ho].
 The annual SST ranges from `r WestrayMin`ºC in March to `r WestrayMax`ºC in August (@fig-SSTs -- blue line).
-
 ```{r SST_plot,fig.pos='H'}
 #| fig-cap: Monthly sea surface temperature values averaged from 2007 to 2023. Values were sourced from @Good2020-nl.
-#| fig-height: 3
+#| fig-height: 2.5
 #| fig-width: 6
@@
 ```
-
 We also studied 7 archaeological specimens of *P.
@@
 In short, all shells were sectioned along the maximum growth axis to expose the internal growth structure.
-From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic layers of the shells.
+From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic layers of the shells (m+2 and m+3, see [@Fenger2007-gf]).
 The exact location of the sampled carbonate powder is not available apart from the distance to the growth edge, where reported.
@@
 Together they show a range of lifespans, with some comparatively long lived for their size (QG1-7246-1, 12 years within 16 mm of growth) making them particularly susceptible to within-shell time-averaging because of slow growth rates.
-That said, the majority are under 5 years old yielding sufficiently fast growth to nearly capture seasonal amplitudes of SST.
+That said, the majority yielded sufficiently fast growth to nearly capture seasonal amplitudes of SST with ages of 2--3 years (n=3), 3--4 years (n=2), and 5 years (n=2).
 
@@
 Mg/Ca elemental imaging was carried out using LIBS at the Leibniz Zentrum für Archäologie (Mainz-Germany) following previously published methods [@Hausmann2023-ih].
-The imaging involves the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume.
+The imaging involves the ablation of the shell section using an infrared (1064 nm) laser (1–-2 mJ, 100 Hz) onto a surface area of 20–-30 µm to generate a plasma plume.
 This plume emits light, which is measured using a synchronised spectrometer (delay time = 0.5 µs, integration time = 1 µs).
-The resulting light spectrum quantifies the emission lines of Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine their intensity ratio.
+The resulting light spectrum quantifies the emission lines of magnesium (MgII: 279.553 nm) and calcium (CaII: 315.887 nm) to determine their intensity ratio.
 While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g.
-[@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca variation within the shell carbonate.
+[@Martinez-Minchero2023-pq]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca variation within the shell carbonate.
 All values reported here are in arbitrary units based on the ratio of intensities of the peaks above.
@@
 This can involve selecting appropriate distance measures and constraints to ensure meaningful alignment.
-We used the default settings for symmetric step patterns and euclidean distance measures (see code in [@Hausmann_Surge_Godino_2024]).
+We used the default settings for symmetric step patterns and euclidean distance measures (see code in Hausmann et al.[-@hausmann2024confirmation]).
 
@@
 As examples, @fig-Pat_LIBS shows the 2D-distribution of elemental ratios across the entire section of ORK-LT5 and the preserved anterior side of QG1-7188-1 (additional maps can be found in the supplementaries).
-The repeating patterns are found across the calcitic layers that are exterior to the myostracum (layers m+2 and m+3) as well as the layers that are interior (m-2)(see Fenger et al.
+The repeating patterns are found across the calcitic layers that are exterior to the myostracum (layers m+2 and m+3) as well as the layers that are interior (m--2)(see Fenger et al.
 [-@Fenger2007-gf] for a detailed explanations of microstructural layers in *P.
@@
 Both specimens in @fig-Pat_LIBS show increased Mg/Ca intensity ratios towards the outside of the shell and some degree of intra-increment variability.
-This is particularly visible in @fig-Pat_LIBS b, where Mg/Ca intensity ratios range from \~0.5–0.7 in the first year of growth (9–12 mm on the y-axis).
-A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0–10 mm on the x-axis).
+This is particularly visible in @fig-Pat_LIBS b, where Mg/Ca intensity ratios range from \~0.5--0.7 in the first year of growth (9-–12 mm on the y-axis).
+A similar pattern is visible in the interior m-2 layer of ORK-LT5 (@fig-Pat_LIBS a), which increase towards the anterior of the shell (0--10 mm on the x-axis).
 
@@
 This shows the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].
-
-@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O values (similar graphs for other specimen can be found in the supplementaries).
-The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior.
-That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.
 
@@
 ```
-
-Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr).
-The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 – 0.065 \* Mg/Ca).
-The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2).
-These specimens are also the oldest and youngest specimens, respectively, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies.
-The mean R^2^ value of the individual *P.
-vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies.
-P-values are overwhelmingly significant and consistently below p<1e^-6^.
-
-```{r Patella_Correlation_Graphs,dev='png',dpi=300}
-#| fig-cap: Correlation graphs for *Patella vulgata* specimens.
+@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O values (similar graphs for other specimen can be found in the supplementaries).
+The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have a range of distances to the edge, depending on where one measures, and a line scan along the interior would be shorter than a line scan along the exterior.
+That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.
+
+```{r Patella_Correlation_Graphs,dev='png',dpi=300,fig.pos='H'}
+#| fig-cap: Correlation graphs for Mg/Ca and $\delta$^18^O values in  *P. vulgata* specimens.
 #| fig-height: 8
@@
 ```
+Aligning the records programmatically using dynamic time warping, allowed us to compare both proxies for each specimen directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr).
+The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 -- 0.065 \* Mg/Ca).
+The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2).
+These specimens are also the oldest and youngest specimens, respectively, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies.
+The mean R^2^ value of the individual *P.
+vulgata* specimens is 0.89, suggesting a generally good fit between the two proxies.
+P-values are overwhelmingly significant and consistently below p\<1e^--6^.
 
@@
 Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
-Note the inset in (*c*) where the shell edge has been resampled using a higher resolution (50 µm instead of 100 µm) to better understand the Mg/Ca ratios at the shell edge.
+Note the inset in (*b*) where the shell edge has been resampled using a higher resolution (50 µm instead of 100 µm) to better understand the Mg/Ca ratios at the shell edge.
 White dashed lines indicate location of line scans.
@@
 [-@Hausmann2023-ih]) using the same emission lines.
-While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen Magnesium to the chosen Calcium peak, the low ratios still indicate lower concentrations than seen elsewhere or seen only in aragonitic parts of e.g.
+While the intensity ratio depends chiefly on the chosen emission lines to calculate the proportion of the chosen magnesium to the chosen calcium peak, the low ratios still indicate lower concentrations than seen elsewhere or seen only in aragonitic parts of e.g.
 *Patella* shells.
@@
 ```{r Nacella_Correlation_Graphs,dev='png',dpi=300}
-#| fig-cap: Correlation graphs for *Nacella* spp. specimens
+#| fig-cap: Correlation graphs for Mg/Ca and $\delta$^18^O values in *Nacella* spp. specimens
 #| fig-height: 5
@@
     coord_fixed(ratio = 6) +
-  geom_text(aes(x = 0.2, y = 0.3, label = "p-value < 0.0001", col = sampleID),size=6,hjust=0) +
+  geom_text(aes(x = 0.2, y = 0.3, label = "p-value < 0.0001", col = sampleID),size=6,hjust=0,show.legend = FALSE) +
 
@@
 Annually repeating patterns of increasing (spring/summer) and decreasing (autum/winter) SST were also visible in the line scans, albeit less straightforward than in the 2D-data.
-Nevertheless, the resulting mean R^2^-values comparin Mg/Ca and δ^18^O values for each species were high (*P.
+Nevertheless, the resulting mean R^2^-values comparing Mg/Ca and δ^18^O values for each species were high (*P.
 vulgata R^2^*=0.89, *N.
@@
 | Species          | Locality         | Specimen   | R^2^                    | Study                             |     |
-|------------|------------|------------|------------|------------|------------|
+|----------|-----------|---------|-------|--------------|-----------------|
 | *P. depressa*    | Northern Spain   | LAN541     | 0.87                    | [@Garcia-Escarzaga2021-ij]        |     |
@@
 Another shell whose results we aim to explain further is ORK-LT5 (see Supplementary Information), which has previously been shown to have no substantial correlation with $\delta$^18^O values or SST by implication [@Graniero2017-io].
-In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20–30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150–300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20–30 µm spots spaced at 10 µm) (@fig-ORK_sub).
+In contrast to other specimens in this study, we were able to resample this specimen and acquire Mg/Ca data on a slightly higher resolution (20--30 µm spots in a 100 µm raster compared to 50 µm spots spaced at 150--300 µm) and using 2 dimensions, as well as using a continuous high-resolution line-scan at (20--30 µm spots spaced at 10 µm) (@fig-ORK_sub).
 These additional data revealed previously missed summer peaks at around 2--3 mm distance to the shell edge (grey box in @fig-ORK_sub) and revealed additional growth at the very edge (dashed box in @fig-ORK_sub) mirroring the change towards warmer temperatures at the time of collection (August).
@@
 ```{r ORK_LT5}
-#| fig-cap: Comparison of previous and new Mg/Ca data for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca and $\delta$^18^O-values
+#| fig-cap: Comparison of previous and new Mg/Ca data for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca and $\delta$^18^O-values.
 #| fig-height: 4.5
@@
 
-As indicated by the stable isotope values, each shell individually has a good correlation with SST (R^2^ values between 0.93 to 0.96 with p-values consistently below 0.001) (@fig-SST_Corr), which are even better than the correlation between elemental ratios and stable isotope values.
+As indicated by the stable isotope values, no reliable shared relationship between the data can be established, but each shell's Mg/Ca data individually have a good correlation with SST (R^2^ values between 0.93 to 0.96 with p-values consistently below 0.001) (@fig-SST_Corr), which are even better than the correlation between elemental ratios and stable isotope values.
 This improvement is most likely due to the simple and sinusoidal SST curve, which is easier to fit to other data.
+
+As a result of the different correlations between Mg/Ca ratios and SST in the modern shells, we did not attempt to use Mg/Ca as a direct proxy for absolute SST values in ºC for the archaeological shells from Quoygrew, but rather as a proxy for relative SST change (i.e.
+warmer/colder).
+In future analyses, these high resolution and rapidly acquired records of relative SST can guide the sampling for oxygen isotope or clumped isotope analysis and reduce analytical costs per shell, thus increasing the quantity of shells that can be analysed per study.
 
@@
 
-```{r SST-Corr,dev='png',dpi=300}
+```{r SST-Corr,dev='png',dpi=300,fig.pos='H'}
 #| fig-cap: Comparison of Mg/Ca and SST in all modern shells.
@@
     aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
-    formula = y~x,size=6
+    formula = y~x,size=c(5,5,5,6,6)
   ) +
@@
 # Acknowledgements
-
-# References
+Niklas Hausmann was supported by the German Research Foundation under the Emmy Noether Program
+(project number: 439799406). We thank the Applied Photonics Lab of the Institute of Elec-
+tronic Structure and Laser of the Foundation for Research and Technology - Hellas for close support with
+the LIBS system. We are also grateful for two anonymous reviewers who helped to improve the manuscript.
 

### 2024-10-14 Manuscript.tex
Published

@@
 
+\usepackage{lineno}
+\linenumbers
 \makeatletter
@@
 NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
-\affiliation[3]{organization={Department of Evolutionary Biology,
-Ecology and Environmental Science, Biodiversity Research Institute
-(IRBio), University of Barcelona},addressline={Avinguda Diagonal,
-643},city={Barcelona},country={Spain},countrysep={,},postcode={8028},postcodesep={}}
+\affiliation[3]{organization={CONICET-CADIC},addressline={AV. B.
+Houssay,
+200.},city={Ushuaia},country={Argentina},countrysep={,},postcode={9410},postcodesep={}}
 
@@
 \begin{abstract}
-This study provides a reassessment of Magnesium to Calcium (Mg/Ca)
+This study provides a reassessment of magnesium to calcium (Mg/Ca)
 ratios in Atlantic limpet shells to determine past sea surface
 temperatures (SST). While \emph{Patella depressa} along the Spanish
-shoreline has repeatedly produced reliable correlations between SST and
-Mg/Ca ratios, this relationship is not the case for other patelloid
-species. \emph{Patella vulgata}, \emph{Nacella deaurata}, and \emph{N.
-magellanica} have been studied using Mg/Ca ratios with mixed or contrary
-results. In this study, we present elemental maps of these three species
-together with oxygen isotope ratios (\(\delta\)\textsuperscript{18}O)
-that confirm a good relationship with SST. Our dataset also reassesses a
-specimen which was previously unsuccessful in providing significant
-correlations between \(\delta\)\textsuperscript{18}O values and Mg/Ca
-ratios. By reassessing these species and including modern and
-archaeological specimens (n=12) from three patelloid species (\emph{P.
-vulgata}, \emph{N. deaurata}, and \emph{N. magellanica}) we further add
-to the growing set of evidence for the reliable use of Mg/Ca ratios to
-detect palaeotemperature change. As a result, these species can in the
-future serve to determine ontogenetic age and season of capture as well
-as to reveal locations of interest within the growth record (i.e.~annual
-temperature minima and maxima) for the targeted analysis using
-\(\delta\)\textsuperscript{18}O and clumped isotope analysis.
+shoreline and \emph{Patella caerulea} in the Mediterranean have
+repeatedly produced reliable correlations between SST and Mg/Ca ratios,
+this relationship is not the case for other patelloid species.
+Particularly, \emph{Patella vulgata} but also \emph{Nacella deaurata}
+have been studied using Mg/Ca ratios with mixed or contrary results. In
+this study, we present elemental maps of these two species as well as
+\emph{Nacella magellanica} together with oxygen isotope ratios
+(\(\delta\)\textsuperscript{18}O) that confirm a good relationship with
+SST. Our dataset also reassesses a specimen which was previously
+unsuccessful in providing significant correlations between
+\(\delta\)\textsuperscript{18}O values and Mg/Ca ratios. By reassessing
+these species and including modern and archaeological specimens (n=12)
+from three patelloid species (\emph{P. vulgata}, \emph{N. deaurata}, and
+\emph{N. magellanica}) we further add to the growing set of evidence for
+the reliable use of Mg/Ca ratios to detect palaeotemperature change. As
+a result, these species can in the future serve to determine ontogenetic
+age and season of capture as well as to reveal locations of interest
+within the growth record (i.e.~annual temperature minima and maxima) for
+targeted \(\delta\)\textsuperscript{18}O and clumped isotope analysis.
 \end{abstract}
@@
 site occupation on seasonal
-\citep{Shackleton1973-ij, Parker2018-wf, Bosch2018-ud} or long-term
+\citep{Shackleton1973-ij, Parker2020-oy, Bosch2018-ud} or long-term
 scales \citep{Ortiz2015-mr}, as well as archives of palaeotemperature
@@
 are not reliable recorders of palaeotemperature \citep{Graniero2017-io}.
+Using modern reference material Graniero \emph{et al}.
+\citeyearpar{Graniero2017-io} tested multiple elemental ratios (Mg/Ca,
+Sr/Ca, Li/Ca, Li/Mg, and Sr/Li) acquired via LA-ICP-MS measurements with
+no ratios showing correlations with corresponding SST changes nor any
+sinusoidal patterns that would indicate a seasonal environmental control
+of elemental ratios. This study included two species also used in our
+study (\emph{Patella vulgata}, \emph{Nacella deaurata}), with one
+\emph{P. vulgata} specimen (i.e.~specimen ORK-LT5) being analysed by
+both. We also added specimens of the species \emph{Nacella magellanica,}
+which --- similar to the other two species --- is also an established
+palaeothermometer using \(\delta\)\textsuperscript{18}O-values
+\citep{Colonese2012-ct, gutiérrez-zugasti2017, Nicastro2020-ih}.
 
-In this study we will repeat and expand the analysis of Atlantic limpets
-to determine the reliability of Mg/Ca ratios as palaeotemperature
-proxies. To do this, we sampled a set of previously published and
-unpublished limpet specimens dating to modern and archaeological
-contexts using LIBS. LIBS allows us to carry out 2D imaging of entire
-shell sections, which allows navigation of the complex elemental
-structure of the shell and better separate the external,
-temperature-related changes from the internal and less understood
-factors that influence the Mg/Ca ratio. By working with already
-published datasets, we were able to simultaneously avoid costs for new
-high-resolution \(\delta\)\textsuperscript{18}O-data, to use real-world
-examples, and to also provide pilot-data for areas of existing research
-interest. Generally establishing the usefulness and reliability of Mg/Ca
-ratios as SST proxy in limpets provides a platform for future research
-and an important steppingstone to a better understanding of elemental
-ratios in other marine mollusc shells.
+In this study, we repeat and expand upon the analysis of Atlantic
+limpets to determine the reliability of Mg/Ca ratios as
+palaeotemperature proxies. To do this, we sampled a set of previously
+published and unpublished limpet specimens dating to modern and
+archaeological contexts using LIBS. LIBS allows us to carry out 2D
+imaging of entire shell sections, which facilitates navigation of the
+complex elemental structure of the shell and a better separation of the
+external, temperature-related changes from the internal and less
+understood factors that influence the Mg/Ca ratio. By working with
+already published datasets, we were able to simultaneously avoid costs
+for new high-resolution \(\delta\)\textsuperscript{18}O-data, to use
+real-world examples, and to also provide pilot-data for areas of
+existing research interest. Generally establishing the usefulness and
+reliability of Mg/Ca ratios as SST proxy in limpets provides a platform
+for future research and an important steppingstone to a better
+understanding of elemental ratios in other marine mollusc shells.
 
@@
 \hyperref[Table_1]{1}. Here we will briefly summarise their contextual
-information, which can be accessed in more detail at the respective
+information, which can be accessed in more detail from the respective
 studies \citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1389}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2361}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
 \caption{Overview of the modern and archaeological limpet specimens
@@
 & & & Westray (UK) & & analysis; \\
-& & & & & Mg, Li, Sr, Ca. \\
+& & & & & Mg, Li, Sr, Ca \\
 & & & & & \\
@@
 Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
-Bay - & QG1-7188-1 & \(\delta\)\textsuperscript{18}O \\
-& & & Westray (UK) & QG1-7189-2 & and \(\delta\)\textsuperscript{13}C \\
-& & & & QG2-1061-1 & analysis \\
+Bay - & QG1-7188-1 & \(\delta\)\textsuperscript{18}O and
+\(\delta\)\textsuperscript{13}C \\
+& & & Westray (UK) & QG1-7189-2 & analysis \\
+& & & & QG2-1061-1 & \\
 & & & & QG2-1064-1 & \\
@@
 Specimens were sourced from two main locations: the Beagle Channel in
-Tierra del Fuego (Argentina) and the island of West Ray in the Orkney
+Tierra del Fuego (Argentina) and the island of Westray in the Orkney
 archipelago of Scotland (UK). Four modern specimens were collected from
 the Beagle Channel in Cambaceres Bay in October 2016. The area receives
-around 570 mm of precipitation, but the channel waters are predominantly
-influenced by marine currents and in the period of October 2015 to
-October 2016 stayed around 30.7 ± 0.7 psu (practical salinity units)
-\citep{Nicastro2020-ih}. SSTs range from 4.7ºC in August to 10ºC in
-February (Figure~\ref{fig-SSTs} -- red line).
+around 570 mm of annual precipitation, but the channel waters are
+predominantly influenced by marine currents and in the period of October
+2015 to October 2016 stayed around 30.7 ± 0.7 psu (practical salinity
+units) \citep{Nicastro2020-ih}. SSTs range from 30.5ºC in August to
+31.9ºC in February (Figure~\ref{fig-SSTs} -- red line).
 
@@
 \(\delta\)\textsuperscript{18}O and \(\delta\)\textsuperscript{13}C
-values, as well as LA-ICP-MS \citep{Graniero2017-io}. Westray lies about
-70 km north of the Scottish mainland and experiences virtually no
-freshwater input with salinity values of 34.0 -- 34.5 psu
-\citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in March to
-13.8ºC in August (Figure~\ref{fig-SSTs} -- blue line).
+values, as well as for elemental ratios via LA-ICP-MS
+\citep{Graniero2017-io}. Westray lies about 70 km north of the Scottish
+mainland and experiences virtually no freshwater input with salinity
+values of 34.0 -- 34.5 psu \citep{Inall2009-ho}. The annual SST ranges
+from 34.4ºC in March to 35.3ºC in August (Figure~\ref{fig-SSTs} -- blue
+line).
 
@@
 
-\includegraphics{fig-SSTs-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-SSTs-1.pdf}
 
@@
 
-\caption{\label{fig-SSTs}Monthly Sea Surface Temperature values averaged
-from years. SST values were sourced from \citet{Good2020-nl}.}
+\caption{\label{fig-SSTs}Monthly sea surface temperature values averaged
+from 2007 to 2023. Values were sourced from \citet{Good2020-nl}.}
 
@@
 during the Medieval Climate Anomaly (810 -- 1229 CE)
-\citep{Surge2012-ba}. Together they show a range of lifespans, with some
-comparatively long lived for their size (QG1-7246-1, 12 years within 16
-mm of growth) making them particularly susceptible to within-shell
-time-averaging because of slow growth rates. That said, the majority are
-under 5 years old yielding sufficiently fast growth to nearly capture
-seasonal amplitudes of SST.
+\citep{Surge2012-ba}.
 
 \subsection{Methods}\label{methods}
+
+\subsubsection{Oxygen isotope ratios}\label{oxygen-isotope-ratios}
+
+High-quality \(\delta\)\textsuperscript{18}O values were acquired from
+previous publications on \emph{P. vulgata}
+\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaurata} and \emph{N.
+magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
+the methodology for the stable isotope analyses can be found. In short,
+all shells were sectioned along the maximum growth axis to expose the
+internal growth structure. From the exposed section carbonate powder
+samples were acquired using a micromill with resolutions of 100--200 µm
+and by targeting the calcitic layers of the shells (m+2 and m+3, see
+\citep{Fenger2007-gf}). The exact location of the sampled carbonate
+powder is not available apart from the distance to the growth edge,
+where reported. Analytical precision of the isotope ratios was
+consistently at 0.05--0.10‰ and values are reported in per mil units (‰)
+relative to the VPDB (Vienna Pee Dee Belemnite) standard.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.png}
+
+}
+
+\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} spp. limpet
+shells from Cambaceres Bay.}
+
+\end{figure}%
+
+Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
+previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
+limpet shells by species (\emph{Nacella} spp., and \emph{Patella
+vulgata}, respectively). Due to the high-resolution sampling, all shells
+show seasonally resolved and multi-year quasi-sinusoidal records of SST
+change, with \emph{Nacella} spp. shells living 2--3 years and \emph{P.
+vulgata} living regularly over 3 years. In those instances where longer
+lifespans resulted in slow growth rates below 1 mm/yr (i.e.~specimen
+QG1-7246-1, Figure~\ref{fig-Pat_iso} b), the temporal-resolution is
+reduced from more than a dozen of samples per year, to only around 7.
+Together they show a range of lifespans, with some comparatively long
+lived for their size (QG1-7246-1, 12 years within 16 mm of growth)
+making them particularly susceptible to within-shell time-averaging
+because of slow growth rates. That said, the majority yielded
+sufficiently fast growth to nearly capture seasonal amplitudes of SST
+with ages of 2--3 years (n=3), 3--4 years (n=2), and 5 years (n=2).
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_iso-1.png}
+
+}
+
+\caption{\label{fig-Pat_iso}Isotope values of \emph{Patella vulgata}
+limpet shells from Quoygrew (QG, \emph{a-g}) and Rack Wick Bay (ORK,
+\emph{h}) on Westray, Orkney.}
+
+\end{figure}%
 
@@
 methods \citep{Hausmann2023-ih}. The imaging involves the ablation of
-the shell section using an infrared (1064 nm) laser (1--2 mJ, 100 Hz)
-onto a surface area of 20--30 µm to generate a plasma plume. This plume
+the shell section using an infrared (1064 nm) laser (1---2 mJ, 100 Hz)
+onto a surface area of 20---30 µm to generate a plasma plume. This plume
 emits light, which is measured using a synchronised spectrometer (delay
 time = 0.5 µs, integration time = 1 µs). The resulting light spectrum
-quantifies the emission lines of Magnesium (MgII; 279.553 nm) and
-Calcium (CaII; 315.887 nm) to determine their intensity ratio. While
+quantifies the emission lines of magnesium (MgII: 279.553 nm) and
+calcium (CaII: 315.887 nm) to determine their intensity ratio. While
 these two peaks alone do not represent the molar concentrations of both
 elements (as opposed to calibration free LIBS; e.g.
-\citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
+\citep{Martinez-Minchero2023-pq}), the intensity ratio is linearly
 correlated to the molar concentration \citep{Hausmann2017-oa} and can be
@@
 
-\subsubsection{Oxygen isotope ratios}\label{oxygen-isotope-ratios}
-
-High-quality \(\delta\)\textsuperscript{18}O values were acquired from
-previous publications on \emph{P. vulgata}
-\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaurata} and \emph{N.
-magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
-the methodology for the stable isotope analyses can be found. In short,
-all shells were sectioned along the maximum growth axis to expose the
-internal growth structure. From the exposed section carbonate powder
-samples were acquired using a micromill with resolutions of 100--200 µm
-and by targeting the calcitic M+2 layer of the shells. Analytical
-precision of the isotope ratios was consistently at 0.05--0.10‰ and
-values are reported in per mil units (‰) relative to the VPDB (Vienna
-Pee Dee Belemnite) standard.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{fig-Nac_iso-1.png}
-
-}
-
-\caption{\label{fig-Nac_iso}Isotope values of \emph{Nacella} spp. limpet
-shells from Cambaceres Bay.}
-
-\end{figure}%
-
-Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
-previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
-limpet shells by species (\emph{Nacella} spp., and \emph{Patella
-vulgata}, respectively). Due to the high-resolution sampling, all shells
-show seasonally resolved and multi-year quasi-sinusoidal records of SST
-change, with \emph{Nacella} spp. shells living 2--3 years and \emph{P.
-vulgata} living regularly over 3 years. In those instances where longer
-lifespans resulted in slow growth rates below 1 mm/yr (i.e.~specimen
-QG1-7246-1, Figure~\ref{fig-Pat_iso} b), the temporal-resolution is
-reduced from more than a dozen of samples per year, to only around 7.
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics{fig-Pat_iso-1.png}
-
-}
-
-\caption{\label{fig-Pat_iso}Isotope values of \emph{Patella vulgata}
-limpet shells from Quoygrew (QG, \emph{a-g}) and Rack Wick Bay (ORK,
-\emph{h}) on Westray, Orkney.}
-
-\end{figure}%
-
 \subsection{Dynamic Time Warping}\label{dynamic-time-warping}
@@
 R \citep{Giorgino2009-sj, R_Core_Team2020-mk}, which provides a robust
-framework for aligning time series data. This involved selecting
+framework for aligning time series data. This can involve selecting
 appropriate distance measures and constraints to ensure meaningful
-alignment (see code in \citep{Hausmann_2024OSF}). The process involved
-iterative adjustments to minimise the overall distance between
-corresponding points in the two data sets.
+alignment. We used the default settings for symmetric step patterns and
+euclidean distance measures (see code in Hausmann et
+al.\citeyearpar{hausmann2024confirmation}).
 
@@
 exterior to the myostracum (layers m+2 and m+3) as well as the layers
-that are interior (m-2)(see Fenger et al. \citeyearpar{Fenger2007-gf}
+that are interior (m--2)(see Fenger et al. \citeyearpar{Fenger2007-gf}
 for a detailed explanations of microstructural layers in \emph{P.
-vulgaata}). The layer m-1, which is also interior, consists of aragonite
+vulgata}). The layer m-1, which is also interior, consists of aragonite
 and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its Mg/Ca intensity
@@
 
-\includegraphics{fig-Pat_LIBS-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_LIBS-1.pdf}
 
@@
 identical but are chosen arbitrarily to increase contrast. Areas in grey
-consist of low-Mg aragonite and fall outside the chosen colour-range.}
+consist of low-Mg aragonite and fall outside the chosen colour-range.
+White dashed lines indicate location of line scans.}
 
@@
 intra-increment variability. This is particularly visible in
-Figure~\ref{fig-Pat_LIBS} c, where Mg/Ca intensity ratios range from
-\textasciitilde0.5--0.7 in the first year of growth (9--12 mm on the
+Figure~\ref{fig-Pat_LIBS} b, where Mg/Ca intensity ratios range from
+\textasciitilde0.5--0.7 in the first year of growth (9-\/--12 mm on the
 y-axis). A similar pattern is visible in the interior m-2 layer of
@@
 
+\begin{figure}
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Comp-1.png}
+
+}
+
+\caption{\label{fig-Pat_Comp}Comparison of LIBS line-scan and sequential
+\(\delta\)\textsuperscript{18}O-values of specimen ORK-LT5}
+
+\end{figure}%
+
 Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
@@
 
-\includegraphics{fig-Pat_Comp-1.png}
+\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Corr-1.png}
 
@@
 
-\caption{\label{fig-Pat_Comp}Comparison of LIBS line-scan and sequential
-\(\delta\)\textsuperscript{18}O-values of specimen ORK-LT5}
+\caption{\label{fig-Pat_Corr}Correlation graphs for Mg/Ca and
+\(\delta\)\textsuperscript{18}O values in \emph{P. vulgata} specimens.}
 
@@
 vulgata} specimens is 0.89, suggesting a generally good fit between the
-two proxies.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{fig-Pat_Corr-1.png}
-
-}
-
-\caption{\label{fig-Pat_Corr}Correlation graphs for \emph{Patella
-vulgata} specimens.}
-
-\end{figure}%
+two proxies. P-values are overwhelmingly significant and consistently
+below p\textless1e\textsuperscript{--6}.
 
@@
 
-\includegraphics{fig-Nac_LIBS-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_LIBS-1.pdf}
 
@@
 contrast. Areas in grey consist of low-Mg aragonite and fall outside the
-chosen colour-range. Note the inset in (\emph{c}) where the shell edge
+chosen colour-range. Note the inset in (\emph{b}) where the shell edge
 has been resampled using a higher resolution (50 µm instead of 100 µm)
-to better understand the Mg/Ca ratios at the shell edge.}
+to better understand the Mg/Ca ratios at the shell edge. White dashed
+lines indicate location of line scans.}
 
@@
 emission lines. While the intensity ratio depends chiefly on the chosen
-emission lines to calculate the proportion of the chosen Magnesium to
-the chosen Calcium peak, the low ratios still indicate lower
+emission lines to calculate the proportion of the chosen magnesium to
+the chosen calcium peak, the low ratios still indicate lower
 concentrations than seen elsewhere or seen only in aragonitic parts of
@@
 
-\includegraphics{fig-Nac_Comp-1.png}
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Comp-1.png}
 
@@
 mean R\textsuperscript{2} value for each \emph{N.} species is 0.86 for
-\emph{N. deaurata} and 0.88 for \emph{N. magellanica}. Why two specimens
-of different species are so well aligned is not entirely certain and
-randomness cannot for now be ruled out. That said, since they both have
-a high R\textsuperscript{2} value, it might just be that they were both
+\emph{N. deaurata} and 0.88 for \emph{N. magellanica} with p-values for
+all speciemns being far below 0.0001. Why two specimens of different
+species are so well aligned is not entirely certain and randomness
+cannot for now be ruled out. That said, since they both have a high
+R\textsuperscript{2} value, it might just be that they were both
 minimally affected by factors other than SST, including the sampling
@@
 
-\includegraphics{fig-Nac_Corr-1.png}
+\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Corr-1.png}
 
@@
 
-\caption{\label{fig-Nac_Corr}Correlation graphs for \emph{Nacella} spp.
-specimens}
+\caption{\label{fig-Nac_Corr}Correlation graphs for Mg/Ca and
+\(\delta\)\textsuperscript{18}O values in \emph{Nacella} spp. specimens}
 
@@
 
-The Mg/Ca ratios from the 2D elemental imaging present positive results
+The Mg/Ca ratios from the 2D elemental imaging present results
 suggesting that SST is the main factor influencing variations in Mg/Ca
@@
 visible in the line scans, albeit less straightforward than in the
-2D-data. Nevertheless, the resulting R\textsuperscript{2}-values for
-each species were high (\emph{P. vulgata mean
-R\textsuperscript{2}}=0.89, \emph{N. deaurata mean
-R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
+2D-data. Nevertheless, the resulting mean R\textsuperscript{2}-values
+comparing Mg/Ca and δ\textsuperscript{18}O values for each species were
+high (\emph{P. vulgata R\textsuperscript{2}}=0.89, \emph{N. deaurata
+R\textsuperscript{2}}=0.86\emph{, N. magellanica
 R\textsuperscript{2}}=0.88). These results are similar to previously
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2449}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1429}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1633}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1224}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2041}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1471}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1618}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1324}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1029}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2059}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2500}}@{}}
 \caption{Correlations of Mg/Ca ratios with
-\(\delta\)\textsuperscript{18}O values and SST found in other studies.
-Bold R\textsuperscript{2} values are interpreted as outliers. *
-indicates that only SST data and no other geochemical data was used. **
-these samples were used to determine a shared R\textsuperscript{2}-value
+\(\delta\)\textsuperscript{18}O values found in other studies. Bold
+R\textsuperscript{2} values are interpreted as outliers. * indicates
+samples that were used to determine a shared R\textsuperscript{2}-value
 of 0.79.}\tabularnewline
@@
 \citep{Ferguson2011-zl} & \\
-& & JL2 & 0.8** & & \\
-\emph{P. caerulea} & Gibraltar & JM00 & 0.69** & & \\
-& & JM30 & 0.83** & & \\
+& & JL2 & 0.8* & & \\
+\emph{P. caerulea} & Gibraltar & JM00 & 0.69* & & \\
+& & JM30 & 0.83* & & \\
 \emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
 \citep{Graniero2017-io} and this study & \\
+& & QG1-7188-1 & 0.88 & this study & \\
+& & QG1-7189-2 & 0.95 & & \\
+& & QG1-7246-1 & 0.81 & & \\
+& & QG2-1061-1 & 0.90 & & \\
+& & QG2-1064-1 & 0.90 & & \\
+& & QG2-7180-1 & 0.87 & & \\
+& & QG2-7180-2 & 0.90 & & \\
+\emph{N. deaurata} & Tierra del Fuego & ND-1016-3 & 0.95 & & \\
+& & ND-1016-4 & 0.76 & & \\
+\emph{N. magellanica} & Tierra del Fuego & NM-1016-1 & 0.92 & & \\
+& & NM-1016-3 & 0.84 & & \\
 \end{longtable}
@@
 R\textsuperscript{2}-values of 0.02 and 0.80, respectively
-\citep{Ferguson2011-zl} (Figure~\ref{fig-Ferg}). Together with the data
-from two \emph{P. caerulea} shells (JM00 and JM30 with
+\citep{Ferguson2011-zl} (see also Supplementaries). Together with the
+data from two \emph{P. caerulea} shells (JM00 and JM30 with
 R\textsuperscript{2}-values of 0.69 and 0.83), the data from JL2 was
@@
 affected JL1 in the same way, albeit stronger, is not clear but it might
-well be possible.
+well be possible. Previous studies of \emph{P. caerulea} found similar
+patterns of anomalous Mg/Ca values in the m+3 layer
+\citep{Hausmann2019-fi}. These areas can be avoided using a
+2-dimensional approach and if the growth increment is sufficiently wide.
+These heterogeneities within growth increments are little understood but
+some research points towards a zonation of the mollusc's mantle and
+along the growth edge \citep{Lazareth2013-jf}. This zonation could
+differentially influence the transport of Mg ions to the extrapallial
+fluid and into the shell.
+
+Another shell whose results we aim to explain further is ORK-LT5 (see
+Supplementary Information), which has previously been shown to have no
+substantial correlation with \(\delta\)\textsuperscript{18}O values or
+SST by implication \citep{Graniero2017-io}. In contrast to other
+specimens in this study, we were able to resample this specimen and
+acquire Mg/Ca data on a slightly higher resolution (20--30 µm spots in a
+100 µm raster compared to 50 µm spots spaced at 150--300 µm) and using 2
+dimensions, as well as using a continuous high-resolution line-scan at
+(20--30 µm spots spaced at 10 µm) (Figure~\ref{fig-ORK_sub}). These
+additional data revealed previously missed summer peaks at around 2--3
+mm distance to the shell edge (grey box in Figure~\ref{fig-ORK_sub}) and
+revealed additional growth at the very edge (dashed box in
+Figure~\ref{fig-ORK_sub}) mirroring the change towards warmer
+temperatures at the time of collection (August). Including these missing
+parts lead to a better alignment of both records and a resulting
+R\textsuperscript{2}-value of 0.88 (Figure~\ref{fig-ORK_sub}).
 
@@
 
-\includegraphics{fig-Ferg-1.png}
-
-}
-
-\caption{\label{fig-Ferg}Correlation graphs for Ferguson et al., (2011)
-specimens.}
-
-\end{figure}%
-
-Another shell whose results we aim to explain further is ORK-LT5, which
-has previously been shown to have no substantial correlation with
-\(\delta\)\textsuperscript{18}O values or SST by implication
-\citep{Graniero2017-io}. In contrast to other specimens in this study,
-we were able to resample this specimen and acquire Mg/Ca data on a
-slightly higher resolution (20--30 µm spots in a 100 µm raster compared
-to 50 µm spots spaced at 150--300 µm) and using 2 dimensions, as well as
-using a continuous high-resolution line-scan at (20--30 µm spots spaced
-at 10 µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
-previously missed summer peaks at around 2--3 mm distance to the shell
-edge (grey box in Figure~\ref{fig-ORK_sub}) and revealed additional
-growth at the very edge (dashed box in Figure~\ref{fig-ORK_sub})
-mirroring the change towards warmer temperatures at the time of
-collection (August). Including these missing parts lead to a better
-alignment of both records and a resulting R\textsuperscript{2}-value of
-0.88 (Figure~\ref{fig-ORK_sub}).
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{fig-ORK_sub-1.pdf}
+\includegraphics{Manuscript_files/figure-pdf/fig-ORK_sub-1.pdf}
 
@@
 for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca
-and \(\delta\)\textsuperscript{18}O-values}
+and \(\delta\)\textsuperscript{18}O-values.}
 
@@
 between Mg/Ca ratios and SST.
+
+\subsection{Temperature correlation}\label{temperature-correlation}
+
+We further related the Mg/Ca ratios of the modern shells to actual SST
+data from the respective localities (Figure~\ref{fig-SSTs}). The data
+for ORK-LT5 from Orkney were cut to the time period between the start of
+our available SST data (2007) and the time of collection (August 2009).
+For the \emph{Nacella} spp. from Cambaceres Bay we chose an overlapping
+time period, to make the results more comparable between the specimens,
+which is the lifespan of the youngest specimen NM-1016-3 from January
+2015 to October 2016, the time of collection for all 4 specimens.
+
+As indicated by the stable isotope values, no reliable shared
+relationship between the data can be established, but each shell's Mg/Ca
+data individually have a good correlation with SST (R\textsuperscript{2}
+values between 0.93 to 0.96 with p-values consistently below 0.001)
+(Figure~\ref{fig-SST_Corr}), which are even better than the correlation
+between elemental ratios and stable isotope values. This improvement is
+most likely due to the simple and sinusoidal SST curve, which is easier
+to fit to other data.
+
+As a result of the different correlations between Mg/Ca ratios and SST
+in the modern shells, we did not attempt to use Mg/Ca as a direct proxy
+for absolute SST values in ºC for the archaeological shells from
+Quoygrew, but rather as a proxy for relative SST change
+(i.e.~warmer/colder). In future analyses, these high resolution and
+rapidly acquired records of relative SST can guide the sampling for
+oxygen isotope or clumped isotope analysis and reduce analytical costs
+per shell, thus increasing the quantity of shells that can be analysed
+per study.
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{Manuscript_files/figure-pdf/fig-SST_Corr-1.png}
+
+}
+
+\caption{\label{fig-SST_Corr}Comparison of Mg/Ca and SST in all modern
+shells.}
+
+\end{figure}%
 
@@
 \citep{Colonese2011-ab, Villagran2011-ld, Zangrando2016-yl}. By using
-LIBS as a screening tool for targeted \delta\^{}\{18\}O analysis,
-focusing on annual minima and maxima revealed in the Mg/Ca record,
-researchers can improve cost and time efficiency. This novel method will
-enable them to increase sample numbers and produce larger, more in-depth
-climatic studies than previously possible.
+LIBS as a screening tool for targeted \(\delta\)\textsuperscript{18}O
+analysis, focusing on annual minima and maxima revealed in the Mg/Ca
+record, researchers can improve cost and time efficiency. This novel
+method will enable them to increase sample numbers and produce larger,
+more in-depth climatic studies than previously possible.
 
-\renewcommand\refname{References}
- \bibliographystyle{spmpsci}
-\bibliography{bibliography}
+\section{Acknowledgements}\label{acknowledgements}
+
+Niklas Hausmann was supported by the German Research Foundation under
+the Emmy Noether Program (project number: 439799406). We thank the
+Applied Photonics Lab of the Institute of Elec- tronic Structure and
+Laser of the Foundation for Research and Technology - Hellas for close
+support with the LIBS system. We are also grateful for two anonymous
+reviewers who helped to improve the manuscript.
+
+  \bibliography{bibliography.bib}
 
